---
title: "Adding Recruiting to CFB Elo Ratings"
author: "Phil Henrickson"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE #adds a Table of Contents
    theme: cerulean
    number_sections: TRUE #number your headings/sections
    toc_float: TRUE #let your ToC follow you as you scroll
    keep_md: no
    fig.caption: yes
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = F,
                      error = F,
                      warning=F,
                      dev="png",
                      fig.width = 10,
                      fig.height = 6)

options(knitr.duplicate.label = "allow")

options(scipen=999)

```

```{r connect to snowflake}

library(DBI)
library(odbc)
library(RODBC)
library(keyring)

# connect to snowflake
myconn <- DBI::dbConnect(odbc::odbc(),
                         "SnowflakeDSII",
                         Database = "CFB_DEMO",
                         warehouse = "DEMO_WH",
                         uid="phil.henrickson",
                         pwd=keyring::key_get("AE_Snowflake"))

```

```{r packages, include=F} 

source(here::here("scripts/load_packages.R"))
library(rstan)
library(rstanarm)
library(rstantools)
library(jsonlite)
library(forcats)
library(teamcolors)
conflict_prefer("lag", "dplyr")

# load teamcolors
teamcolors = teamcolors %>%
        mutate(location = case_when(location == 'Ole' ~ name,
                                    location == 'Southern California Trojans' ~ 'USC',
                                    location == 'Miami' & name == 'Miami (OH)' ~ name,
                                    TRUE ~ location)) %>%
        bind_rows(., 
                  teams_raw %>%
                          filter(TEAM == 'SMU' | TEAM == 'USC') %>%
                          mutate(name = TEAM,
                                 location = TEAM) %>%
                          rename(primary = COLOR,
                                 secondary = ALT_COLOR))

```

```{r flextable settings}

set_flextable_defaults(theme_fun = theme_alafoli,
                       font.color = "grey10",
                       font.size=8,
                       padding.bottom = 6, 
                       padding.top = 6,
                       padding.left = 6,
                       padding.right = 6,
                       background.color = "white")

```

```{r load functions}

# integer plotting function
int_breaks <- function(x, n = 5) {
  l <- pretty(x, n)
  l[abs(l %% 1) < .Machine$double.eps ^ 0.5] 
}

source(here::here("functions/theme_phil.R"))
source(here::here("functions/calc_elo_ratings.R"))
source(here::here("functions/get_expected_score.R"))
source(here::here("functions/get_new_elos.R"))
source(here::here("functions/calc_elo_ratings_with_recruiting.R"))
source(here::here("functions/plot_elo_historical_team.R"))
source(here::here("functions", "simulateX.R"))
source(here::here("functions", "sim_game_margin.R"))
source(here::here("functions", "sim_elo_ratings.R"))
rm(a)

```

```{r get the games data}

# data on teams with logos and colors
teams_raw = bind_rows(
        DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.TEAMS')) %>%
        as_tibble())  %>%
        rename(TEAM = SCHOOL,
               TEAM_ID = ID,
               TEAM_MASCOT = MASCOT,
               TEAM_ABBREVIATION = ABBREVIATION)

# get games
games_raw = DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.ANALYSIS_GAMES')) %>%
        as_tibble() %>%
        mutate(CONFERENCE_GAME = case_when(CONFERENCE_GAME == 'true' ~ T,
                                           CONFERENCE_GAME == 'false' ~ F)) %>%
        mutate(GAME_DATE = as.Date(START_DATE)) %>%
        mutate(START_DATE = as_datetime(START_DATE)) %>%
        arrange(START_DATE) %>%
        group_by(SEASON, SEASON_TYPE, HOME_CONFERENCE) %>%
        mutate(LAST_GAME = START_DATE == max(START_DATE)) %>% 
        ungroup() %>%
        # logic for flagging conference championship games
        mutate(CONFERENCE_CHAMPIONSHIP = case_when(CONFERENCE_GAME == T & 
                                                           (HOME_CONFERENCE != 'FBS Independents') & 
                                                           (HOME_CONFERENCE != 'FCS Independents') &
                                                           !(HOME_CONFERENCE %in% 'Big Ten' & SEASON < 2014) &
                                                           !(HOME_CONFERENCE %in% 'Western Athletic') &
                                                           (NEUTRAL_SITE == T | is.na(VENUE) |
                                                                    VENUE == 'Glass Bowl' |
                                                                    VENUE == 'Bright House Networks Stadium' |
                                                                    VENUE == 'Alamodome' |
                                                                    VENUE == 'Georgia Dome') &
                                                           SEASON > 2000 &
                                                           SEASON_TYPE == 'regular' &
                                                           WEEK > 13 &
                                                           LAST_GAME == T ~ T,
                                                   TRUE ~ F)) %>%
        mutate(HOME_WIN = factor(case_when(HOME_POINTS > AWAY_POINTS ~ 'yes',
                                    HOME_POINTS < AWAY_POINTS ~ 'no',
                                    HOME_POINTS == AWAY_POINTS ~ 'no'),
               levels = c("no", "yes"))) %>%
        mutate(HOME_DIVISION  = replace_na(HOME_DIVISION, "missing"),
               AWAY_DIVISION = replace_na(AWAY_DIVISION, "missing")) %>%
        left_join(., teams_raw %>%
                          select(TEAM, TEAM_ABBREVIATION) %>%
                          rename(HOME_TEAM = TEAM,
                                 HOME_TEAM_ABBR = TEAM_ABBREVIATION),
                  by = c("HOME_TEAM")) %>%
        left_join(., teams_raw %>%
                          select(TEAM, TEAM_ABBREVIATION) %>%
                          rename(AWAY_TEAM = TEAM,
                                 AWAY_TEAM_ABBR = TEAM_ABBREVIATION),
                  by = c("AWAY_TEAM"))

# make a table of team conference by season
team_conference_season = games_raw %>% 
        select(SEASON, HOME_TEAM, HOME_CONFERENCE, HOME_DIVISION) %>% 
        rename(TEAM = HOME_TEAM, 
               CONFERENCE = HOME_CONFERENCE,
               DIVISION = HOME_DIVISION) %>% 
        bind_rows(.,
            games_raw %>%
                    select(SEASON, AWAY_TEAM, AWAY_CONFERENCE, AWAY_DIVISION) %>% 
                    rename(TEAM = AWAY_TEAM, 
                           CONFERENCE = AWAY_CONFERENCE,
                           DIVISION = AWAY_DIVISION)) %>%
        select(SEASON, TEAM, CONFERENCE, DIVISION) %>%
        unique()

```

```{r get recruiting data}

# recruiting
recruiting_teams_raw = DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.RECRUITING_TEAMS')) %>%
        as_tibble() %>%
        mutate(POINTS = as.numeric(POINTS)) %>%
        rename(RECRUITING_POINTS = POINTS)

```

```{r get elo functions}

source(here::here("functions/get_expected_score.R"))
source(here::here("functions/get_new_elos.R"))
source(here::here("functions/calc_elo_ratings.R"))
source(here::here("functions/calc_elo_ratings_with_recruiting"))
source(here::here("functions", "simulateX.R"))
source(here::here("functions", "sim_game_margin.R"))
source(here::here("functions", "sim_elo_ratings.R"))

# with recruiting
source(here::here("functions", "calc_elo_ratings_with_recruiting.R"))
source(here::here("functions", "sim_elo_ratings_with_recruiting.R"))

```

# Simulating Regular Season Games

I'll now do this for every season through 2020. This means I'll compute Elo ratings in the run up to each season, then simulate the season based on the ratings going into the season.

```{r initialize elo, include=F}

# set of games on which to learn baseline elo
initial_games = games_data_tidied %>%
                filter(SEASON >=1970) %>%
                filter(SEASON < 2000) %>%
                arrange(GAME_DATE) %>%
                select(GAME_ID, 
                       SEASON, 
                       WEEK,
                       SEASON_TYPE,
                       GAME_DATE,
                       NEUTRAL_SITE, 
                       HOME_TEAM,
                       AWAY_TEAM,
                       HOME_CONFERENCE,
                       AWAY_CONFERENCE,
                       HOME_POINTS,
                       AWAY_POINTS)

# set pars
selected_pars = tibble(reversion = .1,
                       home_field_advantage = 60,
                       #recruiting_adjustment = .5,
                       k = 35,
                       v = 400)

# now get the elo ratings based on this training set
elo_initial = calc_elo_ratings(
        games = initial_games,
        home_field_advantage = selected_pars$home_field_advantage,
        reversion = selected_pars$reversion,
        k = selected_pars$k,
        v = selected_pars$v,
        verbose = T)

```

```{r set parallel, include=F}

library(future.apply)
plan(multisession, workers = 7)

```

```{r look at performance through 2022, include=F, eval=F}

rm(sim_seasons,
   train_games,
   valid_games,
   elo_train,
   elo_valid)

# run 
seasons = seq(2001, 2022, by = 1)

# loop over season
simulated_seasons = foreach(i = 1:length(seasons)) %do% {
        
        # get elo ratings in the run up to a season
        # filter to games that occured 
        train_games = games_data_tidied %>%
                filter(SEASON > max(initial_games$SEASON)) %>%
                filter(SEASON < seasons[i]) %>%
                arrange(GAME_DATE) %>%
                select(GAME_ID, 
                       SEASON, 
                       WEEK,
                       SEASON_TYPE,
                       GAME_DATE,
                       NEUTRAL_SITE, 
                       HOME_TEAM,
                       AWAY_TEAM,
                       HOME_CONFERENCE,
                       AWAY_CONFERENCE,
                       HOME_POINTS,
                       AWAY_POINTS)
        
        # valid
        sim_games = games_data_tidied %>%
                filter(SEASON == seasons[i]) %>%
                filter(SEASON <= max(seasons)) %>%
                arrange(GAME_DATE) %>%
                select(GAME_ID, 
                       SEASON, 
                       WEEK,
                       SEASON_TYPE,
                       GAME_DATE,
                       NEUTRAL_SITE, 
                       HOME_TEAM,
                       AWAY_TEAM,
                       HOME_CONFERENCE,
                       AWAY_CONFERENCE,
                       HOME_POINTS,
                       AWAY_POINTS)
        
        # now get the elo ratings based on this training set
        elo_train = calc_elo_ratings(train_games,
                                     teams = elo_initial$teams,
                                     team_seasons = elo_initial$team_seasons,
                          home_field_advantage = selected_pars$home_field_advantage,
                          reversion = selected_pars$reversion,
                          k = selected_pars$k,
                          v = selected_pars$v,
                          verbose = F)
        
        # regress point differential on elo differential
        # fit linear model after first few years in training
        points_model = bind_rows(elo_initial$game_outcomes,
                                 elo_train$game_outcomes) %>%
                filter(SEASON >=1975) %>%
                mutate(HOME_ELO_DIFF = case_when(NEUTRAL_SITE == F ~ HOME_PREGAME_ELO + selected_pars$home_field_advantage - AWAY_PREGAME_ELO,
                                                 TRUE ~ HOME_PREGAME_ELO - AWAY_PREGAME_ELO)) %>%
                mutate(HOME_SCORE_DIFF = HOME_POINTS-AWAY_POINTS) %>%
                nest() %>%
                # now fit linear models, i'll do so using stan
                mutate(lm = map(data, ~ lm(HOME_SCORE_DIFF ~ HOME_ELO_DIFF,
                                     data = .x))) %>%
                mutate(tidied = map(lm, tidy, conf.int=T)) %>%
                mutate(glanced = map(lm, glance)) %>%
                pluck("lm",1)
        
        # then, simulate the next season
        set.seed(1999)
        sim_seasons = future_replicate(1000,
                          sim_elo_ratings(sim_games,
                                          teams = elo_train$teams,
                                          team_seasons = elo_train$team_seasons,
                                          home_field_advantage = selected_pars$home_field_advantage,
                                          reversion = selected_pars$reversion,
                                          k = selected_pars$k,
                                          v = selected_pars$v,
                                          verbose=F,
                                          ties =F,
                                          points_model = points_model))
        
        # grab what we need
        # out = list("sim_team_outcomes" = rbindlist(sim_seasons['team_outcomes',],
        #                               idcol = T),
        #            "sim_game_outcomes" = rbindlist(sim_seasons['game_outcomes',],
        #                               idcol = T))
        
        # get the simulated game outcomes
        out = rbindlist(sim_seasons['game_outcomes',],
                                      idcol = T) %>%
                mutate(SIMULATION_FROM_SEASON = seasons[i],
                       SIMULATION_FROM_WEEK = 0) %>%
                mutate(home_field_advantage = selected_pars$home_field_advantage,
                       reversion = selected_pars$reversion,
                       k = selected_pars$k,
                       v = selected_pars$v)
        
        rm(sim_seasons)
        
        # track progress
        cat("\r", i, "of", length(seasons), "seasons completed");  flush.console()
        
        # return output
        out

}

rm(out)

# set the names
names(simulated_seasons) = seasons

# save each season
for(i in 1:length(seasons)) {
        
        season = names(simulated_seasons)[i]
        sims = simulated_seasons[i]
        simulation_from = 0
        save(sims,
             file = here::here("simulations", paste("sims_seasons", "_week_", simulation_from, ".Rdata", sep="")))
        
}

# save all seasons
save(simulated_seasons,
             file = here::here("simulations", paste("sims_seasons", "_week_", simulation_from, ".Rdata", sep="")))

```

```{r load seasons back in}

load(here::here("simulations", paste("sims_seasons", "_week_", simulation_from, ".Rdata", sep="")))

```


```{r make team outcomes}

sim_game_outcomes = rbindlist(simulated_seasons)

sim_team_outcomes = sim_game_outcomes %>% 
        mutate(HOME_OPPONENT = AWAY_TEAM) %>%
        select(.id, 
               GAME_ID, 
               SEASON,
               WEEK,
               GAME_DATE,
               starts_with("HOME_")) %>%
        set_names(., gsub("HOME_", "", names(.))) %>%
        bind_rows(.,
                  sim_game_outcomes %>% 
                          mutate(AWAY_OPPONENT = HOME_TEAM) %>%
                          mutate(AWAY_MARGIN = -HOME_MARGIN) %>%
                          select(.id, GAME_ID, 
                                 SEASON,
                                 WEEK,
                                 GAME_DATE,
                                 starts_with("AWAY_")) %>%
                          set_names(., gsub("AWAY_", "", names(.)))) %>%
        arrange(GAME_DATE) %>%
        mutate(home_field_advantage = selected_pars$home_field_advantage,
               reversion = selected_pars$reversion,
               k = selected_pars$k,
               v = selected_pars$v) %>%
        left_join(., games_data_tidied %>%
                          select(GAME_ID, CONFERENCE_GAME, CONFERENCE_CHAMPIONSHIP),
                  by = c("GAME_ID"))

```

## Assess Historical Performance

I'll get season by season accuracy/log-loss.

```{r simulating all seasons}

assess_metrics = metric_set(yardstick::mn_log_loss,
                           yardstick::accuracy)

overall_assess = sim_game_outcomes %>%
        filter(SEASON < 2022) %>%
        mutate(HOME_PRED = case_when(HOME_MARGIN >0 ~ 1,
                                     TRUE ~ 0)) %>%
        group_by(GAME_ID, SEASON, WEEK, GAME_DATE, NEUTRAL_SITE, HOME_TEAM, AWAY_TEAM, HOME_POINTS, AWAY_POINTS) %>%
        summarize(HOME_PROB = sum(HOME_PRED) / n(),
                  PRED_MARGIN = mean(HOME_MARGIN),
                  .groups = 'drop') %>%
        mutate(AWAY_PROB = 1-HOME_PROB) %>%
        mutate(HOME_WIN = factor(case_when(HOME_POINTS > AWAY_POINTS  ~ 'yes',
                                       TRUE ~'no'))) %>%
        mutate(HOME_PRED = factor(case_when(HOME_PROB >=.5 ~ 'yes',
                                       TRUE ~ 'no'))) %>%
        mutate(PRED_CORRECT = case_when(HOME_WIN == HOME_PRED ~ 'yes',
                                        TRUE ~ 'no')) %>%
        mutate(yes = HOME_PROB) %>%
        group_by(SEASON) %>%
        mutate(games = n()) %>%
        group_by(SEASON, games) %>%
        assess_metrics(truth = HOME_WIN,
                       yes,
                       estimate = HOME_PRED,
                       event_level = 'second')

#
overall_assess %>%
        ggplot(., aes(x=SEASON,
                      y=.estimate))+
        geom_line()+
        geom_point()+
        facet_wrap(.metric ~.,
                   scales = "free_y",
                   ncol = 1)+
        theme_bw()+
        ggtitle("Assessing Elo Simulations by Season")

```

Interesting - I can understand not doing so well on the 2020 season, but 2016 was evidently also a tough one to predict too. Generally speaking, our preseason accuracy hovers around 70% and our log loss is around .55.

I'll also look at this by week for each season.

```{r performance by week of season}

sim_game_outcomes %>%
        filter(SEASON < 2022) %>%
        mutate(HOME_PRED = case_when(HOME_MARGIN >0 ~ 1,
                                     TRUE ~ 0)) %>%
        group_by(GAME_ID, SEASON, WEEK, GAME_DATE, NEUTRAL_SITE, HOME_TEAM, AWAY_TEAM, HOME_POINTS, AWAY_POINTS) %>%
        summarize(HOME_PROB = sum(HOME_PRED) / n(),
                  PRED_MARGIN = mean(HOME_MARGIN),
                  .groups = 'drop') %>%
        mutate(AWAY_PROB = 1-HOME_PROB) %>%
        mutate(HOME_WIN = factor(case_when(HOME_POINTS > AWAY_POINTS  ~ 'yes',
                                       TRUE ~'no'))) %>%
        mutate(HOME_PRED = factor(case_when(HOME_PROB >=.5 ~ 'yes',
                                       TRUE ~ 'no'))) %>%
        mutate(PRED_CORRECT = case_when(HOME_WIN == HOME_PRED ~ 'yes',
                                        TRUE ~ 'no')) %>%
        mutate(yes = HOME_PROB) %>%
        group_by(SEASON, WEEK) %>%
        mutate(games = n()) %>%
        group_by(SEASON, WEEK, games) %>%
        assess_metrics(truth = HOME_WIN,
                       yes,
                       estimate = HOME_PRED,
                       event_level = 'second') %>%
        mutate(SEASON = factor(SEASON)) %>%
        ggplot(., aes(x=WEEK,
                      color = SEASON,
                      y = .estimate))+
         geom_line(lwd = 1.05,
                  stat = 'smooth',
                  method = 'loess',
                  formula = 'y ~ x',
                  span = 0.2)+
        facet_wrap(.metric ~.,
                   ncol =1)+
        scale_color_viridis_d()+
        theme_bw()

```

### Examine Seasons

Now that we have preseason simulations for the selected seasons, we can start diving into the results and examine them.

```{r source functions again}

source(here::here("functions", "plot_elo_conference_season.R"))
source(here::here("functions", "plot_elo_team_season.R"))
source(here::here("functions", "plot_wins_conference_season.R"))
source(here::here("functions", "table_wins_conference_season.R"))
source(here::here("functions", "table_wins_team_season.R"))
source(here::here("functions", "table_wins_season.R"))

```


```{r wins season}
# 
# table_wins_season(sim_team_outcomes,
#                   season = 2005,
#                   14)

table_wins_conference_season(sim_team_outcomes %>%
                                     filter(CONFERENCE_GAME == T),
                             conference = 'SEC',
                             season = 2022)

```



```{r look at specific teams}

plot_elo_conference_season(sim_team_outcomes,
                           season = 2022,
                           conference = 'Big Ten',
                           15)


plot_wins_conference_season(sim_team_outcomes,
                           season = 2022,
                           conference = 'Big Ten',
                           15)

plot_elo_team_season(sim_team_outcomes %>%
                           season = 2022,
                           'Wisconsin',
                           15)

```

```{r looking at graphs}

plot_elo_team_season(sim_team_outcomes,
                           season = 2022,
                           team = 'Wisconsin',
                           15)

table_wins_team_season(sim_team_outcomes,
                       2022,
                       'Wisconsin',
                       15)

```

```{r plot win totals selected}

plot_wins_conference_season(sim_team_outcomes,
                           season = 2022,
                           conference = 'ACC',
                           week = 15)

table_wins_team_season(sim_team_outcomes,
                       season = 2022,
                       team = 'Ohio State',
                       week = 15)

table_wins_conference_season(sim_team_outcomes %>%
                                     filter(CONFERENCE_GAME == T),
                           season = 2022,
                           conference = 'ACC')

plot_elo_team_season(sim_team_outcomes,
                     season = 2022,
                     team = 'Ohio State',
                     week = 15)

plot_elo_conference_season(sim_team_outcomes,
                     season = 2019,
                     conference = 'Big 12',
                     week = 15
                     )

```


# Simulating Conference Championships
