---
title: "Examining CFB Elo Ratings"
author: "Phil Henrickson"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE #adds a Table of Contents
    theme: cerulean
    number_sections: TRUE #number your headings/sections
    toc_float: TRUE #let your ToC follow you as you scroll
    keep_md: no
    fig.caption: yes
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = F,
                      error = F,
                      warning=F,
                      dev="png",
                      fig.width = 10,
                      fig.height = 6)

options(knitr.duplicate.label = "allow")

options(scipen=999)

```

```{r connect to snowflake}

library(DBI)
library(odbc)
library(RODBC)
library(keyring)

# connect to snowflake
myconn <- DBI::dbConnect(odbc::odbc(),
                         "SnowflakeDSII",
                         Database = "CFB_DEMO",
                         warehouse = "DEMO_WH",
                         uid="phil.henrickson",
                         pwd=keyring::key_get("AE_Snowflake"))

```

```{r packages, include=F} 

source(here::here("scripts/load_packages.R"))
library(rstan)
library(rstanarm)
library(rstantools)
library(jsonlite)
library(forcats)
library(teamcolors)
library(shiny)
conflict_prefer("lag", "dplyr")

```

```{r get colors for ncaa teams}

# look at colors for ncca
ncaa_colors = teamcolors %>%
        filter(league == 'ncaa') 

# define palette for ncaa
#league_pal("ncaa", 1)

```

```{r flextable settings}

library(flextable)
set_flextable_defaults(theme_fun = theme_alafoli,
                       font.color = "grey10",
                       font.size=8,
                       padding.bottom = 6, 
                       padding.top = 6,
                       padding.left = 6,
                       padding.right = 6,
                       background.color = "white")

```

```{r load functions}

# integer plotting function
int_breaks <- function(x, n = 5) {
  l <- pretty(x, n)
  l[abs(l %% 1) < .Machine$double.eps ^ 0.5] 
}

source(here::here("functions/theme_phil.R"))
source(here::here("functions/calc_elo_ratings.R"))
source(here::here("functions/get_expected_score.R"))
source(here::here("functions/get_new_elos.R"))
source(here::here("functions/calc_elo_ratings_with_recruiting.R"))
source(here::here("functions/plot_elo_historical_team.R"))

rm(a)

```

```{r get the games data}

# data on teams with logos and colors
teams_raw = bind_rows(
        DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.TEAMS')) %>%
        as_tibble())  %>%
        rename(TEAM = SCHOOL,
               TEAM_ID = ID,
               TEAM_MASCOT = MASCOT,
               TEAM_ABBREVIATION = ABBREVIATION)

teamcolors = teamcolors::teamcolors %>%
        filter(league == 'ncaa')

# get teamcolors
teamcolors = teams_raw %>%
        select(TEAM, COLOR, ALT_COLOR) %>%
        left_join(.,
                  teamcolors %>%
                          filter(league == 'ncaa') %>%
                          mutate(location = case_when(location == 'Ole' ~ name,
                                    location == 'Southern California Trojans' ~ 'USC',
                                    location == 'Miami' & name == 'Miami (OH)' ~ name,
                                    TRUE ~ location)) %>%
                          mutate(TEAM = location),
                  by = c("TEAM")) %>%
        mutate(name = case_when(is.na(name) ~ TEAM,
                                TRUE ~ name)) %>%
        mutate(location = case_when(is.na(location) ~ TEAM)) %>%
        mutate(primary = case_when(is.na(primary) ~ COLOR,
                                   TRUE ~ primary)) %>%
        mutate(secondary = case_when(is.na(secondary) ~ COLOR,
                                   TRUE ~ secondary))

# get games
games_raw = DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.ANALYSIS_GAMES')) %>%
        as_tibble() %>%
        mutate(CONFERENCE_GAME = case_when(CONFERENCE_GAME == 'true' ~ T,
                                           CONFERENCE_GAME == 'false' ~ F)) %>%
        mutate(GAME_DATE = as.Date(START_DATE)) %>%
        mutate(START_DATE = as_datetime(START_DATE)) %>%
        arrange(START_DATE) %>%
        group_by(SEASON, SEASON_TYPE, HOME_CONFERENCE) %>%
        mutate(LAST_GAME = START_DATE == max(START_DATE)) %>% 
        ungroup() %>%
        # logic for flagging conference championship games
        mutate(CONFERENCE_CHAMPIONSHIP = case_when(CONFERENCE_GAME == T & 
                                                           (HOME_CONFERENCE != 'FBS Independents') & 
                                                           (HOME_CONFERENCE != 'FCS Independents') &
                                                           !(HOME_CONFERENCE %in% 'Big Ten' & SEASON < 2014) &
                                                           !(HOME_CONFERENCE %in% 'Western Athletic') &
                                                           (NEUTRAL_SITE == T | is.na(VENUE) |
                                                                    VENUE == 'Glass Bowl' |
                                                                    VENUE == 'Bright House Networks Stadium' |
                                                                    VENUE == 'Alamodome' |
                                                                    VENUE == 'Georgia Dome') &
                                                           SEASON > 2000 &
                                                           SEASON_TYPE == 'regular' &
                                                           WEEK > 13 &
                                                           LAST_GAME == T ~ T,
                                                   TRUE ~ F)) %>%
        mutate(HOME_WIN = factor(case_when(HOME_POINTS > AWAY_POINTS ~ 'yes',
                                    HOME_POINTS < AWAY_POINTS ~ 'no',
                                    HOME_POINTS == AWAY_POINTS ~ 'no'),
               levels = c("no", "yes"))) %>%
        mutate(HOME_DIVISION  = replace_na(HOME_DIVISION, "missing"),
               AWAY_DIVISION = replace_na(AWAY_DIVISION, "missing")) %>%
        left_join(., teams_raw %>%
                          select(TEAM, TEAM_ABBREVIATION) %>%
                          rename(HOME_TEAM = TEAM,
                                 HOME_TEAM_ABBR = TEAM_ABBREVIATION),
                  by = c("HOME_TEAM")) %>%
        left_join(., teams_raw %>%
                          select(TEAM, TEAM_ABBREVIATION) %>%
                          rename(AWAY_TEAM = TEAM,
                                 AWAY_TEAM_ABBR = TEAM_ABBREVIATION),
                  by = c("AWAY_TEAM"))

# make a table of team conference by season
team_conference_season = games_raw %>% 
        select(SEASON, HOME_TEAM, HOME_CONFERENCE, HOME_DIVISION) %>% 
        rename(TEAM = HOME_TEAM, 
               CONFERENCE = HOME_CONFERENCE,
               DIVISION = HOME_DIVISION) %>% 
        bind_rows(.,
            games_raw %>%
                    select(SEASON, AWAY_TEAM, AWAY_CONFERENCE, AWAY_DIVISION) %>% 
                    rename(TEAM = AWAY_TEAM, 
                           CONFERENCE = AWAY_CONFERENCE,
                           DIVISION = AWAY_DIVISION)) %>%
        select(SEASON, TEAM, CONFERENCE, DIVISION) %>%
        unique()

```

```{r get recruiting data}

# recruiting
recruiting_teams_raw = DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.RECRUITING_TEAMS')) %>%
        as_tibble() %>%
        mutate(POINTS = as.numeric(POINTS)) %>%
        rename(RECRUITING_POINTS = POINTS)

# expand
library(TTR)
recruiting_teams = expand.grid(TEAM = unique(recruiting_teams_raw$TEAM),
             YEAR = seq(min(recruiting_teams_raw$YEAR),
                        max(recruiting_teams_raw$YEAR))) %>%
        left_join(., recruiting_teams_raw,
                  by = c("YEAR", "TEAM")) %>%
        mutate(RECRUITING_POINTS = replace_na(RECRUITING_POINTS, 0)) %>%
        filter(YEAR > 2001) %>%
        left_join(., team_conference_season,
                  by = c("YEAR"="SEASON", "TEAM")) %>%
        filter(DIVISION == 'fbs') %>%
        group_by(TEAM) %>%
        mutate(n = n()) %>%
        filter(n > 4) %>%
        group_by(TEAM) %>%
        select(-n) %>%
        mutate(EMA = EMA(RECRUITING_POINTS, n =4)) %>%
        mutate(WMA = WMA(RECRUITING_POINTS, n =4, wts = c(.35, .35, .15, .15))) %>%
        # now fill with actual in the event that a team doesn't have yet have enough data
        mutate(EMA = case_when(is.na(EMA) & !is.na(RECRUITING_POINTS) ~ RECRUITING_POINTS,
                                  TRUE ~ EMA)) %>%
        mutate(WMA = case_when(is.na(WMA) & !is.na(RECRUITING_POINTS) ~ RECRUITING_POINTS,
                                  TRUE ~ WMA)) %>%
        # mutate_at(c("EMA", "WMA"),
        #           replace_na, 0) %>%
        arrange(TEAM, YEAR)  %>%
        mutate(SEASON = YEAR) %>%
        ungroup() %>%
        select(TEAM, SEASON, CONFERENCE, DIVISION, RANK, RECRUITING_POINTS, EMA, WMA)

# get what we need
recruiting = recruiting_teams %>%
        select(TEAM, SEASON, CONFERENCE, RECRUITING_POINTS, EMA) %>%
        rename(RECRUITING_SCORE = EMA)

```


```{r join recruiting with games}

# specify games to run function
games = games_raw %>%
        filter(SEASON >=1970) %>%
        arrange(GAME_DATE) %>%
        select(GAME_ID, 
               SEASON, 
               WEEK,
               SEASON_TYPE,
               START_DATE,
               GAME_DATE,
               NEUTRAL_SITE, 
               HOME_TEAM,
               AWAY_TEAM,
               HOME_CONFERENCE,
               AWAY_CONFERENCE,
               HOME_DIVISION,
               AWAY_DIVISION,
               HOME_POINTS,
               AWAY_POINTS) %>%
        left_join(.,
                  recruiting %>%
                          select(SEASON,
                                 TEAM,
                                 RECRUITING_SCORE) %>%
                          rename(HOME_TEAM = TEAM,
                                 HOME_RECRUITING_SCORE = RECRUITING_SCORE),
                  by = c("SEASON",
                         "HOME_TEAM")) %>%
        left_join(.,
                  recruiting %>%
                          select(SEASON,
                                 TEAM,
                                 RECRUITING_SCORE) %>%
                          rename(AWAY_TEAM = TEAM,
                                 AWAY_RECRUITING_SCORE = RECRUITING_SCORE),
                  by = c("SEASON",
                         "AWAY_TEAM"))

```



```{r calc elo scores both ways}

# set pars
selected_pars = tibble(k = 35,
                       v = 400,
                       reversion = 0.2,
                       home_field_advantage = 75,
                       recruiting_weight = 0)

# now get the elo ratings based on this training set
elo_games= calc_elo_ratings(games %>%
                                    filter(SEASON < 2022) %>%
                                    filter(!is.na(HOME_POINTS) & !is.na(AWAY_POINTS)),
                          home_field_advantage = selected_pars$home_field_advantage,
                          reversion = selected_pars$reversion,
                          k = selected_pars$k,
                          v = selected_pars$v,
                          verbose = T)

# # get with recruiting
# elo_games_recruiting= calc_elo_ratings_with_recruiting(games %>%
#                                     filter(SEASON < 2022) %>%
#                                     filter(!is.na(HOME_POINTS) & !is.na(AWAY_POINTS)),
#                                                        recruiting_weight = selected_pars$recruiting_weight,
#                                                           home_field_advantage = selected_pars$home_field_advantage,
#                                                           reversion = selected_pars$reversion,
#                                                           k = selected_pars$k,
#                                                           v = selected_pars$v,
#                                                           verbose = T)

# fbs teams
teams = elo_games$team_outcomes %>%
        filter(SEASON == 2021) %>%
        filter(DIVISION == 'fbs') %>%
        distinct(TEAM) %>%
        arrange(TEAM)

```


```{r select a team, fig.height=8, include =F, echo=FALSE}

# shinyApp(
# 
#   ui = fluidPage(
#     selectInput("team", "Team:",
#                 selected = 'Alabama',
#                 choices = teams),
#     plotOutput("phonePlot")
#   ),
# 
#   server = function(input, output) {
#     output$phonePlot = renderPlot({
#             
#             plot_elo_historical_team(elo_games$team_outcomes,
#                          team = input$team)
#     })
#   },
#   options = list(height = 1000)
# )

```

# Team Historical Elo Ratings

Having developed Elo ratings for all teams back to 1970, I can now just explore the results a bit. 

## Strongest Teams

Based on ratings over this time period, which teams have had the highest average?

```{r get highest average elo}

top_teams = elo_games$team_outcomes %>%
        filter(DIVISION == 'fbs') %>%
        group_by(TEAM) %>%
        summarize(Elo_Mean= mean(POSTGAME_ELO)) %>%
        mutate_if(is.numeric, round,0) %>%
        arrange(desc(Elo_Mean))

top_teams %>%
        head(10) %>%
        flextable() %>%
        autofit()
        
```

```{r plot these teams}

plot_elo_historical_team(
        elo_games$team_outcomes,
        team = top_teams %>%
        head(9) %>%
        pull(TEAM))+
        ggtitle("Historical Elo Ratings for Top CFB Teams")+
        facet_wrap(SELECTED_TEAM ~.)

```


## Rivalries

I can look at rivals side by side to see how the two teams typically compared in strength.

```{r plot elo historical michigan}

plot_elo_historical_team(elo_games$team_outcomes,
                         team = c('Michigan', 'Ohio State'))

```

```{r texas tamu}

plot_elo_historical_team(elo_games$team_outcomes,
                         team = c('Texas', 'Texas A&M'))

```


```{r plot elo historical Michigan}

plot_elo_historical_team(elo_games$team_outcomes,
                         team = c('Army', 'Navy'))

```

## Highest Variance Teams

Which teams have had the highest variance in the last 40 odd years?

```{r get the variance of all teams}

highest_variance_teams = elo_games$team_outcomes %>%
        filter(DIVISION == 'fbs') %>%
        group_by(TEAM) %>%
        summarize(Elo_Variance = var(POSTGAME_ELO)) %>%
        mutate_if(is.numeric, round,0) %>%
        arrange(desc(Elo_Variance))

highest_variance_teams %>%
        head(10) %>%
        flextable() %>%
        autofit()
```


```{r most volatile teams}

plot_elo_historical_team(
        elo_games$team_outcomes,
        team = highest_variance_teams %>%
        head(9) %>%
        pull(TEAM))+
        ggtitle("Historical Elo Ratings for Most Volatile CFB Teams")+
        facet_wrap(SELECTED_TEAM ~.)

```



## Biggest Season Changes in Elo

Which teams have changed the most between seasons?

```{r which teams have changed most year over year}

year_over_year_changes = elo_games$team_outcomes %>%
        filter(DIVISION == 'fbs') %>%
        arrange(SEASON, TEAM) %>%
        group_by(SEASON, TEAM) %>%
        filter(GAME_DATE == max(GAME_DATE)) %>%
        group_by(TEAM) %>%
        rename(SEASON_END_ELO = POSTGAME_ELO) %>%
        select(SEASON, TEAM, SEASON_END_ELO) %>%
        mutate(PREVIOUS_SEASON_END_ELO = lag(SEASON_END_ELO,1)) %>%
        ungroup() %>%
        mutate(DIFF = SEASON_END_ELO - PREVIOUS_SEASON_END_ELO) %>%
        arrange(DIFF)


```

Here are the biggest drops from the end of one season to the end of the next.

```{r biggest drops}

diff_func = 
                function(x) {
                        
                        breaks = c(-1000, -700, seq(-500, 500, 50), 700, 1000)
                        colorRamp=colorRampPalette(c("red", "white", "deepskyblue1"))
                        col_palette <- colorRamp(length(breaks))
                        mycut <- cut(x, 
                                     breaks = breaks,
                                     include.lowest = TRUE, 
                                     right=T,
                                     label = FALSE)
                        col_palette[mycut]
                        
                }

# biggest drops
year_over_year_changes %>%
        mutate_if(is.numeric, round, 0) %>%
        mutate(SEASON = factor(SEASON)) %>%
        rename(Season = SEASON,
               Team = TEAM,
               Season_End_Elo = SEASON_END_ELO,
               Previous_End_Elo = PREVIOUS_SEASON_END_ELO,
               Elo_Diff = DIFF) %>%
        arrange(Elo_Diff) %>%
        head(25) %>%
        flextable() %>%
        autofit() %>%
        bg(j='Elo_Diff',
           bg = diff_func)

```

Here are the teams that gained the most in Elo over the course of one season.

```{r team elo rating gaines}


# biggest drops
year_over_year_changes %>%
        mutate_if(is.numeric, round, 0) %>%
        mutate(SEASON = factor(SEASON)) %>%
        rename(Season = SEASON,
               Team = TEAM,
               Season_End_Elo = SEASON_END_ELO,
               Previous_End_Elo = PREVIOUS_SEASON_END_ELO,
               Elo_Diff = DIFF) %>%
        arrange(desc(Elo_Diff)) %>%
        head(25) %>%
        flextable() %>%
        autofit() %>%
        bg(j='Elo_Diff',
           bg = diff_func)

```

## Biggest Upsets

Which games were the biggest upsets?

```{r team upsets}

elo_games$team_outcomes %>%
        filter(OUTCOME == 'win') %>%
        left_join(., elo_games$game_outcomes %>% 
                          select(GAME_ID, HOME_TEAM, AWAY_TEAM, HOME_POINTS, AWAY_POINTS, HOME_PROB, AWAY_PROB),
                  by = c("GAME_ID")) %>%
        mutate(PROB = case_when(TEAM == HOME_TEAM ~ HOME_PROB,
                                TEAM == AWAY_TEAM ~ AWAY_PROB)) %>%
        mutate(WIN = case_when(OUTCOME == 'win' ~ 1,
                               OUTCOME == 'loss' ~ 0)) %>%
        mutate(DIFF = WIN - PROB) %>%
        mutate(WINNER = case_when(OUTCOME == 'win' ~ TEAM,
                                  OUTCOME == 'loss' ~ OPPONENT)) %>%
        arrange(desc(DIFF)) %>%
        mutate(Season = factor(SEASON),
               Date = GAME_DATE,
               Opponent = case_when(HOME_TEAM == OPPONENT ~ paste("@", OPPONENT),
                                    AWAY_TEAM == OPPONENT ~ paste("vs", OPPONENT)),
               Team = TEAM,
               Opponent = Opponent,
               `Pr(Win)` = round(PROB, 3), 
               Result = paste(POINTS, "-", OPPONENT_POINTS, " ", WINNER, sep="")) %>%
        select(Season, Date, Team, Opponent, `Pr(Win)`, Result) %>%
        head(25) %>%
        flextable() %>%
        autofit() %>%
        flextable::align(., j = c("Season", "Date", "Pr(Win)", "Result"),
              align = 'center', part = 'all')

```


## Biggest Game Changes in Elo

Which games saw the largest changes in Elo? This is a combination of being an upset as well as the margin of victory. Teams where the underdog won by a lot will see the a big increase.

```{r biggest wins by underdog}

elo_games$team_outcomes %>%
        mutate(ELO_DIFF = POSTGAME_ELO - PREGAME_ELO) %>%
        arrange(ELO_DIFF) %>%
        left_join(., elo_games$game_outcomes %>% 
                          select(GAME_ID, HOME_TEAM, AWAY_TEAM, HOME_POINTS, AWAY_POINTS, HOME_PROB, AWAY_PROB),
                  by = c("GAME_ID")) %>%
        mutate(PROB = case_when(TEAM == HOME_TEAM ~ HOME_PROB,
                                TEAM == AWAY_TEAM ~ AWAY_PROB)) %>%
        mutate(WIN = case_when(OUTCOME == 'win' ~ 1,
                               OUTCOME == 'loss' ~ 0)) %>%
        mutate(DIFF = WIN - PROB) %>%
        mutate(WINNER = case_when(OUTCOME == 'win' ~ TEAM,
                                  OUTCOME == 'loss' ~ OPPONENT)) %>%
        arrange(desc(ELO_DIFF)) %>%
        mutate(Season = factor(SEASON),
               Date = GAME_DATE,
               Opponent = case_when(HOME_TEAM == OPPONENT ~ paste("@", OPPONENT),
                                    AWAY_TEAM == OPPONENT ~ paste("vs", OPPONENT)),
               Team = TEAM,
               Pre_Elo = PREGAME_ELO,
               Post_Elo = POSTGAME_ELO,
               Diff = ELO_DIFF,
               Opponent = Opponent,
               `Pr(Win)` = round(PROB, 3), 
               Result = paste(POINTS, "-", OPPONENT_POINTS, " ", WINNER, sep="")) %>%
        mutate_at(c("Pre_Elo",
                    "Post_Elo", 
                    "Diff"),
                  round, 0) %>%
        head(25) %>%
        select(Season, Date, Team, Opponent, Pre_Elo, Post_Elo, Result, Diff) %>%
        flextable() %>%
        autofit() %>%
        flextable::align(., j = c("Season", "Date", "Result", "Diff"),
              align = 'center', part = 'all')
        
```

# Examining Specific Teams

## Texas A&M

### Best Teams

```{r season end elo tamu}

elo_games$team_outcomes %>%
        filter(DIVISION == 'fbs') %>%
        filter(TEAM == 'Texas A&M') %>%
        arrange(SEASON, TEAM) %>%
        group_by(SEASON, TEAM) %>%
        filter(GAME_DATE == max(GAME_DATE)) %>%
        rename(SEASON_END_ELO = POSTGAME_ELO) %>%
        mutate(SEASON_END_ELO = round(SEASON_END_ELO, 0)) %>%
        select(SEASON, TEAM, SEASON_END_ELO) %>%
        arrange(desc(SEASON_END_ELO)) %>%
        ungroup() %>%
        head(10) %>%
        mutate(SEASON = factor(SEASON)) %>%
        flextable() %>%
        autofit()

```

### Biggest Upsets

```{r tamu upsets}

elo_games$team_outcomes %>%
        filter(TEAM == 'Texas A&M' & OUTCOME == 'win') %>%
        left_join(., elo_games$game_outcomes %>% 
                          select(GAME_ID, HOME_TEAM, AWAY_TEAM, HOME_POINTS, AWAY_POINTS, HOME_PROB, AWAY_PROB),
                  by = c("GAME_ID")) %>%
        mutate(PROB = case_when(TEAM == HOME_TEAM ~ HOME_PROB,
                                TEAM == AWAY_TEAM ~ AWAY_PROB)) %>%
        mutate(WIN = case_when(OUTCOME == 'win' ~ 1,
                               OUTCOME == 'loss' ~ 0)) %>%
        mutate(DIFF = WIN - PROB) %>%
        mutate(WINNER = case_when(OUTCOME == 'win' ~ TEAM,
                                  OUTCOME == 'loss' ~ OPPONENT)) %>%
        arrange(desc(DIFF)) %>%
        mutate(Season = factor(SEASON),
               Date = GAME_DATE,
               Opponent = case_when(HOME_TEAM == OPPONENT ~ paste("@", OPPONENT),
                                    AWAY_TEAM == OPPONENT ~ paste("vs", OPPONENT)),
               Team = TEAM,
               Opponent = Opponent,
               `Pr(Win)` = round(PROB, 3), 
               Result = paste(POINTS, "-", OPPONENT_POINTS, " ", WINNER, sep="")) %>%
        mutate(Rank = row_number()) %>%
        select(Rank, Season, Date, Team, Opponent, `Pr(Win)`, Result) %>%
        head(25) %>%
        flextable() %>%
        autofit() %>%
        flextable::align(., j = c("Rank", "Season", "Date", "Pr(Win)", "Result"),
              align = 'center', part = 'all')


```

### Worst Losses

```{r wrost losses} 

elo_games$team_outcomes %>%
        filter(TEAM == 'Texas A&M' & OUTCOME == 'loss') %>%
        left_join(., elo_games$game_outcomes %>% 
                          select(GAME_ID, HOME_TEAM, AWAY_TEAM, HOME_POINTS, AWAY_POINTS, HOME_PROB, AWAY_PROB),
                  by = c("GAME_ID")) %>%
        mutate(PROB = case_when(TEAM == HOME_TEAM ~ HOME_PROB,
                                TEAM == AWAY_TEAM ~ AWAY_PROB)) %>%
        mutate(WIN = case_when(OUTCOME == 'win' ~ 1,
                               OUTCOME == 'loss' ~ 0)) %>%
        mutate(DIFF = WIN - PROB) %>%
        mutate(WINNER = case_when(OUTCOME == 'win' ~ TEAM,
                                  OUTCOME == 'loss' ~ OPPONENT)) %>%
        mutate(Season = factor(SEASON),
               Date = GAME_DATE,
               Opponent = case_when(HOME_TEAM == OPPONENT ~ paste("@", OPPONENT),
                                    AWAY_TEAM == OPPONENT ~ paste("vs", OPPONENT)),
               Team = TEAM,
               Opponent = Opponent,
               `Pr(Win)` = round(PROB, 3), 
               Result = paste(POINTS, "-", OPPONENT_POINTS, " ", WINNER, sep="")) %>%
        arrange(DIFF) %>%
        mutate(Rank = row_number()) %>%
        select(Rank, Season, Date, Team, Opponent, `Pr(Win)`, Result) %>%
        head(25) %>%
        flextable() %>%
        autofit() %>%
        flextable::align(., j = c("Rank", "Season", "Date", "Pr(Win)", "Result"),
              align = 'center', part = 'all')

```

### Biggest Game Changes

```{r elo swings tamu}

elo_games$team_outcomes %>%
        mutate(ELO_DIFF = POSTGAME_ELO - PREGAME_ELO) %>%
        filter(TEAM == 'Texas A&M') %>%
        arrange(ELO_DIFF) %>%
        left_join(., elo_games$game_outcomes %>% 
                          select(GAME_ID, HOME_TEAM, AWAY_TEAM, HOME_POINTS, AWAY_POINTS, HOME_PROB, AWAY_PROB),
                  by = c("GAME_ID")) %>%
        mutate(PROB = case_when(TEAM == HOME_TEAM ~ HOME_PROB,
                                TEAM == AWAY_TEAM ~ AWAY_PROB)) %>%
        mutate(WIN = case_when(OUTCOME == 'win' ~ 1,
                               OUTCOME == 'loss' ~ 0)) %>%
        mutate(DIFF = WIN - PROB) %>%
        mutate(WINNER = case_when(OUTCOME == 'win' ~ TEAM,
                                  OUTCOME == 'loss' ~ OPPONENT)) %>%
        mutate(Season = factor(SEASON),
               Date = GAME_DATE,
               Opponent = case_when(HOME_TEAM == OPPONENT ~ paste("@", OPPONENT),
                                    AWAY_TEAM == OPPONENT ~ paste("vs", OPPONENT)),
               Team = TEAM,
               Pre_Elo = PREGAME_ELO,
               Post_Elo = POSTGAME_ELO,
               Diff = ELO_DIFF,
               Opponent = Opponent,
               `Pr(Win)` = round(PROB, 3), 
               Result = paste(POINTS, "-", OPPONENT_POINTS, " ", WINNER, sep="")) %>%
        mutate_at(c("Pre_Elo",
                    "Post_Elo", 
                    "Diff"),
                  round, 0) %>%
        arrange(desc(abs(ELO_DIFF))) %>%
        head(25) %>%
        select(Season, Date, Team, Opponent, Pre_Elo, Post_Elo, Result, Diff) %>%
        flextable() %>%
        autofit() %>%
        flextable::align(., j = c("Season", "Date", "Result", "Diff"),
              align = 'center', part = 'all') %>%
        bg(., j = c("Diff"),
           bg = diff_func)
        

```


## Ohio State

### Best Teams

### Biggest Upsets

```{r ohio state upsets}

elo_games$team_outcomes %>%
        filter(TEAM == 'Ohio State' & OUTCOME == 'win') %>%
        left_join(., elo_games$game_outcomes %>% 
                          select(GAME_ID, HOME_TEAM, AWAY_TEAM, HOME_POINTS, AWAY_POINTS, HOME_PROB, AWAY_PROB),
                  by = c("GAME_ID")) %>%
        mutate(PROB = case_when(TEAM == HOME_TEAM ~ HOME_PROB,
                                TEAM == AWAY_TEAM ~ AWAY_PROB)) %>%
        mutate(WIN = case_when(OUTCOME == 'win' ~ 1,
                               OUTCOME == 'loss' ~ 0)) %>%
        mutate(DIFF = WIN - PROB) %>%
        mutate(WINNER = case_when(OUTCOME == 'win' ~ TEAM,
                                  OUTCOME == 'loss' ~ OPPONENT)) %>%
        arrange(desc(DIFF)) %>%
        mutate(Season = factor(SEASON),
               Date = GAME_DATE,
               Opponent = case_when(HOME_TEAM == OPPONENT ~ paste("@", OPPONENT),
                                    AWAY_TEAM == OPPONENT ~ paste("vs", OPPONENT)),
               Team = TEAM,
               Opponent = Opponent,
               `Pr(Win)` = round(PROB, 3), 
               Result = paste(POINTS, "-", OPPONENT_POINTS, " ", WINNER, sep="")) %>%
        mutate(Rank = row_number()) %>%
        select(Rank, Season, Date, Team, Opponent, `Pr(Win)`, Result) %>%
        head(25) %>%
        flextable() %>%
        autofit() %>%
        flextable::align(., j = c("Rank", "Season", "Date", "Pr(Win)", "Result"),
              align = 'center', part = 'all')


```

### Worst Losses

```{r wrost losses} 

elo_games$team_outcomes %>%
        filter(TEAM == 'Ohio State' & OUTCOME == 'loss') %>%
        left_join(., elo_games$game_outcomes %>% 
                          select(GAME_ID, HOME_TEAM, AWAY_TEAM, HOME_POINTS, AWAY_POINTS, HOME_PROB, AWAY_PROB),
                  by = c("GAME_ID")) %>%
        mutate(PROB = case_when(TEAM == HOME_TEAM ~ HOME_PROB,
                                TEAM == AWAY_TEAM ~ AWAY_PROB)) %>%
        mutate(WIN = case_when(OUTCOME == 'win' ~ 1,
                               OUTCOME == 'loss' ~ 0)) %>%
        mutate(DIFF = WIN - PROB) %>%
        mutate(WINNER = case_when(OUTCOME == 'win' ~ TEAM,
                                  OUTCOME == 'loss' ~ OPPONENT)) %>%
        mutate(Season = factor(SEASON),
               Date = GAME_DATE,
               Opponent = case_when(HOME_TEAM == OPPONENT ~ paste("@", OPPONENT),
                                    AWAY_TEAM == OPPONENT ~ paste("vs", OPPONENT)),
               Team = TEAM,
               Opponent = Opponent,
               `Pr(Win)` = round(PROB, 3), 
               Result = paste(POINTS, "-", OPPONENT_POINTS, " ", WINNER, sep="")) %>%
        arrange(DIFF) %>%
        mutate(Rank = row_number()) %>%
        select(Rank, Season, Date, Team, Opponent, `Pr(Win)`, Result) %>%
        head(25) %>%
        flextable() %>%
        autofit() %>%
        flextable::align(., j = c("Rank", "Season", "Date", "Pr(Win)", "Result"),
              align = 'center', part = 'all')

```


```{r elo swings tamu}

elo_games$team_outcomes %>%
        mutate(ELO_DIFF = POSTGAME_ELO - PREGAME_ELO) %>%
        filter(TEAM == 'Texas A&M') %>%
        arrange(ELO_DIFF) %>%
        left_join(., elo_games$game_outcomes %>% 
                          select(GAME_ID, HOME_TEAM, AWAY_TEAM, HOME_POINTS, AWAY_POINTS, HOME_PROB, AWAY_PROB),
                  by = c("GAME_ID")) %>%
        mutate(PROB = case_when(TEAM == HOME_TEAM ~ HOME_PROB,
                                TEAM == AWAY_TEAM ~ AWAY_PROB)) %>%
        mutate(WIN = case_when(OUTCOME == 'win' ~ 1,
                               OUTCOME == 'tie' ~ 0)) %>%
        mutate(DIFF = WIN - PROB) %>%
        mutate(WINNER = case_when(OUTCOME == 'win' ~ TEAM,
                                  OUTCOME == 'loss' ~ OPPONENT)) %>%
        mutate(Season = factor(SEASON),
               Date = GAME_DATE,
               Opponent = case_when(HOME_TEAM == OPPONENT ~ paste("@", OPPONENT),
                                    AWAY_TEAM == OPPONENT ~ paste("vs", OPPONENT)),
               Team = TEAM,
               Pre_Elo = PREGAME_ELO,
               Post_Elo = POSTGAME_ELO,
               Diff = ELO_DIFF,
               Opponent = Opponent,
               `Pr(Win)` = round(PROB, 3), 
               Result = paste(POINTS, "-", OPPONENT_POINTS, " ", WINNER, sep="")) %>%
        mutate_at(c("Pre_Elo",
                    "Post_Elo", 
                    "Diff"),
                  round, 0) %>%
        arrange(desc(ELO_DIFF)) %>%
        head(25) %>%
        select(Season, Date, Team, Opponent, Pre_Elo, Post_Elo, Result, Diff) %>%
        flextable() %>%
        autofit() %>%
        flextable::align(., j = c("Season", "Date", "Result", "Diff"),
              align = 'center', part = 'all') %>%
        bg(., j = c("Diff"),
           bg = diff_func)
        

```

## James Madison

```{r plot historical elo for jmu}

plot_elo_historical_team(elo_games$team_outcomes,
                         min_year = 1970,
                         team = 'James Madison')

```

