---
title: "Simulating CFB Seasons"
author: "Phil Henrickson"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE #adds a Table of Contents
    theme: cerulean
    number_sections: TRUE #number your headings/sections
    toc_float: TRUE #let your ToC follow you as you scroll
    keep_md: no
    fig.caption: yes
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = F,
                      error = F,
                      warning=F,
                      dev="png",
                      fig.width = 10,
                      fig.height = 6)

options(knitr.duplicate.label = "allow")

options(scipen=999)

```

```{r connect to snowflake}

library(DBI)
library(odbc)
library(RODBC)
library(keyring)

# connect to snowflake
myconn <- DBI::dbConnect(odbc::odbc(),
                         "SnowflakeDSII",
                         Database = "CFB_DEMO",
                         warehouse = "DEMO_WH",
                         uid="phil.henrickson",
                         pwd=keyring::key_get("AE_Snowflake"))

```

```{r packages, include=F} 

source(here::here("scripts/load_packages.R"))
library(rstan)
library(rstanarm)
library(rstantools)
library(jsonlite)
library(forcats)
library(teamcolors)
library(shiny)
conflict_prefer("lag", "dplyr")

```

```{r flextable settings}

library(flextable)
set_flextable_defaults(theme_fun = theme_alafoli,
                       font.color = "grey10",
                       font.size=8,
                       padding.bottom = 6, 
                       padding.top = 6,
                       padding.left = 6,
                       padding.right = 6,
                       background.color = "white")

```

```{r load functions}

# integer plotting function
int_breaks <- function(x, n = 5) {
  l <- pretty(x, n)
  l[abs(l %% 1) < .Machine$double.eps ^ 0.5] 
}

source(here::here("functions/get_expected_score.R"))
source(here::here("functions/get_new_elos.R"))
source(here::here("functions/calc_elo_ratings.R"))
source(here::here("functions", "simulateX.R"))
source(here::here("functions", "sim_game_margin.R"))
source(here::here("functions", "sim_elo_ratings.R"))

# with recruiting
source(here::here("functions", "calc_elo_ratings_with_recruiting.R"))
source(here::here("functions", "sim_elo_ratings_with_recruiting.R"))
source(here::here("functions", "sim_rest_of_season.R"))

rm(a)

```

```{r get the games data}

# data on teams with logos and colors
teams_raw = bind_rows(
        DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.TEAMS')) %>%
        as_tibble())  %>%
        rename(TEAM = SCHOOL,
               TEAM_ID = ID,
               TEAM_MASCOT = MASCOT,
               TEAM_ABBREVIATION = ABBREVIATION)

# get games
games_raw = DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.ANALYSIS_GAMES')) %>%
        as_tibble() %>%
        mutate(CONFERENCE_GAME = case_when(CONFERENCE_GAME == 'true' ~ T,
                                           CONFERENCE_GAME == 'false' ~ F)) %>%
        mutate(GAME_DATE = as.Date(START_DATE)) %>%
        mutate(START_DATE = as_datetime(START_DATE)) %>%
        arrange(START_DATE) %>%
        group_by(SEASON, SEASON_TYPE, HOME_CONFERENCE) %>%
        mutate(LAST_GAME = START_DATE == max(START_DATE)) %>% 
        ungroup() %>%
        # logic for flagging conference championship games
        mutate(CONFERENCE_CHAMPIONSHIP = case_when(CONFERENCE_GAME == T & 
                                                           (HOME_CONFERENCE != 'FBS Independents') & 
                                                           (HOME_CONFERENCE != 'FCS Independents') &
                                                           !(HOME_CONFERENCE %in% 'Big Ten' & SEASON < 2014) &
                                                           !(HOME_CONFERENCE %in% 'Western Athletic') &
                                                           (NEUTRAL_SITE == T | 
                                                                    is.na(VENUE) |
                                                                    VENUE == 'Glass Bowl' |
                                                                    VENUE == 'Bright House Networks Stadium' |
                                                                    VENUE == 'Alamodome' |
                                                                    VENUE == 'Georgia Dome' |
                                                                    HOME_CONFERENCE == 'Sun Belt' |
                                                                    HOME_CONFERENCE == 'American Athletic') &
                                                           SEASON > 2000 &
                                                           SEASON_TYPE == 'regular' &
                                                           WEEK > 13 &
                                                           LAST_GAME == T ~ T,
                                                   TRUE ~ F)) %>%
        mutate(HOME_WIN = factor(case_when(HOME_POINTS > AWAY_POINTS ~ 'yes',
                                    HOME_POINTS < AWAY_POINTS ~ 'no',
                                    HOME_POINTS == AWAY_POINTS ~ 'no'),
               levels = c("no", "yes"))) %>%
        mutate(HOME_DIVISION  = replace_na(HOME_DIVISION, "missing"),
               AWAY_DIVISION = replace_na(AWAY_DIVISION, "missing")) %>%
        left_join(., teams_raw %>%
                          select(TEAM, TEAM_ABBREVIATION) %>%
                          rename(HOME_TEAM = TEAM,
                                 HOME_TEAM_ABBR = TEAM_ABBREVIATION),
                  by = c("HOME_TEAM")) %>%
        left_join(., teams_raw %>%
                          select(TEAM, TEAM_ABBREVIATION) %>%
                          rename(AWAY_TEAM = TEAM,
                                 AWAY_TEAM_ABBR = TEAM_ABBREVIATION),
                  by = c("AWAY_TEAM"))

# make a table of team conference by season
team_conference_season = games_raw %>% 
        select(SEASON, HOME_TEAM, HOME_CONFERENCE, HOME_DIVISION) %>% 
        rename(TEAM = HOME_TEAM, 
               CONFERENCE = HOME_CONFERENCE,
               DIVISION = HOME_DIVISION) %>% 
        bind_rows(.,
            games_raw %>%
                    select(SEASON, AWAY_TEAM, AWAY_CONFERENCE, AWAY_DIVISION) %>% 
                    rename(TEAM = AWAY_TEAM, 
                           CONFERENCE = AWAY_CONFERENCE,
                           DIVISION = AWAY_DIVISION)) %>%
        select(SEASON, TEAM, CONFERENCE, DIVISION) %>%
        unique()

```

```{r teamcolors}

teamcolors = teamcolors::teamcolors

# # load teamcolors
# teamcolors = teamcolors %>%
#         mutate(location = case_when(location == 'Ole' ~ name,
#                                     location == 'Southern California Trojans' ~ 'USC',
#                                     location == 'Miami' & name == 'Miami (OH)' ~ name,
#                                     TRUE ~ location)) %>%
#         
#         bind_rows(., 
#                   teams_raw %>%
#                           filter(TEAM == 'SMU' | TEAM == 'USC') %>%
#                           mutate(name = TEAM,
#                                  location = TEAM) %>%
#                           rename(primary = COLOR,
#                                  secondary = ALT_COLOR))

# get teamcolors
teamcolors = teams_raw %>%
        select(TEAM, COLOR, ALT_COLOR) %>%
        left_join(.,
                  teamcolors %>%
                          filter(league == 'ncaa') %>%
                          mutate(location = case_when(location == 'Ole' ~ name,
                                    location == 'Southern California Trojans' ~ 'USC',
                                    location == 'Miami' & name == 'Miami (OH)' ~ name,
                                    TRUE ~ location)) %>%
                          mutate(TEAM = location),
                  by = c("TEAM")) %>%
        mutate(primary = case_when(is.na(primary) ~ COLOR,
                                   TRUE ~ primary)) %>%
        mutate(secondary = case_when(is.na(secondary) ~ COLOR,
                                   TRUE ~ secondary))

```


```{r get recruiting data}

# recruiting
recruiting_teams_raw = DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.RECRUITING_TEAMS')) %>%
        as_tibble() %>%
        mutate(POINTS = as.numeric(POINTS)) %>%
        rename(RECRUITING_POINTS = POINTS)

# expand
library(TTR)
recruiting_teams = expand.grid(TEAM = unique(recruiting_teams_raw$TEAM),
             YEAR = seq(min(recruiting_teams_raw$YEAR),
                        max(recruiting_teams_raw$YEAR))) %>%
        left_join(., recruiting_teams_raw,
                  by = c("YEAR", "TEAM")) %>%
        mutate(RECRUITING_POINTS = replace_na(RECRUITING_POINTS, 0)) %>%
        filter(YEAR > 2001) %>%
        left_join(., team_conference_season,
                  by = c("YEAR"="SEASON", "TEAM")) %>%
        filter(DIVISION == 'fbs') %>%
        group_by(TEAM) %>%
        mutate(n = n()) %>%
        filter(n > 4) %>%
        group_by(TEAM) %>%
        select(-n) %>%
        mutate(EMA = EMA(RECRUITING_POINTS, n =4)) %>%
        # now fill with actual in the event that a team doesn't have yet have enough data
        mutate(EMA = case_when(is.na(EMA) & !is.na(RECRUITING_POINTS) ~ RECRUITING_POINTS,
                                  TRUE ~ EMA)) %>%
        # mutate_at(c("EMA", "WMA"),
        #           replace_na, 0) %>%
        arrange(TEAM, YEAR)  %>%
        mutate(SEASON = YEAR) %>%
        ungroup() %>%
        select(TEAM, SEASON, CONFERENCE, DIVISION, RANK, RECRUITING_POINTS, EMA)

# get what we need
recruiting = recruiting_teams %>%
        select(TEAM, SEASON, CONFERENCE, RECRUITING_POINTS, EMA) %>%
        rename(RECRUITING_SCORE = EMA)

```

```{r join recruiting with games}

# specify games to run function
games = games_raw %>%
        filter(SEASON >=1970) %>%
        arrange(GAME_DATE) %>%
        filter(!(SEASON < 2022 & is.na(HOME_POINTS) & is.na(AWAY_POINTS))) %>%
      #  filter(!is.na(HOME_POINTS) & !is.na(AWAY_POINTS)) %>%
        select(GAME_ID, 
               SEASON, 
               WEEK,
               SEASON_TYPE,
               START_DATE,
               GAME_DATE,
               NEUTRAL_SITE, 
               CONFERENCE_GAME,
               CONFERENCE_CHAMPIONSHIP,
               HOME_TEAM,
               AWAY_TEAM,
               HOME_CONFERENCE,
               AWAY_CONFERENCE,
               HOME_DIVISION,
               AWAY_DIVISION,
               HOME_POINTS,
               AWAY_POINTS) %>%
        left_join(.,
                  recruiting %>%
                          select(SEASON,
                                 TEAM,
                                 RECRUITING_SCORE) %>%
                          rename(HOME_TEAM = TEAM,
                                 HOME_RECRUITING_SCORE = RECRUITING_SCORE),
                  by = c("SEASON",
                         "HOME_TEAM")) %>%
        left_join(.,
                  recruiting %>%
                          select(SEASON,
                                 TEAM,
                                 RECRUITING_SCORE) %>%
                          rename(AWAY_TEAM = TEAM,
                                 AWAY_RECRUITING_SCORE = RECRUITING_SCORE),
                  by = c("SEASON",
                         "AWAY_TEAM"))

```

```{r set future plan, include=F}

library(future.apply)
plan(multisession, workers = 7)

```

# Simulating Seasons

```{r set pars for simulating and get initial elo, include=F}

# initial season
initial_season = 2022

# set initial games
initial_games = games %>%
        filter(SEASON < initial_season)

# set pars
elo_pars = tibble(k = 35,
                       v = 400,
                       reversion = 0.1,
                       home_field_advantage = 75,
                       recruiting_weight = 0.5)

# now get the elo ratings based on this training set
elo_initial = calc_elo_ratings_with_recruiting(
        games = initial_games,
        recruiting_weight = elo_pars$recruiting_weight,
        home_field_advantage = elo_pars$home_field_advantage,
        reversion = elo_pars$reversion,
        k = elo_pars$k,
        v = elo_pars$v,
        verbose = T)

# train points model
points_model = elo_initial$game_outcomes %>%
        filter(SEASON >=1975) %>% 
        filter(SEASON < initial_season) %>%
                mutate(HOME_ELO_DIFF = case_when(NEUTRAL_SITE == F ~ 
                                                         HOME_PREGAME_ELO + 
                                                         elo_pars$home_field_advantage -
                                                         AWAY_PREGAME_ELO,
                                                 TRUE ~ HOME_PREGAME_ELO - AWAY_PREGAME_ELO)) %>%
                mutate(HOME_SCORE_DIFF = HOME_POINTS-AWAY_POINTS) %>%
                nest() %>%
                # now fit linear models, i'll do so using stan
                mutate(lm = map(data, ~ lm(HOME_SCORE_DIFF ~ HOME_ELO_DIFF,
                                     data = .x))) %>%
                pluck("lm",1)

```

Simulate the rest of each season, week after week.

```{r simulate each week of each season}

season_simulations = sim_rest_of_season(games,
                                        season = initial_season,
                                        elo_initial = elo_initial,
                                        elo_pars = elo_pars,
                                        nsims = 1000,
                                        sim_start_week = 0,
                                        season_start_only = T,
                                        aggregate = F,
                                        points_model = points_model)

```

```{r get result}

source(here::here("functions", "table_wins_season.R"))

# run through
table_wins_season(sim_team_outcomes = season_simulations$team_outcomes %>%
                          filter(CONFERENCE == 'SEC'),
        games = games_raw,
        season = 2022)
        
# season_simulations$team_outcomes %>%
#         filter(DIVISION == 'fbs') %>%
#         group_by(.id, SIM_FROM_WEEK, SEASON, TEAM, SIM_OUTCOME) %>%
#         count() %>% 
#         group_by(SEASON, TEAM, SIM_OUTCOME) %>% 
#         summarize(mean = mean(n),
#                   .groups = 'drop') %>%
#         spread(SIM_OUTCOME, mean) %>% 
#         select(SEASON, TEAM, win, loss) %>% 
#         ungroup() %>% 
#         arrange(desc(win)) %>%
#         mutate_if(is.numeric, round, 2) %>%
#         mutate_if(is.numeric, replace_na, 0)

```

season_simulations$team_outcomes %>%
        filter(TEAM == 'Georgia')
        # mutate(win = case_when(SIM_OUTCOME == 'win' ~ 1,
        #                        TRUE ~ 0)) %>%
        # group_by(SEASON, GAME_DATE, TEAM, OPPONENT) %>%
        # summarize(Win_Prob = sum(win),
        #           Margin = median(SIM_MARGIN)) 
        # filter(DIVISION == 'fbs') %>%
        # group_by(.id, SIM_FROM_WEEK, SEASON, TEAM, SIM_OUTCOME) %>%
        # count() %>% 
        # group_by(SEASON, TEAM, SIM_OUTCOME) %>% 
        # summarize(mean = mean(n),
        #           .groups = 'drop') %>%
        # spread(SIM_OUTCOME, mean) %>% 
        # select(SEASON, TEAM, win, loss) %>% 
        # ungroup() %>% 
        # arrange(desc(win)) %>%
        # mutate_if(is.numeric, round, 2)

```{r table wins}

source(here::here("functions", "table_wins_team_season.R"))

table_wins_team_season(season_simulations$team_outcomes,
                       games_raw,
                       'Florida State',
                       season = 2022)

```

                        
                        
```{r examine result}

season_simulations %$% game_outcomes %>%
        filter(SIM_FROM_WEEK == 0) %>%
        filter(HOME_TEAM == 'Florida' & AWAY_TEAM == 'Florida State') %>%
        select(GAME_ID, SIM_FROM_WEEK, HOME_TEAM, AWAY_TEAM, HOME_SIM_PREGAME_ELO, AWAY_SIM_PREGAME_ELO, HOME_SIM_PROB, HOME_SIM_MARGIN)

season_simulations %$% game_outcomes %>%
        filter(GAME_ID == 213210057)  %>%
        select(SIM_FROM_WEEK, WEEK, HOME_TEAM, AWAY_TEAM, HOME_SIM_PREGAME_ELO, AWAY_SIM_PREGAME_ELO, HOME_SIM_PROB, HOME_SIM_MARGIN) %>%
        mutate_at(c("HOME_SIM_PREGAME_ELO",
                    "AWAY_SIM_PREGAME_ELO"),
                  round, 0) %>%
        flextable() %>%
        autofit()

```


