---
title: "Simulating CFB Seasons"
author: "Phil Henrickson"
date: "7/7/2022"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = F,
                      error = F,
                      warning=F,
                      dev="png",
                      fig.width = 10,
                      fig.height = 8)

options(knitr.duplicate.label = "allow")

options(scipen=999)


```

```{r connect to snowflake}

library(DBI)
library(odbc)
library(RODBC)
library(keyring)

# connect to snowflake
myconn <- DBI::dbConnect(odbc::odbc(),
                         "SnowflakeDSII",
                         Database = "CFB_DEMO",
                         warehouse = "DEMO_WH",
                         uid="phil.henrickson",
                         pwd=keyring::key_get("AE_Snowflake"))

```

```{r packages, include=F} 

source(here::here("scripts/load_packages.R"))
library(rstan)
library(rstanarm)
library(rstantools)
library(jsonlite)
library(forcats)
library(teamcolors)
conflict_prefer("lag", "dplyr")

# load teamcolors
teamcolors = teamcolors %>%
        mutate(location = case_when(location == 'Ole' ~ name,
                                    location == 'Miami' & name == 'Miami (OH)' ~ name,
                                    TRUE ~ location))

```

```{r get colors for ncaa teams}

# look at colors for ncca
ncaa_colors = teamcolors %>%
        filter(league == 'ncaa') 

# define palette for ncaa
#league_pal("ncaa", 1)

```


```{r flextable settings}

set_flextable_defaults(theme_fun = theme_alafoli,
                       font.color = "grey10",
                       font.size=8,
                       padding.bottom = 6, 
                       padding.top = 6,
                       padding.left = 6,
                       padding.right = 6,
                       background.color = "white")

```

```{r load functions}

# integer plotting function
int_breaks <- function(x, n = 5) {
  l <- pretty(x, n)
  l[abs(l %% 1) < .Machine$double.eps ^ 0.5] 
}

source(here::here("functions/theme_phil.R"))
rm(a)

```

```{r load and tidy games}

# get games
games_data_raw = DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.GAMES')) %>%
        as_tibble() %>%
        mutate(CONFERENCE_GAME = case_when(CONFERENCE_GAME == 'true' ~ T,
                                           CONFERENCE_GAME == 'false' ~ F)) %>%
        # remove extraneous conferences
        mutate(ID = as.numeric(ID)) %>%
        rename(GAME_ID = ID) %>%
        mutate(GAME_DATE = as.Date(START_DATE))

# get list of conferences that only appear in 2022
fbs_conferences = c(
        games_data_raw %>% 
                filter(SEASON <=2022) %>%
                group_by(SEASON, HOME_CONFERENCE) %>%
                count() %>% group_by(HOME_CONFERENCE) %>%
                mutate(seasons = n_distinct(SEASON)) %>%
                ungroup() %>%
                filter(seasons != 1) %>% 
                filter(!is.na(HOME_CONFERENCE)) %>%
                pull(HOME_CONFERENCE), 
        games_data_raw %>% 
                filter(SEASON <=2022) %>%
                group_by(SEASON, AWAY_CONFERENCE) %>%
                count() %>% group_by(AWAY_CONFERENCE) %>%
                mutate(seasons = n_distinct(SEASON)) %>%
                ungroup() %>%
                filter(seasons != 1) %>%
                filter(!is.na(AWAY_CONFERENCE)) %>%
                pull(AWAY_CONFERENCE)) %>%
        unique()

# list of new conferences
fcs_conferences = c(games_data_raw %>% 
                filter(SEASON <=2022) %>%
                group_by(SEASON, HOME_CONFERENCE) %>%
                count() %>% group_by(HOME_CONFERENCE) %>%
                mutate(seasons = n_distinct(SEASON)) %>%
                ungroup() %>%
                filter(seasons == 1) %>%
                pull(HOME_CONFERENCE), 
        games_data_raw %>% 
                filter(SEASON <=2022) %>%
                group_by(SEASON, AWAY_CONFERENCE) %>%
                count() %>% group_by(AWAY_CONFERENCE) %>%
                mutate(seasons = n_distinct(SEASON)) %>%
                ungroup() %>%
                filter(seasons == 1) %>%
                pull(AWAY_CONFERENCE)) %>%
        unique()

# add flag for fbs game
games_data_tidied = games_data_raw %>%
        mutate(GAME_DATE = as.Date(START_DATE)) %>%
        mutate(GAME_MONTH = lubridate::month(GAME_DATE)) %>%
        arrange(GAME_DATE) %>%
        # group_by(SEASON, HOME_CONFERENCE) %>%
        # mutate(CONFERENCE_LAST_WEEK = max(WEEK)) %>%
        # ungroup() %>%
        # mutate(CONFERENCE_CHAMPIONSHIP = case_when(WEEK == LAST_WEEK 
        #                                                    CONFERENCE_GAME == T
        mutate(CONFERENCE_CHAMPIONSHIP = case_when(CONFERENCE_GAME == T &
                                                           (HOME_CONFERENCE != 'FBS Independents') &
                                                           NEUTRAL_SITE == T &
                                                           WEEK > 11 ~ T, TRUE ~ F)) %>%
        mutate(FBS_GAME = (HOME_CONFERENCE%in% fbs_conferences | AWAY_CONFERENCE %in% fbs_conferences) | SEASON <= 1954) %>%
        # filter(!(HOME_CONFERENCE %in% fcs_conferences)) %>%
        # filter(!(AWAY_CONFERENCE %in% fcs_conferences)) %>%
        mutate(HOME_CONFERENCE = case_when(HOME_CONFERENCE %in% fcs_conferences ~ NA_character_,
                                           TRUE ~ HOME_CONFERENCE)) %>%
        mutate(AWAY_CONFERENCE = case_when(HOME_CONFERENCE %in% fcs_conferences ~ NA_character_,
                                           TRUE ~ AWAY_CONFERENCE))

# flag conference championships
games_data_tidied %>%
        filter(SEASON > 1998 & SEASON < 2022) %>%
        filter(!((HOME_CONFERENCE %in% 'Big Ten') & SEASON < 2011)) %>%
        group_by(SEASON, HOME_CONFERENCE) %>%
        mutate(CONFERENCE_LAST_WEEK = max(WEEK)) %>%
        ungroup() %>%
        mutate(CONFERENCE_CHAMPIONSHIP = case_when(CONFERENCE_GAME == T &
                                                           WEEK == CONFERENCE_LAST_WEEK &
                                                           (HOME_CONFERENCE != 'FBS Independents') &
                                                           (NEUTRAL_SITE == T | (is.na(VENUE_ID) & SEASON > 2000)) & 
                                                           WEEK > 11 ~ T, TRUE ~ F)) %>%
        filter(CONFERENCE_CHAMPIONSHIP == T) %>%
        rename(CONFERENCE = HOME_CONFERENCE) %>%
        select(GAME_ID, SEASON, WEEK, CONFERENCE_GAME, CONFERENCE, HOME_TEAM, AWAY_TEAM)

```


```{r get games}

# recruiting
recruiting_teams_raw = DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.RECRUITING_TEAMS')) %>%
        as_tibble() %>%
        mutate(POINTS = as.numeric(POINTS)) %>%
        rename(RECRUITING_POINTS = POINTS)


# make a table of team conference by season
team_conference_season = games_data_raw %>% 
  select(SEASON, HOME_TEAM, HOME_CONFERENCE) %>% 
  rename(TEAM = HOME_TEAM, CONFERENCE = HOME_CONFERENCE) %>% 
  bind_rows(.,
            games_data_raw %>%
                    select(SEASON, AWAY_TEAM, AWAY_CONFERENCE) %>% 
                    rename(TEAM = AWAY_TEAM, CONFERENCE = AWAY_CONFERENCE)) %>%
  select(SEASON, TEAM, CONFERENCE) %>%
  unique()

# now make our recruiting table
recruiting = recruiting_teams_raw %>%
        filter(YEAR > 2001) %>%
        left_join(., team_conference_season,
                  by = c("YEAR"="SEASON", "TEAM")) %>%
        # remove fcs teams
        filter(!is.na(CONFERENCE)) %>%
        # remove teams for which we have less than 4 years data
        group_by(TEAM) %>%
        mutate(n = n()) %>%
        filter(n > 4) %>%
        select(-n) %>%
        # compute ema over a 4 year window
        mutate(RECRUITING_EMA= EMA(RECRUITING_POINTS, n =4)) %>%
        # if a team is missing over this time period but it is obdserved, then fill it in with their actual value
        mutate(RECRUITING_EMA = case_when(is.na(RECRUITING_EMA) & !is.na(RECRUITING_POINTS) ~ RECRUITING_POINTS,
                                  TRUE ~ RECRUITING_EMA)) %>%
        arrange(TEAM, YEAR) %>%
        # now make our score lag by a year; i'll accomplish this by making SEASON = YEAR + 1
        group_by(TEAM) %>%
        # lag recruiting by a year
        mutate(SEASON = dplyr::lead(YEAR, 1)) %>%
        # now fill in the missing seasons, as this will be the year +1
        mutate(SEASON = case_when(is.na(SEASON) ~ dplyr::lag(SEASON+1, 1),
                                  TRUE ~ SEASON)) %>%
        # this was such a dumb way to do this just to avoid having to type "SEASON" = "YEAR"
        ungroup() %>%
        # now normalize by season, so recruiting score is how far above or below average a team is going into that year
        group_by(SEASON) %>%
        mutate(RECRUITING_SCORE = RECRUITING_EMA - mean(RECRUITING_EMA)) %>%
        ungroup()

```

```{r rm what we dont need}

rm(games_data_raw,
   recruiting_teams_raw)

```



```{r get elo functions}

source(here::here("functions/get_expected_score.R"))
source(here::here("functions/get_new_elos.R"))
source(here::here("functions/calc_elo_ratings.R"))
source(here::here("functions", "simulateX.R"))
source(here::here("functions", "sim_game_margin.R"))
source(here::here("functions", "sim_elo_ratings.R"))

# with recruiting
source(here::here("functions", "calc_elo_ratings_with_recruiting.R"))
source(here::here("functions", "sim_elo_ratings_with_recruiting.R"))


```

# Simulating Regular Season Games

I'll now do this for every season through 2020. This means I'll compute Elo ratings in the run up to each season, then simulate the season based on the ratings going into the season.

```{r initialize elo}

# set of games on which to learn baseline elo
initial_games = games_data_tidied %>%
                filter(SEASON >=1970) %>%
                filter(SEASON < 2000) %>%
                arrange(GAME_DATE) %>%
                select(GAME_ID, 
                       SEASON, 
                       WEEK,
                       SEASON_TYPE,
                       GAME_DATE,
                       NEUTRAL_SITE, 
                       HOME_TEAM,
                       AWAY_TEAM,
                       HOME_CONFERENCE,
                       AWAY_CONFERENCE,
                       HOME_POINTS,
                       AWAY_POINTS)

# set pars
selected_pars = tibble(reversion = .1,
                       home_field_advantage = 60,
                       #recruiting_adjustment = .5,
                       k = 35,
                       v = 400)

# now get the elo ratings based on this training set
elo_initial = calc_elo_ratings(
        games = initial_games,
        home_field_advantage = selected_pars$home_field_advantage,
        reversion = selected_pars$reversion,
        k = selected_pars$k,
        v = selected_pars$v,
        verbose = T)

```

```{r set parallel}

library(future.apply)
plan(multisession, workers = 7)

```


```{r look at performance through 2022, include=F, eval=T}

rm(sim_seasons,
   train_games,
   valid_games,
   elo_train,
   elo_valid)

# run 
seasons = seq(2001, 2022, by = 1)

# loop over season
simulated_seasons = foreach(i = 1:length(seasons)) %do% {
        
        # get elo ratings in the run up to a season
        # filter to games that occured 
        train_games = games_data_tidied %>%
                filter(SEASON > max(initial_games$SEASON)) %>%
                filter(SEASON < seasons[i]) %>%
                arrange(GAME_DATE) %>%
                select(GAME_ID, 
                       SEASON, 
                       WEEK,
                       SEASON_TYPE,
                       GAME_DATE,
                       NEUTRAL_SITE, 
                       HOME_TEAM,
                       AWAY_TEAM,
                       HOME_CONFERENCE,
                       AWAY_CONFERENCE,
                       HOME_POINTS,
                       AWAY_POINTS)
        
        # valid
        sim_games = games_data_tidied %>%
                filter(SEASON == seasons[i]) %>%
                filter(SEASON <= max(seasons)) %>%
                arrange(GAME_DATE) %>%
                select(GAME_ID, 
                       SEASON, 
                       WEEK,
                       SEASON_TYPE,
                       GAME_DATE,
                       NEUTRAL_SITE, 
                       HOME_TEAM,
                       AWAY_TEAM,
                       HOME_CONFERENCE,
                       AWAY_CONFERENCE,
                       HOME_POINTS,
                       AWAY_POINTS)
        
        # now get the elo ratings based on this training set
        elo_train = calc_elo_ratings(train_games,
                                     teams = elo_initial$teams,
                                     team_seasons = elo_initial$team_seasons,
                          home_field_advantage = selected_pars$home_field_advantage,
                          reversion = selected_pars$reversion,
                          k = selected_pars$k,
                          v = selected_pars$v,
                          verbose = F)
        
        # regress point differential on elo differential
        # fit linear model after first few years in training
        points_model = bind_rows(elo_initial$game_outcomes,
                                 elo_train$game_outcomes) %>%
                filter(SEASON >=1975) %>%
                mutate(HOME_ELO_DIFF = case_when(NEUTRAL_SITE == F ~ HOME_PREGAME_ELO + selected_pars$home_field_advantage - AWAY_PREGAME_ELO,
                                                 TRUE ~ HOME_PREGAME_ELO - AWAY_PREGAME_ELO)) %>%
                mutate(HOME_SCORE_DIFF = HOME_POINTS-AWAY_POINTS) %>%
                nest() %>%
                # now fit linear models, i'll do so using stan
                mutate(lm = map(data, ~ lm(HOME_SCORE_DIFF ~ HOME_ELO_DIFF,
                                     data = .x))) %>%
                mutate(tidied = map(lm, tidy, conf.int=T)) %>%
                mutate(glanced = map(lm, glance)) %>%
                pluck("lm",1)
        
        # then, simulate the next season
        set.seed(1999)
        sim_seasons = future_replicate(1000,
                          sim_elo_ratings(sim_games,
                                          teams = elo_train$teams,
                                          team_seasons = elo_train$team_seasons,
                                          home_field_advantage = selected_pars$home_field_advantage,
                                          reversion = selected_pars$reversion,
                                          k = selected_pars$k,
                                          v = selected_pars$v,
                                          verbose=F,
                                          ties =F,
                                          points_model = points_model))
        
        # grab what we need
        # out = list("sim_team_outcomes" = rbindlist(sim_seasons['team_outcomes',],
        #                               idcol = T),
        #            "sim_game_outcomes" = rbindlist(sim_seasons['game_outcomes',],
        #                               idcol = T))
        
        # get the simulated game outcomes
        out = rbindlist(sim_seasons['game_outcomes',],
                                      idcol = T) %>%
                mutate(SIMULATION_FROM_SEASON = seasons[i],
                       SIMULATION_FROM_WEEK = 0) %>%
                mutate(home_field_advantage = selected_pars$home_field_advantage,
                       reversion = selected_pars$reversion,
                       k = selected_pars$k,
                       v = selected_pars$v)
        
        rm(sim_seasons)
        
        # track progress
        cat("\r", i, "of", length(seasons), "seasons completed");  flush.console()
        
        # return output
        out

}

names(simulated_seasons) = seasons

```


```{r save each season, eval=F}

for(i in 1:length(seasons)) {
        
        season = names(simulated_seasons)[i]
        sims = simulated_seasons[i]
        simulation_from = 0
        save(sims,
             file = here::here("simulations", paste("sims_", season, "_week_", simulation_from, ".Rdata", sep="")))
        
}

```

```{r flip to team out comes}

sim_game_outcomes = rbindlist(simulated_seasons)

sim_team_outcomes = sim_game_outcomes %>% 
        mutate(HOME_OPPONENT = AWAY_TEAM) %>%
        select(.id, 
               GAME_ID, 
               SEASON,
               WEEK,
               GAME_DATE,
               starts_with("HOME_")) %>%
        set_names(., gsub("HOME_", "", names(.))) %>%
        bind_rows(.,
                  sim_game_outcomes %>% 
                          mutate(AWAY_OPPONENT = HOME_TEAM) %>%
                          mutate(AWAY_MARGIN = -HOME_MARGIN) %>%
                          select(.id, GAME_ID, 
                                 SEASON,
                                 WEEK,
                                 GAME_DATE,
                                 starts_with("AWAY_")) %>%
                          set_names(., gsub("AWAY_", "", names(.)))) %>%
        arrange(GAME_DATE) %>%
        mutate(home_field_advantage = selected_pars$home_field_advantage,
               reversion = selected_pars$reversion,
               k = selected_pars$k,
               v = selected_pars$v)

```


```{r load season sims}

# load(file = here::here("data/simulated_seasons.Rdata"))
# 
# # get the games
# sim_game_outcomes = rbindlist(lapply(simulated_seasons,'[[','sim_game_outcomes'))
# 
# # get by team
# sim_team_outcomes = rbindlist(lapply(simulated_seasons,'[[','sim_team_outcomes'))
# 
# rm(simulated_seasons)

```

I'll get season by season accuracy/log-loss.

```{r simulating all seasons}

assess_metrics = metric_set(yardstick::mn_log_loss,
                           yardstick::accuracy)

overall_assess = sim_game_outcomes %>%
        filter(SEASON < 2022) %>%
        mutate(HOME_PRED = case_when(HOME_MARGIN >0 ~ 1,
                                     TRUE ~ 0)) %>%
        group_by(GAME_ID, SEASON, WEEK, GAME_DATE, NEUTRAL_SITE, HOME_TEAM, AWAY_TEAM, HOME_POINTS, AWAY_POINTS) %>%
        summarize(HOME_PROB = sum(HOME_PRED) / n(),
                  PRED_MARGIN = mean(HOME_MARGIN),
                  .groups = 'drop') %>%
        mutate(AWAY_PROB = 1-HOME_PROB) %>%
        mutate(HOME_WIN = factor(case_when(HOME_POINTS > AWAY_POINTS  ~ 'yes',
                                       TRUE ~'no'))) %>%
        mutate(HOME_PRED = factor(case_when(HOME_PROB >=.5 ~ 'yes',
                                       TRUE ~ 'no'))) %>%
        mutate(PRED_CORRECT = case_when(HOME_WIN == HOME_PRED ~ 'yes',
                                        TRUE ~ 'no')) %>%
        mutate(yes = HOME_PROB) %>%
        group_by(SEASON) %>%
        mutate(games = n()) %>%
        group_by(SEASON, games) %>%
        assess_metrics(truth = HOME_WIN,
                       yes,
                       estimate = HOME_PRED,
                       event_level = 'second')

#
overall_assess %>%
        ggplot(., aes(x=SEASON,
                      y=.estimate))+
        geom_line()+
        geom_point()+
        facet_wrap(.metric ~.,
                   scales = "free_y",
                   ncol = 1)+
        theme_bw()+
        ggtitle("Assessing Elo Simulations by Season")

```

Interesting - I can understand not doing so well on the 2020 season, but 2016 was evidently also a tough one to predict too. Generally speaking, our preseason accuracy hovers around 70% and our log loss is around .55.

I'll also look at this by week for each season.

```{r performance by week of season}

sim_game_outcomes %>%
        filter(SEASON < 2022) %>%
        mutate(HOME_PRED = case_when(HOME_MARGIN >0 ~ 1,
                                     TRUE ~ 0)) %>%
        group_by(GAME_ID, SEASON, WEEK, GAME_DATE, NEUTRAL_SITE, HOME_TEAM, AWAY_TEAM, HOME_POINTS, AWAY_POINTS) %>%
        summarize(HOME_PROB = sum(HOME_PRED) / n(),
                  PRED_MARGIN = mean(HOME_MARGIN),
                  .groups = 'drop') %>%
        mutate(AWAY_PROB = 1-HOME_PROB) %>%
        mutate(HOME_WIN = factor(case_when(HOME_POINTS > AWAY_POINTS  ~ 'yes',
                                       TRUE ~'no'))) %>%
        mutate(HOME_PRED = factor(case_when(HOME_PROB >=.5 ~ 'yes',
                                       TRUE ~ 'no'))) %>%
        mutate(PRED_CORRECT = case_when(HOME_WIN == HOME_PRED ~ 'yes',
                                        TRUE ~ 'no')) %>%
        mutate(yes = HOME_PROB) %>%
        group_by(SEASON, WEEK) %>%
        mutate(games = n()) %>%
        group_by(SEASON, WEEK, games) %>%
        assess_metrics(truth = HOME_WIN,
                       yes,
                       estimate = HOME_PRED,
                       event_level = 'second') %>%
        mutate(SEASON = factor(SEASON)) %>%
        ggplot(., aes(x=WEEK,
                      color = SEASON,
                      y = .estimate))+
         geom_line(lwd = 1.05,
                  stat = 'smooth',
                  method = 'loess',
                  formula = 'y ~ x',
                  span = 0.2)+
        facet_wrap(.metric ~.,
                   ncol =1)+
        scale_color_viridis_d()+
        theme_bw()

```

### Examine Seasons

```{r source functions again}

source(here::here("functions", "plot_elo_conference_season.R"))
source(here::here("functions", "plot_elo_team_season.R"))
source(here::here("functions", "plot_wins_conference_season.R"))
source(here::here("functions", "table_wins_conference_season.R"))
source(here::here("functions", "table_wins_team_season.R"))
source(here::here("functions", "table_wins_season.R"))

```


```{r wins season}

table_wins_season(sim_team_outcomes,
                  season = 2005,
                  14)

table_wins_conference_season(sim_team_outcomes %>%
                                     filter(CONFERENCE_GAME == T),
                          conference = 'SEC',
                  season = 2022,
                  14)

```



```{r look at specific teams}

plot_elo_conference_season(sim_team_outcomes,
                           season = 2021,
                           'Sun Belt',
                           16)

plot_wins_conference_season(sim_team_outcomes,
                           season = 2022,
                           'Big Ten',
                           14)

```

```{r looking at graphs}

plot_elo_team_season(sim_team_outcomes,
                           season = 2022,
                           team = 'Wisconsin',
                           15)

table_wins_team_season(sim_team_outcomes,
                       2022,
                       'Wisconsin',
                       15)

```

```{r plot win totals selected}

plot_wins_conference_season(sim_team_outcomes,
                           season = 2021,
                           conference = 'ACC',
                           week = 15)

table_wins_conference_season(sim_team_outcomes,
                           season = 2021,
                           conference = 'Big Ten',
                           week = 14)

plot_elo_team_season(sim_team_outcomes,
                     season = 2013,
                     team = 'Florida State',
                     week = 15)

plot_elo_conference_season(sim_team_outcomes,
                     season = 2019,
                     conference = 'Big 12',
                     week = 15
                     )

```


# Simulating Conference Championships
