---
title: "Ranking Teams by Croots"
author: "Phil Henrickson"
date: "6/7/2022"
output: 
  html_document:
    toc: TRUE #adds a Table of Contents
    theme: cerulean
    number_sections: TRUE #number your headings/sections
    toc_float: TRUE #let your ToC follow you as you scroll
    keep_md: no
    fig.caption: yes
    css: "styles.css"
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = F,
                      error = F,
                      warning=F,
                      dev="png",
                      fig.width = 10,
                      fig.height = 6)

options(knitr.duplicate.label = "allow")

options(scipen=999)

```


```{r connect to snowflake}

library(DBI)
library(odbc)
library(RODBC)
library(keyring)


# connect to snowflake
myconn <- DBI::dbConnect(odbc::odbc(),
                         "SnowflakeDSII",
                         Database = "CFB_DEMO",
                         warehouse = "DEMO_WH",
                         uid="phil.henrickson",
                         pwd=keyring::key_get("AE_Snowflake"))

```

```{r packages, include=F} 

source(here::here("scripts/load_packages.R"))
library(jsonlite)
library(forcats)
library(rstan)
library(rstanarm)
library(rstantools)
library(broom)
library(broom.mixed)
library(tidybayes)

conflict_prefer("lag", "dplyr")

```

```{r load functions}

source(here::here("functions/theme_phil.R"))
source(here::here("functions/clean_plays_func.R"))
source(here::here("functions/make_time_features_func.R"))
rm(a)
```

```{r get games and croots data}

# get games
games_data_raw = DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.GAMES')) %>%
        as_tibble() %>%
        mutate(ID = as.numeric(ID)) %>%
        rename(GAME_ID = ID)

# get recruit position rankings
recruiting_players_raw = DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.RECRUITING_PLAYERS')) %>%
        as_tibble()

# get recruiting teams
recruiting_teams_raw = DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.RECRUITING_TEAMS')) %>%
        as_tibble() %>%
        mutate(POINTS = as.numeric(POINTS))

```

```{r get team conference year table}

team_conference = games_data_raw %>% 
  select(SEASON, HOME_TEAM, HOME_CONFERENCE) %>% 
  rename(TEAM = HOME_TEAM, CONFERENCE = HOME_CONFERENCE) %>% 
  bind_rows(.,
            games_data_raw %>%
                    select(SEASON, AWAY_TEAM, AWAY_CONFERENCE) %>% 
                    rename(TEAM = AWAY_TEAM, CONFERENCE = AWAY_CONFERENCE)) %>%
  select(SEASON, TEAM, CONFERENCE) %>%
  unique()

```


# The Data

The data we have on recruits is at the player level.

```{r look at the recruiting data}

set.seed(1)
recruiting_players_raw %>%
        sample_n(5)

```

We have a bunch of info on players, but mostly we want to look at a team's overall recruiting rating by position.

```{r teams by season, warning=F, message=F}

recruits_table = recruiting_players_raw %>%
        select(YEAR, NAME, COMMITTEDTO, POSITION, RANKING, RATING, STARS) 

recruits_table %>%
        select(YEAR, NAME, COMMITTEDTO, POSITION, RANKING, RATING, STARS) %>%
        datatable(escape=F,
                  filter = 'top',
                  options = list(pageLength = 25,
                                 scrollX=F,
                                 autowidth=T,
                                 columnDefs = list(list(className = 'dt-center', targets = which(names(recruits_table)!='NAME')-1))),
                             #    columnDefs=list(list(width = '30px', targets = c(which(names(prob_table)!='Name')-1)))),
                                                # list(width = '400px', targets = c(which(names(prob_table)=='Name')-1)))),
                                       #  targets = seq(3, ncol(prob_table)-1))),
                  rownames = F)

```

## Team Recruiting Composite

We also have team's overall recruiting rankings, which looks to be the team's composite score from 247.

```{r summarize by teams over this time period}

recruiting_teams_raw %>%
        filter(YEAR > 2001) %>%
        filter(YEAR == 2021)

```

For each team, I'll compute an exponentially weighted moving average and a weighted moving average over four year windows.

```{r weighted class}

library(TTR)

max_year = max(recruiting_teams_raw$YEAR)

recruiting_teams_average = 
        recruiting_teams_raw %>%
        filter(YEAR > 2001) %>%
        group_by(TEAM) %>%
        mutate(n = n()) %>%
        filter(n > 4) %>%
        group_by(TEAM) %>%
        select(-n) %>%
        mutate(EMA = EMA(POINTS, n=4)) %>%
        mutate(WMA = WMA(POINTS, n =4, wts = c(.35, .25, .25, .15))) %>%
        ungroup()


recruiting_teams_average %>%
        gather("variable", "value",
               -YEAR, -RANK, -TEAM) %>%
        left_join(., team_conference %>%
                          rename(YEAR = SEASON),
                  by = c("TEAM", "YEAR")) %>%
        filter(CONFERENCE == 'SEC') %>%
        filter(variable %in% c("WMA", "POINTS")) %>%
        group_by(TEAM) %>%
        mutate(LAST_YEAR = max(YEAR)) %>%
        mutate(TEAM_LABEL = case_when(YEAR == LAST_YEAR ~ TEAM)) %>%
        ggplot(., aes(x=YEAR,
                      label = TEAM_LABEL,
                      color = TEAM,
                      y = value))+
        geom_line(stat = 'smooth',
                  method = 'loess',
                  formula = 'y ~ x',
                  lwd = 1.1,
                  span = 0.25)+
        theme_phil()+
        facet_wrap(variable ~.,
                   ncol = 2)+
        geom_text_repel(
                fontface = "bold",
                size = 3,
                direction = "y",
                hjust = 0,
                segment.size = .7,
                segment.alpha = .5,
                segment.linetype = "dotted",
                box.padding = .4,
                segment.curvature = -0.1,
                segment.ncp = 3,
                segment.angle = 20
        )+
        coord_cartesian(clip = 'off',
                        xlim = c(NA, max_year+2))+
        guides(color = "none")

```

# Predicting Games via Team Recruiting Scores

I'll now join up each team's recruiting rankings with the games data. I want to see how the difference in teach team's composite rating affects the outcome of each game.

```{r join games with recruiting rankings}

games_and_recruiting = games_data_raw %>%
        left_join(.,
                  recruiting_teams_average %>%
                          ungroup() %>%
                          filter(!is.na(EMA)) %>%
                          mutate(HOME_TEAM = TEAM,
                                 HOME_RECRUITING_EMA = EMA,
                                 HOME_RECRUITING_WMA = WMA) %>%
                          select(YEAR, starts_with("HOME")) %>%
                          rename(SEASON = YEAR),
                  by = c("SEASON", "HOME_TEAM"))  %>%
        left_join(., 
                  recruiting_teams_average %>%
                          ungroup() %>%
                          filter(!is.na(EMA)) %>%
                          mutate(AWAY_TEAM = TEAM,
                                 AWAY_RECRUITING_EMA = EMA,
                                 AWAY_RECRUITING_WMA = WMA) %>%
                          select(YEAR, starts_with("AWAY")) %>%
                          rename(SEASON = YEAR),
                  by = c("SEASON", "AWAY_TEAM")) %>%
        filter(SEASON > 2006)

```

## Splitting the Data

```{r split into train and test}

games_full = games_and_recruiting %>%
        mutate(HOME_WIN = case_when(HOME_POINTS > AWAY_POINTS ~ 'yes',
                                    AWAY_POINTS > HOME_POINTS ~ 'no')) %>%
        mutate(HOME_WIN = factor(HOME_WIN,
                                 levels = c("no", "yes"))) %>%
        mutate(HOME_FIELD_ADVANTAGE = case_when(NEUTRAL_SITE == F ~ 1,
                                                TRUE ~ 0)) %>%
        mutate_at(c("HOME_RECRUITING_WMA",
                    "HOME_RECRUITING_EMA",
                    "AWAY_RECRUITING_WMA",
                    "AWAY_RECRUITING_EMA"),
                  ~ replace_na(., 0)) %>%
        mutate(HOME_RECRUITING_DIFF_WMA  = HOME_RECRUITING_WMA - AWAY_RECRUITING_WMA,
               HOME_RECRUITING_DIFF_EMA = HOME_RECRUITING_EMA - AWAY_RECRUITING_EMA)

games_train = games_full %>%
        filter(SEASON < 2015)

games_valid = games_full %>%
        filter(SEASON >=2015 & SEASON < 2019)

games_test = games_full %>%
        filter(SEASON >=2019)

```

Looking at games in the training set, at what rate does the home team win as a function of the difference in the home-away recruiting ranings?

```{r look at outcomes by recruiting diff}

games_train %>%
        mutate(HOME_RECRUITING_DIFF = HOME_RECRUITING_EMA - AWAY_RECRUITING_EMA) %>%
        ggplot(., aes(x=HOME_RECRUITING_DIFF,
                      fill = HOME_WIN))+
        geom_histogram(bins = 100)+
        facet_wrap(HOME_WIN ~.,
                   ncol = 1)+
        theme_phil()+
        geom_vline(xintercept = 0,
                   linetype = 'dashed')

games_train %>%
        mutate(HOME_RECRUITING_EMA_BIN = plyr::round_any(HOME_RECRUITING_DIFF_EMA,
                                                         10)) %>%
        mutate(HOME_RECRUITING_EMA_BIN = case_when(HOME_RECRUITING_EMA_BIN < -200 ~ 200,
                                                   HOME_RECRUITING_EMA_BIN > 200 ~ 100,
                                                   TRUE ~ HOME_RECRUITING_EMA_BIN)) %>%
        group_by(HOME_RECRUITING_EMA_BIN, HOME_WIN) %>%
        count() %>%
        group_by(HOME_RECRUITING_EMA_BIN) %>%
        mutate(prop = n / sum(n)) %>%
        mutate(games = sum(n)) %>%
        filter(HOME_WIN == 'yes') %>%
        ggplot(., aes(x=HOME_RECRUITING_EMA_BIN,
                      y=prop))+
        geom_point(aes(size=games))+
        geom_line()+
        theme_phil()+
        geom_hline(yintercept = c(0, 0.5,0.9),
                   linetype='dashed')+
        geom_vline(xintercept = 0,
                   linetype = 'dashed') +
        theme(legend.title = element_text())+
        guides(size = guide_legend(title = 'Number of Games',
                                   title.position = 'top'))+
        xlab("Home Weighted Recruiting Rating - Away Weighted Recruiting Rating")+
        ylab("Pr(Home Team Wins)")+
        geom_smooth(method = 'loess',
                    color = 'grey60',
                    formula = 'y ~ x')
        
```

Just from looking at the training data, we can see that a home team with an advantage of 100 points in their team's recruiting rating (using a weighted average over the last four years) wins games a little under 90% of the time. 

## Training Models

I'll now fit a logistic regression on home win as a function of the two different recruiting differentials, as well as a dummy for whether the game was neutral site.

```{r now fit a simple model, include=F}

train_fit = games_train %>%
        nest() %>%
        mutate(glm_wma = map(data, ~ stan_glm(HOME_WIN ~ HOME_FIELD_ADVANTAGE + HOME_RECRUITING_DIFF_WMA + HOME_RECRUITING_DIFF_WMA*HOME_FIELD_ADVANTAGE,
                                          family = binomial(link = "logit"),
                                          data = .x))) %>%
        mutate(glm_ema = map(data, ~ stan_glm(HOME_WIN ~ HOME_FIELD_ADVANTAGE + HOME_RECRUITING_DIFF_WMA + HOME_RECRUITING_DIFF_WMA*HOME_FIELD_ADVANTAGE,
                                          family = binomial(link = "logit"),
                                          data = .x)))

```

```{r examine results} 

train_fit %>%
        mutate(tidied_wma = map(glm_wma, ~ .x %>% 
                                    tidy(se = "robust", conf.int=T))) %>%
        mutate(tidied_ema = map(glm_ema, ~ .x %>% 
                                    tidy(se = "robust", conf.int=T))) %>%
        select(tidied_wma, tidied_ema)  %>%
        gather("model", "value") %>%
        unnest(value) %>%
        mutate_if(is.numeric, round, 4)

```
Both models return pretty similar results - a 100 point difference in recruiting rankings maps to about 30% increase in win probability?

```{r look at the predicted probabilities}

train_preds = train_fit %>%
        mutate(preds_wma = map(glm_wma, ~ .x %>% predict(newdata = games_train, 
                                                         type = 'response') %>%
                                       as_tibble() %>%
                                       bind_cols(., games_train))) %>%
        mutate(preds_ema = map(glm_ema, ~ .x %>%  
                                       predict(newdata = games_train, type = 'response') %>%
                                       as_tibble() %>%
                                       bind_cols(., games_train))) %>%
        select(preds_wma, preds_ema) %>%
        gather("method", "preds") %>%
        mutate(method = gsub("preds_", "", method)) %>%
        unnest(preds)  %>%
        rename(.pred = value)

train_preds %>%
        select(method, .pred, HOME_WIN, GAME_ID, SEASON, HOME_TEAM, AWAY_TEAM) %>%
        arrange(GAME_ID) %>%
       # spread(method, .pred) %>%
        ggplot(., aes(x=.pred,
                      fill = HOME_WIN))+
        geom_histogram(bins = 100)+
        facet_wrap(HOME_WIN ~ method)+
        theme_phil()+
        geom_vline(xintercept = 0.5,
                   linetype = 'dotted')+
        scale_fill_manual(values = c("red",
                                     "navy"))+
        theme(legend.title = element_text())
```


```{r plot predicted probabilites by recruiting diff}

# log loss
train_preds %>%
        bind_rows(.,
        bind_cols(train_preds %>%
                          select(-method, -.pred),
                          tibble(method = 'home',
                                 .pred = games_train %>% 
                                         group_by(HOME_WIN) %>% 
                                         count() %>%
                                         ungroup() %>%
                                         mutate(prop = n / sum(n)) %>% 
                                         filter(HOME_WIN == 'yes') %>% 
                                         pull(prop)
                          ))) %>%
        group_by(method, SEASON) %>%
        yardstick::mn_log_loss(truth = HOME_WIN,
                               estimate = .pred,
                               event_level = 'second') %>%
        mutate_if(is.numeric, round, 3) %>%
        spread(method, .estimate) %>%
        select(SEASON, .metric, .estimator, home, ema, wma)

# roc auc
train_preds %>%
        bind_rows(.,
        bind_cols(train_preds %>%
                          select(-method, -.pred),
                          tibble(method = 'home',
                                 .pred = games_train %>% 
                                         group_by(HOME_WIN) %>% 
                                         count() %>%
                                         ungroup() %>%
                                         mutate(prop = n / sum(n)) %>% 
                                         filter(HOME_WIN == 'yes') %>% 
                                         pull(prop)
                          ))) %>%
        group_by(method, SEASON) %>%
        yardstick::roc_auc(truth = HOME_WIN,
                               estimate = .pred,
                               event_level = 'second') %>%
        mutate_if(is.numeric, round, 3) %>%
        spread(method, .estimate) %>%
        select(SEASON, .metric, .estimator, home, ema, wma)

# accuracy
train_preds %>%
        bind_rows(.,
        bind_cols(train_preds %>%
                          select(-method, -.pred),
                          tibble(method = 'home',
                                 .pred = games_train %>% 
                                         group_by(HOME_WIN) %>% 
                                         count() %>%
                                         ungroup() %>%
                                         mutate(prop = n / sum(n)) %>% 
                                         filter(HOME_WIN == 'yes') %>% 
                                         pull(prop)
                          ))) %>%
        mutate(.pred_class = factor(case_when(.pred > .5 ~ 'yes',
                                       TRUE ~ 'no'),
                                    levels = c("no", "yes"))) %>%
        group_by(method, SEASON) %>%
        yardstick::accuracy(truth = HOME_WIN,
                               estimate = .pred_class,
                               event_level = 'second') %>%
        mutate_if(is.numeric, round, 3) %>%
        spread(method, .estimate) %>%
        select(SEASON, .metric, .estimator, home, ema, wma)


```

Now predict the validation set with each model.

```{r predict validation set, warning=T,}

valid_preds = train_fit %>%
        mutate(preds_wma = map(glm_wma, ~ .x %>% predict(newdata = games_valid, 
                                                         type = 'response') %>%
                                       as_tibble() %>%
                                       bind_cols(., games_valid))) %>%
        mutate(preds_ema = map(glm_ema, ~ .x %>%  
                                       predict(newdata = games_valid, type = 'response') %>%
                                       as_tibble() %>%
                                       bind_cols(., games_valid))) %>%
        select(preds_wma, preds_ema) %>%
        gather("method", "preds") %>%
        mutate(method = gsub("preds_", "", method)) %>%
        unnest(preds)  %>%
        rename(.pred = value)

# validation set
valid_preds %>%
        select(method, .pred, HOME_WIN, GAME_ID, SEASON, HOME_TEAM, AWAY_TEAM) %>%
        arrange(GAME_ID) %>%
       # spread(method, .pred) %>%
        ggplot(., aes(x=.pred,
                      fill = HOME_WIN))+
        geom_histogram(bins = 100)+
        facet_wrap(HOME_WIN ~ method)+
        theme_phil()+
        geom_vline(xintercept = 0.5,
                   linetype = 'dotted')+
        scale_fill_manual(values = c("red",
                                     "navy"))+
        theme(legend.title = element_text())

```

```{r scatter plot of validation}

valid_preds %>%
        select(method, .pred, HOME_WIN, GAME_ID, SEASON, HOME_TEAM, AWAY_TEAM) %>%
        arrange(GAME_ID) %>%
        spread(method, .pred) %>%
        mutate(GAME_LABEL = paste(HOME_TEAM,
                                  AWAY_TEAM,
                                  SEASON)) %>%
        ggplot(., aes(x=wma,
                      y=ema,
                      label = GAME_LABEL,
                      color = HOME_WIN))+
        geom_point()+
        theme_phil()+
        scale_color_manual(values = c("red",
                                      "navy"))+
        geom_vline(xintercept = 0.5,
                   linetype = 'dashed')+
        geom_hline(yintercept=0.5,
                   linetype = 'dashed')+
        facet_wrap(HOME_WIN~.)+
        geom_text(check_overlap = T,
                  vjust=-1,
                  size=2.5)

```

Assessment on validation set

```{r validation set assessment}

valid_preds %>%
        bind_rows(.,
        bind_cols(valid_preds %>%
                          select(-method, -.pred),
                          tibble(method = 'home',
                                 .pred = games_train %>% 
                                         group_by(HOME_WIN) %>% 
                                         count() %>%
                                         ungroup() %>%
                                         mutate(prop = n / sum(n)) %>% 
                                         filter(HOME_WIN == 'yes') %>% 
                                         pull(prop)
                          ))) %>%
        group_by(method, SEASON) %>%
        yardstick::mn_log_loss(truth = HOME_WIN,
                               estimate = .pred,
                               event_level = 'second') %>%
        mutate_if(is.numeric, round, 3) %>%
        spread(method, .estimate) %>%
        select(SEASON, .metric, .estimator, home, ema, wma)

```

