---
title: "`r paste(params$team, params$input_season)` Predictions"
#date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE #adds a Table of Contents
    theme: cerulean
  #  number_sections: TRUE #number your headings/sections
    toc_float: TRUE #let your ToC follow you as you scroll
    keep_md: no
    fig.caption: yes
params:
        team: "LSU"
        input_season: 2021
        sim_week: 0
        nsims: 1000
---

```{r setup, include=F}

knitr::opts_chunk$set(echo = F,
                      error = F,
                      warning=F,
                      dev="png",
                      fig.width = 10,
                      fig.height = 6)
options(knitr.duplicate.label = "allow")
options(scipen=999)

```

```{r packages, include=F}

source(here::here("scripts/load_packages.R"))
library(jsonlite)
library(forcats)
library(TTR)
library(flextable)
conflict_prefer("lag", "dplyr")

# parallelization
library(future.apply)
plan(multisession, workers = 7)

set_flextable_defaults(theme_fun = theme_alafoli,
                       font.color = "grey10",
                       font.size=10,
                       padding.bottom = 6, 
                       padding.top = 6,
                       padding.left = 6,
                       padding.right = 6,
                       background.color = "white")

```

```{r connect to snowflake}

library(DBI)
library(odbc)
library(RODBC)
library(keyring)

### connect to DBS
# connect to snowflake
myconn <- DBI::dbConnect(odbc::odbc(),
                         "SnowflakeDSII",
                         Database = "CFB_DEMO",
                         warehouse = "DEMO_WH",
                         uid="phil.henrickson",
                         pwd=keyring::key_get("AE_Snowflake"))

```

```{r functions and colors}

### functions
source(here::here("functions/get_expected_score.R"))
source(here::here("functions/get_new_elos.R"))
source(here::here("functions/calc_elo_ratings.R"))
source(here::here("functions", "simulateX.R"))
source(here::here("functions", "sim_game_margin.R"))
source(here::here("functions", "calc_elo_ratings.R"))
source(here::here("functions", "sim_elo_ratings.R"))
source(here::here("functions", "sim_rest_of_season.R"))
source(here::here("functions", "theme_phil.R"))

# color palettes
load(here::here("data", "team_colors.Rdata"))
custom_color_teams <- scale_colour_manual(name = "TEAM", values = team_colors)
custom_fill_teams <- scale_fill_manual(name = "TEAM", values = team_colors)

```

```{r load in more}

### get data sources
# teams raw
teams_raw = DBI::dbGetQuery(myconn,
                            paste('SELECT * FROM CFB_DEMO.CFD_RAW.TEAMS')) %>%
        as_tibble() %>%
        rename(TEAM = SCHOOL,
               TEAM_ID = ID,
               TEAM_MASCOT = MASCOT,
               TEAM_ABBREVIATION = ABBREVIATION)

# get games
games_raw = DBI::dbGetQuery(myconn,
                            paste('SELECT * FROM CFB_DEMO.CFD_RAW.ANALYSIS_GAMES')) %>%
        as_tibble() %>%
        mutate(CONFERENCE_GAME = case_when(CONFERENCE_GAME == 'true' ~ T,
                                           CONFERENCE_GAME == 'false' ~ F)) %>%
        mutate(GAME_DATE = as.Date(START_DATE)) %>%
        mutate(START_DATE = as_datetime(START_DATE)) %>%
        arrange(START_DATE) %>%
        # logic for flagging conference championship games
        group_by(SEASON, SEASON_TYPE, HOME_CONFERENCE) %>%
        mutate(LAST_GAME = START_DATE == max(START_DATE)) %>% 
        ungroup() %>%
        mutate(CONFERENCE_CHAMPIONSHIP = case_when(CONFERENCE_GAME == T & 
                                                           (HOME_CONFERENCE != 'FBS Independents') & 
                                                           (HOME_CONFERENCE != 'FCS Independents') &
                                                           !(HOME_CONFERENCE %in% 'Big Ten' & SEASON < 2014) &
                                                           !(HOME_CONFERENCE %in% 'Western Athletic') &
                                                           (NEUTRAL_SITE == T | 
                                                                    is.na(VENUE) |
                                                                    VENUE == 'Glass Bowl' |
                                                                    VENUE == 'Bright House Networks Stadium' |
                                                                    VENUE == 'Alamodome' |
                                                                    VENUE == 'Georgia Dome' |
                                                                    HOME_CONFERENCE == 'Sun Belt' |
                                                                    HOME_CONFERENCE == 'American Athletic') &
                                                           SEASON > 2000 &
                                                           SEASON_TYPE == 'regular' &
                                                           WEEK > 13 &
                                                           LAST_GAME == T ~ T,
                                                   TRUE ~ F)) %>%
        mutate(HOME_WIN = factor(case_when(HOME_POINTS > AWAY_POINTS ~ 'yes',
                                           HOME_POINTS < AWAY_POINTS ~ 'no',
                                           HOME_POINTS == AWAY_POINTS ~ 'no'),
                                 levels = c("no", "yes"))) %>%
        mutate(HOME_DIVISION  = replace_na(HOME_DIVISION, "missing"),
               AWAY_DIVISION = replace_na(AWAY_DIVISION, "missing")) %>%
        left_join(., teams_raw %>%
                          select(TEAM, TEAM_ABBREVIATION) %>%
                          rename(HOME_TEAM = TEAM,
                                 HOME_TEAM_ABBR = TEAM_ABBREVIATION),
                  by = c("HOME_TEAM")) %>%
        left_join(., teams_raw %>%
                          select(TEAM, TEAM_ABBREVIATION) %>%
                          rename(AWAY_TEAM = TEAM,
                                 AWAY_TEAM_ABBR = TEAM_ABBREVIATION),
                  by = c("AWAY_TEAM"))

# get team conference mapping
team_conference_season = DBI::dbGetQuery(myconn,
                                         paste('SELECT * FROM CFB_DEMO.CFD_RAW.ANALYSIS_TEAM_CONFERENCE_SEASON')) %>%
        as_tibble()

# team mapping function
team_mapping = teams_raw %>%
        mutate(TEAM_ABBR = gsub(" ", "\\.", gsub("[[:punct:]]", "\\.", TEAM))) %>%
        select(TEAM, TEAM_ABBR, TEAM_ABBREVIATION, COLOR, ALT_COLOR) %>%
        mutate(COLOR = case_when(TEAM == 'Toledo' ~ '#003E7E',
                                 TRUE ~ COLOR))

### get historical elo
# games
elo_games = DBI::dbGetQuery(myconn,
                            paste('SELECT * FROM CFB_DEMO.CFD_ANALYTICS.ACTIVE_ELO_GAMES')) %>%
        as_tibble()

# teams
elo_teams = DBI::dbGetQuery(myconn,
                            paste('SELECT * FROM CFB_DEMO.CFD_ANALYTICS.ACTIVE_ELO_TEAMS')) %>%
        as_tibble() %>%
        select(-LOAD_DATE)

### recruiting 
recruiting_teams_raw = DBI::dbGetQuery(myconn,
                                       paste('SELECT * FROM CFB_DEMO.CFD_RAW.RECRUITING_TEAMS ')) %>%
        as_tibble() %>%
        mutate(POINTS = as.numeric(POINTS)) %>%
        rename(RECRUITING_POINTS = POINTS)

# expand
recruiting_teams = expand.grid(TEAM = unique(recruiting_teams_raw$TEAM),
                               YEAR = seq(min(recruiting_teams_raw$YEAR),
                                          max(recruiting_teams_raw$YEAR))) %>%
        left_join(., recruiting_teams_raw,
                  by = c("YEAR", "TEAM")) %>%
        mutate(RECRUITING_POINTS = replace_na(RECRUITING_POINTS, 0)) %>%
        filter(YEAR > 2001) %>%
        left_join(., team_conference_season,
                  by = c("YEAR"="SEASON", "TEAM")) %>%
        filter(DIVISION == 'fbs') %>%
        group_by(TEAM) %>%
        mutate(n = n()) %>%
        filter(n > 4) %>%
        group_by(TEAM) %>%
        select(-n) %>%
        mutate(EMA = EMA(RECRUITING_POINTS, n =4)) %>%
        # now fill with actual in the event that a team doesn't have yet have enough data
        mutate(EMA = case_when(is.na(EMA) & !is.na(RECRUITING_POINTS) ~ RECRUITING_POINTS,
                               TRUE ~ EMA)) %>%
        # mutate_at(c("EMA", "WMA"),
        #           replace_na, 0) %>%
        arrange(TEAM, YEAR)  %>%
        mutate(SEASON = YEAR) %>%
        ungroup() %>%
        select(TEAM, SEASON, CONFERENCE, DIVISION, RANK, RECRUITING_POINTS, EMA)

# get what we need
recruiting = recruiting_teams %>%
        select(TEAM, SEASON, CONFERENCE, RECRUITING_POINTS, EMA) %>%
        rename(RECRUITING_SCORE = EMA)


### efficiency
# season level efficiency rankings
season_team_rankings = DBI::dbGetQuery(myconn,
                                       paste('SELECT * FROM CFB_DEMO.CFD_ANALYTICS.EFFICIENCY_TEAM_SEASON')) %>%
        as_tibble()

### combine datasets to get games
games = games_raw %>%
        filter(SEASON_TYPE == 'regular') %>%
        filter(CONFERENCE_CHAMPIONSHIP == F) %>%
        mutate(HOME_DIVISION = case_when(is.na(HOME_DIVISION) & SEASON < 1900 ~ 'fbs',
                                         TRUE ~ HOME_DIVISION)) %>%
        mutate(AWAY_DIVISION = case_when(is.na(AWAY_DIVISION) & SEASON < 1900 ~ 'fbs',
                                         TRUE ~ AWAY_DIVISION)) %>%
        mutate(FBS = HOME_DIVISION == 'fbs' | AWAY_DIVISION == 'fbs') %>%
        filter((FBS == T) | (FBS==F & SEASON < 1981)) %>%
        arrange(GAME_DATE) %>%
        select(GAME_ID, 
               SEASON, 
               WEEK,
               SEASON_TYPE,
               START_DATE,
               GAME_DATE,
               NEUTRAL_SITE, 
               HOME_TEAM,
               AWAY_TEAM,
               CONFERENCE_GAME,
               CONFERENCE_CHAMPIONSHIP,
               HOME_CONFERENCE,
               AWAY_CONFERENCE,
               HOME_DIVISION,
               AWAY_DIVISION,
               HOME_POINTS,
               AWAY_POINTS) %>%
        left_join(.,
                  recruiting %>%
                          select(SEASON,
                                 TEAM,
                                 RECRUITING_SCORE) %>%
                          rename(HOME_TEAM = TEAM,
                                 HOME_RECRUITING_SCORE = RECRUITING_SCORE),
                  by = c("SEASON",
                         "HOME_TEAM")) %>%
        left_join(.,
                  recruiting %>%
                          select(SEASON,
                                 TEAM,
                                 RECRUITING_SCORE) %>%
                          rename(AWAY_TEAM = TEAM,
                                 AWAY_RECRUITING_SCORE = RECRUITING_SCORE),
                  by = c("SEASON",
                         "AWAY_TEAM"))


# data sources
load(here::here("data", "season_team_rankings.Rdata"))
load(here::here("data", "season_team_rankings_type.Rdata"))
load(here::here("data", "weekly_team_rankings.Rdata"))
load(here::here("data", "weekly_team_rankings_type.Rdata"))

```

```{r set team}

# set
team = params$team

# set pattern
url_pattern <- "http[s]?://(?:[a-zA-Z]|[0-9]|[$-_@.&+]|[!*\\(\\),]|(?:%[0-9a-fA-F][0-9a-fA-F]))+"

# make color function
team_color = team_mapping %>%
        mutate(COLOR = case_when(TEAM == 'West Virginia' |
                                         TEAM == 'LSU' ~ ALT_COLOR,
                                 TRUE ~ COLOR)) %>%
        filter(TEAM == team) %>%
                pull(COLOR)

# team logo
team_logos = teams_raw %>%
        filter(TEAM == team) %>%
        select(LOGOS) %>%
        separate(LOGOS, into = c("logo_1", "logo_2"), sep=",") %>%
        mutate(logo_1 = str_extract(logo_1, url_pattern),
               logo_2 = str_extract(logo_2, url_pattern))
        
# define a function based on that team's color
team_col_func =
        function(x) {

                breaks = seq(0, 2, 0.1)
                colorRamp=colorRampPalette(c("white", team_color))
                col_palette <- colorRamp(length(breaks))
                mycut <- cut(x,
                             breaks = breaks,
                             include.lowest = TRUE,
                             right=T,
                             label = FALSE)
                col_palette[mycut]
        }

# team outcome func
team_outcome_func =
        function(x) {

                breaks = c("loss", "win")
                colorRamp=c("white", team_color)
                col_palette <- colorRamp(length(breaks))
                mycut <- cut(x,
                             breaks = breaks,
                             include.lowest = TRUE,
                             right=T,
                             label = FALSE)
                col_palette[mycut]
        }


my_caption = list(labs(caption = paste("data from collegefootballdata.com",
                                       "analysis and model by Phil Henrickson",
                                       "github.com/phenrickson/cfb",
                                       sep = "\n")))

```



```{r load simulations}

# get files from season folder
files = list.files(here::here("simulations",
                      "season",
                      params$input_season))

# get objects from each week
files_weeks =  files[grepl("sims_week", files)]

# load in simulations
week_list = foreach(i = 1:length(files_weeks)) %do% {
        
        load(here::here("simulations",
                        "season",
                        params$input_season,
                        files_weeks[i]))
        
        foo = all_simulations
        
        rm(all_simulations)
        
        foo
        
}

rm(foo)

```

```{r load sims data}

# pull predictions table
team_predictions = do.call(rbind, lapply(week_list, function (x) x$team_predictions)) %>%
        filter(SIM_FROM_WEEK <= params$sim_week)

# pull game predictions
game_predictions = do.call(rbind, lapply(week_list, function (x) x$game_predictions)) %>%
        filter(SIM_FROM_WEEK <= params$sim_week)

# pull remaining games
sim_remaining_games = do.call(rbind, lapply(week_list, function (x) x$sim_remaining_games)) %>%
        filter(SIM_FROM_WEEK == params$sim_week)

# conference championships
sim_conference_championship_games = do.call(rbind, lapply(week_list, function (x) x$sim_conference_championship_games)) %>%
        filter(SIM_FROM_WEEK <= params$sim_week)

# playoff games
sim_playoff_games = do.call(rbind, lapply(week_list, function (x) x$sim_playoff_games)) %>%
        filter(SIM_FROM_WEEK <= params$sim_week)

# playoff games
sim_nc_games= do.call(rbind, lapply(week_list, function (x) x$sim_nc_games)) %>%
        filter(SIM_FROM_WEEK <= params$sim_week)

# get start of season simulations
sim_from_start =  do.call(rbind, lapply(week_list, function (x) x$sim_remaining_games)) %>%
        filter(SIM_FROM_WEEK == 0)

```

```{r get predictions}

games_and_outcomes = games %>%
        filter(SEASON == params$input_season) %>%
        mutate(HOME_POINTS = case_when(WEEK > params$sim_week ~ NA_real_,
                                       TRUE~  HOME_POINTS),
               AWAY_POINTS = case_when(WEEK > params$sim_week ~ NA_real_,
                                       TRUE ~ AWAY_POINTS)) %>%
        left_join(., game_predictions %>%
                          group_by(GAME_ID) %>%
                          filter(SIM_FROM_WEEK == max(SIM_FROM_WEEK)),
                          by = c("GAME_ID")) %>%
    #    select(WEEK, SIM_FROM_WEEK, SEASON, NEUTRAL_SITE, HOME_TEAM, AWAY_TEAM, HOME_SIM_PROB, HOME_SIM_MARGIN, HOME_PREGAME_ELO, AWAY_PREGAME_ELO, 
              # HOME_POINTS, AWAY_POINTS) %>%
        mutate(QUALITY = case_when(NEUTRAL_SITE == T ~ 2 / ((1/(HOME_PREGAME_ELO+75) + (1/AWAY_PREGAME_ELO))),
                                   NEUTRAL_SITE == F ~ 2 / ((1/(HOME_PREGAME_ELO) + (1/AWAY_PREGAME_ELO))))) %>%
        mutate(QUALITY = round(100*((QUALITY - min(QUALITY)) / ((max(QUALITY) - min(QUALITY)))),0)) %>%
      #  mutate(QUALITY = 2 / ((1/HOME_PREGAME_ELO) + (1/AWAY_PREGAME_ELO))) %>%
        rename(HOME = HOME_TEAM,
               AWAY = AWAY_TEAM) %>%
        mutate(HOME_SIM = case_when(HOME_SIM_PROB >=.5 ~ 'win',
                                     HOME_SIM_PROB < .5 ~ 'loss')) %>%
        mutate(HOME_OUTCOME = case_when(WEEK > params$sim_week ~ NA_character_,
                                        HOME_POINTS > AWAY_POINTS ~ 'win',
                                        HOME_POINTS < AWAY_POINTS ~ 'loss'),
               HOME_MARGIN = HOME_POINTS - AWAY_POINTS) %>%
        mutate(PRED = case_when(HOME_SIM == 'win' ~ paste(HOME, "by", abs(HOME_SIM_MARGIN)),
                                HOME_SIM == 'loss' ~paste(AWAY, "by", abs(HOME_SIM_MARGIN)))) %>%
        mutate(OUTCOME = case_when(WEEK > params$sim_week ~ NA_character_,
                                   HOME_OUTCOME == 'win' ~ paste(HOME, "by", abs(HOME_MARGIN)),
                                   HOME_OUTCOME == 'loss' ~ paste(AWAY, "by", abs(HOME_MARGIN)))) %>%
        mutate(HOME_PROB = HOME_SIM_PROB) %>%
        mutate(HOME_WIN = factor(case_when(HOME_POINTS > AWAY_POINTS ~ 'yes',
                                           HOME_POINTS < AWAY_POINTS ~ 'no'),
                                 levels = c('no', 'yes'))) %>%
        mutate(HOME_PRED = factor(case_when(HOME_SIM_PROB >= .5 ~ 'yes',
                                            HOME_SIM_PROB < 5 ~ 'no'),
                                  levels = c("no", "yes"))) %>%
        mutate(HOME_SIM_OUTCOME = HOME_SIM) %>%
        mutate(AWAY_MARGIN = -HOME_MARGIN) %>%
        mutate(AWAY_OUTCOME = case_when(WEEK > params$sim_week ~ NA_character_,
                                        HOME_POINTS < AWAY_POINTS ~ 'win',
                                        HOME_POINTS > AWAY_POINTS ~ 'loss'))
games_schedule = games_and_outcomes %>%
        select(GAME_ID,
               GAME_DATE,
               SIM_FROM_WEEK,
                WEEK,
                SEASON, 
                SEASON_TYPE,
                HOME_CONFERENCE, 
                AWAY_CONFERENCE, 
               HOME_DIVISION,
                HOME,
                AWAY,
                HOME_PREGAME_ELO,
                AWAY_PREGAME_ELO,
                HOME_POSTGAME_ELO, 
                AWAY_POSTGAME_ELO,
               HOME_SIM_PROB,
               HOME_MARGIN,
               HOME_OUTCOME,
                HOME_SIM_MARGIN,
                HOME_SIM_OUTCOME) %>%
        rename(HOME_TEAM = HOME,
               AWAY_TEAM = AWAY) %>%
        rename(OPPONENT = AWAY_TEAM,
               OPPONENT_CONFERENCE = AWAY_CONFERENCE,
               OPPONENT_PREGAME_ELO = AWAY_PREGAME_ELO,
               OPPONENT_POSTGAME_ELO = AWAY_POSTGAME_ELO) %>%
        set_names(., gsub("HOME_", "", names(.))) %>%
        bind_rows(.,
                  games_and_outcomes %>%
                          mutate(AWAY_SIM_OUTCOME =case_when(HOME_SIM_PROB >=.5 ~ 'loss',
                                                               HOME_SIM_PROB < .5 ~ 'win')) %>%
                          mutate(AWAY_SIM_PROB = 1-HOME_SIM_PROB) %>%
                          select(GAME_ID,
                                 GAME_DATE,
                                 SIM_FROM_WEEK,
                                  WEEK,
                                  SEASON, 
                                  SEASON_TYPE,
                                  AWAY_CONFERENCE, 
                                  HOME_CONFERENCE, 
                                 AWAY_DIVISION,
                                  AWAY,
                                  HOME,
                                  AWAY_PREGAME_ELO,
                                  HOME_PREGAME_ELO,
                                  AWAY_POSTGAME_ELO, 
                                  HOME_POSTGAME_ELO,
                                 AWAY_SIM_PROB,
                                 AWAY_MARGIN,
                                 AWAY_OUTCOME,
                                  HOME_SIM_MARGIN,
                                  AWAY_SIM_OUTCOME) %>%
                          rename(HOME_TEAM = HOME,
                                 AWAY_TEAM = AWAY) %>%
                          mutate(AWAY_SIM_MARGIN = -HOME_SIM_MARGIN) %>%
                          select(-HOME_SIM_MARGIN) %>%
                          rename(OPPONENT = HOME_TEAM,
                                 OPPONENT_CONFERENCE = HOME_CONFERENCE,
                                 OPPONENT_PREGAME_ELO = HOME_PREGAME_ELO,
                                 OPPONENT_POSTGAME_ELO = HOME_POSTGAME_ELO) %>%
                          set_names(., gsub("AWAY_", "", names(.)))) %>%
        arrange(WEEK, GAME_DATE, GAME_ID) %>%
        group_by(GAME_ID) %>%
        filter(SIM_FROM_WEEK == max(SIM_FROM_WEEK)) %>%
        ungroup()

# get strength of schedule played so far
sos = games_schedule %>%
        filter(DIVISION == 'fbs') %>%
        filter(WEEK <= params$sim_week) %>%
        group_by(TEAM) %>%
        summarize(mean_opponent = mean(OPPONENT_PREGAME_ELO),
                  median_opponent = median(OPPONENT_PREGAME_ELO)) %>%
        arrange(desc(mean_opponent)) %>%
        mutate(SOS = row_number()) %>%
        rename(MEAN_OPPONENT_RATING = mean_opponent) %>%
        select(TEAM, SOS)
        
# srs 
srs = games_schedule %>%
        filter(DIVISION == 'fbs') %>%
        filter(WEEK > params$sim_week) %>%
        group_by(TEAM) %>%
        summarize(mean_opponent = mean(OPPONENT_PREGAME_ELO),
                  median_opponent = median(OPPONENT_PREGAME_ELO)) %>%
        arrange(desc(mean_opponent)) %>%
        mutate(SRS = row_number()) %>%
        rename(MEAN_OPPONENT_RATING = mean_opponent) %>%
        select(TEAM, SRS)
        
```

# Season Predictions

How many games will a college football team win in the upcoming season? How likely are they to win their conference championship? 

This notebook displays **`r params$input_season`** predictions (as of Week `r params$sim_week`) and historical data for **`r params$team`**.

```{r proejction for teams}

team_preds = team_predictions %>%
        filter(SIM_FROM_WEEK == params$sim_week) %>%
        group_by(SEASON, SIM_FROM_WEEK) %>%
        arrange(desc(RATING)) %>%
        mutate(RANK = row_number()) %>%
        ungroup() %>%
        filter(TEAM == params$team) %>%
        mutate(RATING = round(RATING, 0)) %>%
                mutate(SEASON = factor(SEASON)) %>%
        mutate(mean_wins = format(round(mean_wins, 1), nsmall=1),
               mean_loss = format(round(mean_loss, 1), nsmall=1)) %>%
        mutate_at(c("made_bowl",
                    "made_conf",
                    "win_conf",
                    "made_playoff",
                    "made_nc",
                    "win_nc"),
                  ~ format(., nsmall = 3)) %>%
        mutate(CONFERENCE = case_when(CONFERENCE == 'American Athletic' ~ 'AAC',
                                      CONFERENCE == 'FBS Independents' ~ 'Independent',
                                      TRUE ~ CONFERENCE)) %>%
        select(SIM_FROM_WEEK,
               TEAM,
               SEASON,
               CONFERENCE,
            #   CONFERENCE_DIVISION,
               RANK,
               RATING,
               mean_wins,
               mean_loss,
               made_bowl,
               made_conf,
               win_conf,
               made_playoff,
               made_nc,
               win_nc) %>%
        left_join(.,
                  sos,
                  by = c("TEAM")) %>%
        left_join(.,
                  srs,
                  by = c("TEAM")) %>%
        select(-TEAM) %>%
        mutate(SOS = as.character(SOS),
               SRS = as.character(SRS)) %>%
        mutate(SOS = replace_na(SOS, SRS), 
               SRS = replace_na(SRS, "")) %>%
        rename(Season = SEASON,
                Week = SIM_FROM_WEEK,
             #  Team = TEAM,
               `Rank` = RANK,
               `Rating` = RATING,
                Conference = CONFERENCE,
                `Strength of Schedule`  = SOS,
             `Strength Remaining Schedule` = SRS,
               `Predicted Wins` = mean_wins,
               `Predicted Losses` = mean_loss,
               `Pr(Bowl Game)` = made_bowl,
             `Pr(Make Conf Champ)` = made_conf,
               `Pr(Win Conf Champ)` = win_conf,
             `Pr(Make Playoff)` = made_playoff,
             `Pr(Make National Champ)` = made_nc,
             `Pr(Win National Champ)` = win_nc) %>%
        select(Season,
               Week,
               Conference,
               Rank,
               Rating,
               `Strength of Schedule`,
               `Strength Remaining Schedule`,
               `Predicted Wins`,
               `Predicted Losses`,
               `Pr(Bowl Game)`,
               `Pr(Make Conf Champ)`,
               `Pr(Win Conf Champ)`,
               `Pr(Make Playoff)`,
               `Pr(Make National Champ)`,
               `Pr(Win National Champ)`) %>%
        gather() %>%
        mutate(key = str_pad(key, max(nchar(key))+1, pad=" ", side = "right")) %>%
        mutate(value = str_pad(value, max(nchar(value))+1, pad = " ", side = "left"))
     #   mutate(key = paste(key, ":", sep=""))
        #         
        # flextable() %>%
        # autofit() %>%
        # fontsize(., size = 11) %>%
        # flextable:: align(j = "value",
        #                   align = "center",
        #                   part = "all") %>%
        # delete_part(part = "header") %>%
        # as_raster(zoom = 1.5, expand = 1.5)


# team_preds1 = team_preds %>%
# #         filter(key %in% c("Team:", "Season:", "Conference:")) %>%
#         flextable() %>%
#         autofit() %>%
#         fontsize(., size = 11) %>%
#         flextable:: align(j = "value",
#                           align = "center",
#                           part = "all") %>%
#         delete_part(part = "header") %>%
#         as_raster()
# 
# team_preds2 = team_preds %>%
#         filter(!(key %in% c("Team:", "Season:", "Conference:"))) %>%
#         flextable() %>%
#         autofit() %>%
#         fontsize(., size = 11) %>%
#         flextable:: align(j = "value",
#                           align = "center",
#                           part = "all") %>%
#         delete_part(part = "header") %>%
#         as_raster()


        # mutate(key = paste(key, ": ", sep="")) %>%
        # mutate(key = str_pad(key, max(nchar(key))+1, pad=" ", side = "right")) %>%
        # mutate(value = str_pad(value, max(nchar(value))+1, pad = " ", side = "left")) %>%
        # mutate(together = paste(key, value, sep=" ")) %>%
        # pull(together) %>%
        # paste0(., collapse="\n")

```

```{r make score card scorecard, fig.height=3.5, fig.width=8.5}

# make text
team_text = c(teams_raw %>%
                          filter(TEAM == params$team) %>%                 
                          mutate(name = paste(TEAM, "\n", TEAM_MASCOT)) %>%
                          pull(name))

library(cowplot)
p1 = ggdraw() + draw_image(team_logos$logo_1, 
                           scale = 0.5, 
                           x=-0.1)+
        draw_label(teams_raw %>%
                          filter(TEAM == params$team) %>%                 
                          mutate(name = paste(TEAM, "\n", TEAM_MASCOT)) %>%
                          pull(name),
                  hjust = 0.5,
                  size = 14,
                  fontfamily = 'mono',
                  fontface = 'bold',
                  color = muted(team_color),
                  x = 0.4,
                  y = 0.15)

p2 = ggdraw() +
        draw_label(team_preds %>%
                          mutate(together = paste(key, value, sep=" ")) %>%
                          pull(together) %>%
                          paste0(., collapse="\n"),
                  fontfamily = 'mono',
                  fontface = 'bold',
                  hjust = 0.5,
                  color = muted(team_color))

plot_grid(p1, p2,
          nrow=1,
          ncol = 2,
          rel_widths = c(0.8, 1.3),
          rel_heights = c(1, 0.5))

```

## `r params$input_season` Game Predictions

Using each team's rating, I simulate the result and margin of victory for each remaining game of the `r params$input_season` season `r params$nsims` times. 

The following table shows the percentage of times in which `r params$team` won each of the games on their upcoming schedule. The margin of victory is the median margin of victory across all simulations.

```{r team schedule}

# table
games_schedule %>%
        filter(TEAM == params$team) %>%
        left_join(.,
                  games %>%
                          select(GAME_ID, HOME_TEAM, AWAY_TEAM),
                  by = c("GAME_ID")) %>%
        left_join(., team_mapping %>%
                          rename(OPPONENT = TEAM,
                                 OPPONENT_ABBREVIATION = TEAM_ABBREVIATION),
                  by = c("OPPONENT")) %>%
        mutate(OPPONENT = case_when(OPPONENT == HOME_TEAM ~ paste("@", OPPONENT_ABBREVIATION),
                                    OPPONENT == AWAY_TEAM ~ paste("vs", OPPONENT_ABBREVIATION))) %>%
        mutate(PRED = case_when(SIM_OUTCOME == 'win' ~ paste("win", "by", abs(SIM_MARGIN)),
                                SIM_OUTCOME == 'loss' ~paste("lose", "by", abs(SIM_MARGIN)))) %>%
        mutate(RESULT = case_when(OUTCOME == 'win' ~ paste("won", "by", abs(MARGIN)),
                                OUTCOME == 'loss' ~paste("lost", "by", abs(MARGIN)))) %>%
        mutate(CORRECT = case_when(is.na(OUTCOME) ~ NA_character_,
                                   SIM_OUTCOME == OUTCOME ~ '\u2713',
                                   SIM_OUTCOME != OUTCOME ~ '\u2715')) %>%
        select(SEASON, WEEK, GAME_DATE, TEAM, OPPONENT, SIM_PROB, SIM_MARGIN, PRED, RESULT, CORRECT) %>%
        #       PRED = expected_win) %>%
        mutate(SEASON = factor(SEASON)) %>%
        mutate(WEEK = factor(WEEK)) %>%
        arrange(GAME_DATE) %>%
        mutate(GAME_DATE = format(GAME_DATE, format="%b %d")) %>%
        mutate(GAME_DATE = factor(GAME_DATE)) %>%
        select(-SIM_MARGIN) %>%
        rename(Season = SEASON,
               Week = WEEK,
               Date = GAME_DATE,
               Team = TEAM,
               Opponent = OPPONENT,
               `Pr(Win)` = SIM_PROB,
               `Prediction` = PRED,
               `Result` = RESULT,
               `Correct?` = CORRECT) %>%
        flextable() %>%
        flextable::align(., j=c("Season", "Week", "Opponent", "Pr(Win)", "Prediction", "Result", "Correct?"),
                         align = "center",
                         part = "all") %>%
        bg(., j=c("Pr(Win)"),
           bg = team_col_func) %>%
        color(part = "all",
              color = "grey20") %>%
        color(., j = 'Correct?',
           i = ~ `Correct?` == '\u2713',
           color = team_color) %>%
                color(., j = 'Correct?',
           i = ~ `Correct?` != '\u2713',
           color = 'grey80')

```

It's important to note that within each simulation the results of each game are not independent. The (simulated) results of each week are used to update each team's rating going into the next week, which will then influence how likely they are to win games in the rest of the season. This means that a shocking upset early on in a simulation can have ramifications for other teams on the remaining schedule.

## Simulated Rating for Rest of Season

The table above indicates the aggregated result of every simulated game for the selected team. The visualization below shows the simulated paths for **`r params$team`** as of Week `r params$sim_week` in the `r params$input_season` season. This illustrates the most common paths left in a team's season by showing how the result of one game tends to affect its future trajectory.

```{r show simulated elo path for team}

source(here::here("functions", "plot_elo_team_season.R"))

p = plot_elo_team_season(sim_remaining_games %>%
                             filter(TEAM == params$team),
                     games_raw,
                     params$team,
                     team_color = team_color,
                     params$input_season,
                     alpha = 0.15)

p+
          ggtitle(paste("Simulated Elo Ratings for" , params$team, params$input_season, "as of Week", params$sim_week),
                        subtitle = str_wrap("Each line is one result from simulating a team's regular season. Win probabilities displayed above each matchup. Displaying results from 1000 simulations.", 120))

```

## Rating vs Preseason Expectations

How is `r params$team` performing compared to preseason expectations? The visualization below shows the team's rating so far compared to simulated ratings for the team from the start of the season. The simulated paths from the beginning of the season are shown in **grey** while the team's actual rating so far is shown in the team's color. Teams that are outperforming preseason expectations will be above the majority of the simulated paths; teams that are under performing will be below.

```{r plot from start}

p = plot_elo_team_season(sim_from_start%>%
                             filter(TEAM == params$team),
                     games_raw,
                     params$team,
                     team_color = 'grey70',
                     params$input_season,
                     alpha = 0.05)+
        geom_line(data = games_schedule %>%
                          filter(WEEK <= params$sim_week) %>%
                          filter(TEAM == params$team) %>%
                          group_by(SEASON, TEAM) %>%
                          mutate(GAME = row_number()) %>%
                          mutate(.id = params$nsims+1) %>%
                          bind_rows(.,
                                    games_schedule %>%
                                            filter(WEEK <= params$sim_week) %>%
                                            filter(TEAM == params$team) %>%
                                            group_by(SEASON, TEAM) %>%
                                            mutate(GAME = row_number()) %>%
                                            filter(GAME_DATE == max(GAME_DATE)) %>%
                                            mutate(GAME = GAME+1) %>%
                                            mutate(.id = params$nsims+1) %>%
                                            mutate(PREGAME_ELO = POSTGAME_ELO)),
                  color = team_color,
                  lwd = 1.3)

p+
          ggtitle(paste(params$team, params$input_season, "Rating vs Preseason Simulations"),
                        subtitle = str_wrap("Each line is one result from simulating a team's regular season. Win probabilities displayed above each matchup. Displaying results from 1000 simulations.", 120))+
        my_caption


```

## Conference and Playoff Probabilities

How has `r params$team`'s chance's of winning a conference championship and making/winning the playoffs changed over the course of the season?

```{r plot trajectory of conference}

load(here::here("data", "season_conference_divisions.Rdata"))

source(here::here("functions", "span_func.R"))

conference_divisions = season_conference_divisions %>%
        filter(SEASON == params$input_season)

# set plot max
max_week = max(sim_nc_games$SIM_FROM_WEEK)

# season
season = max(sim_nc_games$SEASON)

# create data
dat = conference_divisions %>%
        filter(TEAM == params$team) %>%
        mutate(CONFERENCE_DIVISION = replace_na(CONFERENCE_DIVISION, "No Divisions")) %>%
       # filter(CONFERENCE != 'FBS Independents') %>%
        nest(-SEASON) %>%
        bind_cols(., tibble(SIM_FROM_WEEK = seq(0, max_week))) %>%
        unnest()  %>%
        left_join(.,
                  sim_nc_games %>%
                          group_by(SEASON,
                                   SIM_FROM_WEEK,
                                   CONFERENCE,
                                   TEAM,
                                   SIM_OUTCOME) %>%
                          count() %>%
                          spread(SIM_OUTCOME, n) %>%
                          mutate_at(c("win", "loss"),
                                    ~ replace_na(., 0)) %>%
                          filter(TEAM == params$team),
                  by = c("SEASON", "TEAM", "SIM_FROM_WEEK", "CONFERENCE")) %>%
                mutate(type = 'Win National Championship') %>%
        bind_rows(.,
                  conference_divisions %>%
                          filter(TEAM == params$team) %>%
                          mutate(CONFERENCE_DIVISION = replace_na(CONFERENCE_DIVISION, "No Divisions")) %>%
                          # filter(CONFERENCE != 'FBS Independents') %>%
                          nest(-SEASON) %>%
                          bind_cols(., tibble(SIM_FROM_WEEK = seq(0, max_week))) %>%
                          unnest()  %>%
                          left_join(.,
                                    sim_playoff_games %>%
                                            group_by(SEASON,
                                                     SIM_FROM_WEEK,
                                                     CONFERENCE,
                                                     TEAM,
                                                     SIM_OUTCOME) %>%
                                            count() %>%
                                            spread(SIM_OUTCOME, n) %>%
                                            mutate_at(c("win", "loss"),
                                                      ~ replace_na(., 0)) %>%
                                            filter(TEAM == params$team),
                                    by = c("SEASON", "TEAM", "SIM_FROM_WEEK", "CONFERENCE")) %>%
                          mutate(type = 'Make Playoffs')
        ) %>%
            bind_rows(.,
                  conference_divisions %>%
                          filter(TEAM == params$team) %>%
                          mutate(CONFERENCE_DIVISION = replace_na(CONFERENCE_DIVISION, "No Divisions")) %>%
                          # filter(CONFERENCE != 'FBS Independents') %>%
                          nest(-SEASON) %>%
                          bind_cols(., tibble(SIM_FROM_WEEK = seq(0, max_week))) %>%
                          unnest()  %>%
                          left_join(.,
                                    sim_nc_games %>%
                                            group_by(SEASON,
                                                     SIM_FROM_WEEK,
                                                     CONFERENCE,
                                                     TEAM,
                                                     SIM_OUTCOME) %>%
                                            count() %>%
                                            spread(SIM_OUTCOME, n) %>%
                                            mutate_at(c("win", "loss"),
                                                      ~ replace_na(., 0)) %>%
                                            filter(TEAM == params$team),
                                    by = c("SEASON", "TEAM", "SIM_FROM_WEEK", "CONFERENCE")) %>%
                          mutate(type = 'Make National Championship')
        ) %>%
        bind_rows(.,
                    conference_divisions %>%
                          filter(TEAM == params$team) %>%
                          mutate(CONFERENCE_DIVISION = replace_na(CONFERENCE_DIVISION, "No Divisions")) %>%
                          # filter(CONFERENCE != 'FBS Independents') %>%
                          nest(-SEASON) %>%
                          bind_cols(., tibble(SIM_FROM_WEEK = seq(0, max_week))) %>%
                          unnest()  %>%
                          left_join(.,
                                    sim_conference_championship_games %>%
                                            group_by(SEASON,
                                                     SIM_FROM_WEEK,
                                                     CONFERENCE,
                                                     TEAM,
                                                     SIM_OUTCOME) %>%
                                            count() %>%
                                            spread(SIM_OUTCOME, n) %>%
                                            mutate_at(c("win", "loss"),
                                                      ~ replace_na(., 0)) %>%
                                            filter(TEAM == params$team),
                                    by = c("SEASON", "TEAM", "SIM_FROM_WEEK", "CONFERENCE")) %>%
                          mutate(type = 'Win Conference Championship')
        ) %>%
        mutate_at(c("win", "loss"),
                  ~ replace_na(.,0)) %>%
        mutate(perc = case_when(type == 'Make National Championship' |
                                        type == 'Make Playoffs' ~ (win+loss)/params$nsims,
                                TRUE ~ (win) / params$nsims)) %>%
        group_by(TEAM, type) %>%
        mutate(max_perc = max(perc)) %>%
        ungroup() %>%
     #   filter(max_perc >=0.01) %>%
      mutate(TEAM_LABEL = case_when(SIM_FROM_WEEK == max(SIM_FROM_WEEK) ~ 
                                            paste(TEAM, "-", paste(round(perc*100,0),"%", sep=""))))


p = dat %>%
        mutate(type = factor(type,
                             levels = c("Win Conference Championship",
                                        "Make Playoffs",
                                        "Make National Championship",
                                        "Win National Championship"))) %>%
        ggplot(., aes(x=SIM_FROM_WEEK,
                      label = TEAM_LABEL,
                      color = TEAM,
                      y=perc))+
                geom_point(size=2)+
        facet_wrap(type ~.,
                   ncol = 2)+
        geom_line(stat = 'smooth',
                  method = span_func(max_week)$method,
                  se = F,
                  formula = 'y ~ x',
                  lwd = 1,
                 span = span_func(max_week)$span,
                  alpha = 0.85)+
        geom_text_repel(
                fontface = "bold",
                size = 3,
                direction = "y",
                hjust = -1.2,
                segment.size = .7,
                segment.alpha = .5,
                segment.linetype = "dotted",
                box.padding = .4,
                segment.curvature = -0.1,
                segment.ncp = 3,
                segment.angle = 20)+
        custom_color_teams+
        guides(color = "none")+
        coord_cartesian(xlim = c(-3, 18),
                        ylim = c(0,1))+
        theme_phil()+
        xlab("Season Week")+
        ggtitle(paste(params$input_season, params$team, "Conference and Playoff Probabilities as of Week", max_week),
                subtitle = str_wrap(paste("Displaying the percentage of times each team achieved each outcome in simulations of the remaining season, conference championsip games, and playoff games. Simulations run at the beginning of the season and after every week of regular season games."), 120))+
        theme(plot.title = element_text(hjust = 0.5, size = 16),
              plot.subtitle =  element_text(hjust = 0.5),
              strip.text.x = element_text(size = 12))+
        ylab("Probability of Outcome")+
        my_caption

p

```


# Historical Team Data

Simulations for an upcoming season make use of an (adjusted) Elo rating, which is a rating assigned to each team based on wins and losses in previous games. After wins, a team's rating will go up; after losses, a team's rating will go down. The magnitude of these increases/decreases will vary depend on the scoreline and strength of the opponent.

## Elo Rating

The following visualizations shows `r params$team`'s Elo rating from 1900 to `r params$input_season`

```{r plot team historicalhistorical}

# dates
#min_date = min(elo_teams$GAME_DATE)[1]
min_date = as.Date(paste("1900-01-01"))
min_year = year(min_date)
#min_year = 1900

if(params$team == 'Florida State' |
   params$team == 'Connecticut') {smooth = F}else 
           {smooth =T}

all_teams_data = elo_teams %>%
        filter(SEASON < params$input_season) %>%
                filter(SEASON > min_year) %>%
                filter(DIVISION == 'fbs') 
        
all_teams_median = all_teams_data %>%
                filter(SEASON < params$input_season) %>%
                summarize(median = median(POSTGAME_ELO)) %>%
                pull(median)
        
p = all_teams_data %>%
                filter(SEASON < params$input_season) %>%
                ggplot(., aes(x=GAME_DATE,
                              by = TEAM,
                              y= POSTGAME_ELO))+
                geom_point(alpha = 0.005)+
                geom_hline(yintercept = all_teams_median,
                           linetype = 'dashed',
                           color = 'grey60')+
                annotate("text",
                         x = min_date,
                         y=all_teams_median + 40,
                         size = 3,
                         color = 'grey60',
                         label = paste("FBS", "\n", "Median"))
        
team_data = elo_teams %>%
                filter(SEASON < params$input_season) %>%
                filter(SEASON > min_year) %>%
                filter(TEAM %in% team) %>%
                mutate(SELECTED_TEAM = TEAM) %>%
                group_by(TEAM) %>%
                mutate(TEAM_LABEL = case_when(GAME_DATE == max(GAME_DATE) ~ TEAM)) %>%
                ungroup()
        
        
if(smooth == T ) {p = p + geom_line(data =team_data,
                          aes(color = TEAM),
                          stat = 'smooth',
                          formula = 'y ~ x',
                          method = 'loess',
                          alpha =0.85,
                          span = 0.15,
                          lwd = 1.04)} else {p = p}
        
        p + 
                                geom_point(data = team_data,
                           aes(color = TEAM),
                           alpha  =0.4)+
                geom_label_repel(data = team_data %>%
                                        filter(POSTGAME_ELO == min(POSTGAME_ELO)),
                                aes(label = paste("Lowest Rating:", "\n", GAME_DATE, "vs", OPPONENT),
                                    color =TEAM),
                        fontface = "bold",
                        size = 3,
                  #      direction = "y",
                        # vjust = 
                        # hjust = -1.2,
                      #  segment.size = .7,
                       # segment.alpha = .5,
                        # segment.linetype = "dotted",
                        # box.padding = .4,
                        # segment.curvature = -0.1,
                        # segment.ncp = 3,
                        # segment.angle = 20
                ) +
                  geom_label_repel(data = team_data %>%
                                        filter(POSTGAME_ELO == max(POSTGAME_ELO)),
                                aes(label = paste("Highest Rating:", "\n", GAME_DATE, "vs", OPPONENT),
                                    color =TEAM),
                        fontface = "bold",
                        size = 3,
                      #  direction = "y",
                        # vjust =
                        # hjust = -1.2,
                      #  segment.size = .7,
                       # segment.alpha = .5,
                        # segment.linetype = "dotted",
                        # box.padding = .4,
                        # segment.curvature = -0.1,
                        # segment.ncp = 3,
                        # segment.angle = 20
                ) +
                theme_phil()+
                ggtitle(paste("Historical Elo Rating for", paste(team, collapse="-")),
                        subtitle = str_wrap(paste('Displaying (postgame) Elo Ratings for all FBS teams from',
                                                  min_year, 'to', params$input_season, '.Highlighted points/line indicate selected team.'), 120))+
                theme(plot.title = element_text(hjust = 0.5, size = 16),
                      plot.subtitle =  element_text(hjust = 0.5),
                      strip.text.x = element_text(size = 12))+
                custom_color_teams+
                guides(color = 'none')+
                xlab("Game Date")+
                ylab("Postgame Elo Rating")+
                scale_x_date(breaks = as.Date(paste(seq(min_year, 2020, 10), "01", "01", sep="-")),
                             #date_breaks = c("10 year"),
                             date_labels = c("%Y"))+
                xlab("Season (Game Date)")+
                my_caption

```

<!-- ## Best Seasons -->

<!-- `r params$team`'s 10 best teams by end of season Elo rating -->

<!-- ```{r best teams for team based on elo} -->

<!-- elo_teams %>% -->
<!--         left_join(., elo_games %>% -->
<!--                           select(GAME_ID, NEUTRAL_SITE), -->
<!--                   by = c("GAME_ID")) %>% -->
<!--         filter(SEASON >= 1950) %>% -->
<!--         filter(TEAM == team) %>% -->
<!--         group_by(SEASON) %>% -->
<!--         filter(GAME_DATE == max(GAME_DATE)) %>% -->
<!--         ungroup() %>% -->
<!--         rename(ELO = POSTGAME_ELO) %>% -->
<!--         select(SEASON, TEAM, ELO) %>% -->
<!--         mutate_if(is.numeric, round, 2) %>% -->
<!--         arrange(desc(ELO)) %>% -->
<!--         mutate(SEASON = factor(SEASON)) %>% -->
<!--         mutate_if(is.numeric, round, 0) %>% -->
<!--         mutate(Rank = row_number()) %>% -->
<!--         rename(Season = SEASON, -->
<!--                Team = TEAM, -->
<!--                Rating = ELO) %>% -->
<!--         select(Rank, everything()) %>% -->
<!--         head(15) %>% -->
<!--         flextable() %>% -->
<!--         autofit() -->

<!-- ``` -->

<!-- ## Biggest Upsets for `r params$team` -->

<!-- Top 15 games that `r params$team` **won** in which were they were favored to win (since 1950). -->

<!-- ```{r historical upsets games should not have won} -->

<!-- # upset -->
<!-- elo_teams %>% -->
<!--         left_join(., elo_games %>% -->
<!--                           select(GAME_ID, NEUTRAL_SITE), -->
<!--                   by = c("GAME_ID")) %>% -->
<!--         filter(TEAM == team) %>% -->
<!--         filter(SEASON >=1950) %>% -->
<!--         filter(OUTCOME == 'win') %>% -->
<!--         arrange(WIN_PROB) %>% -->
<!--         rename(DATE = GAME_DATE) %>% -->
<!--         mutate(SCORE = paste(POINTS, "-", OPPONENT_POINTS)) %>% -->
<!--         mutate(OPPONENT = case_when(HOME == F ~ paste("@", OPPONENT), -->
<!--                                     HOME == T ~ paste("vs", OPPONENT))) %>% -->
<!--         mutate(RANK = row_number()) %>% -->
<!--         rename(ELO = PREGAME_ELO, -->
<!--                OPPONENT_ELO = OPPONENT_PREGAME_ELO) %>% -->
<!--         mutate(ELO_DIFF = case_when(HOME == T & NEUTRAL_SITE ==F ~ ((75 + ELO) - OPPONENT_ELO), -->
<!--                                     HOME == F & NEUTRAL_SITE == F~ -((OPPONENT_ELO + 75) - ELO), -->
<!--                                     NEUTRAL_SITE==T~ ELO - OPPONENT_ELO)) %>% -->
<!--         mutate(PRED_MARGIN = round(.05 * ELO_DIFF, 1))  %>% -->
<!--                 mutate(OPPONENT_ELO = round(OPPONENT_ELO, 0), -->
<!--                ELO = round(ELO, 0)) %>% -->
<!--         select(RANK, SEASON, WEEK, TEAM, OPPONENT, ELO, OPPONENT_ELO, WIN_PROB, PRED_MARGIN, OUTCOME, SCORE) %>% -->
<!--         mutate(SEASON = factor(SEASON)) %>% -->
<!--         rename(Rank = RANK, -->
<!--                Season = SEASON, -->
<!--                Week = WEEK, -->
<!--                Team = TEAM, -->
<!--                Team.Rating = ELO, -->
<!--                Opponent = OPPONENT, -->
<!--                Opp.Rating = OPPONENT_ELO, -->
<!--                Prob.Win = WIN_PROB, -->
<!--                Pred.Margin = PRED_MARGIN, -->
<!--                Outcome = OUTCOME, -->
<!--                Score = SCORE) %>% -->
<!--         mutate_if(is.numeric, round, 3) %>% -->
<!--         head(15) %>% -->
<!--         flextable() %>% -->
<!--         autofit() %>% -->
<!--         bg(., j= c("Prob.Win"), -->
<!--            bg = team_col_func) %>% -->
<!--         flextable::align(., j=c("Prob.Win", "Outcome", "Score"), -->
<!--                       align = "center", -->
<!--                       part = "all") %>% -->
<!--                 bg(., j=c("Prob.Win"), -->
<!--                    bg = team_col_func) %>% -->
<!--                 color(part = "all", -->
<!--                       color = "grey20") -->

<!-- ``` -->


<!-- ## Biggest Upsets against `r params$team` -->
<!-- Top 15 games that `r params$team` **lost** in which were they were favored to win (since 1950). -->

<!-- ```{r historical upsets} -->

<!-- # upset -->
<!-- elo_teams %>% -->
<!--         left_join(., elo_games %>% -->
<!--                           select(GAME_ID, NEUTRAL_SITE), -->
<!--                   by = c("GAME_ID")) %>% -->
<!--         filter(TEAM == team) %>% -->
<!--                 filter(SEASON >=1950) %>% -->
<!--         filter(OUTCOME == 'loss') %>% -->
<!--         arrange(desc(WIN_PROB)) %>% -->
<!--         rename(DATE = GAME_DATE) %>% -->
<!--         mutate(SCORE = paste(POINTS, "-", OPPONENT_POINTS)) %>% -->
<!--         mutate(OPPONENT = case_when(HOME == F ~ paste("@", OPPONENT), -->
<!--                                     HOME == T ~ paste("vs", OPPONENT))) %>% -->
<!--         mutate(RANK = row_number()) %>% -->
<!--         rename(ELO = PREGAME_ELO, -->
<!--                OPPONENT_ELO = OPPONENT_PREGAME_ELO) %>% -->
<!--         mutate(ELO_DIFF = case_when(HOME == T & NEUTRAL_SITE ==F ~ ((75 + ELO) - OPPONENT_ELO), -->
<!--                                     HOME == F & NEUTRAL_SITE == F~ -((OPPONENT_ELO + 75) - ELO), -->
<!--                                     NEUTRAL_SITE==T~ ELO - OPPONENT_ELO)) %>% -->
<!--         mutate(PRED_MARGIN = round(.05 * ELO_DIFF, 1))  %>% -->
<!--                 mutate(OPPONENT_ELO = round(OPPONENT_ELO, 0), -->
<!--                ELO = round(ELO, 0)) %>% -->
<!--         select(RANK, SEASON, WEEK, TEAM, OPPONENT, ELO, OPPONENT_ELO, WIN_PROB, PRED_MARGIN, OUTCOME, SCORE) %>% -->
<!--         mutate(SEASON = factor(SEASON)) %>% -->
<!--         rename(Rank = RANK, -->
<!--                Season = SEASON, -->
<!--                Week = WEEK, -->
<!--                Team = TEAM, -->
<!--                Team.Rating = ELO, -->
<!--                Opponent = OPPONENT, -->
<!--                Opp.Rating = OPPONENT_ELO, -->
<!--                Prob.Win = WIN_PROB, -->
<!--                Pred.Margin = PRED_MARGIN, -->
<!--                Outcome = OUTCOME, -->
<!--                Score = SCORE) %>% -->
<!--         mutate_if(is.numeric, round, 3) %>% -->
<!--         head(15) %>% -->
<!--         flextable() %>% -->
<!--         autofit() %>% -->
<!--         bg(., j= c("Prob.Win"), -->
<!--            bg = team_col_func) %>% -->
<!--         flextable::align(., j=c("Prob.Win", "Outcome", "Score"), -->
<!--                       align = "center", -->
<!--                       part = "all") %>% -->
<!--                 bg(., j=c("Prob.Win"), -->
<!--                    bg = team_col_func) %>% -->
<!--                 color(part = "all", -->
<!--                       color = "grey20") -->

<!-- ``` -->


## Efficiency

In addition to a team's historical Elo rating, I also compute and use each team's offensive/defensive efficiency metrics from recent seasons to predict their expected future Elo rating. These efficiency metrics come from an expected points model I trained on on play by play data in order to identify the value of individual plays within a game. 

I score every play with the model and then aggregate these results to the game and season level and adjust for opponent quality. These results are only available starting from the season of 2007, as this is when reliable play by play data becomes available. 

For more details on what goes into creating these ratings, go to my [description of the expected points model](https://phenrickson.github.io/cfb/expected_points_01_model.html) and my [methodology for adjusting for opponent quality](https://phenrickson.github.io/cfb/expected_points_03_adjusted.html).

```{r table form for overall}

## make color func
col_type_func = 
        function(x) {
                
                breaks = c(-1, -0.9, -.75, -.5, -.4, -.35, -.3, -.25, -.2, -.15, -.1, -0.05, -0.025, -0.01, -0.001, 0, 0.001, 0.01, 0.025, 0.05, 0.1, 0.15, 0.2, 0.25, 0.3, 0.35, 0.4, 0.5, 0.75, 0.9, 1)
                
                colorRamp=colorRampPalette(c("red", "white", "deepskyblue1"))
                col_palette <- colorRamp(length(breaks))
                mycut <- cut(x, 
                             breaks = breaks,
                             include.lowest = TRUE, 
                             right=T,
                             label = FALSE)
                col_palette[mycut]
                
        }

# overall
season_team_rankings %>%
        filter(TYPE == 'adjusted') %>%
        filter(TEAM == team) %>% 
        arrange(SEASON) %>%
        select(SEASON, TEAM, OFFENSE, DEFENSE, OVERALL) %>%
        mutate(SEASON = factor(SEASON)) %>%
        flextable() %>%
        autofit() %>%
        bg(., j= c("OFFENSE",
                   "DEFENSE",
                   "OVERALL"),
           bg = col_type_func) %>%
        flextable::align(j = c("OFFENSE",
                               "DEFENSE",
                               "OVERALL"),
                         align = 'center',
                         part = 'all')

```

These efficiency ratings indicate a team's **expected points per play** when its offense or defense is on the field, adjusted for opponent quality. A team's overall rating is a combination of its offense and defense ratings, and indicates the net points per play a team would expect when playing an average opponent. 

For offenses, this indicates the average points the team scores against opponents per play. For defenses, this indicates the average points the team prevents opposing offenses from scoring per play. In both cases, I have set the scale of the variable to mean that positive is good while negative is bad.

The visualization below shows how a team has changed over time as well as their season end efficiency ranking (out of all FBS teams).

```{r efficiency, fig.height=9}

plot_dat = season_team_rankings %>%
        filter(SEASON < params$input_season) %>%
        select(-MARGIN) %>%
        filter(TYPE == 'adjusted') %>%
        gather("metric", "value",
               -SEASON, -TEAM, -TYPE) %>%
        left_join(., team_mapping,
                  by = c("TEAM")) %>%
        left_join(., team_conference_season,
                  by = c("TEAM", "SEASON")) %>%
        mutate(SEASON = factor(SEASON)) %>%
        mutate(metric = factor(metric, levels = c("OVERALL", "OFFENSE", "DEFENSE"))) %>%
        group_by(SEASON, TYPE, metric) %>%
        arrange(desc(value)) %>%
        mutate(RANK = row_number()) %>%
        ungroup() %>%
        group_by(metric) %>%
        mutate(min = min(value)) %>%
        ungroup()

all_teams = plot_dat %>%
        ggplot(., aes(x=SEASON,
                      y=value))+
        geom_point(position = position_auto(scale=F,
                                            seed = 1),
                           #position_jitternormal(sd_x= 0.12,
                            #                        sd_y = 0),
            #       fill = 'grey60',
             #      color = 'white',
            color = 'grey60',
            alpha = 0.4)+
        facet_wrap(metric ~.,
                   scales = "free_y",
                   ncol = 1)+
        theme_phil()+
        geom_hline(yintercept =0)+
        guides(color = 'none')+
        ylab("Team Net Predicted Points per Play")

all_teams + 
        geom_point(data = plot_dat %>%
                          filter(TEAM == team),
                  aes(color = TEAM,
                      group = TEAM),
                  lwd = 1.2)+
        geom_line(data = plot_dat %>%
                          filter(TEAM == team),
                  aes(color = TEAM,
                      group = TEAM),
                  lwd = 1.2)+
        geom_text(data = plot_dat %>%
                          filter(TEAM == team),
                  aes(color = TEAM,
                      x = SEASON,
                      label = RANK,
                      y = min),
                  size = 4)+
        custom_color_teams+
        guides(color = 'none')+
        theme(plot.title = element_text(hjust = 0.5,
                                                size = 18),
              axis.text.x = element_text(size = 12),
              plot.subtitle =  element_text(hjust = 0.5),
              strip.text.x = element_text(size = 12,
                                                  hjust = 0.5))+
        ggtitle(paste("College Football Efficiency Metrics by Season - ", team),
                subtitle = str_wrap(paste("Displaying opponent adjusted ratings for all teams based on net points added per play. The background points in grey show the distribution of all teams in a given season. The highlighted line shows ", team, "'s rating by season with their season rank above the year.", sep=""), 140))+
        xlab("")+
        my_caption

```

### Efficiency by Play Type

In addition to examinig how a team performs overall, we can examine each team's efficiency based on the play type. How has `r params$team` performed when running/passing on offense vs defending the run/pass on defense?

```{r efficiency by type table}

# by type
season_team_rankings_type %>%
        filter(SEASON < params$input_season) %>%
        filter(TEAM == team) %>%
        filter(TYPE == 'adjusted') %>%
        arrange(SEASON) %>%
        mutate(SEASON = factor(SEASON)) %>%
        rename(Season = SEASON,
               Team = TEAM,
               Pass_Offense = OFFENSE_PASS,
               Pass_Defense = DEFENSE_PASS,
               Run_Offense = OFFENSE_RUN,
               Run_Defense = DEFENSE_RUN) %>%
        select(Season, Team, Pass_Offense, Run_Offense, Pass_Defense, Run_Defense) %>%
      #  select(SEASON, TEAM, OFFENSE_PASS, OFFENSE_RUN, DEFENSE_PASS, DEFENSE_RUN) %>%
        flextable() %>%
        autofit() %>%
        bg(., j= c("Run_Offense", "Pass_Offense",
                   "Run_Defense", "Pass_Defense"),
           bg = col_type_func) %>%
        flextable::align(j = c("Run_Offense", "Pass_Offense",
                   "Run_Defense", "Pass_Defense"),
              align = 'center')
```

The following visualization shows how efficient `r params$team`'s **offense** has been running/passing the ball each season since 2007.

```{r look at offense defense by type, fig.height=8}

plot_dat_type = season_team_rankings_type %>%
        filter(SEASON < params$input_season) %>%
        filter(TYPE == 'adjusted') %>%
        gather("metric", "value",
               -SEASON, -TEAM, -TYPE) %>%
        filter(metric %in% c("OFFENSE_PASS", "OFFENSE_RUN","DEFENSE_PASS", "DEFENSE_RUN")) %>%
        separate(metric, into = c("SIDE", "PLAY"), sep="_") %>%
        left_join(., team_mapping,
                  by = c("TEAM")) %>%
        left_join(., team_conference_season,
                  by = c("TEAM", "SEASON")) %>%
        mutate(SEASON = factor(SEASON)) %>%
        mutate(PLAY = factor(PLAY, levels = c("PASS", "RUN")),
               SIDE = factor(SIDE, levels = c("OFFENSE", "DEFENSE"))) %>%
        group_by(SEASON, TYPE, SIDE, PLAY) %>%
        arrange(desc(value)) %>%
        mutate(RANK = row_number()) %>%
        ungroup() %>%
        group_by(TYPE, SIDE, PLAY) %>%
        mutate(min = min(value)) %>%
        ungroup()

all_teams_offense_type = plot_dat_type %>%
        filter(SIDE == 'OFFENSE') %>%
        ggplot(., aes(x=SEASON,
                      y=value))+
        geom_point(position = position_auto(scale=F,
                                            seed = 1),
                           #position_jitternormal(sd_x= 0.12,
                            #                        sd_y = 0),
            #       fill = 'grey60',
             #      color = 'white',
            color = 'grey60',
            alpha = 0.4)+
        facet_wrap(SIDE + PLAY ~.,
                   scales = "free_y",
                   ncol = 1)+
        theme_phil()+
        geom_hline(yintercept =0)+
        guides(color = 'none')+
        ylab("Team Net Predicted Points per Play")

all_teams_offense_type + 
       geom_point(data = plot_dat_type %>%
                          filter(SIDE == 'OFFENSE') %>%
                          filter(TEAM == team),
                  aes(color = TEAM,
                      group = TEAM),
                  lwd = 1.2)+
        geom_line(data = plot_dat_type %>%
                          filter(SIDE == 'OFFENSE') %>%
                          filter(TEAM == team),
                  aes(color = TEAM,
                      group = TEAM),
                  lwd = 1.2)+
        geom_text(data = plot_dat_type %>%
                          filter(SIDE == 'OFFENSE') %>%
                          filter(TEAM == team),
                  aes(color = TEAM,
                      x = SEASON,
                      label = RANK,
                      y = min))+
        custom_color_teams+
        guides(color = 'none')+
        theme(plot.title = element_text(hjust = 0.5,
                                                size = 18),
              axis.text.x = element_text(size = 12),
              plot.subtitle =  element_text(hjust = 0.5),
              strip.text.x = element_text(size = 12,
                                                  hjust = 0.5))+
        ggtitle(paste("College Football Offense Efficiency by Season - ", team),
                subtitle = str_wrap(paste("Displaying opponent adjusted ratings for all teams based on net points added per play. The background points in grey show the distribution of all teams in a given season. The highlighted line shows ", team, "'s rating by season with their season rank above the year.", sep=""), 140))+
        xlab("")+
        my_caption
```

The following visualization shows how efficient `r params$team`'s **defense** has been in stopping the opponent's run/pass in each season since 2007.

```{r defense efficiency type, fig.height=8}

plot_dat_type = season_team_rankings_type %>%
        filter(SEASON < params$input_season) %>%
        filter(TYPE == 'adjusted') %>%
        gather("metric", "value",
               -SEASON, -TEAM, -TYPE) %>%
        filter(metric %in% c("OFFENSE_PASS", "OFFENSE_RUN","DEFENSE_PASS", "DEFENSE_RUN")) %>%
        separate(metric, into = c("SIDE", "PLAY"), sep="_") %>%
        left_join(., team_mapping,
                  by = c("TEAM")) %>%
        left_join(., team_conference_season,
                  by = c("TEAM", "SEASON")) %>%
        mutate(SEASON = factor(SEASON)) %>%
        mutate(PLAY = factor(PLAY, levels = c("PASS", "RUN")),
               SIDE = factor(SIDE, levels = c("OFFENSE", "DEFENSE"))) %>%
        group_by(SEASON, TYPE, SIDE, PLAY) %>%
        arrange(desc(value)) %>%
        mutate(RANK = row_number()) %>%
        ungroup() %>%
        group_by(TYPE, SIDE, PLAY) %>%
        mutate(min = min(value)) %>%
        ungroup()

all_teams_defense_type = plot_dat_type %>%
        filter(SIDE == 'DEFENSE') %>%
        ggplot(., aes(x=SEASON,
                      y=value))+
        geom_point(position = position_auto(scale=F,
                                            seed = 1),
                           #position_jitternormal(sd_x= 0.12,
                            #                        sd_y = 0),
            #       fill = 'grey60',
             #      color = 'white',
            color = 'grey60',
            alpha = 0.4)+
        facet_wrap(SIDE + PLAY ~.,
                   scales = "free_y",
                   ncol = 1)+
        theme_phil()+
        geom_hline(yintercept =0)+
        guides(color = 'none')+
        ylab("Team Net Predicted Points per Play")

all_teams_defense_type + 
       geom_point(data = plot_dat_type %>%
                          filter(SIDE == 'DEFENSE') %>%
                          filter(TEAM == team),
                  aes(color = TEAM,
                      group = TEAM),
                  lwd = 1.2)+
        geom_line(data = plot_dat_type %>%
                          filter(SIDE == 'DEFENSE') %>%
                          filter(TEAM == team),
                  aes(color = TEAM,
                      group = TEAM),
                  lwd = 1.2)+
        geom_text(data = plot_dat_type %>%
                          filter(SIDE == 'DEFENSE') %>%
                          filter(TEAM == team),
                  aes(color = TEAM,
                      x = SEASON,
                      label = RANK,
                      y = min))+
        custom_color_teams+
        guides(color = 'none')+
        theme(plot.title = element_text(hjust = 0.5,
                                                size = 18),
              axis.text.x = element_text(size = 12),
              plot.subtitle =  element_text(hjust = 0.5),
              strip.text.x = element_text(size = 12,
                                                  hjust = 0.5))+
        ggtitle(paste("College Football Defense Efficiency by Season - ", team),
                subtitle = str_wrap(paste("Displaying opponent adjusted ratings for all teams based on net points added per play. The background points in grey show the distribution of all teams in a given season. The highlighted line shows ", team, "'s rating by season with their season rank above the year.", sep=""), 140))+
        xlab("")+
        my_caption

```

```{r run gc, include=F}

rm(p)
gc()

```

