---
title: "model_sequences"
author: "Phil Henrickson"
date: "6/7/2022"
output: 
  html_document:
    toc: TRUE #adds a Table of Contents
    theme: cerulean
    number_sections: TRUE #number your headings/sections
    toc_float: TRUE #let your ToC follow you as you scroll
    keep_md: no
    fig.caption: yes
    css: "styles.css"
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = F,
                      error = F,
                      warning=F,
                      dev="png",
                      fig.width = 8,
                      fig.height = 4)

options(knitr.duplicate.label = "allow")

options(scipen=999)

```


```{r connect to snowflake}

library(DBI)
library(odbc)
library(RODBC)
library(keyring)

# connect to snowflake
myconn <- DBI::dbConnect(odbc::odbc(),
                         "SnowflakeDSII",
                         Database = "CFB_DEMO",
                         warehouse = "DEMO_WH",
                         uid="phil.henrickson",
                         pwd=keyring::key_get("AE_Snowflake"))

```

```{r packages, include=F} 

source(here::here("scripts/load_packages.R"))
library(jsonlite)
library(forcats)
conflict_prefer("lag", "dplyr")

```

```{r load functions}

source(here::here("functions/theme_phil.R"))
source(here::here("functions/clean_plays_func.R"))
source(here::here("functions/make_time_features_func.R"))
source(here::here("functions/plot_scoreline_func.R"))
rm(a)
```


```{r query data from snowflake}

### pull from snowflake
# get plays
plays_data_raw = DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.PLAY_BY_PLAY')) %>%
        as_tibble() %>%
        mutate(TIME = gsub('\\{|}|"', '', CLOCK)) %>%
        separate(TIME, into=c("MINUTES", "SECONDS"), sep=",") %>%
        mutate(MINUTES = as.numeric(gsub("minutes:", "", MINUTES))) %>% 
        mutate(SECONDS = as.numeric(gsub("seconds:", "", SECONDS))) %>%
        mutate(MINUTES_IN_HALF = case_when(PERIOD == 1 ~ MINUTES+ 15,
                                           PERIOD == 2 ~ MINUTES,
                                           PERIOD == 3 ~ MINUTES + 15,
                                           PERIOD == 4 ~ MINUTES)) %>%
        mutate(SECONDS_IN_HALF = MINUTES_IN_HALF*60 + SECONDS) %>%
        # flag if seconds in half is invalid
        mutate(FLAG_SECONDS_IN_HALF = case_when(SECONDS_IN_HALF > 1800 ~ 1,
                                                TRUE ~ 0)) %>%
        # convert IDs to numeric
        mutate(ID = as.numeric(ID),
               DRIVE_ID = as.numeric(DRIVE_ID)) %>%
        rename(PLAY_ID = ID) %>%
        group_by(GAME_ID) %>%
        # sort each game by drive
        arrange(PLAY_ID) %>% 
        ungroup() %>%
        # clean up PERIOD by game
        mutate(STATUS_PERIOD = case_when(PERIOD == 0 & nchar(PLAY_TEXT) <=1 ~ 'Drop',
                                        PERIOD == 0 ~ 'Take Previous Value',
                                        TRUE ~ 'Valid')) %>%
        filter(STATUS_PERIOD != 'Drop') %>%
        mutate(PERIOD = case_when(STATUS_PERIOD == 'Take Previous Value' ~ lag(PERIOD, 1),
                                  TRUE ~ PERIOD)) %>%
        ungroup() %>% 
        mutate(HALF = case_when(PERIOD == 1 ~ 'First Half',
                                PERIOD == 2 ~ 'First Half',
                                PERIOD == 3 ~ 'Second Half',
                                PERIOD == 4 ~ 'Second Half',
                                TRUE ~ 'OT')) %>%
        # then clean up the down
        mutate(FLAG_DOWN = case_when(DOWN %in% c(1, 2, 3, 4) ~ 0,
                                       TRUE ~ 1)) %>%
        mutate(DOWN = case_when(DOWN %in% c(1, 2, 3, 4) ~ DOWN,
                                TRUE ~ -1)) %>%
        # then flag the distance
        mutate(FLAG_DISTANCE = case_when((DISTANCE < 0 | DISTANCE >=99) ~ 1,
                                         TRUE ~ 0)) %>%
        # flag yard lines that are outside correct
        mutate(FLAG_YARD_LINE = case_when(YARD_LINE < 0 | YARD_LINE > 100 ~ 1,
                                          TRUE ~ 0))

# get games
games_data_raw = DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.GAMES')) %>%
        as_tibble() %>%
        mutate(ID = as.numeric(ID)) %>%
        rename(GAME_ID = ID)
        
# get drives
drives_data_raw = DBI::dbGetQuery(myconn,
                              paste('SELECT * FROM CFB_DEMO.CFD_RAW.DRIVES')) %>%
        as_tibble() %>%
        mutate(ID = as.numeric(ID),
               GAME_ID = as.numeric(GAME_ID)) %>%
        rename(DRIVE_ID = ID) %>%
        # get drive time features
        # start of drive
        mutate(START_TIME = gsub('\\{|}|"', '', START_TIME)) %>%
        separate(START_TIME, into=c("START_MINUTES", "START_SECONDS"), sep=",") %>%
        mutate(START_MINUTES = as.numeric(gsub("minutes:", "", START_MINUTES))) %>% 
        mutate(START_SECONDS = as.numeric(gsub("seconds:", "", START_SECONDS))) %>%
        # end of drive
        mutate(END_TIME = gsub('\\{|}|"', '', END_TIME)) %>%
        separate(END_TIME, into=c("END_MINUTES", "END_SECONDS"), sep=",") %>%
        mutate(END_MINUTES = as.numeric(gsub("minutes:", "", END_MINUTES))) %>% 
        mutate(END_SECONDS = as.numeric(gsub("seconds:", "", END_SECONDS))) %>%
        # duration of drive
        mutate(ELAPSED = gsub('\\{|}|"', '', ELAPSED)) %>%
        separate(ELAPSED, into=c("ELAPSED_MINUTES", "ELAPSED_SECONDS"), sep=",") %>%
        mutate(ELAPSED_MINUTES = as.numeric(gsub("minutes:", "", ELAPSED_MINUTES))) %>% 
        mutate(ELAPSED_SECONDS = as.numeric(gsub("seconds:", "", ELAPSED_SECONDS))) 
        
```

```{r apply clean up functions}

# now get scoring events for the drives data
drives_data_score_events = drives_data_raw %>%
        group_by(GAME_ID) %>%
        arrange(DRIVE_ID) %>%
        mutate(HALF = case_when(START_PERIOD == 1 | START_PERIOD == 2 ~ 'First Half',
                                START_PERIOD == 3 | START_PERIOD == 4 ~ 'Second Half',
                                TRUE ~ 'OT')) %>%
        mutate(HOME = case_when(IS_HOME_OFFENSE == T ~ OFFENSE,
                                IS_HOME_OFFENSE ==F ~ DEFENSE),
               AWAY = case_when(IS_HOME_OFFENSE == T  ~ DEFENSE,
                                IS_HOME_OFFENSE ==F ~ OFFENSE)) %>%
        mutate(END_HOME_SCORE = case_when(IS_HOME_OFFENSE == T ~ END_OFFENSE_SCORE,
                                          IS_HOME_OFFENSE == F ~ END_DEFENSE_SCORE),
               END_AWAY_SCORE = case_when(IS_HOME_OFFENSE == F ~ END_OFFENSE_SCORE,
                                          IS_HOME_OFFENSE == T ~ END_DEFENSE_SCORE)) %>%
        group_by(GAME_ID, HALF) %>%
        mutate(LAST_DRIVE_HALF = case_when(DRIVE_NUMBER == max(DRIVE_NUMBER) ~ T,
                                           TRUE ~ F)) %>%
        ungroup() %>%
        group_by(GAME_ID) %>%
        mutate(HOME_SCORE_DIFF = END_HOME_SCORE - lag(END_HOME_SCORE, 1, default=0),
               AWAY_SCORE_DIFF = END_AWAY_SCORE - lag(END_AWAY_SCORE, 1, default=0)) %>%
        mutate(MULTIPLE_SCORE_CHANGES = HOME_SCORE_DIFF & AWAY_SCORE_DIFF > 0) %>%
        mutate(PROBLEM_AWAY_CHANGE = grepl("rush td|pass td|rushing td|passing td|fg good", tolower(DRIVE_RESULT)) & IS_HOME_OFFENSE == T & AWAY_SCORE_DIFF > 0,
               PROBLEM_HOME_CHANGE = grepl("rush td|pass td|rushing td|passing td|fg good", tolower(DRIVE_RESULT)) & IS_HOME_OFFENSE == F & HOME_SCORE_DIFF > 0) %>%
        mutate(SCORE_EVENT = case_when((tolower(DRIVE_RESULT) == 'td' | grepl("rush td|pass td|rushing td|passing td", tolower(DRIVE_RESULT))) & IS_HOME_OFFENSE == T ~ 'HOME TD',
                                       (tolower(DRIVE_RESULT) == 'td' | grepl("rush td|pass td|rushing td|passing td", tolower(DRIVE_RESULT))) & IS_HOME_OFFENSE == F ~ 'AWAY TD',
                                       grepl("rush td|pass td|rushing td|passing td", tolower(DRIVE_RESULT)) & IS_HOME_OFFENSE == T ~ 'HOME TD',
                                       grepl("rush td|pass td|rushing td|passing td", tolower(DRIVE_RESULT)) & IS_HOME_OFFENSE == F ~ 'AWAY TD',
                                       grepl("kickoff return td|kickoff td|kick return td|punt return td", tolower(DRIVE_RESULT)) & IS_HOME_OFFENSE ==T ~ 'AWAY TD',
                                       grepl("kickoff return td|kickoff td|kick return td|punt return td", tolower(DRIVE_RESULT)) & IS_HOME_OFFENSE ==F ~ 'HOME TD',
                                       (tolower(DRIVE_RESULT) == 'fg' | grepl("made fg|fg good", tolower(DRIVE_RESULT))) & IS_HOME_OFFENSE == T ~ 'HOME FG',
                                       (tolower(DRIVE_RESULT) == 'fg' | grepl("made fg|fg good", tolower(DRIVE_RESULT))) & IS_HOME_OFFENSE == F ~ 'AWAY FG',
                                       HOME_SCORE_DIFF == 6 | HOME_SCORE_DIFF == 7 | HOME_SCORE_DIFF == 8 ~ 'HOME TD',
                                       HOME_SCORE_DIFF == 3 ~ 'HOME FG',
                                       HOME_SCORE_DIFF == 2 ~ 'HOME Safety',
                                       AWAY_SCORE_DIFF == 6 | AWAY_SCORE_DIFF == 7 | AWAY_SCORE_DIFF == 8 ~ 'AWAY TD',
                                       AWAY_SCORE_DIFF == 3 ~ 'AWAY FG',
                                       AWAY_SCORE_DIFF == 2 ~ 'AWAY Safety', 
                                       (HOME_SCORE_DIFF == 0 & AWAY_SCORE_DIFF == 0) & LAST_DRIVE_HALF == T ~ 'No_Score')) %>%
        mutate(MISSING_SCORING_EVENT = SCORING == T & is.na(SCORE_EVENT)) %>%
        #  filter(MISSING_SCORING_EVENT == T) %>%
        mutate(NEXT_SCORE_EVENT = SCORE_EVENT) %>%
        # this fills in previous drives with no scoring event with the next observed scoring event in a half
        fill(NEXT_SCORE_EVENT, .direction = "up") %>%
        ungroup() 

```

```{r now join this with plays}

# now join with plays
plays_data_score_events = plays_data_raw %>%
        left_join(.,
                  drives_data_score_events %>%
                          select(GAME_ID,
                                 DRIVE_ID,
                                 HOME, 
                                 AWAY,
                                 LAST_DRIVE_HALF,
                                 END_HOME_SCORE,
                                 END_AWAY_SCORE,
                                 SCORE_EVENT,
                                 NEXT_SCORE_EVENT),
                  by = c("GAME_ID",
                         "DRIVE_ID",
                         "HOME", 
                         "AWAY")) %>%
        # now join with games to get seasons
        left_join(.,
                  games_data_raw %>%
                          select(GAME_ID,
                                 SEASON,
                                 HOME_TEAM,
                                 AWAY_TEAM),
                  by = c("GAME_ID")) %>%
        rename(NEXT_SCORE_EVENT_HOME = NEXT_SCORE_EVENT) %>%
        mutate(NEXT_SCORE_EVENT_OFFENSE = case_when(OFFENSE == HOME & NEXT_SCORE_EVENT_HOME == 'HOME TD' ~ 'TD',
                                                    OFFENSE == HOME & NEXT_SCORE_EVENT_HOME == 'AWAY TD' ~ 'Opp_TD',
                                                    OFFENSE == HOME & NEXT_SCORE_EVENT_HOME == 'HOME FG' ~ 'FG',
                                                    OFFENSE == HOME & NEXT_SCORE_EVENT_HOME == 'AWAY FG' ~ 'Opp_FG',
                                                    OFFENSE == HOME & NEXT_SCORE_EVENT_HOME == 'HOME Safety' ~ 'Safety',
                                                    OFFENSE == HOME & NEXT_SCORE_EVENT_HOME == 'AWAY Safety' ~ 'Opp_Safety',
                                                    OFFENSE == AWAY & NEXT_SCORE_EVENT_HOME == 'HOME TD' ~ 'Opp_TD',
                                                    OFFENSE == AWAY & NEXT_SCORE_EVENT_HOME == 'AWAY TD' ~ 'TD',
                                                    OFFENSE == AWAY & NEXT_SCORE_EVENT_HOME == 'HOME FG' ~ 'Opp_FG',
                                                    OFFENSE == AWAY & NEXT_SCORE_EVENT_HOME == 'AWAY FG' ~ 'FG',
                                                    OFFENSE == AWAY & NEXT_SCORE_EVENT_HOME == 'HOME Safety' ~ 'Opp_Safety',
                                                    OFFENSE == AWAY & NEXT_SCORE_EVENT_HOME == 'AWAY Safety' ~ 'Safety',
                                                    NEXT_SCORE_EVENT_HOME == 'No_Score' ~ 'No_Score')) %>%
        mutate(NEXT_SCORE_EVENT_HOME_DIFF = case_when(NEXT_SCORE_EVENT_HOME == 'HOME TD' ~ 7,
                                              NEXT_SCORE_EVENT_HOME == 'HOME FG' ~ 3,
                                              NEXT_SCORE_EVENT_HOME == 'HOME Safety' ~ 2,
                                              NEXT_SCORE_EVENT_HOME == 'No_Score' ~ 0,
                                              NEXT_SCORE_EVENT_HOME == 'HOME Safety' ~ -2,
                                              NEXT_SCORE_EVENT_HOME == 'HOME FG' ~ -3,
                                              NEXT_SCORE_EVENT_HOME == 'HOME TD' ~ -7)) %>%
        mutate(NEXT_SCORE_EVENT_OFFENSE_DIFF = case_when(NEXT_SCORE_EVENT_OFFENSE == 'TD' ~ 7,
                                              NEXT_SCORE_EVENT_OFFENSE == 'FG' ~ 3,
                                              NEXT_SCORE_EVENT_OFFENSE == 'Safety' ~ 2,
                                              NEXT_SCORE_EVENT_OFFENSE == 'No_Score' ~ 0,
                                              NEXT_SCORE_EVENT_OFFENSE == 'Opp_Safety' ~ -2,
                                              NEXT_SCORE_EVENT_OFFENSE == 'Opp_FG' ~ -3,
                                              NEXT_SCORE_EVENT_OFFENSE == 'Opp_TD' ~ -7)) %>%
        mutate(NEXT_SCORE_EVENT_OFFENSE = factor(NEXT_SCORE_EVENT_OFFENSE,
                                                     levels = c("TD",
                                                                "FG",
                                                                "Safety",
                                                                "No_Score",
                                                                "Opp_Safety",
                                                                "Opp_FG",
                                                                "Opp_TD")))
                                                 
```

# What is this? {-}

# Exploring The Data

The data used is from college football games from 2000 to present. Each observation represents one play in a game, in which we know the team, the situation (down, time remaining), and the location on the field (yards to go, yards to reach end zone). We have information about the types of plays called as well in a text field.

```{r show skimr}

plays_data_score_events %>%
        filter(SEASON < 2015) %>%
        mutate_at(vars(contains("_ID")),
                  ~ as.character(.)) %>%
        sample_n(1000) %>%
        skimr::skim_without_charts(.)

```

The goal of this analysis is to determine the value of individual plays in terms of expected points. 

A sequence in this case is defined as all plays within a half between kickoffs (either the beginning or scoring events). Every sequence can end in one of 7 different outcomes, which we will define from the perspective of the team on offense. If a team has the ball, for any play, there are seven possible outcomes that play will eventually reach:

* Touchdown (7 points)
* Field goal (3 points)
* Safety (2 points)
* No_Score (0 points)
* Opp_Safety (2 points)
* Opp field goal (-3 points)
* Opp touchdown (-7 points)

For instance, on drive 1, Team A receives the opening kickoff, drives for a few plays, and then punts. Team B takes over, which starts drive 2, and they drive for a few plays before also punting. Team A then manages to put together a drive that finally scores. All plays on these three drives are one sequence. The outcome of this sequence is the points scored by Team A - if they score a toucbhdown, their points from this sequence is 7 (assuming they make the extra point). Team B's points from this sequence is -7 points.

When Team A kicks off to Team B to start drive 4, we start our next sequence, which will end either with one team scoring or at the end of the half. We'll then start over with a new sequence in the second half.

The outcome for this analysis is the NEXT_SCORE_EVENT. Each play in a given sequence contributes to the eventual outcome of the sequence. Here we can see an example of one game and its drives.

```{r show a full game of sequences}

drives_data_score_events %>%
        left_join(.,
                  games_data_raw %>%
                          select(GAME_ID,
                                 SEASON,
                                 HOME_TEAM,
                                 AWAY_TEAM),
                  by = c("GAME_ID")) %>%
        filter(HOME_TEAM == 'Texas A&M' & AWAY_TEAM == 'Florida') %>%
        filter(SEASON == 2012) %>%
        select(SEASON, HOME, AWAY, HALF, DRIVE_NUMBER, END_HOME_SCORE, END_AWAY_SCORE, NEXT_SCORE_EVENT) %>%
        mutate(SEASON = as.character(SEASON)) %>%
        flextable() %>%
        autofit()
        
```

For this game, we can filter to the plays that took place in the lead up to first score event. In this case, the first sequence included one drive and ended when Texas A&M kicked a field goal.

```{r zoom in on a sequence}

plays_data_score_events %>%
        filter(HOME_TEAM == 'Texas A&M' & AWAY_TEAM == 'Florida') %>%
        filter(SEASON == 2012) %>%
        filter(DRIVE_NUMBER == 1) %>%
        select(SEASON, OFFENSE, DEFENSE, HALF, DRIVE_NUMBER,  PLAY_NUMBER, YARD_LINE, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, NEXT_SCORE_EVENT_OFFENSE) %>%
        mutate(SEASON = as.character(SEASON)) %>%
        flextable() %>%
        autofit()
        
```

If we look at another sequence in the second half, there were multiple drives before a team was able to score.

```{r zoom in on another sequence}

plays_data_score_events %>%
        filter(HOME_TEAM == 'Texas A&M' & AWAY_TEAM == 'Florida') %>%
        filter(SEASON == 2012) %>%
        filter(DRIVE_NUMBER>=9 & DRIVE_NUMBER <=14) %>%
        select(SEASON, OFFENSE, DEFENSE, HALF, OFFENSE_SCORE, DEFENSE_SCORE, DRIVE_NUMBER,  PLAY_NUMBER, YARD_LINE, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, NEXT_SCORE_EVENT_OFFENSE) %>%
        mutate(SEASON = as.character(SEASON)) %>%
        flextable() %>%
        autofit()
```

Our goal is to understand how individual plays contribute to a team's **expected points**, or the average points teams should expect to have given their situation (down, time, possession). 

For instance, in the first drive of this game, Texas A&M received the ball at their own 25 yard line - the simplest intuition of expected points is to ask, for teams starting at the 25 yard line at the beginning of a game, how many points do they typically go on to score? The answer is to look at all starting drives with 75 yards to go and see what happened - we take the average of all of the points that followed from this situation.

```{r expected points given situation}

plays_data_score_events %>%
        filter(SEASON >=2010) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        filter(!grepl("kick", tolower(PLAY_TEXT))) %>%
        filter(YARDS_TO_GOAL == 75 & DRIVE_NUMBER ==1) %>%
        group_by(NEXT_SCORE_EVENT_OFFENSE) %>%
        count(sort=T)

plays_data_score_events %>%
        filter(SEASON >=2010) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        filter(YARDS_TO_GOAL == 75 & DRIVE_NUMBER ==1) %>%
        group_by(YARDS_TO_GOAL, DRIVE_NUMBER) %>%
        summarize(expected_points = mean(NEXT_SCORE_EVENT_OFFENSE_DIFF),
                  n = n(),
                  .groups = 'drop') %>%
        mutate_if(is.numeric, round, 2)

```

In this case, this means teams with the ball at their own 25 to start the game generally obtained more points on the ensuing sequence than their opponents, so they have a slightly positive expected points.

But, this is also a function of the down. If we look at the expected points for a team in this situation in first down vs a team in this situation for fourth down, we should see a drop in their expected points - by the time you hit fourth down, if you haven't moved from the 25, your expected points drops into the negatives, as you will now be punting the ball back to your opponent and it becomes more probable that they score than you.

```{r how does this vary as a function of }

plays_data_score_events %>%
        filter(SEASON >= 2010) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(YARDS_TO_GOAL == 75  & DRIVE_NUMBER ==1) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        group_by(YARDS_TO_GOAL, DOWN, NEXT_SCORE_EVENT_OFFENSE) %>%
        count() %>%
        spread(DOWN, n) %>%
        arrange(NEXT_SCORE_EVENT_OFFENSE) %>%
        arrange(desc(`4`))

plays_data_score_events %>%
        filter(SEASON >= 2010) %>%
        filter(YARDS_TO_GOAL == 75  & DRIVE_NUMBER ==1) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        group_by(YARDS_TO_GOAL, DOWN) %>%
        summarize(expected_points = mean(NEXT_SCORE_EVENT_OFFENSE_DIFF, na.rm=T),
                  n = n(),
                  .groups = 'drop') %>%
        mutate_if(is.numeric, round, 2)

```


We can apply this type of thinking to the entire field. If we look at all total plays in a game, how do expected points vary as a function of a team's distance from their opponent's goal line?

```{r expected points as a function of yard line for opening}

plays_data_score_events %>%
        filter(SEASON >= 2010) %>%
        filter(!grepl("kick", tolower(PLAY_TYPE))) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        filter(YARD_LINE < 100 & YARD_LINE > 0) %>%
      #  mutate(YARD_LINE_5 = plyr::round_any(YARDS_TO_GOAL, 5, floor)) %>%
     #   group_by(YARDS_TO_GOAL) %>%
        group_by(YARDS_TO_GOAL) %>%
        summarize(expected_points = mean(NEXT_SCORE_EVENT_OFFENSE_DIFF, na.rm=T),
                  n = n()) %>%
        ggplot(., aes(x=YARDS_TO_GOAL,
                      y=expected_points))+
        geom_line()+
        geom_point(aes(size=n))+
        theme_phil()+
        geom_hline(yintercept = 0,
                   linetype = 'dotted')+
        scale_x_reverse()

```

```{r look at outcomes by yards to goal}

plays_data_score_events %>%
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        filter(YARD_LINE < 100 & YARD_LINE > 0) %>%
      #  mutate(YARD_LINE_5 = plyr::round_any(YARDS_TO_GOAL, 5, floor)) %>%
     #   group_by(YARDS_TO_GOAL) %>%
        group_by(YARDS_TO_GOAL, NEXT_SCORE_EVENT_OFFENSE) %>%
        count() %>%
        ggplot(., aes(x=YARDS_TO_GOAL,
                      fill = NEXT_SCORE_EVENT_OFFENSE,
                      y=n))+
        geom_col(position = 'fill')+
        scale_fill_viridis_d(option = 'B')+
        theme_phil()+
        scale_x_reverse()

```

This should make sense - if you're backed up against your own end zone, your opponent becomes more likely to have the next scoring event, either by gaining good field advantage after you punt or by getting a safety.

But, it's not just position on the field - it's also about the situation. If we look at how expected points varies by the down, we should see that fourth downs always have lower expected points.

```{r look at expected points by down and field position}

plays_data_score_events %>%
        filter(PLAY_TYPE != 'Timeout') %>%
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        filter(YARD_LINE < 100 & YARD_LINE > 0) %>%
      #  mutate(YARD_LINE_5 = plyr::round_any(YARDS_TO_GOAL, 5, floor)) %>%
     #   group_by(YARDS_TO_GOAL) %>%
        group_by(YARDS_TO_GOAL, DOWN) %>%
        summarize(expected_points = mean(NEXT_SCORE_EVENT_OFFENSE_DIFF, na.rm=T),
                  n = n(),
                  .groups = 'drop') %>%
        mutate(DOWN = factor(DOWN)) %>%
        ggplot(., aes(x=YARDS_TO_GOAL,
                      group = DOWN,
                      color = DOWN,
                      y=expected_points))+
        geom_point(aes(size=n))+
        geom_line(stat = 'smooth',
                  formula = 'y ~ x',
                  method = 'loess',
                  span = 0.25)+
        theme_phil()+
        geom_hline(yintercept = 0,
                   linetype = 'dotted')+
        scale_x_reverse()+
        scale_color_viridis_d()

```

We also have other features like distance.

```{r now look at by downs}

plays_data_score_events %>%
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        filter(YARD_LINE < 100 & YARD_LINE > 0) %>%
        filter(!(DISTANCE <0 | DISTANCE >=99)) %>%
        # group_by(DOWN, DISTANCE) %>%
        # count() %>%
        mutate(DOWN = factor(DOWN)) %>%
        ggplot(., aes(x=DISTANCE,
                      fill = DOWN))+
        geom_histogram(bins = 50)+
        facet_wrap(DOWN ~.,
                   ncol = 2,
                   scales = "free_y")+
        scale_fill_viridis_d()+
        theme_phil()
        

plays_data_score_events %>%
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        filter(YARD_LINE < 100 & YARD_LINE > 0) %>%
        filter(!(DISTANCE <0 | DISTANCE >=99)) %>%
        group_by(DISTANCE, DOWN) %>%
        summarize(expected_points = mean(NEXT_SCORE_EVENT_OFFENSE_DIFF, na.rm=T),
                  n = n(),
                  .groups = 'drop') %>%
        mutate(DOWN = factor(DOWN)) %>%
        ggplot(., aes(x=log1p(DISTANCE),
                      group = DOWN,
                      color = DOWN,
                      y=expected_points))+
        geom_point(aes(size=n))+
        geom_line(stat = 'smooth',
                  formula = 'y ~ x',
                  method = 'loess',
                  span = 0.25)+
        theme_phil()+
        geom_hline(yintercept = 0,
                   linetype = 'dotted')+
        scale_color_viridis_d()


```

As well as time remaining in the half.

```{r look at outcomes as a function of time remaining}

plays_data_score_events %>%
        filter(FLAG_SECONDS_IN_HALF != 1) %>%
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        filter(YARD_LINE < 100 & YARD_LINE > 0) %>%
        mutate(SECONDS_IN_HALF = plyr::round_any(SECONDS_IN_HALF, 5, ceiling)) %>%
        group_by(SECONDS_IN_HALF, NEXT_SCORE_EVENT_OFFENSE) %>%
        count() %>%
        ggplot(., aes(x=SECONDS_IN_HALF,
                      fill = NEXT_SCORE_EVENT_OFFENSE,
                      y=n))+
        geom_bar(position = 'fill',
                 stat='identity',
                 color = NA)+
        scale_fill_viridis_d(option = 'B')+
        theme_phil()+
        scale_x_reverse()

```

# Modeling

We'll now proceed to modeling. I'll set up training, validation, and test sets based around the season.

```{r set up split sequence}

# full plays
plays_full = plays_data_score_events %>%
        filter(PLAY_TYPE != 'Kickoff') %>%
        select(GAME_ID,
               DRIVE_ID,
               PLAY_ID,
               SEASON,
               HOME,
               AWAY,
               OFFENSE,
               DEFENSE,
               OFFENSE_SCORE,
               DEFENSE_SCORE,
               SCORING,
               PLAY_TEXT,
               PLAY_TYPE,
               NEXT_SCORE_EVENT_HOME,
               NEXT_SCORE_EVENT_HOME_DIFF,
               NEXT_SCORE_EVENT_OFFENSE,
               NEXT_SCORE_EVENT_OFFENSE_DIFF,
               YARD_LINE,
               HALF,
               PERIOD,
               MINUTES_IN_HALF,
               SECONDS_IN_HALF,
               DOWN,
               DISTANCE,
               YARD_LINE,
               YARDS_TO_GOAL) %>%
        filter(DOWN %in% c(1, 2, 3, 4)) %>%
        filter(PERIOD %in% c(1,2,3,4)) %>%
        filter(!is.na(SECONDS_IN_HALF)) %>%
        filter(DISTANCE >=0 & DISTANCE <=100) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        mutate(NEXT_SCORE_EVENT_OFFENSE = factor(NEXT_SCORE_EVENT_OFFENSE,
                                                 levels = c("No_Score",
                                                            "TD",
                                                            "FG",
                                                            "Safety",
                                                            "Opp_Safety",
                                                            "Opp_FG",
                                                            "Opp_TD"))) %>%
        arrange(SEASON, GAME_ID, PLAY_ID)

# training set
plays_train = plays_full %>%
        filter(SEASON >= 2012 & SEASON <=2019)

# validation set
plays_valid = plays_full %>%
        filter(SEASON > 2019 & SEASON <= 2020)

# test
plays_test = plays_full %>%
        filter(SEASON > 2020)

# make an initial split based on previously defined splits
valid_split = make_splits(list(analysis = seq(nrow(plays_train)),
                                 assessment = nrow(plays_train) + seq(nrow(plays_valid))),
                               bind_rows(plays_train,
                                         plays_valid))

# test split
test_split = make_splits(
        list(analysis = seq(nrow(plays_train) + nrow(plays_valid)),
             assessment = nrow(plays_train) + nrow(plays_valid) + seq(nrow(plays_test))),
        bind_rows(plays_train,
                  plays_valid,
                  plays_test))

```

I plan to use the following as features in a baseline model.

I'll now set up a recipe for the baseline model.

```{r create recipes}

baseline_recipe = recipe(NEXT_SCORE_EVENT_OFFENSE ~.,
                         data = plays_train) %>%
        update_role(all_predictors(),
                    new_role = "ID") %>%
        update_role(
                c("GAME_ID",
                  "DRIVE_ID",
                  "PLAY_ID",
                  "SEASON",
                  "HOME",
                  "AWAY",
                  "OFFENSE",
                  "DEFENSE",
                  "SCORING",
                  "OFFENSE_SCORE",
                  "DEFENSE_SCORE",
                  "PLAY_TEXT",
                  "PLAY_TYPE",
                  "NEXT_SCORE_EVENT_HOME",
                  "NEXT_SCORE_EVENT_HOME_DIFF",
                  "NEXT_SCORE_EVENT_OFFENSE_DIFF",
                  "YARD_LINE",
                  "MINUTES_IN_HALF",
                  "HALF"),
                new_role = "ID") %>%
        step_mutate(PERIOD_ID = PERIOD,
                    role = "ID") %>%
        # features we're inheriting
        update_role(
                c("PERIOD", 
                "SECONDS_IN_HALF",
                "DOWN",
                "DISTANCE",
                "YARDS_TO_GOAL"),
                new_role = "predictor") %>%
        # filters for issues
        step_filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        step_filter(YARD_LINE <= 100 & YARD_LINE >=0) %>%
        step_filter(YARDS_TO_GOAL <=100 & YARD_LINE >=0) %>%
        step_filter(DOWN %in% c(1, 2, 3, 4)) %>%
        step_filter(DISTANCE >=0 & DISTANCE <=100) %>%
        step_filter(SECONDS_IN_HALF <=1800) %>%
        step_filter(!is.na(SECONDS_IN_HALF)) %>%
        step_filter(PERIOD_ID == 1 | PERIOD_ID == 2 | PERIOD_ID == 3 | PERIOD_ID == 4) %>%
        # create features
        step_mutate(KICKOFF = case_when(grepl("kickoff", tolower(PLAY_TEXT)) | grepl("kickoff", tolower(PLAY_TYPE))==T ~ 1,
                                        TRUE ~ 0)) %>%
        step_mutate(TIMEOUT = case_when(grepl("timeout", tolower(PLAY_TEXT)) ~ 1,
                                        TRUE ~ 0)) %>%
        step_filter(TIMEOUT != 1) %>%
        step_filter(KICKOFF != 1) %>%
        step_mutate(DOWN_TO_GOAL = case_when(DISTANCE == YARDS_TO_GOAL ~ 1,
                                             TRUE ~ 0)) %>%
        # step_mutate(RED_ZONE = case_when(YARDS_TO_GOAL <= 20 ~ 1,
        #                                  TRUE ~ 0)) %>%
      #  step_mutate(OFFENSE_LEAD = OFFENSE_SCORE - DEFENSE_SCORE) %>%
        step_mutate(DOWN = factor(DOWN)) %>%
        step_mutate(PERIOD = factor(PERIOD)) %>%
        step_log(DISTANCE, offset =1) %>%
        step_dummy(all_nominal_predictors()) %>%
        step_novel(all_nominal_predictors(),
                   new_level = "new") %>%
        step_interact(terms = ~ DISTANCE:(starts_with("DOWN_"))) %>%
        step_interact(terms = ~ YARDS_TO_GOAL:(starts_with("DOWN_"))) %>%
        step_interact(terms = ~ YARDS_TO_GOAL*SECONDS_IN_HALF) %>%
        check_missing(all_predictors()) %>%
        step_normalize(all_numeric_predictors())

```

## Workflows

I'll define the model I'll be using here, which is a multinomial logistic regression.

```{r define model objs}

# from glmnet
multinom_mod = multinom_reg(
  mode = "classification",
  engine = "glmnet",
  penalty = 0,
  mixture = NULL
)

```

I'll then create a workflow.

```{r craete workflow}

# create multinomial workflow
multinom_wf = workflow() %>%
        add_recipe(baseline_recipe) %>%
        add_model(multinom_mod)

# workflow settings
# metrics
class_metrics<-metric_set(yardstick::roc_auc,
                          yardstick::mn_log_loss)

# control for resamples
keep_pred <- control_resamples(save_pred = TRUE, 
                               save_workflow = TRUE,
                               allow_par=T)

```

I'll manually define resamples based on the seasons - rather than doing k-fold cross validation, I'll assign each season to be a fold and train and assess the model leaving one season out at a time.

```{r set up manual resamples by season}

# get unique seasons
seasons = unique(plays_train$SEASON) %>%
        sort

# define indices by seasons
indices = foreach(i = 1:length(seasons)) %do% {
        list(analysis = which(plays_train$SEASON != seasons[i]),
             assessment = which(plays_train$SEASON == seasons[i]))
}

# get the splits
splits <- lapply(indices, make_splits, data = plays_train)

# define rset object
manual_resamples = manual_rset(splits, 
                               paste("holdout:", seasons))

```

```{r register parallel cores}

# register parallel
library(doParallel)
numcores = parallel::detectCores()
registerDoParallel(numcores -2)

```

## Training

Now I'll train and assess on these resamples.

```{r fit on set}

# fit to resamples
resamples_multinom = multinom_wf %>%
        fit_resamples(data = plays_train,
                      metrics = class_metrics,
                      resamples = manual_resamples,
                      control = keep_pred,
                      verbose=T)

# fit the model to the whole training set
fit_multinom = multinom_wf %>%
        fit(data = plays_train)

```

```{r get results}

# collect metrics
resamples_multinom %>%
        collect_metrics(summarize=F) %>%
        mutate_if(is.numeric, round, 3) %>%
        select(id, .metric, .estimator, .estimate) %>%
        arrange(.metric)

```

```{r get coefs, fig.height=6, fig.width=8, warning=F, message=F, eval=F}

tidy(fit_multinom) %>%
        mutate(term = factor(term)) %>%
        mutate(term = relevel(term, ref = "(Intercept)")) %>%
        ggplot(., aes(y=term,
                      x=estimate))+
        geom_point()+
        facet_wrap(class~.)+
        theme_bw()

```

Understanding partial effects from a multinomial logit is already difficult, and I've thrown a bunch of interactions in there to make this even more difficult. I'll look at predicted probabilities using an observed values approach for particular features (using a sample rather than the full dataset to save time). This means taking the model and then altering the feature of interest for every observation.

```{r look at predicted probabilities}

# define min and max for feature
setting = seq(1, 100, 2)

# loop over
pred_settings = foreach(i = 1:length(setting),
        .combine = bind_rows) %do% {
                
                set.seed(1999)
                fit_multinom %>%
                        predict(new_data = plays_train %>% 
                                        sample_n(10000) %>%
                                        mutate(YARDS_TO_GOAL = setting[i]),
                                type = 'prob') %>%
                        apply(., 2, mean) %>%
                        as.data.frame() %>%
                        rownames_to_column("outcome") %>% set_names(., c("outcome", ".pred")) %>%
                        mutate(YARDS_TO_GOAL = setting[i])
}

# pred_settings %>%
#         mutate(outcome = paste(gsub(".pred_", "Pr(", outcome), ")", sep="")) %>%
#         mutate(outcome = factor(outcome,
#                                 levels = paste("Pr(", c("TD",
#                                       "FG",
#                                       "Safety",
#                                       "No_Score",
#                                       "Opp_Safety",
#                                       "Opp_FG",
#                                       "Opp_TD"),
#                                       ")",
#                                       sep =""))) %>%
#         ggplot(., 
#                aes(x=YARDS_TO_GOAL,
#                y = .pred,
#                fill = outcome))+
#         geom_col(position = 'fill')+
#         theme_phil()+
#         scale_fill_viridis_d(option = 'B',
#                            end = 0.8,
#                            begin = 0.2)+
#    #     scale_fill_manual(values = col_func(7))+
#         theme(legend.title = element_text(),
#               legend.position = 'top')+
#         guides(color = guide_legend(title = 'Next Scoring Event'))+
#         xlab("Yards to Opponent End Zone")+
#         ylab("Pr(Next Scoring Event)")
# 

```

How is the probability of the next scoring event influenced by where the offense has possession?

```{r show probabilities by yards to goal}

# now plot this effect
pred_settings %>%
        mutate(outcome = paste(gsub(".pred_", "Pr(", outcome), ")", sep="")) %>%
             mutate(outcome = factor(outcome,
                                levels = paste("Pr(", c("TD",
                                      "FG",
                                      "Safety",
                                      "No_Score",
                                      "Opp_Safety",
                                      "Opp_FG",
                                      "Opp_TD"),
                                      ")",
                                      sep =""))) %>%
        ggplot(., 
               aes(x=YARDS_TO_GOAL,
               y = .pred,
               color = outcome,
               group = outcome))+
        facet_wrap(outcome ~.,
                   ncol = 4)+
        geom_line(lwd=1.1)+
        scale_color_viridis_d(option = 'B',
                           end = 0.8,
                           begin = 0.2)+        
        theme_phil()+
        theme(legend.title = element_text())+
        guides(color = guide_legend(title = 'Next Scoring Event',
                                    title.position = 'top'))+
        xlab("Yards to Opponent End Zone")+
        ylab("Pr(Next Scoring Event)")+
        coord_cartesian(ylim= c(0, 1))+
        geom_vline(xintercept = 50,
                   linetype = 'dotted')+
        ggtitle("Probability of Next Scoring Event by Offensive Field Position",
                subtitle = str_wrap("Predicted probabilities from multinomial logistic regression. Displaying probabilities using observed values approach for a sample of training set. Model trained on regular season college football plays from 2012-2020. ",
                                    120))

# now plot this effect
pred_settings %>%
        mutate(outcome = paste(gsub(".pred_", "Pr(", outcome), ")", sep="")) %>%
             mutate(outcome = factor(outcome,
                                levels = paste("Pr(", c("TD",
                                      "FG",
                                      "Safety",
                                      "No_Score",
                                      "Opp_Safety",
                                      "Opp_FG",
                                      "Opp_TD"),
                                      ")",
                                      sep =""))) %>%
        ggplot(., 
               aes(x=YARDS_TO_GOAL,
               y = .pred,
               fill = outcome,
               group = outcome))+
        geom_col(position = 'fill',
                 lwd=1.1)+
        scale_fill_viridis_d(option = 'B',
                           end = 0.8,
                           begin = 0.2)+        
        theme_phil()+
        theme(legend.title = element_text())+
        guides(fill = guide_legend(title = 'Next Scoring Event',
                                    title.position = 'top'))+
        xlab("Yards to Opponent End Zone")+
        ylab("Pr(Next Scoring Event)")+
        coord_cartesian(ylim= c(0, 1))+
        geom_vline(xintercept = 50,
                   linetype = 'dotted')+
        ggtitle("Probability of Next Scoring Event by Offensive Field Position",
                subtitle = str_wrap("Predicted probabilities from multinomial logistic regression. Displaying probabilities using observed values approach for a sample of training set. Model trained on regular season college football plays from 2012-2020. ",
                                    120))


```

How is this affected by the down?

```{r loop over yards to goal and down}

# define min and max for feature
setting_1 = seq(1, 100, 5)
setting_2 = c(1, 2, 3, 4)

# loop over both
pred_settings_12 = foreach(i = 1:length(setting_1),
                        .combine = bind_rows) %:% 
        foreach(j = 1:length(setting_2)) %do% {
                
                set.seed(1999)
                 fit_multinom %>%
                        predict(new_data = plays_train %>% 
                                        sample_n(10000) %>%
                                        mutate(YARDS_TO_GOAL = setting_1[i],
                                               DOWN = setting_2[j]),
                                type = 'prob') %>%
                        apply(., 2, mean) %>%
                        as.data.frame() %>%
                        rownames_to_column("outcome") %>% 
                        set_names(., c("outcome", ".pred")) %>%
                        mutate(YARDS_TO_GOAL = setting_1[i],
                               DOWN = setting_2[j])
                
        }
                
```

```{r look at predicted probs by down and yards to goal}

# now plot this effect
pred_settings_12 %>%
        mutate(outcome = paste(gsub(".pred_", "Pr(", outcome), ")", sep="")) %>%
             mutate(outcome = factor(outcome,
                                levels = paste("Pr(", c("TD",
                                      "FG",
                                      "Safety",
                                      "No_Score",
                                      "Opp_Safety",
                                      "Opp_FG",
                                      "Opp_TD"),
                                      ")",
                                      sep =""))) %>%
        mutate(DOWN = factor(DOWN)) %>%
        ggplot(., 
               aes(x=YARDS_TO_GOAL,
               y = .pred,
               color = DOWN,
               group = DOWN))+
        facet_wrap(outcome ~.,
                   ncol = 4)+
        geom_line(lwd=1.1)+
        scale_color_viridis_d(option = 'D',
                           begin = 0.2)+        
        theme_phil()+
        theme(legend.title = element_text())+
        guides(color = guide_legend(title = 'Down',
                                    title.position = 'top'))+
        xlab("Yards to Opponent End Zone")+
        ylab("Pr(Next Scoring Event)")+
        ggtitle("Probability of Next Scoring Event by Offensive Field Position and Down",
                subtitle = str_wrap("Predicted probabilities from multinomial logistic regression. Displaying probabilities using observed values approach for a sample of training set. Model trained on regular season college football plays from 2012-2020. ",
                                    120))

```

How does this translate into expected points?

```{r make expected points function}

expected_points_func = function(x) {

        x %>%
                mutate(EP = 0 * `.pred_No_Score` + 
                               7 * .pred_TD + 3 * .pred_FG + 
                               2 * .pred_Safety + 
                               -2*`.pred_Opp_Safety` + 
                               -3*`.pred_Opp_FG` + 
                               -7*`.pred_Opp_TD`)
        
}

```



```{r expected points by distance and down}

# loop over both
point_settings_12 = foreach(i = 1:length(setting_1),
                        .combine = bind_rows) %:% 
        foreach(j = 1:length(setting_2)) %do% {
                
                set.seed(1999)
                 fit_multinom %>%
                        predict(new_data = plays_train %>% 
                                        sample_n(10000) %>%
                                        mutate(YARDS_TO_GOAL = setting_1[i],
                                               DOWN = setting_2[j]),
                                type = 'prob') %>%
                         expected_points_func %>%
                         summarize(EP = mean(EP)) %>%
                        mutate(YARDS_TO_GOAL = setting_1[i],
                               DOWN = setting_2[j])
                
        }

```

```{r plot expected points by distance and down}

# now plot this effect
point_settings_12 %>%
        mutate(DOWN = factor(DOWN)) %>%
        ggplot(., 
               aes(x=YARDS_TO_GOAL,
               y = EP,
               color = DOWN,
               group = DOWN))+
        # facet_wrap(outcome ~.,
        #            ncol = 4)+
        geom_line(lwd=1.1)+
        scale_color_viridis_d(option = 'D',
                           begin = 0.2)+        
        theme_phil()+
        geom_hline(yintercept = 0,
                   linetype = 'dashed')+
        theme(legend.title = element_text())+
        guides(color = guide_legend(title = 'Down',
                                    title.position = 'top'))+
        xlab("Yards to Opponent End Zone")+
        ylab("Expected Points")+
        ggtitle("Expected Points by Offensive Field Position and Down",
                subtitle = str_wrap("Expected points using probabilities from multinomial logistic regression. Displaying expected points a portion of the training set using observed values approach, Model trained on regular season college football plays from 2012-2020. ",
                                    120))

```

## Predicting Validation Set

```{r predict}

preds_multinom = fit_multinom %>%
        predict(new_data = plays_valid,
                type = 'prob') %>%
        bind_cols(., 
                  fit_multinom %>%
                          predict(new_data = plays_valid)) %>%
        bind_cols(plays_valid,
                  .)

```


```{r get null preds}

# baseline rates in training set
train_rates = plays_train %>%
        group_by(NEXT_SCORE_EVENT_OFFENSE) %>%
        count() %>%
        ungroup() %>%
        mutate(prop = n / sum(n))

# predict null for all valid
preds_null = train_rates %>% 
        select(-n) %>%
        spread(NEXT_SCORE_EVENT_OFFENSE, prop) %>% 
        set_names(., paste(".pred_", names(.), sep="")) %>%
        mutate(.pred_class = factor('TD',
                                    levels = c("TD",
                                      "FG",
                                      "Safety",
                                      "No_Score",
                                      "Opp_Safety",
                                      "Opp_FG",
                                      "Opp_TD"))) %>%
        bind_cols(plays_valid,
                  .)
```


```{r yardstick assessment on validation, warning=F, message=F}


# yardstick::precision(preds_multinom, 
#            truth = NEXT_SCORE_EVENT_OFFENSE, 
#            estimate = .pred_class,
#            estimator = "macro")

# roc_auc
bind_rows(preds_null %>%
                  mutate(method = 'null'),
          preds_multinom %>%
                  mutate(method = 'multinom')) %>%
        group_by(method) %>%
        yardstick::roc_auc(
                   NEXT_SCORE_EVENT_OFFENSE,
                   `.pred_No_Score`:`.pred_Opp_TD`,
           #        estimator = 'macro_weighted'
                   ) %>%
        mutate_if(is.numeric, round, 3)

# logloss
bind_rows(preds_null %>%
                  mutate(method = 'null'),
          preds_multinom %>%
                  mutate(method = 'multinom')) %>%
        group_by(method) %>%
        yardstick::mn_log_loss(
                   NEXT_SCORE_EVENT_OFFENSE,
                   `.pred_No_Score`:`.pred_Opp_TD`,
           #        estimator = 'macro_weighted'
                   ) %>%
        mutate_if(is.numeric, round, 3)

```


```{r subpopulation log loss}

# roc_auc
bind_rows(preds_null %>%
                  mutate(method = 'null'),
          preds_multinom %>%
                  mutate(method = 'multinom')) %>%
        group_by(method, NEXT_SCORE_EVENT_OFFENSE) %>%
        yardstick::mn_log_loss(
                   NEXT_SCORE_EVENT_OFFENSE,
                   `.pred_No_Score`:`.pred_Opp_TD`,
                   na_rm=T,
           #        estimator = 'macro_weighted'
                   ) %>%
        mutate_if(is.numeric, round, 3) %>%
        spread(method, .estimate)

```

## Game Examples

```{r look at a game from 2013}

resamples_multinom %>%
        filter(id == 'holdout: 2013') %>% 
        select(.predictions) %>% 
        unnest() %>%
        bind_cols(., plays_train %>%
                          filter(SEASON == 2013) %>%
                          select(-NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(HOME == 'Texas A&M' & AWAY == 'Alabama') %>%
        arrange(PLAY_ID) %>% 
        expected_points_func() %>% 
        select(SEASON, GAME_ID, DRIVE_ID, PLAY_ID, HALF, PERIOD, OFFENSE, DEFENSE, HOME, AWAY, OFFENSE_SCORE, DEFENSE_SCORE, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, EP, PLAY_TYPE, NEXT_SCORE_EVENT_OFFENSE) %>%
        mutate(HOME_SCORE = case_when(OFFENSE == HOME ~ OFFENSE_SCORE,
                                      DEFENSE == HOME ~ DEFENSE_SCORE),
               AWAY_SCORE = case_when(DEFENSE == HOME ~ OFFENSE_SCORE,
                                      OFFENSE == HOME ~ DEFENSE_SCORE)) %>%
        group_by(GAME_ID, OFFENSE, HALF, DRIVE_ID) %>%
        rename(EP_Pre = EP)  %>%
        mutate(Touchdown = case_when(grepl("TOUCHDOWN|TD", PLAY_TEXT) ~ 1,
                                     TRUE ~ 0)) %>%
       # mutate(EP_Added = dplyr::lead(EP_Pre, 1)-EP_Pre) %>%
        mutate(Points_Added = case_when(grepl("TOUCHDOWN|TD", PLAY_TEXT) & NEXT_SCORE_EVENT_OFFENSE == 'TD' ~ 7,
                                              grepl("TOUCHDOWN|TD", PLAY_TEXT) & NEXT_SCORE_EVENT_OFFENSE == 'Opp_TD' ~ -7)) %>%
        mutate(EP_Post = case_when(Touchdown==0 ~ dplyr::lead(EP_Pre, 1),
                                   Touchdown==1 ~ Points_Added)) %>%
        mutate(EP_Added = case_when(Touchdown==1 ~ Points_Added - EP_Pre,
                                    Touchdown==0 ~ EP_Post-EP_Pre)) %>%
        ungroup() %>%
  #      filter(DRIVE_ID == 33257024519)
        bind_cols(.,
                  group_indices(., GAME_ID, HALF, HOME_SCORE, AWAY_SCORE) %>%
                          as_tibble() %>%
                          rename(SEQUENCE = value)) %>%
        group_by(GAME_ID, HALF, HOME_SCORE, AWAY_SCORE) %>%
        mutate(EP_Post = case_when(is.na(EP_Post) ~ -(dplyr::lead(EP_Pre,1)),
                                   TRUE ~ EP_Post)) %>%
        mutate(EP_Added = case_when(is.na(EP_Added) ~ EP_Post-EP_Pre,
                                   TRUE ~ EP_Added)) %>%
        ungroup() %>%
        select(SEASON, GAME_ID, DRIVE_ID, PLAY_ID, HALF, PERIOD, OFFENSE, DEFENSE, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, EP_Pre, EP_Post, EP_Added) %>%
        mutate_at(c("SEASON",
                    "GAME_ID",
                    "DRIVE_ID",
                    "PLAY_ID"),
                  ~ as.character(.)) %>%
 #       select(-DRIVE_ID) %>%
        mutate_if(is.numeric, round, 2) %>%
        arrange(desc(EP_Added)) %>%
        flextable() %>%
        autofit()

```

```{r look at a game from 2014}

resamples_multinom %>%
        filter(id == 'holdout: 2014') %>% 
        select(.predictions) %>% 
        unnest() %>%
        bind_cols(., plays_train %>%
                          filter(SEASON == 2014) %>%
                          select(-NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(HOME == 'Florida State' & AWAY == 'Notre Dame') %>%
        arrange(PLAY_ID) %>% 
        expected_points_func() %>% 
        select(SEASON, GAME_ID, DRIVE_ID, PLAY_ID, HALF, PERIOD, OFFENSE, DEFENSE, HOME, AWAY, OFFENSE_SCORE, DEFENSE_SCORE, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, EP, PLAY_TYPE, NEXT_SCORE_EVENT_OFFENSE) %>%
        mutate(HOME_SCORE = case_when(OFFENSE == HOME ~ OFFENSE_SCORE,
                                      DEFENSE == HOME ~ DEFENSE_SCORE),
               AWAY_SCORE = case_when(DEFENSE == HOME ~ OFFENSE_SCORE,
                                      OFFENSE == HOME ~ DEFENSE_SCORE)) %>%
        group_by(GAME_ID, OFFENSE, HALF, DRIVE_ID) %>%
        rename(EP_Pre = EP)  %>%
        mutate(Touchdown = case_when(grepl("TOUCHDOWN|TD", PLAY_TEXT) ~ 1,
                                     TRUE ~ 0)) %>%
       # mutate(EP_Added = dplyr::lead(EP_Pre, 1)-EP_Pre) %>%
        mutate(Points_Added = case_when(grepl("TOUCHDOWN|TD", PLAY_TEXT) & NEXT_SCORE_EVENT_OFFENSE == 'TD' ~ 7,
                                              grepl("TOUCHDOWN|TD", PLAY_TEXT) & NEXT_SCORE_EVENT_OFFENSE == 'Opp_TD' ~ -7)) %>%
        mutate(EP_Post = case_when(Touchdown==0 ~ dplyr::lead(EP_Pre, 1),
                                   Touchdown==1 ~ Points_Added)) %>%
        mutate(EP_Added = case_when(Touchdown==1 ~ Points_Added - EP_Pre,
                                    Touchdown==0 ~ EP_Post-EP_Pre)) %>%
        ungroup() %>%
  #      filter(DRIVE_ID == 33257024519)
        bind_cols(.,
                  group_indices(., GAME_ID, HALF, HOME_SCORE, AWAY_SCORE) %>%
                          as_tibble() %>%
                          rename(SEQUENCE = value)) %>%
        group_by(GAME_ID, HALF, HOME_SCORE, AWAY_SCORE) %>%
        mutate(EP_Post = case_when(is.na(EP_Post) ~ -(dplyr::lead(EP_Pre,1)),
                                   TRUE ~ EP_Post)) %>%
        mutate(EP_Added = case_when(is.na(EP_Added) ~ EP_Post-EP_Pre,
                                   TRUE ~ EP_Added)) %>%
        ungroup() %>%
        select(SEASON, GAME_ID, DRIVE_ID, PLAY_ID, HALF, PERIOD, OFFENSE, DEFENSE, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, EP_Pre, EP_Post, EP_Added) %>%
        mutate_at(c("SEASON",
                    "GAME_ID",
                    "DRIVE_ID",
                    "PLAY_ID"),
                  ~ as.character(.)) %>%
 #       select(-DRIVE_ID) %>%
        mutate_if(is.numeric, round, 2) %>%
        arrange(desc(EP_Added)) %>%
        flextable() %>%
        autofit()

```


Look at offenses year over year.

```{r look at expected points in valid}

yearly_offense_epa = resamples_multinom %>%
   #     filter(id == 'holdout: 2012') %>% 
        select(.predictions) %>% 
        unnest() %>%
        bind_cols(., plays_train %>%
                      #    filter(SEASON == 2012) %>%
                          select(-NEXT_SCORE_EVENT_OFFENSE)) %>%
        expected_points_func() %>% 
        filter(HOME == 'LSU' |
                       AWAY == 'LSU') %>%
        select(SEASON, GAME_ID, DRIVE_ID, PLAY_ID, HALF, PERIOD, OFFENSE, DEFENSE, HOME, AWAY, OFFENSE_SCORE, DEFENSE_SCORE, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, EP, PLAY_TYPE, NEXT_SCORE_EVENT_OFFENSE) %>%
        mutate(HOME_SCORE = case_when(OFFENSE == HOME ~ OFFENSE_SCORE,
                                      DEFENSE == HOME ~ DEFENSE_SCORE),
               AWAY_SCORE = case_when(DEFENSE == HOME ~ OFFENSE_SCORE,
                                      OFFENSE == HOME ~ DEFENSE_SCORE)) %>%
        group_by(GAME_ID, OFFENSE, HALF, DRIVE_ID) %>%
        rename(EP_Pre = EP)  %>%
        mutate(Touchdown = case_when(grepl("TOUCHDOWN|TD", PLAY_TEXT) ~ 1,
                                     TRUE ~ 0)) %>%
       # mutate(EP_Added = dplyr::lead(EP_Pre, 1)-EP_Pre) %>%
        mutate(Points_Added = case_when(grepl("TOUCHDOWN|TD", PLAY_TEXT) & NEXT_SCORE_EVENT_OFFENSE == 'TD' ~ 7,
                                              grepl("TOUCHDOWN|TD", PLAY_TEXT) & NEXT_SCORE_EVENT_OFFENSE == 'Opp_TD' ~ -7)) %>%
        mutate(EP_Post = case_when(Touchdown==0 ~ dplyr::lead(EP_Pre, 1),
                                   Touchdown==1 ~ Points_Added)) %>%
        mutate(EP_Added = case_when(Touchdown==1 ~ Points_Added - EP_Pre,
                                    Touchdown==0 ~ EP_Post-EP_Pre)) %>%
        ungroup()
  #      filter(DRIVE_ID == 33257024519)
        bind_cols(.,
                  group_indices(., GAME_ID, HALF, HOME_SCORE, AWAY_SCORE) %>%
                          as_tibble() %>%
                          rename(SEQUENCE = value)) %>%
        group_by(GAME_ID, HALF, HOME_SCORE, AWAY_SCORE) %>%
        mutate(EP_Post = case_when(is.na(EP_Post) ~ -(dplyr::lead(EP_Pre,1)),
                                   TRUE ~ EP_Post)) %>%
        mutate(EP_Added = case_when(is.na(EP_Added) ~ EP_Post-EP_Pre,
                                   TRUE ~ EP_Added)) %>%
        ungroup() %>%
        select(SEASON, GAME_ID, DRIVE_ID, PLAY_ID, HALF, PERIOD, OFFENSE, DEFENSE, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, EP_Pre, EP_Post, EP_Added) %>%
        filter(OFFENSE == 'LSU')
        group_by(SEASON, OFFENSE) %>%
        summarize(EPA_per_Play = mean(EP_Added, na.rm=T),
                  plays = n(),
                  .groups = 'drop')

yearly_offense_epa %>%
      #  filter(SEASON == 2018) %>%
        filter(plays > 300) %>%
        mutate(SEASON = as.character(SEASON)) %>%
        select(-plays) %>%
        mutate_if(is.numeric, round, 2) %>%
        arrange(desc(EPA_per_Play)) %>%
        flextable() %>%
        autofit()

```

Same thing but for defenses

```{r now do the same thing for defenses}

yearly_defense_epa = resamples_multinom %>%
   #     filter(id == 'holdout: 2012') %>% 
        select(.predictions) %>% 
        unnest() %>%
        bind_cols(., plays_train %>%
                      #    filter(SEASON == 2012) %>%
                          select(-NEXT_SCORE_EVENT_OFFENSE)) %>%
        expected_points_func() %>% 
        select(SEASON, GAME_ID, DRIVE_ID, PLAY_ID, HALF, PERIOD, OFFENSE, DEFENSE, OFFENSE_SCORE, DEFENSE_SCORE, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, EP, PLAY_TYPE, NEXT_SCORE_EVENT_OFFENSE) %>%
        group_by(GAME_ID, OFFENSE, HALF, DRIVE_ID) %>%
        rename(EP_Pre = EP)  %>%
        mutate(Touchdown = case_when(grepl("TOUCHDOWN|TD", PLAY_TEXT) ~ 1,
                                     TRUE ~ 0)) %>%
       # mutate(EP_Added = dplyr::lead(EP_Pre, 1)-EP_Pre) %>%
        mutate(Points_Added = case_when(grepl("TOUCHDOWN|TD", PLAY_TEXT) & NEXT_SCORE_EVENT_OFFENSE == 'TD' ~ 7,
                                              grepl("TOUCHDOWN|TD", PLAY_TEXT) & NEXT_SCORE_EVENT_OFFENSE == 'Opp_TD' ~ -7)) %>%
        mutate(EP_Post = case_when(Touchdown==0 ~ dplyr::lead(EP_Pre, 1),
                                   Touchdown==1 ~ Points_Added)) %>%
        mutate(EP_Added = case_when(Touchdown==1 ~ Points_Added - EP_Pre,
                                    Touchdown==0 ~ EP_Post-EP_Pre)) %>%
        ungroup() %>%
        select(SEASON, GAME_ID, DRIVE_ID, PLAY_ID, HALF, PERIOD, OFFENSE, DEFENSE, OFFENSE_SCORE, DEFENSE_SCORE, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, PLAY_TYPE, EP_Pre, EP_Post, EP_Added) %>%
        group_by(SEASON, DEFENSE) %>%
        summarize(EPA_per_Play = -mean(EP_Added, na.rm=T),
                  plays = n(),
                  .groups = 'drop')

yearly_defense_epa %>%
      #  filter(SEASON == 2018) %>%
        filter(plays > 300) %>%
        mutate(SEASON = as.character(SEASON)) %>%
        select(-plays) %>%
        mutate_if(is.numeric, round, 2) %>%
        arrange(desc(EPA_per_Play)) %>%
        flextable() %>%
        autofit()

```


Now plot these together

```{r }

```


```{r examine analytic, warning=F, message=F}

plays_valid %>%
        filter(SEASON == 2018) %>%
        filter(HOME == 'Texas A&M') %>%
        filter(AWAY == 'Clemson') %>%
        arrange(PLAY_ID) %>%
        left_join(.,
                  preds_multinom %>%
                          expected_points_func %>%
                          group_by(GAME_ID) %>%
                          arrange(PLAY_ID) %>%
                          mutate(EPA = EP - lag(EP, 1)) %>%
                          select(GAME_ID, PLAY_ID, EP, EPA),
                  by = c("GAME_ID", "PLAY_ID")
                  ) %>%
        select(OFFENSE, DEFENSE, OFFENSE_SCORE, DEFENSE_SCORE, YARDS_TO_GOAL, DOWN, DISTANCE, EP, EPA, PLAY_TYPE, PLAY_TEXT) %>%
        mutate_if(is.numeric, round, 2) %>% 
        View()

```


# Validation

