---
title: "CFB Expected Points Added"
author: "Phil Henrickson"
date: "6/7/2022"
output: 
  html_document:
    toc: TRUE #adds a Table of Contents
    theme: cerulean
    number_sections: TRUE #number your headings/sections
    toc_float: TRUE #let your ToC follow you as you scroll
    keep_md: no
    fig.caption: yes
    css: "styles.css"
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = F,
                      error = F,
                      warning=F,
                      dev="png",
                      fig.width = 10,
                      fig.height = 6)

options(knitr.duplicate.label = "allow")

options(scipen=999)

```


```{r connect to snowflake}

library(DBI)
library(odbc)
library(RODBC)
library(keyring)

# connect to snowflake
myconn <- DBI::dbConnect(odbc::odbc(),
                         "SnowflakeDSII",
                         Database = "CFB_DEMO",
                         warehouse = "DEMO_WH",
                         uid="phil.henrickson",
                         pwd=keyring::key_get("AE_Snowflake"))

```

```{r packages, include=F} 

source(here::here("scripts/load_packages.R"))
library(jsonlite)
library(forcats)
conflict_prefer("lag", "dplyr")

```

```{r load functions}

source(here::here("functions/theme_phil.R"))
source(here::here("functions/clean_plays_func.R"))
source(here::here("functions/make_time_features_func.R"))
source(here::here("functions/plot_scoreline_func.R"))
rm(a)
```

```{r query data from snowflake}

### pull from snowflake
# get plays
plays_data_raw = DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.PLAY_BY_PLAY')) %>%
        as_tibble() %>%
        mutate(TIME = gsub('\\{|}|"', '', CLOCK)) %>%
        separate(TIME, into=c("MINUTES", "SECONDS"), sep=",") %>%
        mutate(MINUTES = as.numeric(gsub("minutes:", "", MINUTES))) %>% 
        mutate(SECONDS = as.numeric(gsub("seconds:", "", SECONDS))) %>%
        mutate(MINUTES_IN_HALF = case_when(PERIOD == 1 ~ MINUTES+ 15,
                                           PERIOD == 2 ~ MINUTES,
                                           PERIOD == 3 ~ MINUTES + 15,
                                           PERIOD == 4 ~ MINUTES)) %>%
        mutate(SECONDS_IN_HALF = MINUTES_IN_HALF*60 + SECONDS) %>%
        # flag if seconds in half is invalid
        mutate(FLAG_SECONDS_IN_HALF = case_when(SECONDS_IN_HALF > 1800 ~ 1,
                                                TRUE ~ 0)) %>%
        # convert IDs to numeric
        mutate(ID = as.numeric(ID),
               DRIVE_ID = as.numeric(DRIVE_ID)) %>%
        rename(PLAY_ID = ID) %>%
        group_by(GAME_ID) %>%
        # sort each game by drive
        arrange(PLAY_ID) %>% 
        ungroup() %>%
        # clean up PERIOD by game
        mutate(STATUS_PERIOD = case_when(PERIOD == 0 & nchar(PLAY_TEXT) <=1 ~ 'Drop',
                                        PERIOD == 0 ~ 'Take Previous Value',
                                        TRUE ~ 'Valid')) %>%
        filter(STATUS_PERIOD != 'Drop') %>%
        mutate(PERIOD = case_when(STATUS_PERIOD == 'Take Previous Value' ~ lag(PERIOD, 1),
                                  TRUE ~ PERIOD)) %>%
        ungroup() %>% 
        mutate(HALF = case_when(PERIOD == 1 ~ 'First Half',
                                PERIOD == 2 ~ 'First Half',
                                PERIOD == 3 ~ 'Second Half',
                                PERIOD == 4 ~ 'Second Half',
                                TRUE ~ 'OT')) %>%
        # then clean up the down
        mutate(FLAG_DOWN = case_when(DOWN %in% c(1, 2, 3, 4) ~ 0,
                                       TRUE ~ 1)) %>%
        mutate(DOWN = case_when(DOWN %in% c(1, 2, 3, 4) ~ DOWN,
                                TRUE ~ -1)) %>%
        # then flag the distance
        mutate(FLAG_DISTANCE = case_when((DISTANCE < 0 | DISTANCE >=99) ~ 1,
                                         TRUE ~ 0)) %>%
        # flag yard lines that are outside correct
        mutate(FLAG_YARD_LINE = case_when(YARD_LINE < 0 | YARD_LINE > 100 ~ 1,
                                          TRUE ~ 0))

# get games
games_data_raw = DBI::dbGetQuery(myconn,
                             paste('SELECT * FROM CFB_DEMO.CFD_RAW.GAMES')) %>%
        as_tibble() %>%
        mutate(ID = as.numeric(ID)) %>%
        rename(GAME_ID = ID)
        
# get drives
drives_data_raw = DBI::dbGetQuery(myconn,
                              paste('SELECT * FROM CFB_DEMO.CFD_RAW.DRIVES')) %>%
        as_tibble() %>%
        mutate(ID = as.numeric(ID),
               GAME_ID = as.numeric(GAME_ID)) %>%
        rename(DRIVE_ID = ID) %>%
        # get drive time features
        # start of drive
        mutate(START_TIME = gsub('\\{|}|"', '', START_TIME)) %>%
        separate(START_TIME, into=c("START_MINUTES", "START_SECONDS"), sep=",") %>%
        mutate(START_MINUTES = as.numeric(gsub("minutes:", "", START_MINUTES))) %>% 
        mutate(START_SECONDS = as.numeric(gsub("seconds:", "", START_SECONDS))) %>%
        # end of drive
        mutate(END_TIME = gsub('\\{|}|"', '', END_TIME)) %>%
        separate(END_TIME, into=c("END_MINUTES", "END_SECONDS"), sep=",") %>%
        mutate(END_MINUTES = as.numeric(gsub("minutes:", "", END_MINUTES))) %>% 
        mutate(END_SECONDS = as.numeric(gsub("seconds:", "", END_SECONDS))) %>%
        # duration of drive
        mutate(ELAPSED = gsub('\\{|}|"', '', ELAPSED)) %>%
        separate(ELAPSED, into=c("ELAPSED_MINUTES", "ELAPSED_SECONDS"), sep=",") %>%
        mutate(ELAPSED_MINUTES = as.numeric(gsub("minutes:", "", ELAPSED_MINUTES))) %>% 
        mutate(ELAPSED_SECONDS = as.numeric(gsub("seconds:", "", ELAPSED_SECONDS))) 
        
```

```{r apply clean up functions}

# now get scoring events for the drives data
drives_data_score_events = drives_data_raw %>%
        group_by(GAME_ID) %>%
        arrange(DRIVE_ID) %>%
        mutate(HALF = case_when(START_PERIOD == 1 | START_PERIOD == 2 ~ 'First Half',
                                START_PERIOD == 3 | START_PERIOD == 4 ~ 'Second Half',
                                TRUE ~ 'OT')) %>%
        mutate(HOME = case_when(IS_HOME_OFFENSE == T ~ OFFENSE,
                                IS_HOME_OFFENSE ==F ~ DEFENSE),
               AWAY = case_when(IS_HOME_OFFENSE == T  ~ DEFENSE,
                                IS_HOME_OFFENSE ==F ~ OFFENSE)) %>%
        mutate(END_HOME_SCORE = case_when(IS_HOME_OFFENSE == T ~ END_OFFENSE_SCORE,
                                          IS_HOME_OFFENSE == F ~ END_DEFENSE_SCORE),
               END_AWAY_SCORE = case_when(IS_HOME_OFFENSE == F ~ END_OFFENSE_SCORE,
                                          IS_HOME_OFFENSE == T ~ END_DEFENSE_SCORE)) %>%
        group_by(GAME_ID, HALF) %>%
        mutate(LAST_DRIVE_HALF = case_when(DRIVE_NUMBER == max(DRIVE_NUMBER) ~ T,
                                           TRUE ~ F)) %>%
        ungroup() %>%
        group_by(GAME_ID) %>%
        mutate(HOME_SCORE_DIFF = END_HOME_SCORE - lag(END_HOME_SCORE, 1, default=0),
               AWAY_SCORE_DIFF = END_AWAY_SCORE - lag(END_AWAY_SCORE, 1, default=0)) %>%
        mutate(MULTIPLE_SCORE_CHANGES = HOME_SCORE_DIFF & AWAY_SCORE_DIFF > 0) %>%
        mutate(PROBLEM_AWAY_CHANGE = grepl("rush td|pass td|rushing td|passing td|fg good", tolower(DRIVE_RESULT)) & IS_HOME_OFFENSE == T & AWAY_SCORE_DIFF > 0,
               PROBLEM_HOME_CHANGE = grepl("rush td|pass td|rushing td|passing td|fg good", tolower(DRIVE_RESULT)) & IS_HOME_OFFENSE == F & HOME_SCORE_DIFF > 0) %>%
        mutate(SCORE_EVENT = case_when((tolower(DRIVE_RESULT) == 'td' | grepl("rush td|pass td|rushing td|passing td", tolower(DRIVE_RESULT))) & IS_HOME_OFFENSE == T ~ 'HOME TD',
                                       (tolower(DRIVE_RESULT) == 'td' | grepl("rush td|pass td|rushing td|passing td", tolower(DRIVE_RESULT))) & IS_HOME_OFFENSE == F ~ 'AWAY TD',
                                       grepl("rush td|pass td|rushing td|passing td", tolower(DRIVE_RESULT)) & IS_HOME_OFFENSE == T ~ 'HOME TD',
                                       grepl("rush td|pass td|rushing td|passing td", tolower(DRIVE_RESULT)) & IS_HOME_OFFENSE == F ~ 'AWAY TD',
                                       grepl("kickoff return td|kickoff td|kick return td|punt return td", tolower(DRIVE_RESULT)) & IS_HOME_OFFENSE ==T ~ 'AWAY TD',
                                       grepl("kickoff return td|kickoff td|kick return td|punt return td", tolower(DRIVE_RESULT)) & IS_HOME_OFFENSE ==F ~ 'HOME TD',
                                       (tolower(DRIVE_RESULT) == 'fg' | grepl("made fg|fg good", tolower(DRIVE_RESULT))) & IS_HOME_OFFENSE == T ~ 'HOME FG',
                                       (tolower(DRIVE_RESULT) == 'fg' | grepl("made fg|fg good", tolower(DRIVE_RESULT))) & IS_HOME_OFFENSE == F ~ 'AWAY FG',
                                       HOME_SCORE_DIFF == 6 | HOME_SCORE_DIFF == 7 | HOME_SCORE_DIFF == 8 ~ 'HOME TD',
                                       HOME_SCORE_DIFF == 3 ~ 'HOME FG',
                                       HOME_SCORE_DIFF == 2 ~ 'HOME Safety',
                                       AWAY_SCORE_DIFF == 6 | AWAY_SCORE_DIFF == 7 | AWAY_SCORE_DIFF == 8 ~ 'AWAY TD',
                                       AWAY_SCORE_DIFF == 3 ~ 'AWAY FG',
                                       AWAY_SCORE_DIFF == 2 ~ 'AWAY Safety', 
                                       (HOME_SCORE_DIFF == 0 & AWAY_SCORE_DIFF == 0) & LAST_DRIVE_HALF == T ~ 'No_Score')) %>%
        mutate(MISSING_SCORING_EVENT = SCORING == T & is.na(SCORE_EVENT)) %>%
        #  filter(MISSING_SCORING_EVENT == T) %>%
        mutate(NEXT_SCORE_EVENT = SCORE_EVENT) %>%
        # this fills in previous drives with no scoring event with the next observed scoring event in a half
        fill(NEXT_SCORE_EVENT, .direction = "up") %>%
        ungroup() 

```

```{r now join this with plays}

# now join with plays
plays_data_score_events = plays_data_raw %>%
        left_join(.,
                  drives_data_score_events %>%
                          select(GAME_ID,
                                 DRIVE_ID,
                                 HOME, 
                                 AWAY,
                                 LAST_DRIVE_HALF,
                                 END_HOME_SCORE,
                                 END_AWAY_SCORE,
                                 SCORE_EVENT,
                                 NEXT_SCORE_EVENT),
                  by = c("GAME_ID",
                         "DRIVE_ID",
                         "HOME", 
                         "AWAY")) %>%
        # now join with games to get seasons
        left_join(.,
                  games_data_raw %>%
                          select(GAME_ID,
                                 SEASON,
                                 HOME_TEAM,
                                 AWAY_TEAM),
                  by = c("GAME_ID")) %>%
        rename(NEXT_SCORE_EVENT_HOME = NEXT_SCORE_EVENT) %>%
        mutate(NEXT_SCORE_EVENT_OFFENSE = case_when(OFFENSE == HOME & NEXT_SCORE_EVENT_HOME == 'HOME TD' ~ 'TD',
                                                    OFFENSE == HOME & NEXT_SCORE_EVENT_HOME == 'AWAY TD' ~ 'Opp_TD',
                                                    OFFENSE == HOME & NEXT_SCORE_EVENT_HOME == 'HOME FG' ~ 'FG',
                                                    OFFENSE == HOME & NEXT_SCORE_EVENT_HOME == 'AWAY FG' ~ 'Opp_FG',
                                                    OFFENSE == HOME & NEXT_SCORE_EVENT_HOME == 'HOME Safety' ~ 'Safety',
                                                    OFFENSE == HOME & NEXT_SCORE_EVENT_HOME == 'AWAY Safety' ~ 'Opp_Safety',
                                                    OFFENSE == AWAY & NEXT_SCORE_EVENT_HOME == 'HOME TD' ~ 'Opp_TD',
                                                    OFFENSE == AWAY & NEXT_SCORE_EVENT_HOME == 'AWAY TD' ~ 'TD',
                                                    OFFENSE == AWAY & NEXT_SCORE_EVENT_HOME == 'HOME FG' ~ 'Opp_FG',
                                                    OFFENSE == AWAY & NEXT_SCORE_EVENT_HOME == 'AWAY FG' ~ 'FG',
                                                    OFFENSE == AWAY & NEXT_SCORE_EVENT_HOME == 'HOME Safety' ~ 'Opp_Safety',
                                                    OFFENSE == AWAY & NEXT_SCORE_EVENT_HOME == 'AWAY Safety' ~ 'Safety',
                                                    NEXT_SCORE_EVENT_HOME == 'No_Score' ~ 'No_Score')) %>%
        mutate(NEXT_SCORE_EVENT_HOME_DIFF = case_when(NEXT_SCORE_EVENT_HOME == 'HOME TD' ~ 7,
                                              NEXT_SCORE_EVENT_HOME == 'HOME FG' ~ 3,
                                              NEXT_SCORE_EVENT_HOME == 'HOME Safety' ~ 2,
                                              NEXT_SCORE_EVENT_HOME == 'No_Score' ~ 0,
                                              NEXT_SCORE_EVENT_HOME == 'HOME Safety' ~ -2,
                                              NEXT_SCORE_EVENT_HOME == 'HOME FG' ~ -3,
                                              NEXT_SCORE_EVENT_HOME == 'HOME TD' ~ -7)) %>%
        mutate(NEXT_SCORE_EVENT_OFFENSE_DIFF = case_when(NEXT_SCORE_EVENT_OFFENSE == 'TD' ~ 7,
                                              NEXT_SCORE_EVENT_OFFENSE == 'FG' ~ 3,
                                              NEXT_SCORE_EVENT_OFFENSE == 'Safety' ~ 2,
                                              NEXT_SCORE_EVENT_OFFENSE == 'No_Score' ~ 0,
                                              NEXT_SCORE_EVENT_OFFENSE == 'Opp_Safety' ~ -2,
                                              NEXT_SCORE_EVENT_OFFENSE == 'Opp_FG' ~ -3,
                                              NEXT_SCORE_EVENT_OFFENSE == 'Opp_TD' ~ -7)) %>%
        mutate(NEXT_SCORE_EVENT_OFFENSE = factor(NEXT_SCORE_EVENT_OFFENSE,
                                                     levels = c("TD",
                                                                "FG",
                                                                "Safety",
                                                                "No_Score",
                                                                "Opp_Safety",
                                                                "Opp_FG",
                                                                "Opp_TD")))
                                                 
```

# Learning Expected Points

In this notebook I develop and explore an expected points model at the play level for evaluating college football offenses and defenses. The goal of this analysis is to place a value on offensive/defensive plays in terms of their contribution's to a team's expected points. 

The data used is from college football games from 2003 to present. Each observation represents one play in a game, in which we know the team, the situation (down, time remaining), and the location on the field (yards to go, yards to reach end zone). We have information about the types of plays called as well in a text field.

## Sequences of Play

For each play in a game, I model the probability of the next scoring event that will occur within the same half. This means the analysis is not at the *drive* level, but at what I dub the *sequence* level. Suppose a team has the ball on offense to start the first half.  The next scoring event can take on one of seven outcomes:

* Touchdown (7 points)
* Field goal (3 points)
* Safety (2 points)
* No Score (0 points)
* Opp safety (2 points)
* Opp field goal (-3 points)
* Opp touchdown (-7 points)

If the team on offense drives down and scores a TD/FG, this will end the sequence. If the team on offense does not score but punts or turns the ball over, the sequence will continue with the other team now on offense. The sequence will continue until either one team scores, or the half comes to an end. From this, a sequence begins at kickoff and ends at the next kick off. 

Suppose we have two teams, A and B, playing in a game. Team A receives the opening kickoff, drives for a few plays, and then punts. Team B takes over, which starts drive 2, and they drive for a few plays before also punting. Team A then manages to put together a drive that finally scores. 

All plays on these three drives are one **sequence**. The outcome of this sequence is the points scored by Team A - if they score a touchdown, their points from this sequence is 7 (assuming for now they make the extra point). Team B's points from this sequence is -7 points.

When Team A kicks off to Team B to start drive 4, we start our next sequence, which will end either with one team scoring or at the end of the half. We'll then start over with a new sequence in the second half.

Why model the outcome of sequences rather than individual drives? Individual plays have the potential to affect both team's chances of scoring, positively or negatively, and we want our model to directly capture this. If an offense turns the ball over at midfield, they are not only hurting their own chances of scoring, they are increasing the other team's chance of scoring. The value of a play in terms of expected points is function of how both team's probabilities are affected by the outcome.

## Defining Expected Points

A team's expected points is sum of the probability of each possible scoring event multiplied by the points of that event. For this analysis, I assume that touchdowns equate to 7 rather than 6 points, assuming that extra points will be made. I can later bake in the actual probability of making extra points, but this will be a simplification for now.

For a given play $i$ for  Team $A$, we can compute Team A's expected points using the following:

$$ {Expected Points}_A = \\Pr(TD)*7 + \\ Pr(FG)*3 + \\Pr(Safety)*2 + \\ Pr(No Score)*0 + \\ Pr(Opp. Safety)*-2 + \\ Pr(Opp. FG) * -3 +\\ Pr(Opp. TD) * -7  $$

How do we get the probabilities of each scoring event? We learn these from historical data by using a model - I train a multinomial logistic regression model on many seasons worth of college football plays to learn how situations on the field affect the probability of the next scoring event. 

## Next Scoring Event

The outcome for our analysis is the **NEXT_SCORE_EVENT**. Each play in a given sequence contributes to the eventual outcome of the sequence. Here we can see an example of one game and its drives:

```{r show a full game of sequences}

drives_data_score_events %>%
        left_join(.,
                  games_data_raw %>%
                          select(GAME_ID,
                                 SEASON,
                                 HOME_TEAM,
                                 AWAY_TEAM),
                  by = c("GAME_ID")) %>%
        filter(HOME_TEAM == 'Texas A&M' & AWAY_TEAM == 'Florida') %>%
        filter(SEASON == 2012) %>%
        select(SEASON, HOME, AWAY, HALF, DRIVE_NUMBER, END_HOME_SCORE, END_AWAY_SCORE, NEXT_SCORE_EVENT) %>%
        rename(HOME_SCORE = END_HOME_SCORE,
               AWAY_SCORE = END_AWAY_SCORE) %>%
        mutate(SEASON = as.character(SEASON)) %>%
        flextable() %>%
        autofit()
        
```

For this game, we can filter to the plays that took place in the lead up to first score event. In this case, the first sequence included one drive and ended when Texas A&M kicked a field goal.

```{r zoom in on a sequence}

plays_data_score_events %>%
        filter(HOME_TEAM == 'Texas A&M' & AWAY_TEAM == 'Florida') %>%
        filter(SEASON == 2012) %>%
        filter(DRIVE_NUMBER == 1) %>%
        select(OFFENSE, DEFENSE, DRIVE_NUMBER, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, NEXT_SCORE_EVENT_OFFENSE) %>%
        rename(DRIVE = DRIVE_NUMBER,
               NEXT_SCORE_EVENT = NEXT_SCORE_EVENT_OFFENSE) %>%
        flextable() %>%
        autofit()
        
```

If we look at another sequence in the second half, there were multiple drives before a team was able to score in that sequence. The next scoring event is always defined from the perspective of the offense.

```{r zoom in on another sequence}

plays_data_score_events %>%
        filter(HOME_TEAM == 'Texas A&M' & AWAY_TEAM == 'Florida') %>%
        filter(SEASON == 2012) %>%
        filter(DRIVE_NUMBER>=9 & DRIVE_NUMBER <=14) %>%
        select(OFFENSE, DEFENSE, DRIVE_NUMBER, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, NEXT_SCORE_EVENT_OFFENSE) %>%
        rename(DRIVE = DRIVE_NUMBER,
               NEXT_SCORE_EVENT = NEXT_SCORE_EVENT_OFFENSE) %>%
        flextable() %>%
        autofit()
        
```

# Modeling Expected Points

Our goal is to understand how individual plays contribute to a team's **expected points**, or the average points teams should expect to have given their situation (down, time, possession). 

For instance, in the first drive of the Texas A&M-Florida game in 2012,  Texas A&M received the ball at their own 25 yard line to open the game. The simplest intuition of expected points is to ask, for teams starting at the 25 yard line at the beginning of a game, how many points do they typically go on to score? The answer is to look at all starting drives with 75 yards to go and see what the eventual next scoring event was for each of these plays - we take the average of all of the points that followed from this situation.

```{r expected points given situation}

plays_data_score_events %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        filter(!grepl("kick", tolower(PLAY_TEXT))) %>%
        filter(YARDS_TO_GOAL == 75 & DRIVE_NUMBER ==1) %>%
        group_by(NEXT_SCORE_EVENT_OFFENSE) %>%
        count(sort=T) %>% 
        ungroup() %>%
        mutate(prop = n / sum(n)) %>%
        mutate_if(is.numeric, round, 3) %>%
        flextable() %>%
        autofit() 

plays_data_score_events %>%
       # filter(SEASON >=2010) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        filter(YARDS_TO_GOAL == 75 & DRIVE_NUMBER ==1) %>%
        group_by(YARDS_TO_GOAL, DRIVE_NUMBER) %>%
        summarize(expected_points = mean(NEXT_SCORE_EVENT_OFFENSE_DIFF),
                  n = n(),
                  .groups = 'drop') %>%
        mutate_if(is.numeric, round, 2) %>%
        flextable() %>%
        autofit()

```

In this case, this means teams with the ball at their own 25 to start the game generally obtained more points on the ensuing sequence than their opponents, so they have a slightly positive expected points.

But, this is also a function of the down. If we look at the expected points for a team in this situation in first down vs a team in this situation for fourth down, we should see a drop in their expected points - by the time you hit fourth down, if you haven't moved from the 25, your expected points drops into the negatives, as you will now be punting the ball back to your opponent and it becomes more probable that they score than you.

```{r how does this vary as a function of }

plays_data_score_events %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(YARDS_TO_GOAL == 75  & DRIVE_NUMBER ==1) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        group_by(YARDS_TO_GOAL, DOWN, NEXT_SCORE_EVENT_OFFENSE) %>%
        count() %>%
        mutate(DOWN = paste("DOWN", DOWN, sep="_")) %>%
        spread(DOWN, n) %>%
        arrange(NEXT_SCORE_EVENT_OFFENSE) %>%
        arrange(desc(`DOWN_4`)) %>%
        flextable() %>%
        autofit()

plays_data_score_events %>%
        filter(YARDS_TO_GOAL == 75  & DRIVE_NUMBER ==1) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        group_by(YARDS_TO_GOAL, DOWN) %>%
        summarize(expected_points = mean(NEXT_SCORE_EVENT_OFFENSE_DIFF, na.rm=T),
                  n = n(),
                  .groups = 'drop') %>%
        mutate_if(is.numeric, round, 2) %>%
        flextable() %>%
        autofit()

```

The fact that the expected point changes based on the down and yard line allows us to look at the difference between expected points from play to play - the difference in expected points based on how the situation changed allows us to compute the Expected Points *Added* from a single play.

For any given play, we get a sense of the expected points a team can expect from their situation. For instance, if we look at all total plays in a game, how do expected points vary as a function of a team's distance from their opponent's goal line?

```{r expected points as a function of yard line for opening}

plays_data_score_events %>%
        filter(!grepl("kick", tolower(PLAY_TYPE))) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        filter(YARD_LINE < 100 & YARD_LINE > 0) %>%
      #  mutate(YARD_LINE_5 = plyr::round_any(YARDS_TO_GOAL, 5, floor)) %>%
     #   group_by(YARDS_TO_GOAL) %>%
        group_by(YARDS_TO_GOAL) %>%
        summarize(expected_points = mean(NEXT_SCORE_EVENT_OFFENSE_DIFF, na.rm=T),
                  n = n()) %>%
        ggplot(., aes(x=YARDS_TO_GOAL,
                      y=expected_points))+
        geom_line()+
        geom_point(aes(size=n))+
        theme_phil()+
        geom_hline(yintercept = 0,
                   linetype = 'dashed')+
        scale_x_reverse()

```

This should make sense - if you're backed up against your own end zone, your opponent has higher expected points because they are, historically, more likely to have the next scoring event, either by gaining good field advantage after you punt or by getting a safety. We can see this if we just look at the proportion of next scoring events based on the offense's position on the field.

```{r look at outcomes by yards to goal}

# make my own color ramp
my_col = colorRampPalette(c("blue", "white", "red"))
score_palette = c(my_col(6)[1:3], "grey80", my_col(6)[4:6])

plays_data_score_events %>%
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        filter(YARD_LINE < 100 & YARD_LINE > 0) %>%
      #  mutate(YARD_LINE_5 = plyr::round_any(YARDS_TO_GOAL, 5, floor)) %>%
     #   group_by(YARDS_TO_GOAL) %>%
        group_by(YARDS_TO_GOAL, NEXT_SCORE_EVENT_OFFENSE) %>%
        count() %>%
        ggplot(., aes(x=YARDS_TO_GOAL,
                      fill = NEXT_SCORE_EVENT_OFFENSE,
                      y=n))+
        geom_col(position = 'fill')+
        scale_fill_manual(values = score_palette)+
       # scale_fill_viridis_d(option = 'B')+
        theme_phil()+
        scale_x_reverse()

```

From this, when we see an offense move the ball up the field on a given play, we will generally see their expected points go up. The difference in expected points before the snap and after the snap is the value added (positively or negatively) by the play.

But, it's not just position on the field - it's also about the situation. If we look at how expected points varies by the down, we should see that fourth downs have lower expected points.

```{r look at expected points by down and field position}

plays_data_score_events %>%
        filter(PLAY_TYPE != 'Timeout') %>%
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        filter(YARD_LINE < 100 & YARD_LINE > 0) %>%
      #  mutate(YARD_LINE_5 = plyr::round_any(YARDS_TO_GOAL, 5, floor)) %>%
     #   group_by(YARDS_TO_GOAL) %>%
        group_by(YARDS_TO_GOAL, DOWN) %>%
        summarize(expected_points = mean(NEXT_SCORE_EVENT_OFFENSE_DIFF, na.rm=T),
                  n = n(),
                  .groups = 'drop') %>%
        mutate(DOWN = factor(DOWN)) %>%
        ggplot(., aes(x=YARDS_TO_GOAL,
                      group = DOWN,
                      color = DOWN,
                      y=expected_points))+
        geom_point(aes(size=n))+
        geom_line(stat = 'smooth',
                  formula = 'y ~ x',
                  method = 'loess',
                  span = 0.25)+
        theme_phil()+
        geom_hline(yintercept = 0,
                   linetype = 'dotted')+
        scale_x_reverse()+
        scale_color_viridis_d()

```

We also have other features like distance to convert the first down (filtering here to plays with a maximum of 30 yards to go, as we start to run out of data at higher values and it looks wonky).

```{r distance remaining}

plays_data_score_events %>%
        filter(FLAG_SECONDS_IN_HALF != 1) %>%
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        filter(YARD_LINE < 100 & YARD_LINE > 0) %>%
        mutate(SECONDS_IN_HALF = plyr::round_any(SECONDS_IN_HALF, 5, ceiling)) %>%
        filter(DISTANCE > 0 & DISTANCE<=30) %>%
        group_by(DOWN, DISTANCE, NEXT_SCORE_EVENT_OFFENSE) %>%
        mutate(DOWN = paste("DOWN", DOWN, sep="_")) %>%
        count() %>%
        ggplot(., aes(x=DISTANCE,
                      fill = NEXT_SCORE_EVENT_OFFENSE,
                      y=n))+
        #geom_col()+
        facet_wrap(DOWN ~., ncol = 2) +
        geom_bar(position = 'fill',
                 stat='identity',
                 color = NA)+
        scale_fill_manual(values = score_palette)+
       # scale_fill_viridis_d(option = 'B')+
        theme_phil()

```

And we also have info on time remaining in the half - as we might expect, the proportion of drives leading to no scoring goes up as the amount of time remaining in the half goes down.

```{r look at outcomes as a function of time remaining}

plays_data_score_events %>%
        filter(FLAG_SECONDS_IN_HALF != 1) %>%
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        filter(DOWN %in% c(1,2,3,4)) %>%
        filter(YARD_LINE < 100 & YARD_LINE > 0) %>%
        mutate(SECONDS_IN_HALF = plyr::round_any(SECONDS_IN_HALF, 5, ceiling)) %>%
        group_by(SECONDS_IN_HALF, NEXT_SCORE_EVENT_OFFENSE) %>%
        count() %>%
        ggplot(., aes(x=SECONDS_IN_HALF,
                      fill = NEXT_SCORE_EVENT_OFFENSE,
                      y=n))+
        geom_bar(position = 'fill',
                 stat='identity',
                 color = NA)+
        scale_fill_manual(values = score_palette)+
       # scale_fill_viridis_d(option = 'B')+
        theme_phil()+
        scale_x_reverse()

```

We use all of this historical data to learn the expected points from a given situation, then look at the difference in expected points from play to play - this is the intuition behind how we will value individual plays, which we can then roll up to the offense/defense/game/season level.

## Building Models

How do these various features like down, distance, yards to goal, and time remaining affect the probability of the next scoring event? We use a model to learn this relationship from historical plays. I'll now proceed to building the model which I'll use for the bulk of the analysis.

I'll set up training, validation, and test sets based around the season. I'm mostly going to build the model using plays from the 2010 season onwards, as the data quality of the play by play data starts to get worse the further back we go, though I'll do some backtesting of the model on older seasons.

I'm going to use the seasons of 2010-2018 as my main training set, building and evaluating the model using a leave-one-season out approach, akin to k-fold cross validation using seasons as the folds. I'll use the 2019-2020 seasons as a validation set, and leave 2021 as my test set which I won't look at till later on.

```{r set up training valid test splits, echo =T}

# full plays
plays_full = plays_data_score_events %>%
        filter(PLAY_TYPE != 'Kickoff') %>%
        select(GAME_ID,
               DRIVE_ID,
               PLAY_ID,
               SEASON,
               HOME,
               AWAY,
               OFFENSE,
               DEFENSE,
               OFFENSE_SCORE,
               DEFENSE_SCORE,
               SCORING,
               PLAY_TEXT,
               PLAY_TYPE,
               NEXT_SCORE_EVENT_HOME,
               NEXT_SCORE_EVENT_HOME_DIFF,
               NEXT_SCORE_EVENT_OFFENSE,
               NEXT_SCORE_EVENT_OFFENSE_DIFF,
               YARD_LINE,
               HALF,
               PERIOD,
               MINUTES_IN_HALF,
               SECONDS_IN_HALF,
               DOWN,
               DISTANCE,
               YARD_LINE,
               YARDS_TO_GOAL) %>%
        filter(DOWN %in% c(1, 2, 3, 4)) %>%
        filter(PERIOD %in% c(1,2,3,4)) %>%
        filter(!is.na(SECONDS_IN_HALF)) %>%
        filter(DISTANCE >=0 & DISTANCE <=100) %>%
        filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        mutate(NEXT_SCORE_EVENT_OFFENSE = factor(NEXT_SCORE_EVENT_OFFENSE,
                                                 levels = c("No_Score",
                                                            "TD",
                                                            "FG",
                                                            "Safety",
                                                            "Opp_Safety",
                                                            "Opp_FG",
                                                            "Opp_TD"))) %>%
        arrange(SEASON, GAME_ID, PLAY_ID)

# training set
plays_train = plays_full %>%
        filter(SEASON >= 2010 & SEASON <2019)

# validation set
plays_valid = plays_full %>%
        filter(SEASON >= 2019 & SEASON <= 2020)

# test
plays_test = plays_full %>%
        filter(SEASON > 2020)

# make an initial split based on previously defined splits
valid_split = make_splits(list(analysis = seq(nrow(plays_train)),
                                 assessment = nrow(plays_train) + seq(nrow(plays_valid))),
                               bind_rows(plays_train,
                                         plays_valid))

# test split
test_split = make_splits(
        list(analysis = seq(nrow(plays_train) + nrow(plays_valid)),
             assessment = nrow(plays_train) + nrow(plays_valid) + seq(nrow(plays_test))),
        bind_rows(plays_train,
                  plays_valid,
                  plays_test))

```

The outcome is the next scoring event, always defined from the perspective of the offense for any given play.

```{r outcome year over year}

plays_train %>%
        group_by(SEASON, NEXT_SCORE_EVENT_OFFENSE) %>%
        count() %>%
        ggplot(., aes(x=n,
                      fill = NEXT_SCORE_EVENT_OFFENSE,
                      y=reorder(NEXT_SCORE_EVENT_OFFENSE, n)))+
        geom_col()+
        theme_phil()+
        facet_wrap(SEASON ~.,
                   ncol = 3)+
        scale_fill_manual(values = score_palette)+
        ylab("Next Score Event")
```

I currently use the following as features for plays in a baseline model: 

* Quarter
* Seconds Remaining in Half 
* Down 
* Distance (logged)
* Yards to opponent's end zone
* Down and goal indicator for whether the offennse is in a 'first and goal' situation

I also include interactions between down and distance, down and yards to end zone, and yards to end zone and seconds remaining. This baseline model doesn't account for things like offense/defense quality or scoring effects, I'll train other models later on with these features to get things like 'Defense Adjusted Expected Points'. But this analysis is focused first and foremost on estimating probabilities 
 
I'll now set up a recipe for the baseline model.

```{r create recipes, echo=T}

baseline_recipe = recipe(NEXT_SCORE_EVENT_OFFENSE ~.,
                         data = plays_train) %>%
        update_role(all_predictors(),
                    new_role = "ID") %>%
        update_role(
                c("GAME_ID",
                  "DRIVE_ID",
                  "PLAY_ID",
                  "SEASON",
                  "HOME",
                  "AWAY",
                  "OFFENSE",
                  "DEFENSE",
                  "SCORING",
                  "OFFENSE_SCORE",
                  "DEFENSE_SCORE",
                  "PLAY_TEXT",
                  "PLAY_TYPE",
                  "NEXT_SCORE_EVENT_HOME",
                  "NEXT_SCORE_EVENT_HOME_DIFF",
                  "NEXT_SCORE_EVENT_OFFENSE_DIFF",
                  "YARD_LINE",
                  "MINUTES_IN_HALF",
                  "HALF"),
                new_role = "ID") %>%
        step_mutate(PERIOD_ID = PERIOD,
                    role = "ID") %>%
        # features we're inheriting
        update_role(
                c("PERIOD", 
                "SECONDS_IN_HALF",
                "DOWN",
                "DISTANCE",
                "YARDS_TO_GOAL"),
                new_role = "predictor") %>%
        # filters for issues
        step_filter(!is.na(NEXT_SCORE_EVENT_OFFENSE)) %>%
        step_filter(YARD_LINE <= 100 & YARD_LINE >=0) %>%
        step_filter(YARDS_TO_GOAL <=100 & YARD_LINE >=0) %>%
        step_filter(DOWN %in% c(1, 2, 3, 4)) %>%
        step_filter(DISTANCE >=0 & DISTANCE <=100) %>%
        step_filter(SECONDS_IN_HALF <=1800) %>%
        step_filter(!is.na(SECONDS_IN_HALF)) %>%
        step_filter(PERIOD_ID == 1 | PERIOD_ID == 2 | PERIOD_ID == 3 | PERIOD_ID == 4) %>%
        # create features
        step_mutate(KICKOFF = case_when(grepl("kickoff", tolower(PLAY_TEXT)) | grepl("kickoff", tolower(PLAY_TYPE))==T ~ 1,
                                        TRUE ~ 0)) %>%
        step_mutate(TIMEOUT = case_when(grepl("timeout", tolower(PLAY_TEXT)) ~ 1,
                                        TRUE ~ 0)) %>%
        step_filter(TIMEOUT != 1) %>%
        step_filter(KICKOFF != 1) %>%
        step_mutate(DOWN_TO_GOAL = case_when(DISTANCE == YARDS_TO_GOAL ~ 1,
                                             TRUE ~ 0)) %>%
        step_mutate(DOWN = factor(DOWN)) %>%
        step_mutate(PERIOD = factor(PERIOD)) %>%
        step_log(DISTANCE, offset =1) %>%
        step_dummy(all_nominal_predictors()) %>%
        step_novel(all_nominal_predictors(),
                   new_level = "new") %>%
        step_interact(terms = ~ DISTANCE:(starts_with("DOWN_"))) %>%
        step_interact(terms = ~ YARDS_TO_GOAL:(starts_with("DOWN_"))) %>%
        step_interact(terms = ~ YARDS_TO_GOAL*SECONDS_IN_HALF) %>%
        check_missing(all_predictors()) %>%
        step_normalize(all_numeric_predictors())

```

## Workflows

I'll define the model I'll be using here, which is a multinomial logistic regression.

```{r define model objs, echo=T}

# from glmnet
multinom_mod = multinom_reg(
  mode = "classification",
  engine = "glmnet",
  penalty = 0,
  mixture = NULL
)

```

I'll then create a workflow.

```{r craete workflow, echo=T}

# create multinomial workflow
multinom_wf = workflow() %>%
        add_recipe(baseline_recipe) %>%
        add_model(multinom_mod)

# workflow settings
# metrics
class_metrics<-metric_set(yardstick::roc_auc,
                          yardstick::mn_log_loss)

# control for resamples
keep_pred <- control_resamples(save_pred = TRUE, 
                               save_workflow = TRUE,
                               allow_par=T)

```

I'll manually define resamples based on the seasons - rather than doing k-fold cross validation, I'll assign each season to be a fold and train and assess the model leaving one season out at a time.

```{r set up manual resamples by season}

# get unique seasons
seasons = unique(plays_train$SEASON) %>%
        sort

# define indices by seasons
indices = foreach(i = 1:length(seasons)) %do% {
        list(analysis = which(plays_train$SEASON != seasons[i]),
             assessment = which(plays_train$SEASON == seasons[i]))
}

# get the splits
splits <- lapply(indices, make_splits, data = plays_train)

# define rset object
manual_resamples = manual_rset(splits, 
                               paste("holdout:", seasons))

```

```{r register parallel cores}

# register parallel
library(doParallel)
numcores = parallel::detectCores()
registerDoParallel(numcores -2)

```

## Training

Now I'll train and assess the model on the training set via leave-one-season-out cross validation, then I'll refit the model to the entire training set.

```{r fit on set, echo=T}

# fit to resamples
resamples_multinom = multinom_wf %>%
        fit_resamples(data = plays_train,
                      metrics = class_metrics,
                      resamples = manual_resamples,
                      control = keep_pred,
                      verbose=T)

# # # save locally so as to not need to retrain everytime
# write_rds(resamples_multinom,
#      file = here::here("models", "resamples_expected_points.Rds"),
#      compress = "gz")

```

```{r load previously run resamples}

# resamples_multinom = read_rds(here::here("models", 
#                                          "resamples_expected_points.Rds"))

```


```{r fit to training set, echo=T}

# fit the model to the whole training set
fit_multinom = multinom_wf %>%
        fit(data = plays_train)

```

## Resampling Performance

How did the model perform on each resample? I'm looking at the logloss and area under the receiver operating characteristic curve (roc_auc). The logloss is only meaningful by comparison, I'll compare how the model does to a null model on the validation set.

```{r get results}

# collect metrics
resamples_multinom %>%
        collect_metrics(summarize=F) %>%
        mutate_if(is.numeric, round, 3) %>%
        select(id, .metric, .estimator, .estimate) %>%
        arrange(.metric) %>%
        flextable() %>%
        autofit()

```

## Inference

```{r get coefs, fig.height=10, fig.width=8, warning=F, message=F}

tidy(fit_multinom) %>%
        mutate(term = factor(term)) %>%
        mutate(term = relevel(term, ref = "(Intercept)")) %>%
        ggplot(., aes(y=term,
                      x=estimate))+
        geom_point()+
        facet_wrap(class~.)+
        theme_bw()+
        ggtitle("Coefficients from Multinomial Logistic Regression")

```

Understanding partial effects from a multinomial logit is already difficult, and I've thrown a bunch of interactions in there to make this even more difficult. 

I'll look at predicted probabilities using an observed values approach for particular features (using a sample rather than the full dataset to save time). This means taking the model and then altering the feature of interest for every observation and taking the average predicted probability for each outcome across all observations.

```{r look at predicted probabilities}

# define min and max for feature
setting = seq(1, 100, 2)

# loop over
pred_settings = foreach(i = 1:length(setting),
        .combine = bind_rows) %do% {
                
                set.seed(1999)
                fit_multinom %>%
                        predict(new_data = plays_train %>% 
                                        sample_n(50000) %>%
                                        mutate(YARDS_TO_GOAL = setting[i]),
                                type = 'prob') %>%
                        apply(., 2, mean) %>%
                        as.data.frame() %>%
                        rownames_to_column("outcome") %>% set_names(., c("outcome", ".pred")) %>%
                        mutate(YARDS_TO_GOAL = setting[i])
}

```

How is the probability of the next scoring event influenced by where the offense has possession?

```{r show probabilities by yards to goal}

# now plot this effect
pred_settings %>%
        mutate(outcome = paste(gsub(".pred_", "Pr(", outcome), ")", sep="")) %>%
             mutate(outcome = factor(outcome,
                                levels = paste("Pr(", c("TD",
                                      "FG",
                                      "Safety",
                                      "No_Score",
                                      "Opp_Safety",
                                      "Opp_FG",
                                      "Opp_TD"),
                                      ")",
                                      sep =""))) %>%
        ggplot(., 
               aes(x=YARDS_TO_GOAL,
               y = .pred,
               color = outcome,
               group = outcome))+
        facet_wrap(outcome ~.,
                   ncol = 4)+
        geom_line(lwd=1.1)+
        # scale_color_viridis_d(option = 'B',
        #                    end = 0.8,
        #                    begin = 0.2)+        
        scale_color_manual(values = score_palette)+
        theme_phil()+
        theme(legend.title = element_text())+
        guides(color = guide_legend(title = 'Next Scoring Event',
                                    title.position = 'top'))+
        xlab("Yards to Opponent End Zone")+
        ylab("Pr(Next Scoring Event)")+
        coord_cartesian(ylim= c(0, 1))+
        geom_vline(xintercept = 50,
                   linetype = 'dotted')+
        ggtitle("Probability of Next Scoring Event by Offensive Field Position",
                subtitle = str_wrap("Predicted probabilities from multinomial logistic regression. Displaying probabilities using observed values approach for a sample of training set. Model trained on regular season college football plays from 2012-2020. ",
                                    120))

# now plot this effect
pred_settings %>%
        mutate(outcome = paste(gsub(".pred_", "Pr(", outcome), ")", sep="")) %>%
             mutate(outcome = factor(outcome,
                                levels = paste("Pr(", c("TD",
                                      "FG",
                                      "Safety",
                                      "No_Score",
                                      "Opp_Safety",
                                      "Opp_FG",
                                      "Opp_TD"),
                                      ")",
                                      sep =""))) %>%
        ggplot(., 
               aes(x=YARDS_TO_GOAL,
               y = .pred,
               fill = outcome,
               group = outcome))+
        geom_col(position = 'fill',
                 lwd=1.1)+
        # scale_fill_viridis_d(option = 'B',
        #                    end = 0.8,
        #                    begin = 0.2)+        
        scale_fill_manual(values = score_palette)+
        theme_phil()+
        theme(legend.title = element_text())+
        guides(fill = guide_legend(title = 'Next Scoring Event',
                                    title.position = 'top'))+
        xlab("Yards to Opponent End Zone")+
        ylab("Pr(Next Scoring Event)")+
        coord_cartesian(ylim= c(0, 1))+
        geom_vline(xintercept = 50,
                   linetype = 'dotted')+
        ggtitle("Probability of Next Scoring Event by Offensive Field Position",
                subtitle = str_wrap("Predicted probabilities from multinomial logistic regression. Displaying probabilities using observed values approach for a sample of training set. Model trained on regular season college football plays from 2012-2020. ",
                                    120))


```

How is this affected by the down?

```{r loop over yards to goal and down}

# define min and max for feature
setting_1 = seq(1, 100, 5)
setting_2 = c(1, 2, 3, 4)

# loop over both
pred_settings_12 = foreach(i = 1:length(setting_1),
                        .combine = bind_rows) %:% 
        foreach(j = 1:length(setting_2)) %do% {
                
                set.seed(1999)
                 fit_multinom %>%
                        predict(new_data = plays_train %>% 
                                        sample_n(50000) %>%
                                        mutate(YARDS_TO_GOAL = setting_1[i],
                                               DOWN = setting_2[j]),
                                type = 'prob') %>%
                        apply(., 2, mean) %>%
                        as.data.frame() %>%
                        rownames_to_column("outcome") %>% 
                        set_names(., c("outcome", ".pred")) %>%
                        mutate(YARDS_TO_GOAL = setting_1[i],
                               DOWN = setting_2[j])
                
        }
                
```

```{r look at predicted probs by down and yards to goal}

# now plot this effect
pred_settings_12 %>%
        mutate(outcome = paste(gsub(".pred_", "Pr(", outcome), ")", sep="")) %>%
             mutate(outcome = factor(outcome,
                                levels = paste("Pr(", c("TD",
                                      "FG",
                                      "Safety",
                                      "No_Score",
                                      "Opp_Safety",
                                      "Opp_FG",
                                      "Opp_TD"),
                                      ")",
                                      sep =""))) %>%
        mutate(DOWN = factor(DOWN)) %>%
        ggplot(., 
               aes(x=YARDS_TO_GOAL,
               y = .pred,
               color = DOWN,
               group = DOWN))+
        facet_wrap(outcome ~.,
                   ncol = 4)+
        geom_line(lwd=1.1)+
        scale_color_viridis_d(option = 'D',
                           begin = 0.2)+        
        theme_phil()+
        theme(legend.title = element_text())+
        guides(color = guide_legend(title = 'Down',
                                    title.position = 'top'))+
        xlab("Yards to Opponent End Zone")+
        ylab("Pr(Next Scoring Event)")+
        ggtitle("Probability of Next Scoring Event by Offensive Field Position and Down",
                subtitle = str_wrap("Predicted probabilities from multinomial logistic regression. Displaying probabilities using observed values approach for a sample of training set. Model trained on regular season college football plays from 2012-2020. ",
                                    120))

```

How does this translate into expected points?

```{r make expected points function}

# define function here
expected_points_func = function(x) {

        x %>%
                mutate(EP = 0 * `.pred_No_Score` + 
                               7 * .pred_TD + 3 * .pred_FG + 
                               2 * .pred_Safety + 
                               -2*`.pred_Opp_Safety` + 
                               -3*`.pred_Opp_FG` + 
                               -7*`.pred_Opp_TD`)
        
}

```

```{r expected points by distance and down}

# loop over both
point_settings_12 = foreach(i = 1:length(setting_1),
                        .combine = bind_rows) %:% 
        foreach(j = 1:length(setting_2)) %do% {
                
                set.seed(1999)
                 fit_multinom %>%
                        predict(new_data = plays_train %>% 
                                        sample_n(50000) %>%
                                        mutate(YARDS_TO_GOAL = setting_1[i],
                                               DOWN = setting_2[j]),
                                type = 'prob') %>%
                         expected_points_func %>%
                         summarize(EP = mean(EP)) %>%
                        mutate(YARDS_TO_GOAL = setting_1[i],
                               DOWN = setting_2[j])
                
        }

```

```{r plot expected points by distance and down}

# now plot this effect
point_settings_12 %>%
        mutate(DOWN = factor(DOWN)) %>%
        ggplot(., 
               aes(x=YARDS_TO_GOAL,
               y = EP,
               color = DOWN,
               group = DOWN))+
        # facet_wrap(outcome ~.,
        #            ncol = 4)+
        geom_line(lwd=1.1)+
        scale_color_viridis_d(option = 'D',
                           begin = 0.2)+        
        theme_phil()+
        geom_hline(yintercept = 0,
                   linetype = 'dashed')+
        theme(legend.title = element_text())+
        guides(color = guide_legend(title = 'Down',
                                    title.position = 'top'))+
        xlab("Yards to Opponent End Zone")+
        ylab("Expected Points")+
        ggtitle("Expected Points by Offensive Field Position and Down",
                subtitle = str_wrap("Expected points using probabilities from multinomial logistic regression. Displaying expected points a portion of the training set using observed values approach, Model trained on regular season college football plays from 2012-2020. ",
                                    120))+
        geom_vline(xintercept = 50,
                   linetype = 'dotted')

```

## Validation Set

We can evaluate the model via our leave-one-out approach, but we'll also predict the validation set as an additional check. I'll compare performance relative to a null model that simply predicts the incidence rate of each outcome in the training set.

```{r predict validation set}

preds_multinom = fit_multinom %>%
        predict(new_data = plays_valid,
                type = 'prob') %>%
        bind_cols(., 
                  fit_multinom %>%
                          predict(new_data = plays_valid)) %>%
        bind_cols(plays_valid,
                  .)

```


```{r get null preds}

# baseline rates in training set
train_rates = plays_train %>%
        group_by(NEXT_SCORE_EVENT_OFFENSE) %>%
        count() %>%
        ungroup() %>%
        mutate(prop = n / sum(n))

# predict null for all valid
preds_null = train_rates %>% 
        select(-n) %>%
        spread(NEXT_SCORE_EVENT_OFFENSE, prop) %>% 
        set_names(., paste(".pred_", names(.), sep="")) %>%
        mutate(.pred_class = factor('TD',
                                    levels = c("TD",
                                      "FG",
                                      "Safety",
                                      "No_Score",
                                      "Opp_Safety",
                                      "Opp_FG",
                                      "Opp_TD"))) %>%
        bind_cols(plays_valid,
                  .)
```


```{r yardstick assessment on validation, warning=F, message=F}

# roc_auc
bind_rows(preds_null %>%
                  mutate(method = 'null'),
          preds_multinom %>%
                  mutate(method = 'multinom')) %>%
        group_by(method) %>%
        yardstick::roc_auc(
                   NEXT_SCORE_EVENT_OFFENSE,
                   `.pred_No_Score`:`.pred_Opp_TD`,
           #        estimator = 'macro_weighted'
                   ) %>%
        mutate_if(is.numeric, round, 3) %>%
        flextable() %>%
        autofit()

# logloss
bind_rows(preds_null %>%
                  mutate(method = 'null'),
          preds_multinom %>%
                  mutate(method = 'multinom')) %>%
        group_by(method) %>%
        yardstick::mn_log_loss(
                   NEXT_SCORE_EVENT_OFFENSE,
                   `.pred_No_Score`:`.pred_Opp_TD`,
           #        estimator = 'macro_weighted'
                   ) %>%
        mutate_if(is.numeric, round, 3) %>%
        flextable() %>%
        autofit()

```

What's the log loss for each outcome?

```{r subpopulation log loss}

# roc_auc
bind_rows(preds_null %>%
                  mutate(method = 'null'),
          preds_multinom %>%
                  mutate(method = 'multinom')) %>%
        group_by(method, NEXT_SCORE_EVENT_OFFENSE) %>%
        yardstick::mn_log_loss(
                   NEXT_SCORE_EVENT_OFFENSE,
                   `.pred_No_Score`:`.pred_Opp_TD`,
                   na_rm=T,
           #        estimator = 'macro_weighted'
                   ) %>%
        mutate_if(is.numeric, round, 3) %>%
        spread(method, .estimate) %>%
        flextable() %>%
        autofit()

```

```{r rm}

rm(splits,
   pred_settings,
   pred_settings_12,
   point_settings_12)

```

# Examining Results

I'll now start diving into the predictions for individual plays as a means to evaluate plays and teams.

It's worth noting that we might see some season-level differences that make comparison across seasons difficult, since the predictions are all coming from slightly different models due to resampling.

```{r combine resamples with validation predictions}

predicted_plays = 
        resamples_multinom %>%
        select(.predictions) %>% 
        unnest() %>%
        bind_cols(., plays_train %>%
                          select(-NEXT_SCORE_EVENT_OFFENSE)) %>%
        bind_rows(.,
                  preds_multinom) %>%
        expected_points_func() # convert to expected points

```

I'll get Expected Points Added for all non scoring plays. This part can be a little wobbly, due to data quality issues with defining sequences. The basic thought here is to say, at the start of a play, we know the expected points for a team in that situation, EP_Pre. We then look to the next play to see the expected points for the team after the result of the previous play, EP_Post. EP_Added is the difference between these two outcomes from the perspective of the offense. 

This means that if the ball is turned over, but not scored, the team on offense becomes the defense and the sign of the expected points on the next play flips for their calculation. For events that produce touchdowns, EP_Added is empty, but I create another feature simply called Points_Added in which I take the difference between EP_Pre and the points scored on the play, 7 for touchdowns, 3 for FGs, 2 for safeties.

```{r compute expected points for all plays}

scored_plays = predicted_plays %>%
        select(SEASON, GAME_ID, DRIVE_ID, PLAY_ID, HALF, PERIOD, OFFENSE, DEFENSE, HOME, AWAY, OFFENSE_SCORE, DEFENSE_SCORE,SECONDS_IN_HALF, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, EP, PLAY_TYPE, NEXT_SCORE_EVENT_OFFENSE) %>%
        # compute home and away score
        mutate(HOME_SCORE = case_when(OFFENSE == HOME ~ OFFENSE_SCORE,
                                      DEFENSE == HOME ~ DEFENSE_SCORE),
               AWAY_SCORE = case_when(DEFENSE == HOME ~ OFFENSE_SCORE,
                                      OFFENSE == HOME ~ DEFENSE_SCORE)) %>%
        # make features to denote scoring plays
        mutate(TOUCHDOWN = case_when(
                (grepl("TOUCHDOWN", PLAY_TEXT) | PLAY_TYPE == 'TD' | grepl("td|touchdown", tolower(PLAY_TYPE))) ~ 1,
                                     TRUE ~ 0)) %>%
        mutate(FG = case_when(
                (grepl("Field Goal", PLAY_TEXT) | PLAY_TYPE == 'FG' | grepl("fg|field goal", tolower(PLAY_TYPE))) ~ 1,
                                     TRUE ~ 0)) %>%
        mutate(SAFETY = case_when(
                grepl("safety", tolower(PLAY_TEXT)) ~ 1,
                 TRUE ~ 0)) %>%
        # compute the actual points after scoring plays
        mutate(Points_Post = case_when(TOUCHDOWN==1 & NEXT_SCORE_EVENT_OFFENSE == 'TD' ~ 7,
                                        TOUCHDOWN==1 & NEXT_SCORE_EVENT_OFFENSE == 'Opp_TD' ~ -7,
                                        FG==1 & NEXT_SCORE_EVENT_OFFENSE == 'FG' ~ 3,
                                        FG==1 & NEXT_SCORE_EVENT_OFFENSE == 'Opp_FG' ~ -3,
                                        SAFETY==1 & NEXT_SCORE_EVENT_OFFENSE == 'Safety' ~ 2,
                                        SAFETY==1 & NEXT_SCORE_EVENT_OFFENSE == 'Opp Safety' ~ -2)) %>%
        # rename expected points for play as EPA_Pre
        rename(EP_Pre = EP)  %>%
        # now, compute EPA_Pre, EPA_Post, EPA_Added for individual drives
        # all of these calculations are keeping plays in the order in which they appear from the API
         group_by(GAME_ID, OFFENSE, HALF, DRIVE_ID) %>%
        # compute EP_Post, the expected points of the situation in the next play. 
        # for scoring plays we'll be computing a slightly different quantity
        mutate(EP_Post = case_when(
                (TOUCHDOWN == 0 & FG ==0 & SAFETY==0)  ~ dplyr::lead(EP_Pre,1))
               ) %>%
        mutate(EP_Added = EP_Post-EP_Pre) %>%
        # unfortunately, this doesn't catch what happens if possession changes before a scoring event
        # define a sequence as plays within a half where the score remains the same
        # add a sequence id
        bind_cols(.,
                  group_indices(., GAME_ID, HALF, HOME_SCORE, AWAY_SCORE) %>% # this creates a group index
                          as_tibble() %>%
                          rename(SEQUENCE = value)) %>%group_by(GAME_ID, HALF, SEQUENCE) %>%
        # now, if EP_Post is NA, it is typically because a drive ended without a scoring event
        # in the event that possession changed, flip the sign of the expected points for the next play (ie, what is the other teams EP now that they have the ball)
        # the difference in this case will be the result of the turnover/change in possession
        mutate(EP_Post = case_when(
                (is.na(EP_Post) & (TOUCHDOWN == 0 & FG ==0 & SAFETY==0) & (OFFENSE!=dplyr::lead(OFFENSE,1))) ~ -1*(dplyr::lead(EP_Pre,1)),
                 (is.na(EP_Post) & (TOUCHDOWN == 0 & FG ==0 & SAFETY==0) & (OFFENSE==dplyr::lead(OFFENSE,1))) ~ 1*(dplyr::lead(EP_Pre,1)),
                                   TRUE ~ EP_Post)) %>%
        mutate(EP_Added = case_when(
                (is.na(EP_Added) & (TOUCHDOWN == 0 & FG ==0 & SAFETY==0)) ~ EP_Post-EP_Pre,
                                   TRUE ~ EP_Added)) %>%
        # Points added by scoring plays
        mutate(Points_Added = Points_Post-EP_Pre) %>%
        ungroup()

```

## Game Results

I'll look at a few games, play by play, to get a sense of how this is looking. I'll pick one game completely at random, in no way influenced by my fandom: Texas A&M vs Alabama in 2012.

What were the most impactful plays in this game in terms of expected points added? I'll look at the 15 plays in this game with the largest absolute change.

```{r look at a game from 2013}

scored_plays %>%
        filter(SEASON == 2012) %>%
        filter(HOME == 'Alabama' & AWAY == 'Texas A&M') %>%
        select(SEASON, GAME_ID, DRIVE_ID, PLAY_ID, PERIOD, OFFENSE, DEFENSE, PERIOD, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, EP_Pre, EP_Post, EP_Added) %>%
        mutate_at(c("SEASON",
                    "GAME_ID",
                    "DRIVE_ID",
                    "PLAY_ID"),
                  ~ as.character(.)) %>%
        rename(YTG = YARDS_TO_GOAL,
               PLAY = PLAY_TEXT) %>%
        select(-DRIVE_ID, -GAME_ID, -PLAY_ID) %>%
        arrange(desc(abs(EP_Added))) %>%
        head(15) %>%
        mutate_if(is.numeric, round, 2) %>%
 #       arrange(desc(EP_Added)) %>%
        flextable() %>%
        autofit() %>%
        color(., i = ~ EP_Added > 0,
           j = "EP_Added",
           color = 'deepskyblue1') %>%
        color(., i = ~ EP_Added < 0,
           j = "EP_Added",
           color = 'red')

```

Kind of interesting - this game is remembered for a lot of plays by Johnny Manziel, but the most impactful plays in the game in terms of expected points changes were actually turnovers forced by the A&M defense.

However, this doesn't include scoring plays. The Points_Added measure looks at the difference between actual points scored vs expected points from the situation.

```{r top plays my expected points}

scored_plays %>%
        filter(SEASON == 2012) %>%
        filter(HOME == 'Alabama' & AWAY == 'Texas A&M') %>%
        select(SEASON, GAME_ID, DRIVE_ID, PLAY_ID, PERIOD, OFFENSE, DEFENSE, PERIOD, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, EP_Pre, Points_Added) %>%
        mutate_at(c("SEASON",
                    "GAME_ID",
                    "DRIVE_ID",
                    "PLAY_ID"),
                  ~ as.character(.)) %>%
        rename(YTG = YARDS_TO_GOAL,
               PLAY = PLAY_TEXT) %>%
        select(-DRIVE_ID, -GAME_ID, -PLAY_ID) %>%
        arrange(desc(abs(Points_Added))) %>%
        head(5) %>%
        mutate_if(is.numeric, round, 2) %>%
 #       arrange(desc(EP_Added)) %>%
        flextable() %>%
        autofit() %>%
        color(., i = ~ Points_Added > 0,
           j = "Points_Added",
           color = 'deepskyblue1') %>%
        color(., i = ~ Points_Added < 0,
           j = "Points_Added",
           color = 'red')

```

In this case, we can still use the expected points of the situation to give us a measure of how many points the play added. In the case of Christine Michael's 1 yard TD run, the points added are pretty low, as you'd expect teams to score from that situation. By comparison, AJ McCarron's 54 yard pass on 3rd and 6 in the 4th quarter had a much higher points added, as they were in a third down situation at the midfield and made a huge play. For this reason, what I'm calling Points Added has been dubbed by others as a measure of a team's offensive explosiveness. For most of my analyses I'm condensing Expected Points Added and points Added into one metric that I'm calling Predicted Points Added.

What's cool is that we can do this (right now) for any play from any game from 2010-2020. For instance, we can look at a team and ask, what were its highest predicted points added plays in each season? This is Wisconsin's top plays for each year.

```{r look at a wisconsin game play by play}

scored_plays %>%
        select(SEASON, GAME_ID, DRIVE_ID, PLAY_ID, PERIOD, OFFENSE, DEFENSE, PERIOD, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, EP_Pre, EP_Post, EP_Added, Points_Added) %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        filter(OFFENSE == 'Wisconsin') %>%
        group_by(SEASON) %>%
        slice_max(., order_by = PPA,
                  n=1) %>%
        mutate_at(c("SEASON",
                    "GAME_ID",
                    "DRIVE_ID",
                    "PLAY_ID"),
                  ~ as.character(.)) %>%
        rename(YTG = YARDS_TO_GOAL,
               PLAY = PLAY_TEXT) %>%
        select(-DRIVE_ID, -GAME_ID, -PLAY_ID) %>%
        ungroup() %>%
        mutate_if(is.numeric, round, 2) %>%
        flextable() %>%
        autofit() %>%
        color(., i = ~ PPA > 0,
           j = "PPA",
           color = 'deepskyblue1') %>%
        color(., i = ~ PPA < 0,
           j = "PPA",
           color = 'red')


```

 What were the highest predicted points added plays (including scoring plays) from each season?

```{r look at a game from 2014}

scored_plays %>%
        select(SEASON, GAME_ID, DRIVE_ID, PLAY_ID, PERIOD, OFFENSE, DEFENSE, PERIOD, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, EP_Pre, EP_Post, EP_Added, Points_Added) %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        group_by(SEASON) %>%
        slice_max(., order_by = PPA,
                  n=1) %>%
        mutate_at(c("SEASON",
                    "GAME_ID",
                    "DRIVE_ID",
                    "PLAY_ID"),
                  ~ as.character(.)) %>%
        rename(YTG = YARDS_TO_GOAL,
               PLAY = PLAY_TEXT) %>%
        select(-DRIVE_ID, -GAME_ID, -PLAY_ID) %>%
        ungroup() %>%
        mutate_if(is.numeric, round, 2) %>%
        flextable() %>%
        autofit() %>%
        color(., i = ~ PPA > 0,
           j = "PPA",
           color = 'deepskyblue1') %>%
        color(., i = ~ PPA < 0,
           j = "PPA",
           color = 'red')

```

Looks like there's a data quality issue with the Iowa vs Northern Iowa game, the yards to go looks to be incorrect there which is making that play look like it was a big yardage gain. We also notice some data quality issues if we look at highest expected points added plays, not including scoring plays.

```{r look for issues}

scored_plays %>%
        select(SEASON, GAME_ID, DRIVE_ID, PLAY_ID, PERIOD, OFFENSE, DEFENSE, PERIOD, DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, EP_Pre, EP_Post, EP_Added, Points_Added) %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        group_by(SEASON) %>%
        slice_max(., order_by = EPA,
                  n=1) %>%
        mutate_at(c("SEASON",
                    "GAME_ID",
                    "DRIVE_ID",
                    "PLAY_ID"),
                  ~ as.character(.)) %>%
        rename(YTG = YARDS_TO_GOAL,
               PLAY = PLAY_TEXT) %>%
        select(-DRIVE_ID, -GAME_ID, -PLAY_ID) %>%
        ungroup() %>%
        mutate_if(is.numeric, round, 2) %>%
        flextable() %>%
        autofit() %>%
        color(., i = ~ EPA > 0,
           j = "EPA",
           color = 'deepskyblue1') %>%
        color(., i = ~ EPA < 0,
           j = "EPA",
           color = 'red')
```

For that 2011 play, it's 4th and 58 with 93 yards to go and the pass is complete for a loss of seven yards, and yet the model considers this a play worthy of 9 expected points? The heck? Turns out this, and most of these, are actually just issues of data quality with ESPNs play by play data.

```{r look at example of a problem}

scored_plays %>%
        filter(SEASON == 2011 & OFFENSE == 'Troy' & DEFENSE == 'Clemson' & PERIOD ==3) %>%
        select(SEASON, GAME_ID, DRIVE_ID, PLAY_ID, PERIOD, OFFENSE, DEFENSE, PERIOD,  DOWN, DISTANCE, YARDS_TO_GOAL, PLAY_TEXT, EP_Pre, EP_Post, EP_Added, Points_Added) %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        mutate_at(c("SEASON",
                    "GAME_ID",
                    "DRIVE_ID",
                    "PLAY_ID"),
                  ~ as.character(.)) %>%
        rename(YTG = YARDS_TO_GOAL,
               PLAY = PLAY_TEXT) %>%
        filter(PLAY_ID %in% c(312460228140,
                              312460228160,
                              312460228163)) %>%
        select(-DRIVE_ID, -GAME_ID, -PLAY_ID) %>%
        ungroup() %>%
        mutate_if(is.numeric, round, 2) %>%
        flextable() %>%
        autofit() %>%
        color(., i = ~ EPA > 0,
           j = "EPA",
           color = 'deepskyblue1') %>%
        color(., i = ~ EPA < 0,
           j = "EPA",
           color = 'red')

```

In this case, the play by play data has a bizarre sequence in which Troy is listed as punting, then completing a loss on 4th and 58, while still having possession on the next play with a 1st and 10. Since the expected points added calculation is based on the difference between plays, that phantom 4th and 58 play throws off the value of the next play. This is unfortunately quite common in ESPNs play by play data, especially as we go further back.

## Team Offenses by Season

Having scored all individual plays, we can now roll this up to whatever level of analysis we're interested in. We can, for instance, look at a team's offense for each season over this time period. Continuing to select a team at random, we'll look at Texas A&M's offense in terms of predicted/expected points game by game over this time period. The line in navy, EPA_Average, looks at a team's expected points added per play without including scoring plays. The line in light blue, PPA_Average, looks at all plays, including scoring.

```{r take a look at a team by season}

int_breaks <- function(x, n = 5) {
  l <- pretty(x, n)
  l[abs(l %% 1) < .Machine$double.eps ^ 0.5] 
}

scored_plays %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        # exclude kickoffs from this
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        # also field goals, not that interested in kicking performance at the minute
        filter(FG == 0) %>% # remove FGs
        filter(OFFENSE == 'Texas A&M') %>%
        group_by(SEASON, OFFENSE) %>%
        summarize(EPA_Average = mean(EPA, na.rm=T),
             #     PA_Average = mean(PA, na.rm=T),
                  PPA_Average = mean(PPA, na.rm=T),
                  plays = n(),
                  .groups = 'drop') %>%
        rename(TEAM = OFFENSE) %>%
        mutate(TYPE = "OFFENSE") %>%
        select(SEASON, 
               TEAM,
               TYPE,
               EPA_Average, 
           #    PA_Average, 
               PPA_Average,
           plays) %>%
        mutate_if(is.numeric, round, 3) %>%
        gather("variable","value",
               -SEASON, -TEAM, -TYPE, -plays) %>%
        mutate(SEASON = as.integer(SEASON)) %>%
      #  mutate(SEASON = factor(SEASON)) %>%
        ggplot(., aes(x=SEASON,
                      label = SEASON,
                      color = variable,
                  #    size = plays,
                      y = value))+
        geom_text(check_overlap = T,
                  size=2.5,
                  vjust=-1)+
        geom_point()+
        geom_line(lwd = 0.8,
                  alpha = 0.8,
                  stat = 'smooth',
                  method = 'loess',
                  span = 0.3,
                  se = F,
                  formula = 'y ~ x')+
        theme_phil()+
        facet_wrap(TEAM ~.,
                   ncol = 1)+
        geom_hline(yintercept =0,
                   linetype = 'dashed')+
        scale_color_manual(values = c("navy", "deepskyblue1"))+
        ylab("Points per Play")+
        ggtitle("Team Offensive Metrics by Season",
                "Average expected points added and predicted points added for a team's offense over the course of a season. Model estimates from multinomial logistic regression.")+
        scale_x_continuous(breaks = int_breaks)

```

If we use these as general measures of the efficiency of an offense in a season/game, where does A&M place compared to all teams? What is good? I'll plot the distribution of all teams with at least 400 plays in a season, then overlay where A&M ranks.

```{r indicate where tamu offense falls, fig.height=8}

library(ggforce)

team = 'Texas A&M'

temp = scored_plays %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        # exclude kickoffs from this
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        # also field goals, not that interested in kicking performance at the minute
        filter(FG == 0) %>% # remove FGs
        group_by(SEASON, OFFENSE) %>%
        summarize(EPA_Average = mean(EPA, na.rm=T),
             #     PA_Average = mean(PA, na.rm=T),
                  PPA_Average = mean(PPA, na.rm=T),
                  plays = n(),
                  .groups = 'drop') %>%
        rename(TEAM = OFFENSE) %>%
        mutate(TYPE = "OFFENSE") %>%
        select(SEASON, 
               TEAM,
               TYPE,
               EPA_Average, 
           #    PA_Average, 
               PPA_Average,
           plays) %>%
        mutate_if(is.numeric, round, 3) %>%
        filter(plays > 400) %>%
        gather("variable","value",
               -SEASON, -TEAM, -TYPE, -plays) %>%
        mutate(SEASON = as.integer(SEASON))


p1 = temp %>%
        ggplot(., aes(x=SEASON,
                      label = TEAM,
                      color = variable,
                  #    size = plays,
                      y = value))+
        facet_wrap(variable~.,
                           ncol =1)+
        geom_point(position = ggforce::position_jitternormal(sd_x = 0.1,
                                                             sd_y = 0),
                   alpha = 0.2)+
        # geom_line(lwd = 0.8,
        #           alpha = 0.8,
        #           stat = 'smooth',
        #           method = 'loess',
        #           span = 0.3,
        #           se = F,
        #           formula = 'y ~ x')+
        theme_phil()+
        geom_hline(yintercept =0,
                   linetype = 'dashed')

p1 +
        geom_point(data = temp %>%
                           filter(TEAM == team),
                   aes(x=SEASON,
                       label = SEASON,
                      color = variable,
                  #    size = plays,
                      y = value))+
        geom_line(data = temp %>%
                           filter(TEAM == team),
                  lwd = 0.8,
                  alpha = 0.8,
                  stat = 'smooth',
                  method = 'loess',
                  span = 0.3,
                  se = F,
                  formula = 'y ~ x')+
        theme_phil()+
        geom_hline(yintercept =0,
                   linetype = 'dashed')+
        scale_color_manual(values = c("navy", "deepskyblue1"))+
        ylab("Points per Play")+
        ggtitle(paste(team, "Offensive Metrics by Season"),
                "Average expected points added and predicted points added for a team's offense over the course of a season. Model estimates from multinomial logistic regression.")+
        scale_x_continuous(breaks = int_breaks)
        

```

Based on either metric, we can see that A&M's offense in 2012 and 2013 were among the best in college football at the time. It also looks like A&M's 2018 and 2020 offense were pretty darn good as well.

Let's look at another team that we would expect to be really strong this entire time period: Ohio State.

```{r look at osu offense, fig.height=8}

team = 'Ohio State'

temp = scored_plays %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        # exclude kickoffs from this
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        # also field goals, not that interested in kicking performance at the minute
        filter(FG == 0) %>% # remove FGs
        group_by(SEASON, OFFENSE) %>%
        summarize(EPA_Average = mean(EPA, na.rm=T),
             #     PA_Average = mean(PA, na.rm=T),
                  PPA_Average = mean(PPA, na.rm=T),
                  plays = n(),
                  .groups = 'drop') %>%
        rename(TEAM = OFFENSE) %>%
        mutate(TYPE = "OFFENSE") %>%
        select(SEASON, 
               TEAM,
               TYPE,
               EPA_Average, 
           #    PA_Average, 
               PPA_Average,
           plays) %>%
        mutate_if(is.numeric, round, 3) %>%
        filter(plays > 400) %>%
        gather("variable","value",
               -SEASON, -TEAM, -TYPE, -plays) %>%
        mutate(SEASON = as.integer(SEASON))


p1 = temp %>%
        ggplot(., aes(x=SEASON,
                      label = TEAM,
                      color = variable,
                  #    size = plays,
                      y = value))+
        facet_wrap(variable~.,
                           ncol =1)+
        geom_point(position = ggforce::position_jitternormal(sd_x = 0.1,
                                                             sd_y = 0),
                   alpha = 0.2)+
        # geom_line(lwd = 0.8,
        #           alpha = 0.8,
        #           stat = 'smooth',
        #           method = 'loess',
        #           span = 0.3,
        #           se = F,
        #           formula = 'y ~ x')+
        theme_phil()+
        geom_hline(yintercept =0,
                   linetype = 'dashed')

p1 +
        geom_point(data = temp %>%
                           filter(TEAM == team),
                   aes(x=SEASON,
                       label = SEASON,
                      color = variable,
                  #    size = plays,
                      y = value))+
        geom_line(data = temp %>%
                           filter(TEAM == team),
                  lwd = 0.8,
                  alpha = 0.8,
                  stat = 'smooth',
                  method = 'loess',
                  span = 0.3,
                  se = F,
                  formula = 'y ~ x')+
        theme_phil()+
        geom_hline(yintercept =0,
                   linetype = 'dashed')+
        scale_color_manual(values = c("navy", "deepskyblue1"))+
        ylab("Points per Play")+
        ggtitle(paste(team, "Offensive Metrics by Season"),
                "Average expected points added and predicted points added for a team's offense over the course of a season. Model estimates from multinomial logistic regression.")+
        scale_x_continuous(breaks = int_breaks)
        
```

What about a team that's been up and down during this time period? Some lowly team, like Texas.

```{r look at texas offense, fig.height=8}

team = 'Texas'

temp = scored_plays %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        # exclude kickoffs from this
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        # also field goals, not that interested in kicking performance at the minute
        filter(FG == 0) %>% # remove FGs
        group_by(SEASON, OFFENSE) %>%
        summarize(EPA_Average = mean(EPA, na.rm=T),
             #     PA_Average = mean(PA, na.rm=T),
                  PPA_Average = mean(PPA, na.rm=T),
                  plays = n(),
                  .groups = 'drop') %>%
        rename(TEAM = OFFENSE) %>%
        mutate(TYPE = "OFFENSE") %>%
        select(SEASON, 
               TEAM,
               TYPE,
               EPA_Average, 
           #    PA_Average, 
               PPA_Average,
           plays) %>%
        mutate_if(is.numeric, round, 3) %>%
        filter(plays > 400) %>%
        gather("variable","value",
               -SEASON, -TEAM, -TYPE, -plays) %>%
        mutate(SEASON = as.integer(SEASON))


p1 = temp %>%
        ggplot(., aes(x=SEASON,
                      label = TEAM,
                      color = variable,
                  #    size = plays,
                      y = value))+
        facet_wrap(variable~.,
                           ncol =1)+
        geom_point(position = ggforce::position_jitternormal(sd_x = 0.1,
                                                             sd_y = 0),
                   alpha = 0.2)+
        # geom_line(lwd = 0.8,
        #           alpha = 0.8,
        #           stat = 'smooth',
        #           method = 'loess',
        #           span = 0.3,
        #           se = F,
        #           formula = 'y ~ x')+
        theme_phil()+
        geom_hline(yintercept =0,
                   linetype = 'dashed')

p1 +
        geom_point(data = temp %>%
                           filter(TEAM == team),
                   aes(x=SEASON,
                       label = SEASON,
                      color = variable,
                  #    size = plays,
                      y = value))+
        geom_line(data = temp %>%
                           filter(TEAM == team),
                  lwd = 0.8,
                  alpha = 0.8,
                  stat = 'smooth',
                  method = 'loess',
                  span = 0.3,
                  se = F,
                  formula = 'y ~ x')+
        theme_phil()+
        geom_hline(yintercept =0,
                   linetype = 'dashed')+
        scale_color_manual(values = c("navy", "deepskyblue1"))+
        ylab("Points per Play")+
        ggtitle(paste(team, "Offensive Metrics by Season"),
                "Average expected points added and predicted points added for a team's offense over the course of a season. Model estimates from multinomial logistic regression.")+
        scale_x_continuous(breaks = int_breaks)

```

What about Florida State? We would expect them to be towards the top in the Jimob/Jameis era and then fall back to Earth during the Taggart era.

```{r look at fsu offense, fig.height=8}

team = 'Florida State'

temp = scored_plays %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        # exclude kickoffs from this
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        # also field goals, not that interested in kicking performance at the minute
        filter(FG == 0) %>% # remove FGs
        group_by(SEASON, OFFENSE) %>%
        summarize(EPA_Average = mean(EPA, na.rm=T),
             #     PA_Average = mean(PA, na.rm=T),
                  PPA_Average = mean(PPA, na.rm=T),
                  plays = n(),
                  .groups = 'drop') %>%
        rename(TEAM = OFFENSE) %>%
        mutate(TYPE = "OFFENSE") %>%
        select(SEASON, 
               TEAM,
               TYPE,
               EPA_Average, 
           #    PA_Average, 
               PPA_Average,
           plays) %>%
        mutate_if(is.numeric, round, 3) %>%
        filter(plays > 400) %>%
        gather("variable","value",
               -SEASON, -TEAM, -TYPE, -plays) %>%
        mutate(SEASON = as.integer(SEASON))


p1 = temp %>%
        ggplot(., aes(x=SEASON,
                      label = TEAM,
                      color = variable,
                  #    size = plays,
                      y = value))+
        facet_wrap(variable~.,
                           ncol =1)+
        geom_point(position = ggforce::position_jitternormal(sd_x = 0.1,
                                                             sd_y = 0),
                   alpha = 0.2)+
        # geom_line(lwd = 0.8,
        #           alpha = 0.8,
        #           stat = 'smooth',
        #           method = 'loess',
        #           span = 0.3,
        #           se = F,
        #           formula = 'y ~ x')+
        theme_phil()+
        geom_hline(yintercept =0,
                   linetype = 'dashed')

p1 +
        geom_point(data = temp %>%
                           filter(TEAM == team),
                   aes(x=SEASON,
                       label = SEASON,
                      color = variable,
                  #    size = plays,
                      y = value))+
        geom_line(data = temp %>%
                           filter(TEAM == team),
                  lwd = 0.8,
                  alpha = 0.8,
                  stat = 'smooth',
                  method = 'loess',
                  span = 0.3,
                  se = F,
                  formula = 'y ~ x')+
        theme_phil()+
        geom_hline(yintercept =0,
                   linetype = 'dashed')+
        scale_color_manual(values = c("navy", "deepskyblue1"))+
        ylab("Points per Play")+
        ggtitle(paste(team, "Offensive Metrics by Season"),
                "Average expected points added and predicted points added for a team's offense over the course of a season. Model estimates from multinomial logistic regression.")+
        scale_x_continuous(breaks = int_breaks)

```

Hmm. That's kind of interesting - I've heard people discuss Jimbo's offense as lacking in explosiveness, which some argue is the difference between the expected points and the predicted points. This definitely wasn't the case in 2013, but generally FSU had a pretty efficient and explosive offense until their collapse in 2017.

## Team Offense and Defense by Game

But this is only looking at one side of the ball at the season level. We can look at team's defense in the same way in terms of the average expected/predicted points allowed by the offenses they face. In this case, you want your defense to have a negative value - this indicates that offenses weren't able to generate points against your defense. I'll look at a team's offense and defense side by side in terms of Predicted Points Added - meaning I am including scoring plays. We would expect a good team to be one whose offense (in blue) generates more points per play than they allow on defense (in red).

```{r look at tamu offense and defense by season by game, fig.height=8}

team = 'Texas A&M'

scored_plays %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        # exclude kickoffs from this
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        # also field goals, not that interested in kicking performance at the minute
        filter(FG == 0) %>% # remove FGs
        filter(OFFENSE == team) %>%
        group_by(SEASON, OFFENSE, GAME_ID) %>%
        summarize(EPA_Average = mean(EPA, na.rm=T),
                  PPA_Average = mean(PPA, na.rm=T),
                  plays = n(),
                  .groups = 'drop') %>%
        mutate(TEAM = OFFENSE) %>%
        mutate(TYPE = "OFFENSE") %>%
        left_join(., games_data_raw %>%
                          select(GAME_ID, HOME_TEAM, AWAY_TEAM),
                  by = c("GAME_ID")) %>%
        mutate(OPPONENT = case_when(HOME_TEAM == TEAM ~ AWAY_TEAM,
                                    AWAY_TEAM == TEAM ~ HOME_TEAM)) %>%
        select(SEASON, GAME_ID, TEAM, OPPONENT, TYPE, EPA_Average, PPA_Average, plays) %>%
        group_by(SEASON) %>%
        mutate(game = row_number()) %>%
        mutate(SEASON = factor(SEASON)) %>%
        bind_rows(.,
                  scored_plays %>%
                rename(EPA = EP_Added,
                       PA = Points_Added) %>%
                mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                                       !is.na(EPA) & is.na(PA) ~ EPA)) %>%
                # exclude kickoffs from this
                filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
                # also field goals, not that interested in kicking performance at the minute
                filter(FG == 0) %>% # remove FGs
                filter(DEFENSE == team) %>%
                group_by(SEASON, DEFENSE, GAME_ID) %>%
                summarize(EPA_Average = mean(EPA, na.rm=T),
                          PPA_Average = mean(PPA, na.rm=T),
                          plays = n(),
                          .groups = 'drop') %>%
                mutate(TEAM = DEFENSE) %>%
                mutate(TYPE = "DEFENSE") %>%
                left_join(., games_data_raw %>%
                                  select(GAME_ID, HOME_TEAM, AWAY_TEAM),
                          by = c("GAME_ID")) %>%
                mutate(OPPONENT = case_when(HOME_TEAM == TEAM ~ AWAY_TEAM,
                                            AWAY_TEAM == TEAM ~ HOME_TEAM)) %>%
                select(SEASON, GAME_ID, TEAM, OPPONENT, TYPE, EPA_Average, PPA_Average, plays) %>%
                        group_by(SEASON) %>%
                mutate(game = row_number()) %>%
                mutate(SEASON = factor(SEASON))) %>%
        mutate(TYPE = paste(TYPE, "PPA_Average")) %>%
        ggplot(., aes(x=game,
                      color = TYPE,
                  #    size = plays,
                      label = OPPONENT,
                      y = PPA_Average))+
        geom_point()+
        geom_line(lwd = 0.8,
                  alpha = 0.8,
                  stat = 'smooth',
                  method = 'loess',
                  span = 0.25,
                  se = F,
                  formula = 'y ~ x')+
        geom_text(check_overlap=T,
                  vjust=-1,
                  size = 2.5)+
        theme_phil()+
        facet_wrap(TEAM ~ SEASON,
                   ncol = 3)+
        geom_hline(yintercept =0,
                   linetype = 'dashed')+
        scale_color_manual(values = c("red", "deepskyblue1"))+
        coord_cartesian(ylim = c(-.8, .8))+
        ggtitle(paste(team, "Offense/Defense Efficency by Game"),
                subtitle = str_wrap("Measuring offensive and defensive efficiency via predicted points added. Model estimates from multinomial logistic regression.",120))
        
     
```

I'll look at a team like Wisconsin. They seem to generally have a pretty good defense, but struggle in terms of their ability to generate points on offense.

```{r look at wisconson offense and defense by season by game, fig.height=8}

team = 'Wisconsin'

scored_plays %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        # exclude kickoffs from this
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        # also field goals, not that interested in kicking performance at the minute
        filter(FG == 0) %>% # remove FGs
        filter(OFFENSE == team) %>%
        group_by(SEASON, OFFENSE, GAME_ID) %>%
        summarize(EPA_Average = mean(EPA, na.rm=T),
                  PPA_Average = mean(PPA, na.rm=T),
                  plays = n(),
                  .groups = 'drop') %>%
        mutate(TEAM = OFFENSE) %>%
        mutate(TYPE = "OFFENSE") %>%
        left_join(., games_data_raw %>%
                          select(GAME_ID, HOME_TEAM, AWAY_TEAM),
                  by = c("GAME_ID")) %>%
        mutate(OPPONENT = case_when(HOME_TEAM == TEAM ~ AWAY_TEAM,
                                    AWAY_TEAM == TEAM ~ HOME_TEAM)) %>%
        select(SEASON, GAME_ID, TEAM, OPPONENT, TYPE, EPA_Average, PPA_Average, plays) %>%
        group_by(SEASON) %>%
        mutate(game = row_number()) %>%
        mutate(SEASON = factor(SEASON)) %>%
        bind_rows(.,
                  scored_plays %>%
                rename(EPA = EP_Added,
                       PA = Points_Added) %>%
                mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                                       !is.na(EPA) & is.na(PA) ~ EPA)) %>%
                # exclude kickoffs from this
                filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
                # also field goals, not that interested in kicking performance at the minute
                filter(FG == 0) %>% # remove FGs
                filter(DEFENSE == team) %>%
                group_by(SEASON, DEFENSE, GAME_ID) %>%
                summarize(EPA_Average = mean(EPA, na.rm=T),
                          PPA_Average = mean(PPA, na.rm=T),
                          plays = n(),
                          .groups = 'drop') %>%
                mutate(TEAM = DEFENSE) %>%
                mutate(TYPE = "DEFENSE") %>%
                left_join(., games_data_raw %>%
                                  select(GAME_ID, HOME_TEAM, AWAY_TEAM),
                          by = c("GAME_ID")) %>%
                mutate(OPPONENT = case_when(HOME_TEAM == TEAM ~ AWAY_TEAM,
                                            AWAY_TEAM == TEAM ~ HOME_TEAM)) %>%
                select(SEASON, GAME_ID, TEAM, OPPONENT, TYPE, EPA_Average, PPA_Average, plays) %>%
                        group_by(SEASON) %>%
                mutate(game = row_number()) %>%
                mutate(SEASON = factor(SEASON))) %>%
        mutate(TYPE = paste(TYPE, "PPA_Average")) %>%
        ggplot(., aes(x=game,
                      color = TYPE,
                  #    size = plays,
                      label = OPPONENT,
                      y = PPA_Average))+
        geom_point()+
        geom_line(lwd = 0.8,
                  alpha = 0.8,
                  stat = 'smooth',
                  method = 'loess',
                  span = 0.25,
                  se = F,
                  formula = 'y ~ x')+
        geom_text(check_overlap=T,
                  vjust=-1,
                  size = 2.5)+
        theme_phil()+
        facet_wrap(TEAM ~ SEASON,
                   ncol = 3)+
        geom_hline(yintercept =0,
                   linetype = 'dashed')+
        scale_color_manual(values = c("red", "deepskyblue1"))+
        coord_cartesian(ylim = c(-.8, .8))+
        ggtitle(paste(team, "Offense/Defense Efficency by Game"),
                subtitle = str_wrap("Measuring offensive and defensive efficiency via predicted points added. Model estimates from multinomial logistic regression.",120))
        
        
```

For a dominant team like Alabama, we should see their offensive efficiency be high and their defensive efficiency be high as well - the blue should almost always outstrip the red.

```{r look at alabama offense and defense by season by game, fig.height=8}

team = 'Alabama'

scored_plays %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        # exclude kickoffs from this
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        # also field goals, not that interested in kicking performance at the minute
        filter(FG == 0) %>% # remove FGs
        filter(OFFENSE == team) %>%
        group_by(SEASON, OFFENSE, GAME_ID) %>%
        summarize(EPA_Average = mean(EPA, na.rm=T),
                  PPA_Average = mean(PPA, na.rm=T),
                  plays = n(),
                  .groups = 'drop') %>%
        mutate(TEAM = OFFENSE) %>%
        mutate(TYPE = "OFFENSE") %>%
        left_join(., games_data_raw %>%
                          select(GAME_ID, HOME_TEAM, AWAY_TEAM),
                  by = c("GAME_ID")) %>%
        mutate(OPPONENT = case_when(HOME_TEAM == TEAM ~ AWAY_TEAM,
                                    AWAY_TEAM == TEAM ~ HOME_TEAM)) %>%
        select(SEASON, GAME_ID, TEAM, OPPONENT, TYPE, EPA_Average, PPA_Average, plays) %>%
        group_by(SEASON) %>%
        mutate(game = row_number()) %>%
        mutate(SEASON = factor(SEASON)) %>%
        bind_rows(.,
                  scored_plays %>%
                rename(EPA = EP_Added,
                       PA = Points_Added) %>%
                mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                                       !is.na(EPA) & is.na(PA) ~ EPA)) %>%
                # exclude kickoffs from this
                filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
                # also field goals, not that interested in kicking performance at the minute
                filter(FG == 0) %>% # remove FGs
                filter(DEFENSE == team) %>%
                group_by(SEASON, DEFENSE, GAME_ID) %>%
                summarize(EPA_Average = mean(EPA, na.rm=T),
                          PPA_Average = mean(PPA, na.rm=T),
                          plays = n(),
                          .groups = 'drop') %>%
                mutate(TEAM = DEFENSE) %>%
                mutate(TYPE = "DEFENSE") %>%
                left_join(., games_data_raw %>%
                                  select(GAME_ID, HOME_TEAM, AWAY_TEAM),
                          by = c("GAME_ID")) %>%
                mutate(OPPONENT = case_when(HOME_TEAM == TEAM ~ AWAY_TEAM,
                                            AWAY_TEAM == TEAM ~ HOME_TEAM)) %>%
                select(SEASON, GAME_ID, TEAM, OPPONENT, TYPE, EPA_Average, PPA_Average, plays) %>%
                        group_by(SEASON) %>%
                mutate(game = row_number()) %>%
                mutate(SEASON = factor(SEASON))) %>%
        mutate(TYPE = paste(TYPE, "PPA_Average")) %>%
        ggplot(., aes(x=game,
                      color = TYPE,
                  #    size = plays,
                      label = OPPONENT,
                      y = PPA_Average))+
        geom_point()+
        geom_line(lwd = 0.8,
                  alpha = 0.8,
                  stat = 'smooth',
                  method = 'loess',
                  span = 0.25,
                  se = F,
                  formula = 'y ~ x')+
        geom_text(check_overlap=T,
                  vjust=-1,
                  size = 2.5)+
        theme_phil()+
        facet_wrap(TEAM ~ SEASON,
                   ncol = 3)+
        geom_hline(yintercept =0,
                   linetype = 'dashed')+
        scale_color_manual(values = c("red", "deepskyblue1"))+
        coord_cartesian(ylim = c(-.8, .8))+
        ggtitle(paste(team, "Offense/Defense Efficency by Game"),
                subtitle = str_wrap("Measuring offensive and defensive efficiency via predicted points added. Model estimates from multinomial logistic regression.",120))
        
        
     
```

Evidently they were so good at stopping Tennesee in 2011 that it breaks my axis.

For a mediocre team, like Kansas, we should see the opposite type of graph here.

```{r look at kansas offense and defense by season by game, fig.height=8}

team = 'Kansas'

scored_plays %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        # exclude kickoffs from this
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        # also field goals, not that interested in kicking performance at the minute
        filter(FG == 0) %>% # remove FGs
        filter(OFFENSE == team) %>%
        group_by(SEASON, OFFENSE, GAME_ID) %>%
        summarize(EPA_Average = mean(EPA, na.rm=T),
                  PPA_Average = mean(PPA, na.rm=T),
                  plays = n(),
                  .groups = 'drop') %>%
        mutate(TEAM = OFFENSE) %>%
        mutate(TYPE = "OFFENSE") %>%
        left_join(., games_data_raw %>%
                          select(GAME_ID, HOME_TEAM, AWAY_TEAM),
                  by = c("GAME_ID")) %>%
        mutate(OPPONENT = case_when(HOME_TEAM == TEAM ~ AWAY_TEAM,
                                    AWAY_TEAM == TEAM ~ HOME_TEAM)) %>%
        select(SEASON, GAME_ID, TEAM, OPPONENT, TYPE, EPA_Average, PPA_Average, plays) %>%
        group_by(SEASON) %>%
        mutate(game = row_number()) %>%
        mutate(SEASON = factor(SEASON)) %>%
        bind_rows(.,
                  scored_plays %>%
                rename(EPA = EP_Added,
                       PA = Points_Added) %>%
                mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                                       !is.na(EPA) & is.na(PA) ~ EPA)) %>%
                # exclude kickoffs from this
                filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
                # also field goals, not that interested in kicking performance at the minute
                filter(FG == 0) %>% # remove FGs
                filter(DEFENSE == team) %>%
                group_by(SEASON, DEFENSE, GAME_ID) %>%
                summarize(EPA_Average = mean(EPA, na.rm=T),
                          PPA_Average = mean(PPA, na.rm=T),
                          plays = n(),
                          .groups = 'drop') %>%
                mutate(TEAM = DEFENSE) %>%
                mutate(TYPE = "DEFENSE") %>%
                left_join(., games_data_raw %>%
                                  select(GAME_ID, HOME_TEAM, AWAY_TEAM),
                          by = c("GAME_ID")) %>%
                mutate(OPPONENT = case_when(HOME_TEAM == TEAM ~ AWAY_TEAM,
                                            AWAY_TEAM == TEAM ~ HOME_TEAM)) %>%
                select(SEASON, GAME_ID, TEAM, OPPONENT, TYPE, EPA_Average, PPA_Average, plays) %>%
                        group_by(SEASON) %>%
                mutate(game = row_number()) %>%
                mutate(SEASON = factor(SEASON))) %>%
        mutate(TYPE = paste(TYPE, "PPA_Average")) %>%
        ggplot(., aes(x=game,
                      color = TYPE,
                  #    size = plays,
                      label = OPPONENT,
                      y = PPA_Average))+
        geom_point()+
        geom_line(lwd = 0.8,
                  alpha = 0.8,
                  stat = 'smooth',
                  method = 'loess',
                  span = 0.25,
                  se = F,
                  formula = 'y ~ x')+
        geom_text(check_overlap=T,
                  vjust=-1,
                  size = 2.5)+
        theme_phil()+
        facet_wrap(TEAM ~ SEASON,
                   ncol = 3)+
        geom_hline(yintercept =0,
                   linetype = 'dashed')+
        scale_color_manual(values = c("red", "deepskyblue1"))+
        coord_cartesian(ylim = c(-.8, .8))+
        ggtitle(paste(team, "Offense/Defense Efficency by Game"),
                subtitle = str_wrap("Measuring offensive and defensive efficiency via predicted points added. Model estimates from multinomial logistic regression.",120))
        
        
     
```

## Ranking Teams by Season

We can then aggregate to a team's offensive performance within a full year to rate their overall offensive/defensive efficiency. We can also break this down by passing vs rushing plays on both sides of the ball.

For the purpose of evaluating a team's defense in this analysis, I'll flip the sign of a defense's points, which are currently scored from the perspective of the offense, so that positive is always good for a team.

```{r look at expected points at the season level}

# offense
yearly_offense_epa = scored_plays %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        filter(FG == 0) %>% # remove FGs
        group_by(SEASON, OFFENSE) %>%
        summarize(EPA_Average = mean(EPA, na.rm=T),
                  PPA_Average = mean(PPA, na.rm=T),
                  plays = n(),
                  .groups = 'drop') %>%
        mutate(TEAM = OFFENSE) %>%
        mutate(TYPE = "OFFENSE") %>%
        select(SEASON, TEAM, TYPE, EPA_Average, PPA_Average, plays)

# offense by play type
yearly_offense_epa_by_play = scored_plays %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        mutate(PLAY = case_when(
                grepl("pass|completion|sack|intercept", tolower(PLAY_TYPE)) ~ 'Pass',
                grepl("run|rush", tolower(PLAY_TYPE)) ~ 'Run',
                TRUE ~ 'Special')) %>%
        filter(FG == 0) %>% # remove FGs
        group_by(SEASON, OFFENSE, PLAY) %>%
        summarize(EPA_Average = mean(EPA, na.rm=T),
                  PPA_Average = mean(PPA, na.rm=T),
                  plays = n(),
                  .groups = 'drop') %>%
        mutate(TEAM = OFFENSE) %>%
        mutate(TYPE = "OFFENSE") %>%
        select(SEASON, TEAM, TYPE, PLAY, EPA_Average, PPA_Average, plays)

# defense
yearly_defense_epa = scored_plays %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        filter(FG == 0) %>% # remove FGs
        group_by(SEASON, DEFENSE) %>%
        summarize(EPA_Average = -1*mean(EPA, na.rm=T),
                  PPA_Average = -1*mean(PPA, na.rm=T),
                  plays = n(),
                  .groups = 'drop') %>%
        mutate(TEAM = DEFENSE) %>%
        mutate(TYPE = "DEFENSE") %>%
        select(SEASON, TEAM, TYPE, EPA_Average, PPA_Average, plays)

# defense by play type
yearly_defense_epa_by_play = scored_plays %>%
        rename(EPA = EP_Added,
               PA = Points_Added) %>%
        mutate(PPA = case_when(is.na(EPA) & !is.na(PA) ~ PA,
                               !is.na(EPA) & is.na(PA) ~ EPA)) %>%
        filter(!grepl("kickoff", tolower(PLAY_TYPE))) %>%
        mutate(PLAY = case_when(
                grepl("pass|completion|sack|intercept", tolower(PLAY_TYPE)) ~ 'Pass',
                grepl("run|rush", tolower(PLAY_TYPE)) ~ 'Run',
                TRUE ~ 'Special')) %>%
        filter(FG == 0) %>% # remove FGs
        group_by(SEASON, DEFENSE, PLAY) %>%
        summarize(EPA_Average = -1*mean(EPA, na.rm=T),
                  PPA_Average = -1*mean(PPA, na.rm=T),
                  plays = n(),
                  .groups = 'drop') %>%
        mutate(TEAM = DEFENSE) %>%
        mutate(TYPE = "DEFENSE") %>%
        select(SEASON, TEAM, TYPE, PLAY, EPA_Average, PPA_Average, plays)
        
```

Putting this all together, I can rank a teams offense/defense by year and then sort to see where the top teams of all time tend to rank. Here are the top 50 teams using a composite score of both, including only offenses and defenses with at least 400 plays in a season.

```{r look at overall offense defense per team season}

yearly_teams = bind_rows(yearly_offense_epa,
          yearly_defense_epa) %>%
        pivot_wider(id_cols = c("SEASON", 
                                "TEAM"),
                    values_from = c("EPA_Average", "PPA_Average", "plays"),
                    names_from = c("TYPE")) %>%
        filter(plays_OFFENSE & plays_DEFENSE > 400) %>%
        rename(OFFENSE_PPA = PPA_Average_OFFENSE,
               DEFENSE_PPA = PPA_Average_DEFENSE) %>%
        select(SEASON, TEAM, OFFENSE_PPA, DEFENSE_PPA) %>%
        mutate(OVERALL = OFFENSE_PPA + DEFENSE_PPA)

yearly_teams %>%
        mutate(SEASON = as.character(SEASON)) %>%
        mutate_if(is.numeric, round, 2) %>%
        arrange(desc(OVERALL)) %>%
        mutate(Rank = row_number()) %>%
        select(Rank, everything()) %>%
        head(50) %>%
        flextable() %>%
        autofit()

```

And here are the 25 worst teams using the same criterion.

```{r bottom overall by season}

yearly_teams %>%
        mutate(SEASON = as.character(SEASON)) %>%
        mutate_if(is.numeric, round, 2) %>%
        arrange(desc(OVERALL)) %>%
        mutate(Rank = row_number()) %>%
        select(Rank, everything()) %>%
        arrange(OVERALL) %>%
        head(25) %>%
        flextable() %>%
        autofit()

```

We can visualize all of this by placing every team based on its overall offensive/defensive efficiency.

```{r make plot for overall, fig.height=8}

yearly_teams %>%
        mutate(SEASON = as.character(SEASON)) %>%
        ggplot(., aes(x=OFFENSE_PPA,
                      color = OVERALL,
                      label = paste(TEAM, SEASON, sep=" "),
                      y=DEFENSE_PPA))+
        geom_vline(xintercept = 0,
                   linetype = 'dashed',
                   color = 'black')+
        geom_hline(yintercept = 0,
                   linetype = 'dashed',
                   color = 'black')+
        geom_point(alpha = 0.5)+
                   #color = 'grey60')+
        geom_text(check_overlap = T,
                  size = 2.5,
                  hjust=0.5,
                  vjust = -1)+
        theme_phil()+
        coord_cartesian(xlim= c(-0.45, 0.45),
                        ylim = c(-0.45, 0.45))+
        annotate("label",
                 x=-0.25,
                 y=0.3,
                 label = "Bad Offense \n Good Defense")+
        annotate("label",
                 x=0.3,
                 y=-0.3,
                 label = "Good Offense \n Bad Defense")+
        annotate("label",
                 x=0.3,
                 y=0.3,
                 label = "Good Offense \n Good Defense")+
        annotate("label",
                 x=-0.25,
                 y=-0.3,
                 label = "Bad Offense \n Bad Defense")+
        ggtitle("College Football Teams by Offense/Defense Efficiency, 2010-2020",
                subtitle = "Teams placed based on regular season offense and defense predicted points added per play.")+
        scale_color_gradient(low = "red", high = "deepskyblue1",
                             limits = c(-0.25, 0.25),
                             oob = scales::squish)+
        theme(legend.title = element_text())+
        guides(color = guide_colorbar(barheight = 0.5,
                                      title = 'Team PPA Overall',
                                      title.position = 'top',
                                     barwidth = 10))

```

We can also break this down by conference year over year. For instance, where do we place SEC teams in each of these seasons?

```{r SEC overall score, fig.height=8}

yearly_teams %>%
        left_join(.,
                  games_data_raw %>% 
                          select(SEASON, HOME_TEAM, HOME_CONFERENCE) %>% 
                          rename(TEAM = HOME_TEAM, CONFERENCE = HOME_CONFERENCE) %>% 
                          bind_rows(.,
                                    games_data_raw %>%
                                            select(SEASON, AWAY_TEAM, AWAY_CONFERENCE) %>% 
                                            rename(TEAM = AWAY_TEAM, CONFERENCE = AWAY_CONFERENCE)) %>%
                          select(SEASON, TEAM, CONFERENCE) %>%
                          unique(),
                  by = c("SEASON", "TEAM")) %>%
        filter(CONFERENCE == 'SEC') %>%
        mutate(SEASON = as.character(SEASON)) %>%
        ggplot(., aes(x=OFFENSE_PPA,
                      color = OVERALL,
                      label = paste(TEAM, SEASON, sep=" "),
                      y=DEFENSE_PPA))+
        geom_vline(xintercept = 0,
                   linetype = 'dashed',
                   alpha = 0.8,
                   color = 'black')+
        geom_hline(yintercept = 0,
                   linetype = 'dashed',
                   alpha = 0.8,
                   color = 'black')+
        geom_point(alpha = 0.5)+
        geom_text_repel(size=3)+
                   #color = 'grey60')+
        # geom_text(check_overlap = T,
        #           size = 2.5,
        #           hjust=0.5,
        #           vjust = -1)+
        theme_phil()+
        coord_cartesian(xlim= c(-0.45, 0.45),
                        ylim = c(-0.45, 0.45))+
        facet_wrap(CONFERENCE + SEASON ~.,
                   ncol =3)+
        # annotate("label",
        #          x=-0.25,
        #          y=0.3,
        #          label = "Bad Offense \n Good Defense")+
        # annotate("label",
        #          x=0.3,
        #          y=-0.3,
        #          label = "Good Offense \n Bad Defense")+
        # annotate("label",
        #          x=0.3,
        #          y=0.3,
        #          label = "Good Offense \n Good Defense")+
        # annotate("label",
        #          x=-0.25,
        #          y=-0.3,
        #          label = "Bad Offense \n Bad Defense")+
        ggtitle("College Football Teams by Offense/Defense Efficiency, 2010-2020",
                subtitle = "Teams placed based on regular season offense and defense predicted points added per play.")+
        scale_color_gradient(low = "red", high = "deepskyblue1",
                             limits = c(-0.25, 0.25),
                             oob = scales::squish)+
        theme(legend.title = element_text())+
        guides(color = guide_colorbar(barheight = 0.5,
                                      title = 'Team PPA Overall',
                                      title.position = 'top',
                                     barwidth = 10))

```

What about the Big Ten?

```{r big ten overall score, fig.height=8}

yearly_teams %>%
        left_join(.,
                  games_data_raw %>% 
                          select(SEASON, HOME_TEAM, HOME_CONFERENCE) %>% 
                          rename(TEAM = HOME_TEAM, CONFERENCE = HOME_CONFERENCE) %>% 
                          bind_rows(.,
                                    games_data_raw %>%
                                            select(SEASON, AWAY_TEAM, AWAY_CONFERENCE) %>% 
                                            rename(TEAM = AWAY_TEAM, CONFERENCE = AWAY_CONFERENCE)) %>%
                          select(SEASON, TEAM, CONFERENCE) %>%
                          unique(),
                  by = c("SEASON", "TEAM")) %>%
        filter(CONFERENCE == 'Big Ten') %>%
        mutate(SEASON = as.character(SEASON)) %>%
        ggplot(., aes(x=OFFENSE_PPA,
                      color = OVERALL,
                      label = paste(TEAM, SEASON, sep=" "),
                      y=DEFENSE_PPA))+
        geom_vline(xintercept = 0,
                   linetype = 'dashed',
                   alpha = 0.8,
                   color = 'black')+
        geom_hline(yintercept = 0,
                   linetype = 'dashed',
                   alpha = 0.8,
                   color = 'black')+
        geom_point(alpha = 0.5)+
        geom_text_repel(size=3,
                        max.overlaps = 20)+
                   #color = 'grey60')+
        # geom_text(check_overlap = T,
        #           size = 2.5,
        #           hjust=0.5,
        #           vjust = -1)+
        theme_phil()+
        coord_cartesian(xlim= c(-0.45, 0.45),
                        ylim = c(-0.45, 0.45))+
        facet_wrap(CONFERENCE + SEASON ~.,
                   ncol =3)+
        # annotate("label",
        #          x=-0.25,
        #          y=0.3,
        #          label = "Bad Offense \n Good Defense")+
        # annotate("label",
        #          x=0.3,
        #          y=-0.3,
        #          label = "Good Offense \n Bad Defense")+
        # annotate("label",
        #          x=0.3,
        #          y=0.3,
        #          label = "Good Offense \n Good Defense")+
        # annotate("label",
        #          x=-0.25,
        #          y=-0.3,
        #          label = "Bad Offense \n Bad Defense")+
        ggtitle("College Football Teams by Offense/Defense Efficiency, 2010-2020",
                subtitle = "Teams placed based on regular season offense and defense predicted points added per play.")+
        scale_color_gradient(low = "red", high = "deepskyblue1",
                             limits = c(-0.25, 0.25),
                             oob = scales::squish)+
        theme(legend.title = element_text())+
        guides(color = guide_colorbar(barheight = 0.5,
                                      title = 'Team PPA Overall',
                                      title.position = 'top',
                                     barwidth = 10))

```

## Ranking Teams by Pass/Run Efficiency

We can break this down further to examine a team's expected points based on pass/run. 

```{r look at offense defense efficiency vs run}

yearly_teams_type = bind_rows(yearly_offense_epa_by_play,
          yearly_defense_epa_by_play) %>%
        pivot_wider(id_cols = c("SEASON", 
                                "TEAM"),
                    values_from = c("EPA_Average", "PPA_Average", "plays"),
                    names_from = c("TYPE", "PLAY")) %>%
        filter(plays_OFFENSE_Pass + plays_OFFENSE_Run + plays_OFFENSE_Special  > 400) %>%
        rename(OFFENSE_PPA_Pass = PPA_Average_OFFENSE_Pass,
               DEFENSE_PPA_Pass = PPA_Average_DEFENSE_Pass,
               OFFENSE_PPA_Run = PPA_Average_OFFENSE_Run,
               DEFENSE_PPA_Run = PPA_Average_DEFENSE_Run,
               OFFENSE_Plays_Pass = plays_OFFENSE_Pass,
               DEFENSE_Plays_Pass = plays_DEFENSE_Pass,
               OFFENSE_Plays_Run = plays_OFFENSE_Run,
               DEFENSE_Plays_Run = plays_DEFENSE_Run) %>%
        select(SEASON, TEAM, starts_with("OFFENSE_PPA"), starts_with("DEFENSE_PPA"), starts_with("OFFENSE_Plays"), starts_with("DEFENSE_Plays"))

```
Filtering to teams with at least 200 such plays in a season, which teams had the most efficient passing offense in terms of predicted points added?

```{r pass efficiency by teams}

yearly_teams_type %>%
        filter(OFFENSE_Plays_Pass > 200) %>%
        arrange(desc(OFFENSE_PPA_Pass)) %>%
        select(SEASON, TEAM, OFFENSE_PPA_Pass, OFFENSE_Plays_Pass) %>%
        mutate_if(is.numeric, round, 2) %>%
        mutate(SEASON = as.character(SEASON)) %>%
        head(25) %>%
        flextable() %>%
        autofit()

```

What about run offense?

```{r run efficiency by teams}

yearly_teams_type %>%
        filter(OFFENSE_Plays_Run > 200) %>%
        arrange(desc(OFFENSE_PPA_Run)) %>%
        select(SEASON, TEAM, OFFENSE_PPA_Run, OFFENSE_Plays_Run) %>%
        mutate_if(is.numeric, round, 2) %>%
        mutate(SEASON = as.character(SEASON)) %>%
        head(25) %>%
        flextable() %>%
        autofit()

```

Flipping this around, which defenses yielded the most points to passing/run?

```{r worst defenses}

yearly_teams_type %>%
        filter(DEFENSE_Plays_Pass > 200) %>%
        arrange(DEFENSE_PPA_Pass) %>%
        select(SEASON, TEAM, DEFENSE_PPA_Pass, DEFENSE_Plays_Pass) %>%
        mutate_if(is.numeric, round, 2) %>%
        mutate(SEASON = as.character(SEASON)) %>%
        head(25) %>%
        flextable() %>%
        autofit()

yearly_teams_type %>%
        filter(DEFENSE_Plays_Run > 200) %>%
        arrange(DEFENSE_PPA_Run) %>%
        select(SEASON, TEAM, DEFENSE_PPA_Run, DEFENSE_Plays_Run) %>%
        mutate_if(is.numeric, round, 2) %>%
        mutate(SEASON = as.character(SEASON)) %>%
        head(25) %>%
        flextable() %>%
        autofit()

```

What about the best defenses? Rutgers in 2012 with the best run defense? Really? Keep in mind that these ratings are unconditional based on opponent strength, though evidently Rutgers did have a bunch of defenders drafted in 2013, so..?

```{r best defenses}

yearly_teams_type %>%
        filter(DEFENSE_Plays_Pass > 200) %>%
        arrange(desc(DEFENSE_PPA_Pass)) %>%
        select(SEASON, TEAM, DEFENSE_PPA_Pass, DEFENSE_Plays_Pass) %>%
        mutate_if(is.numeric, round, 2) %>%
        mutate(SEASON = as.character(SEASON)) %>%
        head(25) %>%
        flextable() %>%
        autofit()

yearly_teams_type %>%
        filter(DEFENSE_Plays_Run > 200) %>%
        arrange(desc(DEFENSE_PPA_Run)) %>%
        select(SEASON, TEAM, DEFENSE_PPA_Run, DEFENSE_Plays_Run) %>%
        mutate_if(is.numeric, round, 2) %>%
        mutate(SEASON = as.character(SEASON)) %>%
        head(25) %>%
        flextable() %>%
        autofit()

```

Putting this all together we can break down teams overall as before, only now explicitly looking at teams based on their pass and run offense/defense efficiency. Florida State 2013 still jumps to the top, evidently their passing defense that year was fantastic in addition to their passing offense. The picture is mostly unchanged from looking at just the overall numbers, but breaking it down can reveal which teams were more balanced while others were pass/run heavy.

```{r look at A&M by pass run efficiency}

ppa_col_func = 
function(x) {
        
        breaks = seq(-1, 1,
                     by = .1)
        
        colorRamp=colorRampPalette(c("red", "white", "deepskyblue1"))
        col_palette <- colorRamp(length(breaks))
        mycut <- cut(x, 
                     breaks = breaks,
                     include.lowest = TRUE, 
                     right=T,
                     label = FALSE)
        col_palette[mycut]
        
}

overall_col_func = 
function(x) {
        
        breaks = seq(-1.6, 1.6,
                     by = .2)
        
        colorRamp=colorRampPalette(c("red", "white", "deepskyblue1"))
        col_palette <- colorRamp(length(breaks))
        mycut <- cut(x, 
                     breaks = breaks,
                     include.lowest = TRUE, 
                     right=T,
                     label = FALSE)
        col_palette[mycut]
        
}

yearly_teams_type %>%
        filter(OFFENSE_Plays_Pass > 200 & DEFENSE_Plays_Pass > 200 &
                       OFFENSE_Plays_Run > 200 & DEFENSE_Plays_Run > 200) %>%
   #     filter(TEAM == 'Texas')  %>%
        mutate_if(is.numeric, round, 2) %>%
        select(-contains("Plays")) %>%
        mutate(Overall_PPA = OFFENSE_PPA_Pass + OFFENSE_PPA_Run + DEFENSE_PPA_Pass + DEFENSE_PPA_Run) %>%
        mutate(SEASON = as.character(SEASON)) %>%
        arrange(desc(Overall_PPA)) %>%
        head(25) %>%
        flextable() %>%  
        autofit() %>%
        bg(., j=c("OFFENSE_PPA_Pass", 
                  "OFFENSE_PPA_Run",
                  "DEFENSE_PPA_Pass",
                  "DEFENSE_PPA_Run"),
           bg = ppa_col_func) %>%
                bg(., j=c("Overall_PPA"),
           bg = overall_col_func) 
        

```

We can also look at individual teams year over year to see how their offense/defense/overall.

For instance, Oregon was great in the early 2010s then had a few years in which they were down.

```{r look at individual teams oregon}

yearly_teams_type %>%
        filter(TEAM == 'Oregon') %>%
        # filter(OFFENSE_Plays_Pass > 200 & DEFENSE_Plays_Pass > 200 &
        #                OFFENSE_Plays_Run > 200 & DEFENSE_Plays_Run > 200) %>%
   #     filter(TEAM == 'Texas')  %>%
        mutate_if(is.numeric, round, 2) %>%
        select(-contains("Plays")) %>%
        mutate(Overall_PPA = OFFENSE_PPA_Pass + OFFENSE_PPA_Run + DEFENSE_PPA_Pass + DEFENSE_PPA_Run) %>%
        mutate(SEASON = as.character(SEASON)) %>%
     #   arrange(desc(Overall_PPA)) %>%
        head(25) %>%
        flextable() %>%  
        autofit() %>%
        bg(., j=c("OFFENSE_PPA_Pass", 
                  "OFFENSE_PPA_Run",
                  "DEFENSE_PPA_Pass",
                  "DEFENSE_PPA_Run"),
           bg = ppa_col_func) %>%
                bg(., j=c("Overall_PPA"),
           bg = overall_col_func) 

```

Florida State had that amazing year, but has really fallen back after Jimbo's bad year and eventual departure.

```{r look at teams florida state}

yearly_teams_type %>%
        filter(TEAM == 'Florida State') %>%
        # filter(OFFENSE_Plays_Pass > 200 & DEFENSE_Plays_Pass > 200 &
        #                OFFENSE_Plays_Run > 200 & DEFENSE_Plays_Run > 200) %>%
   #     filter(TEAM == 'Texas')  %>%
        mutate_if(is.numeric, round, 2) %>%
        select(-contains("Plays")) %>%
        mutate(Overall_PPA = OFFENSE_PPA_Pass + OFFENSE_PPA_Run + DEFENSE_PPA_Pass + DEFENSE_PPA_Run) %>%
        mutate(SEASON = as.character(SEASON)) %>%
     #   arrange(desc(Overall_PPA)) %>%
        head(25) %>%
        flextable() %>%  
        autofit() %>%
        bg(., j=c("OFFENSE_PPA_Pass", 
                  "OFFENSE_PPA_Run",
                  "DEFENSE_PPA_Pass",
                  "DEFENSE_PPA_Run"),
           bg = ppa_col_func) %>%
                bg(., j=c("Overall_PPA"),
           bg = overall_col_func) 

```


Wisconsin has generally been pretty good but not quite elite during this time period, people usually hound them for their passing game lately, how do they compare year over year?

```{r look at individual teams wisconsin}

yearly_teams_type %>%
        filter(TEAM == 'Wisconsin') %>%
        # filter(OFFENSE_Plays_Pass > 200 & DEFENSE_Plays_Pass > 200 &
        #                OFFENSE_Plays_Run > 200 & DEFENSE_Plays_Run > 200) %>%
   #     filter(TEAM == 'Texas')  %>%
        mutate_if(is.numeric, round, 2) %>%
        select(-contains("Plays")) %>%
        mutate(Overall_PPA = OFFENSE_PPA_Pass + OFFENSE_PPA_Run + DEFENSE_PPA_Pass + DEFENSE_PPA_Run) %>%
        mutate(SEASON = as.character(SEASON)) %>%
     #   arrange(desc(Overall_PPA)) %>%
        head(25) %>%
        flextable() %>%  
        autofit() %>%
        bg(., j=c("OFFENSE_PPA_Pass", 
                  "OFFENSE_PPA_Run",
                  "DEFENSE_PPA_Pass",
                  "DEFENSE_PPA_Run"),
           bg = ppa_col_func) %>%
                bg(., j=c("Overall_PPA"),
           bg = overall_col_func) 

```

What about a historically weaker team that has been on the rise lately? Iowa State?

```{r iowa state year over year}

yearly_teams_type %>%
        filter(TEAM == 'Iowa State') %>%
        # filter(OFFENSE_Plays_Pass > 200 & DEFENSE_Plays_Pass > 200 &
        #                OFFENSE_Plays_Run > 200 & DEFENSE_Plays_Run > 200) %>%
   #     filter(TEAM == 'Texas')  %>%
        mutate_if(is.numeric, round, 2) %>%
        select(-contains("Plays")) %>%
        mutate(Overall_PPA = OFFENSE_PPA_Pass + OFFENSE_PPA_Run + DEFENSE_PPA_Pass + DEFENSE_PPA_Run) %>%
        mutate(SEASON = as.character(SEASON)) %>%
     #   arrange(desc(Overall_PPA)) %>%
        head(25) %>%
        flextable() %>%  
        autofit() %>%
        bg(., j=c("OFFENSE_PPA_Pass", 
                  "OFFENSE_PPA_Run",
                  "DEFENSE_PPA_Pass",
                  "DEFENSE_PPA_Run"),
           bg = ppa_col_func) %>%
                bg(., j=c("Overall_PPA"),
           bg = overall_col_func) 


```

And of course the team I keep randomly selecting, Texas A&M.

```{r look at individual teams tamu}

yearly_teams_type %>%
        filter(TEAM == 'Texas A&M') %>%
        # filter(OFFENSE_Plays_Pass > 200 & DEFENSE_Plays_Pass > 200 &
        #                OFFENSE_Plays_Run > 200 & DEFENSE_Plays_Run > 200) %>%
   #     filter(TEAM == 'Texas')  %>%
        mutate_if(is.numeric, round, 2) %>%
        select(-contains("Plays")) %>%
        mutate(Overall_PPA = OFFENSE_PPA_Pass + OFFENSE_PPA_Run + DEFENSE_PPA_Pass + DEFENSE_PPA_Run) %>%
        mutate(SEASON = as.character(SEASON)) %>%
     #   arrange(desc(Overall_PPA)) %>%
        head(25) %>%
        flextable() %>%  
        autofit() %>%
        bg(., j=c("OFFENSE_PPA_Pass", 
                  "OFFENSE_PPA_Run",
                  "DEFENSE_PPA_Pass",
                  "DEFENSE_PPA_Run"),
           bg = ppa_col_func) %>%
                bg(., j=c("Overall_PPA"),
           bg = overall_col_func) 

```

