---
title: "Simulating the Upcoming CFB Season"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE #adds a Table of Contents
    theme: cerulean
    number_sections: TRUE #number your headings/sections
    toc_float: TRUE #let your ToC follow you as you scroll
    keep_md: no
    fig.caption: yes
params:
        input_season: 2022
---

```{r setup, include=F}

knitr::opts_chunk$set(echo = F,
                      error = F,
                      warning=F,
                      dev="png",
                      fig.width = 10,
                      fig.height = 6)
options(knitr.duplicate.label = "allow")
options(scipen=999)

```

```{r packages, include=F}

source(here::here("scripts/load_packages.R"))
library(jsonlite)
library(forcats)
library(TTR)
library(flextable)
conflict_prefer("lag", "dplyr")

# parallelization
library(future.apply)
plan(multisession, workers = 7)

set_flextable_defaults(theme_fun = theme_alafoli,
                       font.color = "grey10",
                       font.size=8,
                       padding.bottom = 6, 
                       padding.top = 6,
                       padding.left = 6,
                       padding.right = 6,
                       background.color = "white")
```

```{r connect to snowflake}

library(DBI)
library(odbc)
library(RODBC)
library(keyring)

### connect to DBS
# connect to snowflake
myconn <- DBI::dbConnect(odbc::odbc(),
                         "SnowflakeDSII",
                         Database = "CFB_DEMO",
                         warehouse = "DEMO_WH",
                         uid="phil.henrickson",
                         pwd=keyring::key_get("AE_Snowflake"))


```

```{r functions and colors}

### functions
source(here::here("functions/get_expected_score.R"))
source(here::here("functions/get_new_elos.R"))
source(here::here("functions/calc_elo_ratings.R"))
source(here::here("functions", "simulateX.R"))
source(here::here("functions", "sim_game_margin.R"))
source(here::here("functions", "calc_elo_ratings.R"))
source(here::here("functions", "sim_elo_ratings.R"))
source(here::here("functions", "sim_rest_of_season.R"))
source(here::here("functions", "theme_phil.R"))

# color palettes
load(here::here("data", "team_colors.Rdata"))
custom_color_teams <- scale_colour_manual(name = "TEAM", values = team_colors)
custom_fill_teams <- scale_fill_manual(name = "TEAM", values = team_colors)

```

```{r data sources}

### get data sources
# teams raw
teams_raw = DBI::dbGetQuery(myconn,
                            paste('SELECT * FROM CFB_DEMO.CFD_RAW.TEAMS')) %>%
        as_tibble() %>%
        rename(TEAM = SCHOOL,
               TEAM_ID = ID,
               TEAM_MASCOT = MASCOT,
               TEAM_ABBREVIATION = ABBREVIATION)

# get games
games_raw = DBI::dbGetQuery(myconn,
                            paste('SELECT * FROM CFB_DEMO.CFD_RAW.ANALYSIS_GAMES')) %>%
        as_tibble() %>%
        mutate(CONFERENCE_GAME = case_when(CONFERENCE_GAME == 'true' ~ T,
                                           CONFERENCE_GAME == 'false' ~ F)) %>%
        mutate(GAME_DATE = as.Date(START_DATE)) %>%
        mutate(START_DATE = as_datetime(START_DATE)) %>%
        arrange(START_DATE) %>%
        # logic for flagging conference championship games
        group_by(SEASON, SEASON_TYPE, HOME_CONFERENCE) %>%
        mutate(LAST_GAME = START_DATE == max(START_DATE)) %>% 
        ungroup() %>%
        mutate(CONFERENCE_CHAMPIONSHIP = case_when(CONFERENCE_GAME == T & 
                                                           (HOME_CONFERENCE != 'FBS Independents') & 
                                                           (HOME_CONFERENCE != 'FCS Independents') &
                                                           !(HOME_CONFERENCE %in% 'Big Ten' & SEASON < 2014) &
                                                           !(HOME_CONFERENCE %in% 'Western Athletic') &
                                                           (NEUTRAL_SITE == T | 
                                                                    is.na(VENUE) |
                                                                    VENUE == 'Glass Bowl' |
                                                                    VENUE == 'Bright House Networks Stadium' |
                                                                    VENUE == 'Alamodome' |
                                                                    VENUE == 'Georgia Dome' |
                                                                    HOME_CONFERENCE == 'Sun Belt' |
                                                                    HOME_CONFERENCE == 'American Athletic') &
                                                           SEASON > 2000 &
                                                           SEASON_TYPE == 'regular' &
                                                           WEEK > 13 &
                                                           LAST_GAME == T ~ T,
                                                   TRUE ~ F)) %>%
        mutate(HOME_WIN = factor(case_when(HOME_POINTS > AWAY_POINTS ~ 'yes',
                                           HOME_POINTS < AWAY_POINTS ~ 'no',
                                           HOME_POINTS == AWAY_POINTS ~ 'no'),
                                 levels = c("no", "yes"))) %>%
        mutate(HOME_DIVISION  = replace_na(HOME_DIVISION, "missing"),
               AWAY_DIVISION = replace_na(AWAY_DIVISION, "missing")) %>%
        left_join(., teams_raw %>%
                          select(TEAM, TEAM_ABBREVIATION) %>%
                          rename(HOME_TEAM = TEAM,
                                 HOME_TEAM_ABBR = TEAM_ABBREVIATION),
                  by = c("HOME_TEAM")) %>%
        left_join(., teams_raw %>%
                          select(TEAM, TEAM_ABBREVIATION) %>%
                          rename(AWAY_TEAM = TEAM,
                                 AWAY_TEAM_ABBR = TEAM_ABBREVIATION),
                  by = c("AWAY_TEAM"))

# get team conference mapping
team_conference_season = DBI::dbGetQuery(myconn,
                                         paste('SELECT * FROM CFB_DEMO.CFD_RAW.ANALYSIS_TEAM_CONFERENCE_SEASON')) %>%
        as_tibble()

# team mapping function
team_mapping = teams_raw %>%
        mutate(TEAM_ABBR = gsub(" ", "\\.", gsub("[[:punct:]]", "\\.", TEAM))) %>%
        select(TEAM, TEAM_ABBR, TEAM_ABBREVIATION, COLOR, ALT_COLOR) %>%
        mutate(COLOR = case_when(TEAM == 'Toledo' ~ '#003E7E',
                                 TRUE ~ COLOR))

### get historical elo
# games
elo_games = DBI::dbGetQuery(myconn,
                                       paste('SELECT * FROM CFB_DEMO.CFD_ANALYTICS.ACTIVE_ELO_GAMES')) %>%
        as_tibble()

# teams
elo_teams = DBI::dbGetQuery(myconn,
                            paste('SELECT * FROM CFB_DEMO.CFD_ANALYTICS.ACTIVE_ELO_TEAMS')) %>%
        as_tibble() %>%
        select(-LOAD_DATE)

### recruiting 
recruiting_teams_raw = DBI::dbGetQuery(myconn,
                                       paste('SELECT * FROM CFB_DEMO.CFD_RAW.RECRUITING_TEAMS ')) %>%
        as_tibble() %>%
        mutate(POINTS = as.numeric(POINTS)) %>%
        rename(RECRUITING_POINTS = POINTS)

# expand
recruiting_teams = expand.grid(TEAM = unique(recruiting_teams_raw$TEAM),
                               YEAR = seq(min(recruiting_teams_raw$YEAR),
                                          max(recruiting_teams_raw$YEAR))) %>%
        left_join(., recruiting_teams_raw,
                  by = c("YEAR", "TEAM")) %>%
        mutate(RECRUITING_POINTS = replace_na(RECRUITING_POINTS, 0)) %>%
        filter(YEAR > 2001) %>%
        left_join(., team_conference_season,
                  by = c("YEAR"="SEASON", "TEAM")) %>%
        filter(DIVISION == 'fbs') %>%
        group_by(TEAM) %>%
        mutate(n = n()) %>%
        filter(n > 4) %>%
        group_by(TEAM) %>%
        select(-n) %>%
        mutate(EMA = EMA(RECRUITING_POINTS, n =4)) %>%
        # now fill with actual in the event that a team doesn't have yet have enough data
        mutate(EMA = case_when(is.na(EMA) & !is.na(RECRUITING_POINTS) ~ RECRUITING_POINTS,
                               TRUE ~ EMA)) %>%
        # mutate_at(c("EMA", "WMA"),
        #           replace_na, 0) %>%
        arrange(TEAM, YEAR)  %>%
        mutate(SEASON = YEAR) %>%
        ungroup() %>%
        select(TEAM, SEASON, CONFERENCE, DIVISION, RANK, RECRUITING_POINTS, EMA)

# get what we need
recruiting = recruiting_teams %>%
        select(TEAM, SEASON, CONFERENCE, RECRUITING_POINTS, EMA) %>%
        rename(RECRUITING_SCORE = EMA)


### efficiency
# season level efficiency rankings
season_team_rankings = DBI::dbGetQuery(myconn,
                paste('SELECT * FROM CFB_DEMO.CFD_ANALYTICS.EFFICIENCY_TEAM_SEASON')) %>%
        as_tibble()

### combine datasets to get games
games = games_raw %>%
        mutate(HOME_DIVISION = case_when(is.na(HOME_DIVISION) & SEASON < 1900 ~ 'fbs',
                                         TRUE ~ HOME_DIVISION)) %>%
        mutate(AWAY_DIVISION = case_when(is.na(AWAY_DIVISION) & SEASON < 1900 ~ 'fbs',
                                         TRUE ~ AWAY_DIVISION)) %>%
        mutate(FBS = HOME_DIVISION == 'fbs' | AWAY_DIVISION == 'fbs') %>%
        filter((FBS == T) | (FBS==F & SEASON < 1981)) %>%
        arrange(GAME_DATE) %>%
        select(GAME_ID, 
               SEASON, 
               WEEK,
               SEASON_TYPE,
               START_DATE,
               GAME_DATE,
               NEUTRAL_SITE, 
               HOME_TEAM,
               AWAY_TEAM,
               CONFERENCE_GAME,
               CONFERENCE_CHAMPIONSHIP,
               HOME_CONFERENCE,
               AWAY_CONFERENCE,
               HOME_DIVISION,
               AWAY_DIVISION,
               HOME_POINTS,
               AWAY_POINTS) %>%
        left_join(.,
                  recruiting %>%
                          select(SEASON,
                                 TEAM,
                                 RECRUITING_SCORE) %>%
                          rename(HOME_TEAM = TEAM,
                                 HOME_RECRUITING_SCORE = RECRUITING_SCORE),
                  by = c("SEASON",
                         "HOME_TEAM")) %>%
        left_join(.,
                  recruiting %>%
                          select(SEASON,
                                 TEAM,
                                 RECRUITING_SCORE) %>%
                          rename(AWAY_TEAM = TEAM,
                                 AWAY_RECRUITING_SCORE = RECRUITING_SCORE),
                  by = c("SEASON",
                         "AWAY_TEAM"))

# data sources
load(file=here::here("data", "season_team_rankings_type.Rdata"))

```

```{r functions for turning elo to list}

## functions for converting elo to list from specified year
teams_to_list = function(x) {
        
        setNames(unlist(x$ELO) %>%
                         split(., seq(nrow(x))),
                 rownames(x))
}

team_seasons_to_list = function(x) {
        
        setNames(unlist(x$SEASON) %>%
                         split(., seq(nrow(x))),
                 rownames(x))
}

```

```{r conference divisions}

conference_divisions = 
        tibble(SEASON = 2022,
               CONFERENCE = "SEC",
               CONFERENCE_DIVISION = c("East"),
               TEAM = c("Florida",
                        "Georgia",
                        "Kentucky",
                        "Missouri",
                        "South Carolina",
                        "Tennessee",
                        "Vanderbilt")) %>%
        bind_rows(.,
                  tibble(SEASON = 2022,
                         CONFERENCE = "SEC",
                         CONFERENCE_DIVISION = "West",
                         TEAM = c("Alabama",
                                  "Arkansas",
                                  "Auburn",
                                  "LSU",
                                  "Mississippi State",
                                  "Ole Miss",
                                  "Texas A&M"))
        ) %>%
           bind_rows(.,
                  tibble(SEASON = 2022,
                         CONFERENCE = "Big Ten",
                         CONFERENCE_DIVISION = "West",
                         TEAM = c("Illinois", 
                                  "Iowa", 
                                  "Minnesota",
                                  "Nebraska",
                                  "Northwestern",
                                  "Purdue",
                                  "Wisconsin"))
           ) %>%
         bind_rows(.,
                  tibble(SEASON = 2022,
                         CONFERENCE = "Big Ten",
                         CONFERENCE_DIVISION = "East",
                         TEAM = c("Indiana",
                                  "Maryland",
                                  "Michigan",
                                  "Michigan State",
                                  "Ohio State",
                                  "Penn State",
                                  "Rutgers"))
         ) %>%
        bind_rows(.,
                  tibble(SEASON = 2022,
                         CONFERENCE = "ACC",
                         CONFERENCE_DIVISION = "Atlantic",
                         TEAM = c("Boston College",
                                  "Clemson",
                                  "Florida State",
                                  "Louisville",
                                  "NC State",
                                  "Syracuse",
                                  "Wake Forest"))
                  ) %>%
                bind_rows(.,
                  tibble(SEASON = 2022,
                         CONFERENCE = "ACC",
                         CONFERENCE_DIVISION = "Coastal",
                         TEAM = c("Duke",
                                  "Georgia Tech",
                                  "Miami",
                                  "North Carolina",
                                  "Pittsburgh",
                                  "Virginia",
                                  "Virginia Tech"))
                ) %>%
        bind_rows(.,
                  tibble(SEASON = 2022,
                         CONFERENCE = "Mid-American",
                         CONFERENCE_DIVISION = "East",
                         TEAM = c("Akron",
                                  "Bowling Green",
                                  "Buffalo", 
                                  "Kent State",
                                  "Miami (OH)",
                                  "Ohio"))
        ) %>%
                bind_rows(.,
                  tibble(SEASON = 2022,
                         CONFERENCE = "Mid-American",
                         CONFERENCE_DIVISION = "West",
                         TEAM = c("Ball State",
                                  "Central Michigan",
                                  "Eastern Michigan",
                                  "Northern Illinois",
                                  "Toledo",
                                  "Western Michigan"))
                  ) %>%
        bind_rows(.,
                  tibble(SEASON = 2022,
                         CONFERENCE = "Mountain West",
                         CONFERENCE_DIVISION = "Mountain",
                         TEAM = c("Air Force",
                                  "Boise State",
                                  "Colorado State",
                                  "New Mexico",
                                  "Utah State",
                                  "Wyoming"))
        ) %>%
          bind_rows(.,
                  tibble(SEASON = 2022,
                         CONFERENCE = "Mountain West",
                         CONFERENCE_DIVISION = "West",
                         TEAM = c("Fresno State",
                                  "Hawai'i",
                                  "Nevada",
                                  "San Diego State",
                                  "San José State",
                                  "UNLV"))) %>%
        bind_rows(.,
                  tibble(SEASON = 2022,
                         CONFERENCE = "Sun Belt",
                         CONFERENCE_DIVISION = "East",
                         TEAM = c("Appalachian State",
                                  "Coastal Carolina",
                                  "Georgia Southern",
                                  "Georgia State",
                                  "James Madison",
                                  "Marshall",
                                  "Old Dominion"))
        ) %>% 
                bind_rows(.,
                  tibble(SEASON = 2022,
                         CONFERENCE = "Sun Belt",
                         CONFERENCE_DIVISION = "West",
                         TEAM = c("Arkansas State",
                                  "Louisiana",
                                  "Louisiana Monroe",
                                  "South Alabama",
                                  "Southern Mississippi",
                                  "Texas State",
                                  "Troy"))
                  ) %>%
        bind_rows(.,
                  tibble(SEASON = 2022,
                         CONFERENCE = 'Big 12',
                         TEAM = c("Baylor",
                                  "Iowa State",
                                  "Kansas State",
                                  "Kansas",
                                  "Oklahoma",
                                  "Oklahoma State",
                                  "TCU",
                                  "Texas",
                                  "Texas Tech",
                                  "West Virginia"))) %>%
        bind_rows(.,
                  tibble(SEASON = 2022,
                         CONFERENCE = 'Pac-12',
                         TEAM = c("Arizona",
                                  "Arizona State",
                                  "California",
                                  "Colorado",
                                  "Oregon",
                                  "Oregon State",
                                  "Stanford",
                                  "UCLA",
                                  "USC",
                                  "Utah",
                                  "Washington",
                                  "Washington State"))
        ) %>%
          bind_rows(.,
                  tibble(SEASON = 2022,
                         CONFERENCE = 'Conference USA',
                         TEAM = c("Charlotte",
                                  "Florida Atlantic",
                                  "Florida International",
                                  "Louisiana Tech",
                                  "Middle Tennessee",
                                  "North Texas",
                                  "Rice",
                                  "UAB",
                                  "UTEP",
                                  "UT San Antonio",
                                  "Western Kentucky"))
          ) %>%
        bind_rows(.,
                  tibble(SEASON = 2022,
                         CONFERENCE = 'American Athletic',
                         TEAM = c("Cincinnati",
                                  "East Carolina",
                                  "Houston",
                                  "Memphis",
                                  "Navy",
                                  "SMU",
                                  "South Florida",
                                  "Temple",
                                  "Tulane",
                                  "Tulsa",
                                  "UCF"
                                  ))) %>%
        bind_rows(.,
                    tibble(SEASON = 2022,
                         CONFERENCE = 'FBS Independents',
                         TEAM = c("Army",
                                  "BYU",
                                  "Liberty",
                                  "New Mexico State",
                                  "Notre Dame",
                                  "UConn",
                                  "UMass")))

```

```{r create data for simulating, include=F}

# 2022 season 
future_seasons = expand.grid(TEAM = season_team_rankings  %>%
                                     select(TEAM) %>%
                                     distinct(TEAM) %>%
                                     pull(),
                             TYPE = c('raw', 'adjusted'),
                             SEASON = params$input_season) %>%
        as_tibble()

# get efficiency and elo for historical + future season
teams_data = season_team_rankings %>%
        bind_rows(., future_seasons) %>%
        filter(TYPE == 'adjusted') %>%
        select(-TYPE) %>%
        # add in season end elo ratings
        left_join(.,
                  elo_teams %>%
                          group_by(SEASON, TEAM) %>%
                          arrange(GAME_DATE) %>%
                          mutate(WEEK = row_number()) %>%
                          mutate(LAST_GAME = GAME_DATE == max(GAME_DATE)) %>%
                          filter(LAST_GAME ==T) %>%
                          select(SEASON, TEAM, POSTGAME_ELO) %>%
                          rename(ELO = POSTGAME_ELO),
                  by = c("SEASON", "TEAM")) %>%
        # add in season recruiting data
        left_join(., 
                  recruiting %>%
                          select(TEAM, SEASON, RECRUITING_SCORE),
                  by = c("SEASON", "TEAM")) %>%
        arrange(SEASON, TEAM) %>%
        group_by(TEAM) %>%
        mutate(PREVIOUS_OVERALL = lag(OVERALL, 1),
               PREVIOUS_OFFENSE = lag(OFFENSE, 1),
               PREVIOUS_DEFENSE = lag(DEFENSE, 1),
               PREVIOUS_ELO = lag(ELO, 1)) %>%
        ungroup() %>%
        filter(SEASON > 2007)


# split into training and test
teams_train  = teams_data %>%
        filter(SEASON < params$input_season)

teams_test = teams_data %>%
        filter(SEASON == params$input_season)

# create recipe
teams_recipe =      
        recipe(ELO ~.,
               data = teams_train) %>%
        update_role(all_predictors(),
                    new_role = "id") %>%
        update_role(PREVIOUS_OFFENSE,
                    PREVIOUS_DEFENSE,
                    PREVIOUS_ELO,
                    RECRUITING_SCORE,
                    new_role = 'predictor') %>%
        # manual missingness
        step_mutate(RECRUITING_SCORE = replace_na(RECRUITING_SCORE, 0),
                    PREVIOUS_ELO = replace_na(PREVIOUS_ELO, 1200),
                    PREVIOUS_OFFENSE = replace_na(PREVIOUS_OFFENSE,-0.1),
                    PREVIOUS_DEFENSE = replace_na(PREVIOUS_DEFENSE, -0.1)) %>%
        step_normalize(all_predictors())

# linear model
lm_mod = linear_reg(mode = "regression",
                    engine = "lm")

# create workflow
teams_workflow = workflow() %>%
        add_recipe(teams_recipe) %>%
        add_model(lm_mod)

# run workflow
adjusted_teams_test = 
        teams_workflow %>%
        fit(bind_rows(teams_train)) %>%
        augment(bind_rows(teams_test %>%
                                  mutate(dataset = 'test'))) %>%
        select(SEASON, dataset, TEAM, starts_with("PREVIOUS"),
               ELO, .pred) %>%
        mutate_if(is.numeric, round, 3)

# set season to simulate
# set initial games
initial_games = games %>%
        filter(SEASON < params$input_season)

# elo pars
elo_pars = tibble(k = 35,
                  v = 400,
                  reversion = 0,
                  home_field_advantage = 75,
                  recruiting_weight =0)

# train points model
points_model = elo_games %>%
        filter(SEASON >=1970) %>% 
        filter(SEASON < params$input_season) %>%
        mutate(HOME_ELO_DIFF = case_when(NEUTRAL_SITE == F ~ 
                                                 HOME_PREGAME_ELO + 
                                                 elo_pars$home_field_advantage -
                                                 AWAY_PREGAME_ELO,
                                         TRUE ~ HOME_PREGAME_ELO - AWAY_PREGAME_ELO)) %>%
        mutate(HOME_SCORE_DIFF = HOME_POINTS-AWAY_POINTS) %>%
        nest() %>%
        # now fit linear models, i'll do so using stan
        mutate(lm = map(data, ~ lm(HOME_SCORE_DIFF ~ HOME_ELO_DIFF +0,
                                   data = .x))) %>%
        pluck("lm",1)

```

```{r workflow and set up}

# base elo, from historical games
base_teams_elo = elo_teams %>%
        filter(SEASON < params$input_season) %>%
        group_by(TEAM) %>%
        filter(GAME_DATE == max(GAME_DATE)) %>%
        slice_max(POSTGAME_ELO, n=1, with_ties = F) %>%
        ungroup() %>%
        select(TEAM, POSTGAME_ELO) %>%
        rename(ELO = POSTGAME_ELO) %>%
        column_to_rownames("TEAM")

# now adjusted
# get computed elo as input
adjusted_teams_elo = adjusted_teams_test %>%
        filter(SEASON == params$input_season) %>%
        select(TEAM, .pred) %>%
        rename(ELO = .pred) %>%
        column_to_rownames("TEAM")

# get team seasons
team_elo_seasons = elo_teams %>%
        filter(SEASON < params$input_season) %>%
        group_by(TEAM) %>%
        slice_max(., order_by = SEASON, n=1, with_ties =F) %>%
        ungroup() %>%
        select(TEAM, SEASON) %>%
        column_to_rownames("TEAM")

# convert to list
base_teams = teams_to_list(base_teams_elo)
adjusted_teams = teams_to_list(adjusted_teams_elo)
team_seasons = team_seasons_to_list(team_elo_seasons)

# in format for function
elo_input = list(teams = c(adjusted_teams,
                           base_teams[!(names(base_teams) %in% names(adjusted_teams))]),
                   team_seasons = team_seasons)

```

# What is this? {-}

This notebook details the results of simulating the `r params$input_season` college football season before any actual games are played. 

# `r params$input_season` Week 0 Team Ratings

The following table displays each team's **Rating**, which is a combination of their previous season's Elo Rating, offensive efficiency, defensive efficiency, and a moving average of their composite recruiting score. These ratings are **Week 0** ratings because they rely only on information from previous seasons. As the season progresses, these ratings will update for each team as we gain new information based on the results of games (with the exception of recruiting, which is computed at the yearly level).

```{r rank teams after adjustment}

# overall 
preseason_ratings = do.call(rbind, adjusted_teams) %>%
        as.data.frame() %>%
        rownames_to_column("TEAM") %>%
        rename(RATING = V1) %>%
        mutate(SEASON = params$input_season) %>%
        left_join(.,
                  team_mapping,
                  by = c("TEAM")) %>%
        left_join(.,
                  team_conference_season,
                  by = c("SEASON", "TEAM")) %>%
        filter(!is.na(CONFERENCE)) %>%
        filter(DIVISION =='fbs') %>%
        left_join(.,
                  teams_data %>%
                          select(SEASON,
                                 TEAM,
                                 PREVIOUS_OFFENSE, 
                                 PREVIOUS_DEFENSE,
                                         RECRUITING_SCORE),
                  by  = c("SEASON", "TEAM")) %>%
        rename(OFFENSE = PREVIOUS_OFFENSE,
               DEFENSE = PREVIOUS_DEFENSE,
               RECRUITING = RECRUITING_SCORE)
                  
# make colors break for overall
blueColorRamp = colorRampPalette(c("white", "deepskyblue1"))
blueRedColorRamp = colorRampPalette(c("red", "white", "deepskyblue1"))

# set up color mapping
breaks_df = data.frame(outcome = 'Rating',
                       breaks = quantile(preseason_ratings %>% 
                                                 pull(RATING), 
                                         probs = seq(0, 1, .01), na.rm=T) %>%
                               as.vector) %>%
        bind_rows(data.frame(outcome = 'Recruiting',
                       breaks = quantile(preseason_ratings %>% 
                                                 pull(RECRUITING), 
                                         probs = seq(0, 1, .01), na.rm=T) %>%
                               as.vector)) %>%
        bind_rows(data.frame(outcome = 'Offense',
                       breaks = quantile(preseason_ratings %>% 
                                                 pull(OFFENSE), 
                                         probs = seq(0, 1, .01), na.rm=T) %>%
                               as.vector)) %>%
        bind_rows(data.frame(outcome = 'Defense',
                       breaks = quantile(preseason_ratings %>% 
                                                 pull(DEFENSE), 
                                         probs = seq(0, 1, .01), na.rm=T) %>%
                               as.vector))

# table
preseason_ratings %>%
        arrange(desc(RATING)) %>%
        mutate(RANK = row_number()) %>%
        select(SEASON, CONFERENCE, TEAM, RANK, RATING, OFFENSE, DEFENSE, RECRUITING) %>%
        mutate(RATING = round(RATING, 0),
               RECRUITING = round(RECRUITING, 0)) %>%
        mutate(SEASON = factor(SEASON),
               CONFERENCE = factor(CONFERENCE)) %>%
        rename(Season = SEASON,
               Conference = CONFERENCE,
               Team = TEAM,
               Rank = RANK,
               Rating = RATING,
               Offense = OFFENSE,
               Defense = DEFENSE,
               Recruiting = RECRUITING) %>%
        datatable(escape=F,
                  rownames = F,
                       filter = 'top',
                  options = list(pageLength = 25,
                                 scrollX=F,
                                 autowidth=T,
                                 columnDefs = list(list(className = 'dt-center',
                                                        targets = c("Season", "Conference",
                                                                    "Team", "Rank", "Rating",
                                                                    "Offense", "Defense",
                                                                    "Recruiting"))))) %>%
             formatStyle("Rating", backgroundColor = styleInterval(breaks_df %>%
                                                                          filter(outcome == 'Rating') %>%
                                                                          pull(breaks),
                                                                  blueRedColorRamp(breaks_df %>%
                                                                          filter(outcome == 'Rating') %>%
                                                                          pull(breaks) %>%
                                                                                  length()+1))) %>%
          formatStyle("Offense", backgroundColor = styleInterval(breaks_df %>%
                                                                          filter(outcome == 'Offense') %>%
                                                                          pull(breaks),
                                                                  blueRedColorRamp(breaks_df %>%
                                                                          filter(outcome == 'Offense') %>%
                                                                          pull(breaks) %>%
                                                                                  length()+1))) %>%
                     formatStyle("Defense", backgroundColor = styleInterval(breaks_df %>%
                                                                          filter(outcome == 'Defense') %>%
                                                                          pull(breaks),
                                                                  blueRedColorRamp(breaks_df %>%
                                                                          filter(outcome == 'Defense') %>%
                                                                          pull(breaks) %>%
                                                                                  length()+1))) %>%
                     formatStyle("Recruiting", backgroundColor = styleInterval(breaks_df %>%
                                                                          filter(outcome == 'Recruiting') %>%
                                                                          pull(breaks),
                                                                  blueRedColorRamp(breaks_df %>%
                                                                          filter(outcome == 'Recruiting') %>%
                                                                          pull(breaks) %>%
                                                                                  length()+1)))

```

<!-- The following visualization shows the distribution of team ratings at the start of the season. -->

<!-- ```{r graph of preseason rankings, fig.height=8} -->

<!-- preseason_ratings %>% -->
<!--         ggplot(., aes(x = "1", -->
<!--                       label = TEAM_ABBREVIATION, -->
<!--                       fill = TEAM, -->
<!--                       y = RATING))+ -->
<!--         # geom_point(position = 'auto', -->
<!--         #            jitter.width = 0.05)+ -->
<!--         geom_label_repel(position = position_auto(seed = 1, -->
<!--                                                   jitter.width=0.5, -->
<!--                                             scale =T), -->
<!--                          max.overlaps = 100, -->
<!--                    color = 'white', -->
<!--                    size = 3, -->
<!--                    alpha = 0.8)+ -->
<!--         custom_fill_teams+ -->
<!--         guides(fill = 'none')+ -->
<!--         theme_phil()+ -->
<!--         xlab("This axis has no meaning")+ -->
<!--         theme(axis.ticks.x=element_blank(), -->
<!--               axis.text.x = element_blank())+ -->
<!--   #      facet_wrap(SEASON + CONFERENCE + metric~.)+ -->
<!--         xlab("This axis has no meaning")+ -->
<!--         ylab("Team Rating")+ -->
<!--         geom_hline(yintercept = 1520, -->
<!--                    linetype = 'dashed', -->
<!--                    color = 'grey60')+ -->
<!--         theme(plot.title = element_text(hjust = 0.5, -->
<!--                                                 size = 16), -->
<!--                       plot.subtitle =  element_text(hjust = 0.5), -->
<!--                       strip.text.x = element_text(size = 12, -->
<!--                                                   hjust = 0.5))+ -->
<!--                 ggtitle(paste(params$input_season, "Week 0 Team Ratings"), -->
<!--                 subtitle = "Displaying preseason team ratings for all teams. Grey line indicates FBS median.") -->

<!-- ``` -->

## Conferences

The following visualization shows the distribution of team ratings for each conference.

```{r conference ratings, fig.height=8}

preseason_ratings %>%
        ggplot(., aes(x = "1",
                      label = TEAM_ABBREVIATION,
                      fill = TEAM,
                      y = RATING))+
        # geom_point(position = 'auto',
        #            jitter.width = 0.05)+
        geom_label_repel(position = position_auto(seed = 5,
                                                  jitter.width=0.15,
                                            scale =T),
                         max.overlaps = 100,
                   color = 'white',
                   size = 3,
                   alpha = 0.8)+
        custom_fill_teams+
        guides(fill = 'none')+
        theme_phil()+
        xlab("This axis has no meaning")+
        theme(axis.ticks.x=element_blank(),
              axis.text.x = element_blank())+
        facet_wrap(CONFERENCE ~.)+
  #      facet_wrap(SEASON + CONFERENCE + metric~.)+
        xlab("This axis has no meaning")+
        ylab("Team Rating")+
        geom_hline(yintercept = 1520,
                   linetype = 'dashed',
                   color = 'grey40')+
        theme(plot.title = element_text(hjust = 0.5,
                                                size = 16),
                      plot.subtitle =  element_text(hjust = 0.5),
                      strip.text.x = element_text(size = 12,
                                                  hjust = 0.5))+
        ggtitle(paste(params$input_season, "Week 0 Team Ratings by Conference"),
                subtitle = "Displaying preseason team ratings for all teams by conference. Grey line indicates FBS median.")

```

# Simulations

```{r simulate, include=F}

# simulate season from the start
# set.seed(1999)
set.seed(1999)
season_simulations = sim_rest_of_season(games,
                                                 season = params$input_season,
                                                 elo_initial = elo_input,
                                                 elo_pars = elo_pars,
                                                 nsims = 1000,
                                                 sim_start_week = 0,
                                                 season_start_only = T,
                                                 aggregate = F,
                                                 points_model = points_model)

```


```{r check on conference}

# conference wins
conference_wins = season_simulations %$% 
        team_outcomes %>%
        left_join(.,
                  team_conference_season,
                  by = c("SEASON", "TEAM", "CONFERENCE")) %>%
        left_join(.,
                  team_conference_season %>%
                  rename(OPPONENT = TEAM,
                         OPPONENT_CONFERENCE = CONFERENCE),
                  by = c("SEASON", "OPPONENT")) %>%
        mutate(CONFERENCE_GAME = CONFERENCE == OPPONENT_CONFERENCE) %>%
        filter(CONFERENCE_GAME == T) %>%
       # filter(CONFERENCE_DIVISION_GAME == T) %>%
        group_by(CONFERENCE) %>%
        group_by(.id, SEASON, TEAM, CONFERENCE, SIM_OUTCOME) %>%
        count() %>%
        ungroup() %>%
        spread(SIM_OUTCOME, n) %>%
        # group_by(.id, SEASON, TEAM, CONFERENCE, CONFERENCE_DIVISION, SIM_OUTCOME) %>%
        # summarize(mean = mean(n),
        #           median = median(n),
        #           .groups = 'drop') %>%
        # select(-mean) %>%
        mutate_at(c("win", "loss"),
                  ~ replace_na(., 0)) %>%
        select(.id, SEASON, TEAM, CONFERENCE, win, loss) 

# division conference wins
division_wins = season_simulations %$% 
        team_outcomes %>%
        left_join(.,
                  conference_divisions,
                  by = c("SEASON", "TEAM", "CONFERENCE")) %>%
        left_join(.,
                  conference_divisions %>%
                  rename(OPPONENT = TEAM,
                         OPPONENT_CONFERENCE = CONFERENCE,
                         OPPONENT_CONFERENCE_DIVISION = CONFERENCE_DIVISION),
                  by = c("SEASON", "OPPONENT")) %>%
        mutate(CONFERENCE_DIVISION_GAME = CONFERENCE_DIVISION == OPPONENT_CONFERENCE_DIVISION) %>%
        mutate(CONFERENCE_GAME = CONFERENCE == OPPONENT_CONFERENCE) %>%
        filter(CONFERENCE_DIVISION_GAME == T) %>%
        group_by(CONFERENCE, CONFERENCE_DIVISION_GAME) %>%
        group_by(.id, SEASON, TEAM, CONFERENCE, CONFERENCE_DIVISION, SIM_OUTCOME) %>%
        count() %>%
        ungroup() %>%
        spread(SIM_OUTCOME, n) %>%
        # group_by(.id, SEASON, TEAM, CONFERENCE, CONFERENCE_DIVISION, SIM_OUTCOME) %>%
        # summarize(mean = mean(n),
        #           median = median(n),
        #           .groups = 'drop') %>%
        # select(-mean) %>%
        mutate_at(c("win", "loss"),
                  ~ replace_na(., 0)) %>%
        select(.id, SEASON, TEAM, CONFERENCE, CONFERENCE_DIVISION, win, loss)

# tie breaker
# head to head winner
head_to_head_wins = season_simulations %$% 
        game_outcomes %>%
        left_join(.,
                  conference_divisions %>%
                          rename(HOME_TEAM = TEAM,
                                 HOME_CONFERENCE = CONFERENCE,
                                 HOME_CONFERENCE_DIVISION = CONFERENCE_DIVISION),
                  by = c("SEASON", "HOME_TEAM", "HOME_CONFERENCE")) %>%
        left_join(.,
                  conference_divisions %>%
                          rename(AWAY_TEAM = TEAM,
                                 AWAY_CONFERENCE = CONFERENCE,
                                 AWAY_CONFERENCE_DIVISION = CONFERENCE_DIVISION),
                  by = c("SEASON", "AWAY_TEAM", "AWAY_CONFERENCE")) %>%
        mutate(CONFERENCE_DIVISION_GAME = HOME_CONFERENCE_DIVISION == AWAY_CONFERENCE_DIVISION) %>%
        mutate(CONFERENCE_GAME = HOME_CONFERENCE == AWAY_CONFERENCE) %>%
        filter(CONFERENCE_GAME == T) %>%
   #     filter(CONFERENCE_DIVISION_GAME == T) %>%
        mutate(winner = case_when(HOME_SIM_OUTCOME == 'win' ~ HOME_TEAM,
                                  HOME_SIM_OUTCOME == 'loss' ~ AWAY_TEAM)) %>%
      #  distinct(.id, SEASON, GAME_ID, CONFERENCE, CONFERENCE_DIVISION, TEAM, OPPONENT, winner) %>%
        rename(CONFERENCE = HOME_CONFERENCE,
               CONFERENCE_DIVISION = HOME_CONFERENCE_DIVISION,
               TEAM = HOME_TEAM,
               OPPONENT = AWAY_TEAM) %>%
        select(.id, GAME_ID, SEASON, CONFERENCE, CONFERENCE_DIVISION, TEAM, OPPONENT, winner) 

head_to_head_winners = bind_rows(head_to_head_wins,
                              head_to_head_wins %>%
                                      rename(OPPONENT = TEAM,
                                             TEAM = OPPONENT))

# for conference championships:
# for sec, big 12, ACC...
# take team from each division with most division wins
# if tied, take team with more conference wins
# if still tied, take team with head to head win
division_winners = division_wins %>%
        group_by(.id, CONFERENCE, CONFERENCE_DIVISION) %>%
        mutate(division_wins = win) %>%
        slice_max(., division_wins, n=1, with_ties=T) %>%
        left_join(.,
                  season_simulations %$% 
                          team_outcomes %>%
                          group_by(.id, SEASON, TEAM) %>%
                          filter(GAME_DATE == max(GAME_DATE)) %>%
                          ungroup() %>%
                          select(.id, SEASON, TEAM, POSTGAME_ELO),
                  by = c(".id", "SEASON", "TEAM")) %>%
        ungroup() %>%
        left_join(.,
                  conference_wins %>%
                          mutate(conference_wins = win) %>%
                          select(.id, SEASON, TEAM, conference_wins),
                  by = c(".id", "SEASON", "TEAM")) %>%
        group_by(.id, SEASON, CONFERENCE, CONFERENCE_DIVISION) %>%
        slice_max(., order_by = conference_wins, with_ties = T) %>%
        ungroup() %>%
        group_by(.id, SEASON, CONFERENCE, CONFERENCE_DIVISION) %>% 
        mutate(division_teams = n()) %>%
        ungroup()


# for other conferences, need to just get top winning percentages
conference_participants =  season_simulations %$% 
        game_outcomes %>%
        filter(
        (HOME_CONFERENCE == 'Pac-12' & AWAY_CONFERENCE == 'Pac-12') |
                (HOME_CONFERENCE == 'Big 12' & AWAY_CONFERENCE == 'Big 12') |
                (HOME_CONFERENCE == 'Conference USA' & AWAY_CONFERENCE == 'Conference USA') |
                (HOME_CONFERENCE == 'American Athletic' & AWAY_CONFERENCE == 'American Athletic')) %>%
        filter(CONFERENCE_GAME == T) %>%
        rename(TEAM = HOME_TEAM,
               CONFERENCE = HOME_CONFERENCE,
               OUTCOME = HOME_SIM_OUTCOME) %>%
        select(.id, SEASON, CONFERENCE, TEAM, OUTCOME) %>%
        bind_rows(.,
                  season_simulations %$%
                          game_outcomes %>%
                                filter(
                                (HOME_CONFERENCE == 'Pac-12' & AWAY_CONFERENCE == 'Pac-12') |
                                        (HOME_CONFERENCE == 'Big 12' & AWAY_CONFERENCE == 'Big 12') |
                                        (HOME_CONFERENCE == 'Conference USA' & AWAY_CONFERENCE == 'Conference USA') |
                                        (HOME_CONFERENCE == 'American Athletic' & AWAY_CONFERENCE == 'American Athletic')) %>%
                                filter(CONFERENCE_GAME == T) %>%
                                rename(TEAM = AWAY_TEAM,
                                       CONFERENCE = AWAY_CONFERENCE,
                                       OUTCOME = AWAY_SIM_OUTCOME) %>%
                                select(.id, SEASON, CONFERENCE, TEAM, OUTCOME)) %>%
        group_by(.id, SEASON, TEAM, CONFERENCE, OUTCOME) %>%
        count() %>%
        ungroup() %>%
        spread(OUTCOME ,n) %>%
        mutate_at(c("win", "loss"),
                  ~ replace_na(., 0)) %>%
        mutate(perc = win / (win + loss)) %>%
        group_by(.id, SEASON, CONFERENCE) %>%
        arrange(desc(perc)) %>%
    #    mutate(rank =  rank(desc(perc)))
        slice_max(perc, n=2, with_ties = T) %>%
        ungroup() %>%
        group_by(.id, SEASON, CONFERENCE) %>%
        mutate(rank = rank(desc(perc))) %>%
        ungroup() %>%
        group_by(.id, SEASON, CONFERENCE) %>%
        mutate(participants = n())
 
# now tiebreak
conference_status = 
        # teams that are in
        conference_participants %>%
        filter(rank <=2 & participants < 3) %>%
        ungroup() %>%
        # teams that need to have a tiebreaker based on head to head
        bind_rows(.,
                  conference_participants %>%
                          filter(participants > 2) %>%
                          left_join(.,
                                      conference_participants %>%
                                            filter(participants > 2) %>%
                                            rename(OPPONENT=TEAM) %>%
                                            select(.id, SEASON, CONFERENCE, OPPONENT),
                                    by = c(".id", "SEASON", "CONFERENCE")
                                    ) %>%
                           left_join(.,
                                     head_to_head_winners %>%
                                             rename(head_to_head = winner)  %>%
                                             select(.id, SEASON, CONFERENCE, TEAM, OPPONENT, head_to_head),
                                     by = c(".id", "SEASON", "CONFERENCE", "TEAM", "OPPONENT")) %>%
                          filter(TEAM != OPPONENT) %>%
                          group_by(.id, SEASON, CONFERENCE, TEAM, loss, win, perc, rank) %>%
                          mutate(head_to_head_win = case_when(TEAM == head_to_head ~ 1,
                                                    TRUE ~ 0)) %>%
                          summarize(head_to_head_wins = sum(head_to_head_win),
                                    .groups = 'drop') %>%
                          group_by(.id, SEASON, CONFERENCE) %>%
                          slice_max(order_by = perc,
                                    n = 2, 
                                    with_ties=T)
                  ) %>%
        group_by(.id, SEASON, CONFERENCE) %>%
        mutate(participants = n(),
               max_perc = max(perc),
               min_perc = min(perc)) %>%
        mutate(status = case_when(participants == 2 ~ 1,
                                  perc == max(perc) & perc > min(perc) ~ 1,
                                  TRUE ~ 0)) %>%
                group_by(.id, SEASON, CONFERENCE) %>%
        mutate(spots_left = 2-sum(status)) 

# conference_booked
conference_booked = conference_status %>%
        select(.id, SEASON, TEAM, CONFERENCE, loss, win, perc, rank, participants, status,  spots_left) %>%
        filter(status == 1) %>%
        bind_rows(.,
                  conference_status %>%
                          select(.id, SEASON, TEAM, CONFERENCE, loss, win, perc, rank, participants, status,  spots_left) %>%
                          filter(status == 0) %>%
                          group_by(.id, SEASON, CONFERENCE) %>%
                          sample_n(size = spots_left)
                  ) %>%
        left_join(.,
          season_simulations %$% 
                          team_outcomes %>%
                          group_by(.id, SEASON, TEAM) %>%
                          filter(GAME_DATE == max(GAME_DATE)) %>%
                          ungroup() %>%
                          select(.id, SEASON, TEAM, POSTGAME_ELO),
                  by = c(".id", "SEASON", "TEAM")) %>%
        ungroup() %>%
        rename(PREGAME_ELO = POSTGAME_ELO)

# division wins
division_booked = division_winners %>%
        left_join(., division_winners %>%
                          filter(division_teams > 1) %>%
                          select(.id, SEASON, CONFERENCE, CONFERENCE_DIVISION, TEAM) %>%
                          rename(OPPONENT = TEAM),
                  by = c(".id", "SEASON", "CONFERENCE", "CONFERENCE_DIVISION")) %>%
        filter(TEAM != OPPONENT | is.na(OPPONENT)) %>%
        rename(PREGAME_ELO = POSTGAME_ELO) %>%
        select(.id, SEASON, CONFERENCE, CONFERENCE_DIVISION, TEAM, OPPONENT,  PREGAME_ELO, division_wins, conference_wins) %>%
        left_join(.,
                  head_to_head_winners %>%
                          rename(head_to_head = winner)  %>%
                          select(.id, SEASON, CONFERENCE, CONFERENCE_DIVISION, TEAM, OPPONENT, head_to_head),
                  by = c(".id", "SEASON", "CONFERENCE", "CONFERENCE_DIVISION", "TEAM", "OPPONENT")) %>%
        filter(TEAM == head_to_head | is.na(head_to_head)) %>%
        # if still more than two, im just doing a damn coin flip and i will fight anyone who disagrees
        group_by(.id, CONFERENCE, CONFERENCE_DIVISION) %>%
        sample_n(1) %>%
        ungroup() 


# combine into conference championships table
conference_championships = bind_rows(division_booked,
                                     conference_booked) %>%
        arrange(.id,  CONFERENCE)

```

```{r simulate the damn conference championships}

# simulate outcome
set.seed(2000)
conference_championship_results = conference_championships %>%
        select(.id, SEASON, TEAM, CONFERENCE, PREGAME_ELO) %>%
        group_by(.id, SEASON, CONFERENCE) %>%
        mutate(DIVISION_ID = row_number()) %>%
        ungroup() %>%
        mutate(HOME = case_when(DIVISION_ID == 1 ~ 'HOME',
                                DIVISION_ID== 2 ~ 'AWAY')) %>%
        pivot_wider(., id_cols = c(".id", "SEASON", "CONFERENCE"),
                    names_from = c("HOME"),
                    values_from = c("TEAM", "PREGAME_ELO")) %>%
        nest(-.id, -SEASON, -CONFERENCE) %>%
        mutate(result=  map(data, ~ sim_game_margin(home_rating = .x$PREGAME_ELO_HOME,
                                                    away_rating = .x$PREGAME_ELO_AWAY,
                                                    points_model) %>%
                                    as_tibble() %>%
                                    rename(HOME_SIM_MARGIN = value) %>%
                                  bind_cols(.x, .) %>%
                                    mutate(HOME_SIM_MARGIN = case_when(HOME_SIM_MARGIN == 0 ~ sample(c(-1, 1), 1),
                                                                       TRUE ~ HOME_SIM_MARGIN)) %>%
                                  mutate(WINNER= case_when(HOME_SIM_MARGIN > 0 ~ TEAM_HOME,
                                                           HOME_SIM_MARGIN <0 ~ TEAM_AWAY),
                                         RUNNER_UP = case_when(HOME_SIM_MARGIN >= 0 ~ TEAM_AWAY,
                                                           HOME_SIM_MARGIN <0 ~ TEAM_HOME)))) %>%
        mutate(update = map2(data, result, ~ get_new_elos(home_rating = .x$PREGAME_ELO_HOME,
                                                         away_rating = .x$PREGAME_ELO_AWAY,
                                                         home_field_advantage = 0,
                                                         home_margin = .y$HOME_SIM_MARGIN,
                                                         k = elo_pars$k,
                                                         v = elo_pars$v) %>%
                                     t() %>%
                                     as_tibble() %>%
                                     set_names(c("HOME_POSTGAME_ELO",
                                                 "AWAY_POSTGAME_ELO",
                                                 "HOME_EXPECTED_SCORE",
                                                 "AWAY_EXPECTED_SCORE")))) %>%
        select(.id, SEASON, CONFERENCE, data, result, update) %>%
        unnest() %>%
        ungroup() %>%
        select(.id, SEASON, CONFERENCE, 
               TEAM_HOME, TEAM_AWAY, 
               PREGAME_ELO_HOME, PREGAME_ELO_AWAY,
               HOME_POSTGAME_ELO, AWAY_POSTGAME_ELO,
               HOME_SIM_MARGIN, 
               WINNER, RUNNER_UP) %>%
        rename(POSTGAME_ELO_HOME = HOME_POSTGAME_ELO,
               POSTGAME_ELO_AWAY = AWAY_POSTGAME_ELO)

# conference championship probs
conference_championship_probs = 
        conference_divisions %>%
        left_join(., 
        conference_championship_results %>%
                gather("OUTCOME", "TEAM", 
                       -.id, -SEASON, -CONFERENCE, -TEAM_HOME, -TEAM_AWAY, -HOME_SIM_MARGIN) %>%
                mutate(sims = n_distinct(.id)) %>%
                group_by(SEASON, CONFERENCE, OUTCOME, TEAM, sims) %>%
                count() %>%
                ungroup() %>%
                mutate(perc = n/sims)  %>%
                select(-n) %>%
                spread(OUTCOME, perc) %>%
                mutate_at(c("WINNER", "RUNNER_UP"),
                          ~ replace_na(., 0)) %>%
                rename(CHAMPION = WINNER) %>%
                # mutate(DIVISION_WIN = RUNNER_UP + WINNER,
                #        CONFERENCE_WIN = WINNER) %>%
        select(SEASON, CONFERENCE, TEAM, CHAMPION, RUNNER_UP),
        by = c("SEASON","CONFERENCE", "TEAM"))

source(here::here("functions", "table_conference_championship.R"))

```


## Strength of Schedule and Predicted Record

```{r schedule difficulty and record}

## make color func
bg_bowl_func = 
        function(x) {
                
                breaks = seq(0, 1, 0.01)
                colorRamp=colorRampPalette(c("white", "deepskyblue1"))
                col_palette <- colorRamp(length(breaks))
                mycut <- cut(x, 
                             breaks = breaks,
                             include.lowest = TRUE, 
                             right=T,
                             label = FALSE)
                col_palette[mycut]
        }

# table of difficulty and bowl
preseason_predictions = games %>%
        filter(SEASON == params$input_season) %>%
        select(SEASON, NEUTRAL_SITE, 
               HOME_TEAM, AWAY_TEAM,
               HOME_CONFERENCE, AWAY_CONFERENCE,
               HOME_DIVISION, AWAY_DIVISION) %>%
        mutate(HOME = T) %>%
        rename(TEAM = HOME_TEAM,
               CONFERENCE = HOME_CONFERENCE,
               DIVISION = HOME_DIVISION,
               OPPONENT = AWAY_TEAM,
               OPPONENT_CONFERENCE = AWAY_CONFERENCE) %>%
        bind_rows(.,
                  games %>%
                filter(SEASON == params$input_season) %>%
                select(SEASON, NEUTRAL_SITE, AWAY_TEAM, HOME_TEAM, AWAY_CONFERENCE, HOME_CONFERENCE, HOME_DIVISION, AWAY_DIVISION) %>%
                mutate(HOME = F) %>%
                rename(TEAM = AWAY_TEAM,
                       CONFERENCE = AWAY_CONFERENCE,
                       DIVISION = AWAY_DIVISION,
                       OPPONENT = HOME_TEAM,
                       OPPONENT_CONFERENCE = HOME_CONFERENCE)) %>%
        left_join(.,
                          do.call(rbind, c(adjusted_teams,
                           base_teams[!(names(base_teams) %in% names(adjusted_teams))])) %>% 
                          as.data.frame() %>% 
                          rownames_to_column("OPPONENT") %>%
                          rename(OPPONENT_ELO = V1),
                  by = c("OPPONENT")) %>%
        left_join(.,
                          do.call(rbind, c(adjusted_teams,
                           base_teams[!(names(base_teams) %in% names(adjusted_teams))])) %>% 
                          as.data.frame() %>% 
                          rownames_to_column("TEAM") %>%
                          rename(TEAM_ELO = V1),
                  by = c("TEAM")) %>%
        mutate(OPPONENT_ELO = case_when(HOME == F ~ OPPONENT_ELO + 75,
                                        HOME == T ~ OPPONENT_ELO - 75)) %>%
        filter(DIVISION == 'fbs') %>%
        mutate_at(c("OPPONENT_ELO"),
                  ~ replace_na(., 0)) %>%
        group_by(SEASON, TEAM, TEAM_ELO) %>%
        summarize(OPPONENT_MEAN_ELO = mean(OPPONENT_ELO),
                  OPPONENT_MEDIAN_ELO = median(OPPONENT_ELO),
                  diff = mean(TEAM_ELO - OPPONENT_ELO),
                  .groups = 'drop') %>%
        arrange(desc(OPPONENT_MEAN_ELO)) %>%
        mutate_if(is.numeric, round, 1) %>%
        mutate(RANK = row_number()) %>%
        select(SEASON, RANK,
               TEAM, 
               TEAM_ELO, 
               OPPONENT_MEAN_ELO, 
               OPPONENT_MEDIAN_ELO) %>%
        select(SEASON, RANK, TEAM, TEAM_ELO, OPPONENT_MEAN_ELO) %>%
        rename(SCHEDULE_RANK = RANK) %>%
        left_join(.,
                  # elo and win totals
        season_simulations %$%
                team_outcomes %>% 
                group_by(.id, SEASON, TEAM, CONFERENCE, SIM_OUTCOME) %>%
                count() %>% 
                group_by(SEASON, TEAM, CONFERENCE, SIM_OUTCOME) %>%
                summarize(mean = mean(n),
                          median = median(n),
                          .groups = 'drop') %>%
                select(-median) %>%
                spread(SIM_OUTCOME, mean) %>%
                mutate_at(c("win", "loss"),
                          ~ replace_na(., 0)) %>%
                select(SEASON, TEAM, win, loss) %>%
                arrange(desc(win)) %>%
                mutate_if(is.numeric, round, 2) %>%
                left_join(.,
                          season_simulations %$%
                team_outcomes %>% 
                group_by(SEASON, TEAM) %>% 
                filter(GAME_DATE == max(GAME_DATE)) %>% 
                summarize(ELO = round(mean(POSTGAME_ELO),0),
                          .groups = 'drop') %>%
                arrange(desc(ELO)),
                by = c("SEASON", "TEAM")) %>%
                select(SEASON, TEAM, win, loss),
        by = c("SEASON", "TEAM")) %>%
        left_join(.,
                  season_simulations %$%
                        team_outcomes %>% 
                        filter(DIVISION == 'fbs') %>%
                        mutate(win = case_when(SIM_OUTCOME == 'win' ~ 1, TRUE ~ 0)) %>% 
                        mutate(sims = n_distinct(.id)) %>% 
                        group_by(.id, SEASON, TEAM, CONFERENCE, SIM_OUTCOME, sims) %>% 
                        summarize(wins = sum(win), .groups = 'drop') %>% 
                        filter(wins >=6) %>%
                        group_by(SEASON, TEAM, sims) %>% 
                        count() %>%
                        ungroup() %>% 
                        mutate(BOWL = n/sims) %>% 
                        select(SEASON, TEAM, BOWL),
                  by = c("SEASON", "TEAM")) %>%
        left_join(.,
                  conference_championship_probs,
                  by = c("TEAM", "SEASON"))

# make datatable
preseason_predictions %>%
        mutate(TEAM = factor(TEAM)) %>%
        mutate(CONFERENCE = factor(CONFERENCE)) %>%
        rename(Season = SEASON,
               Team = TEAM,
               Conference = CONFERENCE,
               Difficulty = SCHEDULE_RANK,
               Wins = win,
               Losses = loss,
               Bowl = BOWL,
               Conf.Ch. = CHAMPION,
               Rating = TEAM_ELO) %>%
        mutate_at(c("Conf.Ch."),
                  ~ replace_na(.,0)) %>%
        select(Season,
               Conference,
               Team,
               Difficulty,
               Wins,
               Losses,
               Bowl, 
               Conf.Ch.) %>%
           datatable(escape=F,
                  rownames = F,
                       filter = 'top',
                  options = list(pageLength = 25,
                                 scrollX=F,
                                 autowidth=T,
                                 columnDefs = list(list(className = 'dt-center',
                                                        targets = c("Season", "Conference",
                                                                    "Team", "Difficulty",
                                                                    "Wins",
                                                                    "Losses",
                                                                    "Conf.Ch.")))))

```

## Win Total Probabilities

```{r table of win probabilities}

source(here::here("functions", "table_wins_season.R"))

table_wins_season(season_simulations %$%
                          team_outcomes,
                  games,
                  2022)

```

# Conference Championship Probabilities

## SEC 

```{r conf champ sec}

table_conference_championship(conference_championship_probs,
                              'SEC')

```

## Big Ten

```{r conf champ big ten}

table_conference_championship(conference_championship_probs,
                              'Big Ten')

```

## Big 12

```{r conference championship tables big 12}

table_conference_championship(conference_championship_probs,
                              'Big 12')

```

## ACC

```{r conference champs acc}

table_conference_championship(conference_championship_probs,
                              'ACC')

```

## Pac-12

```{r conference champs pac 12}

table_conference_championship(conference_championship_probs,
                              'Pac-12')

```

### American Athletic

```{r conference championship tables aac}

table_conference_championship(conference_championship_probs,
                              'American Athletic')

```

## Sun Belt

```{r conference championship tables conf usa}

table_conference_championship(conference_championship_probs,
                              'Sun Belt')

```

## Mountain West

```{r conference champs mountain west}

table_conference_championship(conference_championship_probs,
                              'Mountain West')

```

## Mid-American

```{r conference champs mid america}

table_conference_championship(conference_championship_probs,
                              'Mid-American')

```

## Conference USA

```{r conference champs conf usa}

table_conference_championship(conference_championship_probs,
                              'Conference USA')

```

<!-- # One Simulated Season -->

<!-- ```{r randomly sample and look at results} -->

<!-- set.seed(1999) -->
<!-- samp = sample(1:1000, 1) -->
<!-- max_week = max(season_simulations$game_outcomes$WEEK) -->

<!-- set.seed(1999) -->
<!-- season_simulations %$% -->
<!--         game_outcomes %>% -->
<!--         filter(.id == samp) %>% -->
<!--         rename(HOME_MARGIN = HOME_SIM_MARGIN, -->
<!--                sim = .id) %>% -->
<!--         mutate(CONFERENCE_GAME = case_when(HOME_CONFERENCE == AWAY_CONFERENCE ~ T, -->
<!--                                            TRUE ~ F)) %>% -->
<!--         select(sim, SEASON, WEEK, HOME_TEAM, AWAY_TEAM, HOME_MARGIN) %>% -->
<!--         bind_rows(., -->
<!--                   conference_championship_results %>% -->
<!--                           filter(.id == samp) %>% -->
<!--                           rename(HOME_TEAM = TEAM_HOME, -->
<!--                                  AWAY_TEAM = TEAM_AWAY, -->
<!--                                  HOME_MARGIN = HOME_SIM_MARGIN, -->
<!--                                  sim = .id) %>% -->
<!--                           mutate(WEEK = max_week + 1) %>% -->
<!--                           mutate(CONFERENCE_GAME = T) %>% -->
<!--                           select(sim, SEASON, WEEK, HOME_TEAM, AWAY_TEAM, HOME_MARGIN)) %>% -->

<!--         mutate(WINNER = case_when(HOME_MARGIN > 0 ~ HOME_TEAM, -->
<!--                                   HOME_MARGIN < 0 ~ AWAY_TEAM)) %>% -->
<!--         rename(Sim = sim, -->
<!--                Season = SEASON, -->
<!--                Week = WEEK, -->
<!--                Home = HOME_TEAM, -->
<!--                Away = AWAY_TEAM, -->
<!--                Margin = HOME_MARGIN, -->
<!--                Winner = WINNER) %>% -->
<!--          datatable(escape=F, -->
<!--                   rownames = F) -->

<!-- ``` -->

```{r save output}

# save preseason ratings
save(adjusted_teams,
     file = here::here("simulations",
                       "preseason",
                       params$input_season,
                       "adjusted_teams.Rdata"))

# base
save(base_teams,
     file = here::here("simulations",
                       "preseason",
                       params$input_season,
                       "base_teams.Rdata"))

# save results of simulations
save(season_simulations,
     file = here::here("simulations",
                       "preseason",
                       params$input_season,
                       "season_simulations.Rdata"))

# save conference championship results
save(conference_championship_results,
     file = here::here("simulations",
                       "preseason",
                       params$input_season,
                       "conference_champ_sims.Rdata"))

# table of probabilities
save(conference_championship_probs,
     file = here::here("simulations",
                       "preseason",
                       params$input_season,
                       "conference_champ_probs.Rdata"))

# save 
save(preseason_ratings,
     file = here::here("simulations",
                       "preseason",
                       params$input_season,
                       "preseason_ratings.Rdata"))

# 
save(preseason_predictions,
          file = here::here("simulations",
                       "preseason",
                       params$input_season,
                       "preseason_predictions.Rdata"))

```
